### 2017-02-01 Manuscript/manuscript.tex
Updates

%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with 
% the REVTEX 4.1 document class and the AIP substyles.
% 
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

%\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

%\usepackage{graphicx}

%\draft % marks overfull lines with a black rule on the right
%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}

\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}

\usepackage{rotating} 
\usepackage{times}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[obeyFinal]{easy-todo}
\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{NMRlipids IV: Other than PC headgroups} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{University of Helsinki}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
% insert abstract here
  Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine
  (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between different lipid molecules in bilayers.
\end{abstract}

%\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands

%\label{}
\section{Introduction}
Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between different lipid molecules in bilayers.

%\section{Methods}

%\subsection{Molecular dynamics simulations}

\section{Comparison of order parameters between different headgroups}

Experimental order parameters for different lipid headgroups are
collected in Fig. \ref{HGorderParameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=17.2cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Order parameters for headgroup and glycerol backbone with different headgroups
  }
\end{figure}

\section{Ion binding in bilayers with charged lipids}

Headgroup order parameters as a function of CaCl concentration
from experiments with charged lipids are shown in Figs \ref{OrderParametersWithCaCl}
and \ref{OrderParametersWithCaClBELOW1M}.

Order parameter changes are shown in Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}

\begin{figure}[]
  \centering
  \includegraphics[width=17.2cm]{../Figs/LIPIDSwithCaCl.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration for systems with charged lipids.
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=17.2cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaClBELOW1M}
    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=17.2cm]{../Figs/CHANGESwithCaClBELOW1M.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes in order parameters
  }
\end{figure}

\section{Conclusions}

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}
\newpage
\appendix
\begin{center}
{\bf SUPPLEMENTARY INFORMATION}
\end{center}

% Create the reference section using BibTe
\bibliography{refs.bib}

%\newpage
%\section{APPENDIX: The NMR results reported by Tiago Ferreira}

\listoftodos

\end{document}
%
% ****** End of file aiptemplate.tex ******

### 2017-02-16 Manuscript/manuscript.tex

@@
   Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine
-  (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between different lipid molecules in bilayers.
+  (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between 
+  different lipid molecules in bilayers.
 \end{abstract}
@@
 \section{Introduction}
-Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine (these are discussed in NMRlipids I) headgroup.
-The information will be used to understand structural differences between different lipid molecules in bilayers.
+Phospholipids containing various polar headgroups and acyl
+chains are essential building blocks of biological membranes.
+Atomistic level structural details of lipids and lipid-ion 
+interactions are considered highly important in several
+biological processes. The lipid structure and ion binding
+can be studied in detail with NMR spectroscopy. However,
+the structural interpretation of NMR data requires usage
+of models. The combination of classical molecular dynamics
+simulations with NMR data, especially with C-H bond
+order parameters, can potentially give atomistic resolution
+interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.
+
+Our recent studies concluded that MD models are cabable to
+give structural interpretation for phosphatidylcholine 
+hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
+backbone and cation binding posed a major challenge for current
+force fields \cite{botan15,ollila16,catte16,ferreira16}. 
+These conclusions were reached by reviewing extensively 
+available experimental data from various sources and using
+NMRlipids Open Collaboration project to collect massive 
+amount of simulation data \cite{botan15,catte16}.
+
+Here we apply the same approach in search of MD simulation
+models which would reproduce glycerol backbone and 
+headgroup structures of lipids with PE, PG and PS headgroups.  
+In addition, we attempt to find a MD simulation model
+that would be able to correctly describe cation binding in
+bilayer containing negatively charged PG and PS lipids.
+
+%Several MD simulation models have succesfully described 
+%acyl chain structures and their qaulitative changes in different
+%conditions for phosphatidylcholine lipids (for review see \cite{ollila16}).
+%However, the current simulation models were not found to be accurate
+%enough for full structural interpretation of phosphatidylcholine headgroup and
+%glycerol backbone \cite{botan15}. Also Na$ +$  binding affinities
+%were found to be significantly overestimated in several models and 
+%none of the available models was able to Ca$2+$
 
@@
 
-\section{Comparison of order parameters between different headgroups}
+\section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}
 
@@
   \centering
-  \includegraphics[width=17.2cm]{../Figs/HGorderparameters.eps}
+  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
   \caption{\label{HGorderParameters}
@@
 
-\section{Ion binding in bilayers with charged lipids}
+\section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}
 
-Headgroup order parameters as a function of CaCl concentration
-from experiments with charged lipids are shown in Figs \ref{OrderParametersWithCaCl}
-and \ref{OrderParametersWithCaClBELOW1M}.
+\todo{Simulation data for lipids with different headgroups to be collected}
 
-Order parameter changes are shown in Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}
+\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids}
 
+Ion binding affinity in lipid bilayer containing PC lipids can be 
+measured by using C-H bond order parameters for headgroup $\alpha$ and
+$\beta$ carbons by using the electrometer concept. The decrease of these 
+order parameters is linearly proportional to the amount of bound charge 
+in bilayer \cite{??,catte16}. The electrometer concept can be used also
+for bilayers containing PC lipids mixed with charged lipids \cite{macdonal87,roux90,??}.
+This is demonstrated in Figs \ref{OrderParametersWithCaCl},
+\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M}, where PC headgroup order parameters 
+as a function of CaCl$_2$ concentration are shown from experiments with POPC and PS or PG.
 \begin{figure}[]
   \centering
-  \includegraphics[width=17.2cm]{../Figs/LIPIDSwithCaCl.eps}
+  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
   \caption{\label{OrderParametersWithCaCl}
-    PC headgroup order parameters as a function of CaCl concentration for systems with charged lipids.
+    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
+    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
+    POPC:POPS mixture data from \cite{roux90} and POPC:POPG mixture data from \cite{macdonald87}.
+  }
+  \todo{Check the NaCl concentrations in the samples.}
+\end{figure}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
+  \caption{\label{OrderParametersWithCaClBELOW1M}
+    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
   }
@@
   \centering
-  \includegraphics[width=17.2cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
-  \caption{\label{OrderParametersWithCaClBELOW1M}
-    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
-  }
-\end{figure}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=17.2cm]{../Figs/CHANGESwithCaClBELOW1M.eps}
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClBELOW1M.eps}
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
@@
 \end{figure}
+
+Order parameters increase when PS or PG are mixed with PC 
+in the absence of additional ions, as seen from Figs. 
+\ref{OrderParametersWithCaCl} and \ref{OrderParametersWithCaClBELOW1M}.
+In electrometer concept this is explained by the tilting of
+headgroup more parallel to membrane normal \cite{??}.
+The order parameter decrease with increasing CaCl$_2$ concentration 
+is explained by the tilting of headgroup more perpendicular
+to membrane normal. This decrease is shown to be a good measure
+for the amount of bound ions in lipid bilayer with PC lipids \cite{??,catte16}.
+
+The order parameter decrease as a function of added CaCl$_2$
+for systems with different amount of negative charge are shown 
+in Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}.
+As expected, the order parameter decrease is more 
+pronounced with increasing amount of negatively charged lipids in
+bilayer. This is explained by the increase of cation concentration
+in proximity of bilayer containing negatively charged lipids \cite{??}.
+
+In addition to the changes in PC headgroup order parameters with 
+ion binding, also changes in PS and PG headgroup order parameters
+are measured.
+\todo{Add this data}.
+
+\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids in simulations}
+
+\todo{Simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
 

### 2017-02-16 Manuscript/manuscript.tex
Lot of updates

@@
 
-Experimental order parameters for different lipid headgroups are
-collected in Fig. \ref{HGorderParameters}.
+Absolute values of experimental order parameters for different lipid headgroups are
+collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
+thus only absolute values are used for now.
 \begin{figure}[]
@@
   \caption{\label{HGorderParameters}
-    Order parameters for headgroup and glycerol backbone with different headgroups
+    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
+    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
+    POPG from \cite{borle85} contains 10nM PIPES, DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
   }
@@
 in bilayer \cite{??,catte16}. The electrometer concept can be used also
-for bilayers containing PC lipids mixed with charged lipids \cite{macdonal87,roux90,??}.
+for bilayers containing PC lipids mixed with charged lipids \cite{macdonald87,roux90,??}.
 This is demonstrated in Figs \ref{OrderParametersWithCaCl},
@@
     Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-    POPC:POPS mixture data from \cite{roux90} and POPC:POPG mixture data from \cite{macdonald87}.
+    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
+    and POPC:POPG mixture data without NaCl from \cite{borle85}.
   }
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClBELOW1M.eps}
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}

### 2017-02-21 Manuscript/manuscript.tex
Updates

@@
 
+Based on superficial reading, the conclusions in the literature are roughly
+1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81},
+2) glycerol backbone and headgroup strucures are similar in model membranes and in bacteria \cite{gally81,??},
+3) headgroup structure is phosphatidylserine more rigid than in other phospholipids, which are quite similar to
+each others \cite{??}.
+In contrast to PC lipids, extensive discussion on glycerol backbone and headgroup structural details do not
+exist (as far as I know).
+
 \section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}
 
-\todo{Simulation data for lipids with different headgroups to be collected}
+Several simulations containing PE, PG and PS lipids have been published \cite{??},
+however, glycerol backbone and headgroup order parameters are not compared to
+the experiments (based on superficial reading of literature).
+
+Glycerol backbone and headgroup order parameters for PS lipids from experiments
+and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
+The preliminary test with CHARMM GUI simulations show suggest that the model
+overestimates magnitudes in $\beta$ and g$_3$ order parameters and forking in $\alpha$.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
+  \caption{\label{HGorderParametersPOPS}
+    Order parameters for POPS headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
+    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
+    Absolute values are shown for experimental data, because signs are not known.
+    Simulations values are -S$_{CH}$
+  }
+  \todo{More simulation data for lipids with different headgroups to be collected}
+  \todo{CHARMM GUI simulation contains only counter ions as potassium.
+    All experiments here contain some amount of sodium salt. The best
+    ion concentrations for comparison should be figured out.}
+  \todo{Experimental signs of the order parameters would highly useful.}
+\end{figure}
 
@@
 are measured.
-\todo{Add this data}.
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
+  \caption{\label{PSPGchangesWITHCaCl}
+    Changes in order parameters
+  }
+  \todo{Get the small concentration data from the inserts}
+\end{figure}
 

### 2017-02-22 Manuscript/manuscript.tex
Updates

@@
     from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
-    POPG from \cite{borle85} contains 10nM PIPES, DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
+    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
   }
@@
     Simulations values are -S$_{CH}$
+  }
+  \todo{More simulation data for lipids with different headgroups to be collected}
+  \todo{CHARMM GUI simulation contains only counter ions as potassium.
+    All experiments here contain some amount of sodium salt. The best
+    ion concentrations for comparison should be figured out.}
+  \todo{Experimental signs of the order parameters would highly useful.}
+\end{figure}
+
+Glycerol backbone and headgroup order parameters for PS lipids from experiments
+and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
+The preliminary test with CHARMM GUI simulations show suggest that the model
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
+  \caption{\label{HGorderParametersPOPG}
+    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
+    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
+    E.Coli PG results from \cite{gally81}).
+    Absolute values are shown, because signs are not known for experimental data.
+    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
+    smaller $\gamma$ are positive.
   }

### 2017-03-07 Manuscript/manuscript.tex
Update manuscript

@@
 \section{Introduction}
-Phospholipids containing various polar headgroups and acyl
-chains are essential building blocks of biological membranes.
-Atomistic level structural details of lipids and lipid-ion 
-interactions are considered highly important in several
-biological processes. The lipid structure and ion binding
-can be studied in detail with NMR spectroscopy. However,
-the structural interpretation of NMR data requires usage
-of models. The combination of classical molecular dynamics
-simulations with NMR data, especially with C-H bond
-order parameters, can potentially give atomistic resolution
-interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.
 
-Our recent studies concluded that MD models are cabable to
-give structural interpretation for phosphatidylcholine 
-hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
-backbone and cation binding posed a major challenge for current
-force fields \cite{botan15,ollila16,catte16,ferreira16}. 
-These conclusions were reached by reviewing extensively 
-available experimental data from various sources and using
-NMRlipids Open Collaboration project to collect massive 
-amount of simulation data \cite{botan15,catte16}.
+In NMRlipids I and II project we were looking for a MD model
+which would correctly reproduce headgroup and glycerol
+backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
+Here we extend the same goal for other than PC lipids.
+Currently the focus is on PE, PG and PS bilayers and their
+mixtures with PC. Experimental data with different amounts of 
+added salt is now collected and presented in this manuscript.
+Simulations for bilayers containing PG and PS with low salt conditions
+are ran with parameters from CHARMM GUI
+and compared to experiments. Some order parameters from
+simulations seems to be off from experimental values.
+Simulation data from other models would be highly useful 
+to see if some of the existing models would reproduce the 
+experimental order parameters and to analyze different 
+conformations predicted by different models respect to 
+experiments.
 
-Here we apply the same approach in search of MD simulation
-models which would reproduce glycerol backbone and 
-headgroup structures of lipids with PE, PG and PS headgroups.  
-In addition, we attempt to find a MD simulation model
-that would be able to correctly describe cation binding in
-bilayer containing negatively charged PG and PS lipids.
+%Phospholipids containing various polar headgroups and acyl
+%chains are essential building blocks of biological membranes.
+%Atomistic level structural details of lipids and lipid-ion 
+%interactions are considered highly important in several
+%biological processes. The lipid structure and ion binding
+%can be studied in detail with NMR spectroscopy. However,
+%the structural interpretation of NMR data requires usage
+%of models. The combination of classical molecular dynamics
+%simulations with NMR data, especially with C-H bond
+%order parameters, can potentially give atomistic resolution
+%interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.
+
+%Our recent studies concluded that MD models are cabable to
+%give structural interpretation for phosphatidylcholine 
+%hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
+%backbone and cation binding posed a major challenge for current
+%force fields \cite{botan15,ollila16,catte16,ferreira16}. 
+%These conclusions were reached by reviewing extensively 
+%available experimental data from various sources and using
+%NMRlipids Open Collaboration project to collect massive 
+%amount of simulation data \cite{botan15,catte16}.
+
+%Here we apply the same approach in search of MD simulation
+%models which would reproduce glycerol backbone and 
+%headgroup structures of lipids with PE, PG and PS headgroups.  
+%In addition, we attempt to find a MD simulation model
+%that would be able to correctly describe cation binding in
+%bilayer containing negatively charged PG and PS lipids.
 

### 2017-03-08 Manuscript/manuscript.tex
Update text

@@
 \usepackage[obeyFinal]{easy-todo}
+\usepackage{csquotes}
+
 \begin{document}
@@
 
-Based on superficial reading, the conclusions in the literature are roughly
-1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81},
-2) glycerol backbone and headgroup strucures are similar in model membranes and in bacteria \cite{gally81,??},
-3) headgroup structure is phosphatidylserine more rigid than in other phospholipids, which are quite similar to
-each others \cite{??}.
-In contrast to PC lipids, extensive discussion on glycerol backbone and headgroup structural details do not
-exist (as far as I know).
+Based on superficial reading, the conclusions in the literature are roughly \\ 
+1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
+2) glycerol backbone and headgroup strucure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
+3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
+Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
+In contrast to PC lipids (see \cite{botan15} and references therein).
 
@@
 
-Several simulations containing PE, PG and PS lipids have been published \cite{??},
+Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
 however, glycerol backbone and headgroup order parameters are not compared to
@@
 
-Glycerol backbone and headgroup order parameters for PS lipids from experiments
-and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
-The preliminary test with CHARMM GUI simulations show suggest that the model
-overestimates magnitudes in $\beta$ and g$_3$ order parameters and forking in $\alpha$.
+Comparison between experiments and simulations with CHARMM GUI
+parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
+the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
+and forking in $\alpha$.
 \begin{figure}[]
@@
 
-Glycerol backbone and headgroup order parameters for PS lipids from experiments
-and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
-The preliminary test with CHARMM GUI simulations show suggest that the model
+Comparison between experiments and simulations with CHARMM GUI
+parameters for PG lipids in Fig. \ref{HGorderParametersPOPG} suggest that
+the model overestimates forking in $\gamma$ and $\alpha$ order parameters,
+and magnitudes of $\beta$ and $g_1$ order parameters.
 \begin{figure}[]
@@
 
-Ion binding affinity in lipid bilayer containing PC lipids can be 
-measured by using C-H bond order parameters for headgroup $\alpha$ and
-$\beta$ carbons by using the electrometer concept. The decrease of these 
-order parameters is linearly proportional to the amount of bound charge 
-in bilayer \cite{??,catte16}. The electrometer concept can be used also
-for bilayers containing PC lipids mixed with charged lipids \cite{macdonald87,roux90,??}.
+PC lipid headgroup order parameters can used to measure ion binding
+affinity, because their magnitude is linearly proportional
+to the amount of bound charge in bilayer \cite{seelig87,catte16}.
+This molecular electrometer concept can be used also
+for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
 This is demonstrated in Figs \ref{OrderParametersWithCaCl},
-\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M}, where PC headgroup order parameters 
-as a function of CaCl$_2$ concentration are shown from experiments with POPC and PS or PG.
+\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
+showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
+as a function of CaCl$_2$ concentration in the presence of different amounts of
+negatively charged PS or PG lipids.
 \begin{figure}[]
@@
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    Changes in order parameters
+    The change of PC headgroup order parameters
+    in the presence of different amount of negatively charged lipids
+    respect to the values without added CaCl$_2$.
+    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
   }
@@
 
-Order parameters increase when PS or PG are mixed with PC 
-in the absence of additional ions, as seen from Figs. 
-\ref{OrderParametersWithCaCl} and \ref{OrderParametersWithCaClBELOW1M}.
-In electrometer concept this is explained by the tilting of
-headgroup more parallel to membrane normal \cite{??}.
-The order parameter decrease with increasing CaCl$_2$ concentration 
-is explained by the tilting of headgroup more perpendicular
-to membrane normal. This decrease is shown to be a good measure
-for the amount of bound ions in lipid bilayer with PC lipids \cite{??,catte16}.
+PC headgroup order parameters increase when negatively charged
+PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
+as expected based on electrometer concept \cite{seelig87}
+(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
+%In electrometer concept this is explained by the tilting of
+%headgroup more parallel to membrane normal \cite{??}.
+Further, the order parameters decrease with the addition
+of CaCl$_2$ and the decrease is more pronounced for systems with more
+negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
+the Ca2+ binding is presumably fully cancels the charge from negative lipids and
+overcharging occurs above these concenterations.
+The interpretation of this data and some other results has been that \cite{seelig90}
+\begin{displayquote}
+  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
+  (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; (ii) the free Ca$^{2+}$
+  concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
+  charge, either due to protein or to lipid; (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
+  of bound Ca$^{2+}$ at neutral and charged lipids; (iv) the actual binding step can be described by a Langmuir
+  adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
+  used to describe the chemical binding equilibrium.''}
+\end{displayquote}
 
-The order parameter decrease as a function of added CaCl$_2$
-for systems with different amount of negative charge are shown 
-in Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}.
-As expected, the order parameter decrease is more 
-pronounced with increasing amount of negatively charged lipids in
-bilayer. This is explained by the increase of cation concentration
-in proximity of bilayer containing negatively charged lipids \cite{??}.
-
-In addition to the changes in PC headgroup order parameters with 
-ion binding, also changes in PS and PG headgroup order parameters
-are measured.
-
+Also the experimental order parameters for PS and PG headgroups
+as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
+\todo{These should be compared to simulations for potential structural interpretation of the changes.}
 \begin{figure}[]
@@
   \caption{\label{PSPGchangesWITHCaCl}
-    Changes in order parameters
+    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
   }

### 2017-03-09 Manuscript/manuscript.tex
Updates

@@
 
-\title{NMRlipids IV: Other than PC headgroups} %Title of paper
+\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper
 
@@
 %\homepage[]{Your web page}
-%\thanks{}
-%\altaffiliation{}
-\affiliation{University of Helsinki}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\affiliation{Institute of Biotechnology, University of Helsinki}
 
@@
 1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
-2) glycerol backbone and headgroup strucure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
+2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
 3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
@@
   \caption{\label{HGorderParametersPOPS}
-    Order parameters for POPS headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
+    Order parameters for POPS headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
     (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
@@
   \caption{\label{HGorderParametersPOPG}
-    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
+    Order parameters for PG headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
     (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
@@
 
-\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids}
+\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}
 
@@
 At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
-the Ca2+ binding is presumably fully cancels the charge from negative lipids and
+the Ca2+ binding presumably fully cancels the charge from negative lipids and
 overcharging occurs above these concenterations.
@@
   {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
-  (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; (ii) the free Ca$^{2+}$
+    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
+    (ii) the free Ca$^{2+}$
     concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
-  charge, either due to protein or to lipid; (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
-  of bound Ca$^{2+}$ at neutral and charged lipids; (iv) the actual binding step can be described by a Langmuir
+    charge, either due to protein or to lipid; \\
+    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
+    of bound Ca$^{2+}$ at neutral and charged lipids; \\
+    (iv) the actual binding step can be described by a Langmuir
     adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is

### 2017-04-18 Manuscript/manuscript.tex
Add results by Thomas Piggot, reported in http://nmrlipids.blogspot.fi/2017/03/nmrlipids-iv-headgroup-glycerol.html

@@
   \caption{\label{HGorderParametersPOPS}
-    Order parameters for POPS headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
+    Order parameters for POPS (top) and DOPS (bottom) headgroup and glycerol
+    backbone from simulations with different models and experiments without CaCl$_2$ 
     (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
@@
   }
-  \todo{More simulation data for lipids with different headgroups to be collected}
-  \todo{CHARMM GUI simulation contains only counter ions as potassium.
+  \todo{More simulation data for lipids with different headgroups to be collected} \\
+  \todo{Some simulations contain potassium as counterions, while some sodium.
     All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.}
-  \todo{Experimental signs of the order parameters would highly useful.}
+    ion concentrations for comparison should be figured out.} \\
+  \todo{Experimental signs of the order parameters would highly useful.} \\
+  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
+  pure POPS?}
 \end{figure}

### 2017-05-18 Manuscript/manuscript.tex
Add link to the issue
Add results analyzed from from DPPE data delivered by Fernando Favela at https://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1492449148750#c3107105122336970292

@@
 
+Absolute values of order parameters between Slipids simulations \cite{??}
+and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
+Glycerol bakcbone order parameters in Slipids are off from experiments,
+as systematically observed in this and previous studies \cite{??}.
+Alpha order parameter is too low, but beta shows apparent agreement
+with experiments. However, the sign of beta order parameter is positive,
+in contrast to PC where negative sign was measured. Thus, the the beta order
+parameter agrees with experiment with the assumption that its sign is opposite
+than for PC.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.eps}
+  \caption{\label{HGorderParametersPE}
+    Order parameters for DPPE headgroup and glycerol
+    backbone from simulations with Slipids \cite{??} and experiments
+    (DPPE from \cite{seelig76} and E.coliPE from \cite{gally81}).
+    Absolute values are shown, because signs are not known experimentally.
+  }
+  \todo{Experimental signs of the order parameters would be highly useful.} \\
+\end{figure}
+
 Comparison between experiments and simulations with CHARMM GUI
@@
   \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
-  pure POPS?}
+  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
 \end{figure}

### 2017-07-27 Manuscript/manuscript.tex
Add POPG data with Calcium, reported by Jesper Madsen

@@
 
-\todo{Simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
+Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
+is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
+is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
+The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
+It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
+but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
+Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
+suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
+POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
+parameter of bound charge \todo{Should we check this against cationic surfactant experiments \cite{scherer89}?}.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPG.eps}
+  \caption{\label{changesWITHCaClPG}
+    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
+    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
+  }
+\end{figure}
+
+Also dependence of $\beta$-carbon of PG on CaCl$_2$ concentration is compared with
+experiments \cite{borle85} in Fig. \ref{PSPGchangesWITHCaCl}. Absolute value of
+the order parameter is too large without ions, but rapid decrease due to addition of
+CaCl$_2$ is observed in agreement with experiments for systems with 1:1 mixture of POPC and POPG.
+In addition, absolute value in systems with CaCl$_2$ is in agreement with experiments.
+However, system with 4:1 mixture of POPC and POPG behaves differently, but experimental
+data is not available for comparison for this mixture.
+
+\todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
 

### 2017-08-15 Manuscript/manuscript.tex
Add results delivered by Buslaev and some updates in the manuscript

@@
 
+\subsection{PE headgroup}
 Absolute values of order parameters between Slipids simulations \cite{??}
@@
 
-Comparison between experiments and simulations with CHARMM GUI
-parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
-the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
-and forking in $\alpha$.
+\subsection{PS headgroup}
+Experimental order parameters for bilayers with PS headgroups are 
+available in literature only for DOPS  \cite{browning80}. These are shown together
+with the results from different simulation models in Fig. \ref{HGorderParametersDOPS}.
+The signs are not yet know experimentally.
+Before the experimental data becomes available we have denoted 
+negative signs for $\beta$ and $\alpha$ segment order parameters,
+because this selection of signs is in best agreement with the most simulation
+results. With this interpretation the CHARMM36, CHARMM36ua and Slipids
+models would be in relatively good agreement with experimental data in
+the headgroup region. However, if the experimental signs turn out to
+be different than expected, this conclusion will change.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersDOPS.eps}
+  \caption{\label{HGorderParametersDOPS}
+    Order parameters for DOPS headgroup and glycerol
+    backbone from simulations with different models and experiments without CaCl$_2$ 
+    Experimental data from \cite{browning80} contains 0.1M of NaCl.
+    Signs are not known experimentally. They chosen to give best agreement simulations.
+  }
+  \todo{Experimental signs of the order parameters would highly useful.} \\
+\end{figure}
+
+The glycerol backbone order parameters for Slipids model are, however,
+far from experimental values, as previously observed also for PC 
+lipids \cite{botan15}. The glycerol backbone structures between PC and
+PS lipids simulated with CHARMM36 are compared with the structures
+simulated with CHARMM36 in Fig. \ref{glycerol_buslaev}.
+The differences in sampled conformation leading to the order parameter
+differences are clearly visible in the figure.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
+  \caption{\label{glycerol_buslaev}
+    Snapshots overlayed from different simulations for glycerol backbone region
+    by Pavel Buslaev.
+  }
+\end{figure}
+
+%Comparison between experiments and simulations with CHARMM GUI
+%parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
+%the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
+%and forking in $\alpha$.
 \begin{figure}[]
@@
 
+\subsection{PG headgroup}
 Comparison between experiments and simulations with CHARMM GUI

### 2017-08-23 Manuscript/manuscript.tex
Add experimental results for the signs of PS headgroup order parameters

@@
 \subsection{PS headgroup}
-Experimental order parameters for bilayers with PS headgroups are 
-available in literature only for DOPS  \cite{browning80}. These are shown together
-with the results from different simulation models in Fig. \ref{HGorderParametersDOPS}.
-The signs are not yet know experimentally.
-Before the experimental data becomes available we have denoted 
-negative signs for $\beta$ and $\alpha$ segment order parameters,
-because this selection of signs is in best agreement with the most simulation
-results. With this interpretation the CHARMM36, CHARMM36ua and Slipids
-models would be in relatively good agreement with experimental data in
-the headgroup region. However, if the experimental signs turn out to
-be different than expected, this conclusion will change.
+Order parameters for bilayers with PS headgroups
+from different experiments and simulations are shown in
+Figs. \ref{HGorderParametersDOPS} and  \ref{HGorderParametersPOPS}.
+None of the tested models gives satisfactory agreement with
+experiments for order parameters in headgroup $\alpha$ and $\beta$
+carbons.
 \begin{figure}[]
@@
     Experimental data from \cite{browning80} contains 0.1M of NaCl.
-    Signs are not known experimentally. They chosen to give best agreement simulations.
+    Signs are taken from experiments for POPS described in Supplementary Information.
   }
-  \todo{Experimental signs of the order parameters would highly useful.} \\
+\end{figure}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
+  \caption{\label{HGorderParametersPOPS}
+    Headgroup and glycerol backbone order parameters for systems with POPS 
+    from simulations with different models and experiments \cite{roux90}.
+    Signs are taken from experiments described in Supplementary Information.
+  }
+  \todo{More simulation data for lipids with different headgroups to be collected} \\
+  \todo{Some simulations contain potassium as counterions, while some sodium.
+    All experiments here contain some amount of sodium salt. The best
+    ion concentrations for comparison should be figured out.} \\
+  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
+  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
 \end{figure}
 
-The glycerol backbone order parameters for Slipids model are, however,
-far from experimental values, as previously observed also for PC 
-lipids \cite{botan15}. The glycerol backbone structures between PC and
+Glycerol backbone order parameters seems similar in all models,
+except in Slipids. Even thought glycerol backbone order parameter
+values are not yet experimentally available for PS lipids,
+the comparison with the results for PC lipids suggest that
+Slipid model do not correctly capture the glycerol backbone
+structure \cite{botan15}. The glycerol backbone structures between PC and
 PS lipids simulated with CHARMM36 are compared with the structures
@@
 %and forking in $\alpha$.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
-  \caption{\label{HGorderParametersPOPS}
-    Order parameters for POPS (top) and DOPS (bottom) headgroup and glycerol
-    backbone from simulations with different models and experiments without CaCl$_2$ 
-    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
-    Absolute values are shown for experimental data, because signs are not known.
-    Simulations values are -S$_{CH}$
-  }
-  \todo{More simulation data for lipids with different headgroups to be collected} \\
-  \todo{Some simulations contain potassium as counterions, while some sodium.
-    All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.} \\
-  \todo{Experimental signs of the order parameters would highly useful.} \\
-  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
-  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
-\end{figure}
 
@@
 
+\section{Measurements of order parameter sign}
+
+Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
+measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
+In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
+identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
+splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
+of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
+calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
+random collection of slices from noise such that it gets clear what is significant. The S-DROSS
+slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
+the higher order parameter is positive and suggests that the smaller order parameter is negative
+(from the deviation towards negative values in the longer t1 times).
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
+  \caption{\label{PShgSIGNS}
+    Experimental results for sign measurement for POPS sample
+  }
+\end{figure}
+
+The results updated with SIMPSON simulations for the SDROSS profiles
+are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
+alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
+because resolution in 13C NMR experiments was nor high enough to determine
+numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
+the following. The error bars and points are the experimental SDROSS data.
+The thick lines are SIMPSON simulations. The simulations were done by using
+the order parameter for beta equal to -0.12 and for alpha one order parameter
+equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
+Since the black lines agree with experimental data, we conlude that
+the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
+order parameters are 0.09 and -0.02.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
+  \caption{\label{PShgSIGNSsimpson}
+    Experimental results for sign measurement for POPS sample
+  }
+\end{figure}
+
 % Create the reference section using BibTe

### 2017-08-23 Manuscript/manuscript.tex
All simulation results (at least wihtout ions) are now reported with signs. Experimental signs are then set to give best agreement with simulations, when data is not available (PE and PG).

@@
 \subsection{PE headgroup}
-Absolute values of order parameters between Slipids simulations \cite{??}
+Order parameters from Slipids simulations 
 and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
-Glycerol bakcbone order parameters in Slipids are off from experiments,
-as systematically observed in this and previous studies \cite{??}.
-Alpha order parameter is too low, but beta shows apparent agreement
+Glycerol backbone order parameters in Slipids are off from experiments,
+as already observed previously for PC lipids \cite{botan15}.
+Order parameter signs for PE are not experimentally measured yet.
+For headgroup the signs are set to give best agreement with simulations
+and for glycerol to be consistent with experimental signs for PC.
+Order parameter for $\beta$ carbon shows apparent agreement
 with experiments. However, the sign of beta order parameter is positive,
@@
 parameter agrees with experiment with the assumption that its sign is opposite
-than for PC.
+than for PC. This is yet to be confirmed by experiments.
+Order parameter for $\alpha$ carbon is too close to zero, even if the sign
+would be correct.
+
 \begin{figure}[]
@@
 \subsection{PG headgroup}
-Comparison between experiments and simulations with CHARMM GUI
-parameters for PG lipids in Fig. \ref{HGorderParametersPOPG} suggest that
-the model overestimates forking in $\gamma$ and $\alpha$ order parameters,
-and magnitudes of $\beta$ and $g_1$ order parameters.
+Comparison between experiments and simulations for PG lipids is shown
+in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
+experimentally. They are set to give the best argeement with experiments.
+This would suggest that the $\beta$ order parameter would be positive,
+in contrast to PC and PS headgroups, were negative signs were measured.
+Even thought the signs turned out to be correct, the tested models would
+not give a very good argeement with the experiments.
 \begin{figure}[]
@@
   \caption{\label{HGorderParametersPOPG}
-    Order parameters for PG headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
+    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
     (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
     E.Coli PG results from \cite{gally81}).
-    Absolute values are shown, because signs are not known for experimental data.
-    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
-    smaller $\gamma$ are positive.
+    Signs are not known for experimental data. They are determined to give best agreement
+    with simulations. This is not reliable and should be corrected when experimental
+    data becomes available.
   }
-  \todo{More simulation data for lipids with different headgroups to be collected}
+  \todo{More simulation data for lipids with different headgroups to be collected} \\
   \todo{CHARMM GUI simulation contains only counter ions as potassium.
     All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.}
+    ion concentrations for comparison should be figured out.} \\
   \todo{Experimental signs of the order parameters would highly useful.}

### 2017-08-23 Manuscript/manuscript.tex
Add table with simulated systems. Not ready yet.

@@
 
+\section{Simulated systems}
+\begin{table*}[!p]
+%\begin{sidewaystable*}[!p]
+\centering
+\caption{List of MD simulations. The salt concentrations calculated as 
+   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+%   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+%   The lipid force fields named as in our previous work~\cite{botan15}.
+}\label{IONsystems}
+\begin{minipage}[t]{\textwidth}
+\begin{tabular}{l c c r r r r r r c c}
+  %\hline
+  % some footnotes are not visible in typeset-MS (pdf)
+ lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  \hline
+  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
+    \hline
+    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
+We need to decide the switching version or discuss this somehow.}  \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 			   		& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+\end{tabular}
+\end{minipage}
+%\end{sidewaystable*} 
+\end{table*}
+
 \section{Measurements of order parameter sign}

### 2017-08-30 Manuscript/manuscript.tex
Simulation without additional salt added to the table

--- 2017-08-23 Manuscript/manuscript.tex
@@
 
+\section{Simulated systems}
+\begin{table*}[!p]
+%\begin{sidewaystable*}[!p]
+\centering
+\caption{List of MD simulations. The salt concentrations calculated as 
+   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+%   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+%   The lipid force fields named as in our previous work~\cite{botan15}.
+}\label{IONsystems}
+\begin{minipage}[t]{\textwidth}
+\begin{tabular}{l c c r r r r r r c c}
+  %\hline
+  % some footnotes are not visible in typeset-MS (pdf)
+ lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  \hline
+  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
+    \hline
+    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
+We need to decide the switching version or discuss this somehow.}  \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 			   		& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    \hline
+    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    \hline
+    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
+    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
+    \hline
+    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
+    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
+    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
+    \hline
+    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
+    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+\end{tabular}
+\end{minipage}
+%\end{sidewaystable*} 
+\end{table*}
+
 \section{Measurements of order parameter sign}

### 2017-10-23 Manuscript/manuscript.tex
Update PS figure

@@
 \subsection{PS headgroup}
-Order parameters for bilayers with PS headgroups
+The headgroup order parameters of DOPS and POPS bilayers 
 from different experiments and simulations are shown in
-Figs. \ref{HGorderParametersDOPS} and  \ref{HGorderParametersPOPS}.
+Figs. \ref{HGorderParametersPS}.
 None of the tested models gives satisfactory agreement with
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersDOPS.eps}
-  \caption{\label{HGorderParametersDOPS}
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.eps}
+  \caption{\label{HGorderParametersPS}
     Order parameters for DOPS headgroup and glycerol
@@
   }
-\end{figure}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
-  \caption{\label{HGorderParametersPOPS}
-    Headgroup and glycerol backbone order parameters for systems with POPS 
-    from simulations with different models and experiments \cite{roux90}.
-    Signs are taken from experiments described in Supplementary Information.
-  }
-  \todo{More simulation data for lipids with different headgroups to be collected} \\
-  \todo{Some simulations contain potassium as counterions, while some sodium.
-    All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.} \\
-  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
-  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
+  \todo{Check and report all the counterions.}
+  \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
 \end{figure}
@@
   \todo{Experimental signs of the order parameters would highly useful.}
+\end{figure}
+
+\section{Lipid bilayers with mixed compositions}
+
+\subsection{PC/PS mixtures}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
+  \caption{\label{HGorderParametersPOPS}
+    Headgroup and glycerol backbone order parameters for systems with POPS 
+    from simulations with different models and experiments \cite{roux90}.
+    Signs are taken from experiments described in Supplementary Information.
+  }
+  \todo{Some simulations contain potassium as counterions, while some sodium.
+    All experiments here contain some amount of sodium salt. The best
+    ion concentrations for comparison should be figured out.} \\
+  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
+  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
 \end{figure}

### 2017-10-27 Manuscript/manuscript.tex
Add data and figures about PC, PE, PS and PG mixtures

@@
 
-\subsection{PC/PS mixtures}
-\begin{figure}[]
+\subsection{Effect of PE, PS and PG on PC headgroup}
+The headgroup order parameters for PC lipids (POPC and DOPC)
+mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
+from different simulation model and experiments \cite{scherer87} with different
+mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
+follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
+parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
+remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
+This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
+The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
+order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
+in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
+headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
+smaller than in experiments.
+\begin{figure*}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
-  \caption{\label{HGorderParametersPOPS}
-    Headgroup and glycerol backbone order parameters for systems with POPS 
-    from simulations with different models and experiments \cite{roux90}.
-    Signs are taken from experiments described in Supplementary Information.
+  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPSPG.eps}
+  \caption{\label{HGorderparametersPCvsPEPSPG}
+    PC headgroup order parameters from mixtures with PE, PS and PG
+    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
+    Signs are determined as discussed in \cite{botan15,ollila16}.
+  }
+  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
+\end{figure*}
+
+\subsection{Effect of PC on PS and PG headgroups}
+The headgroup order parameters for PS and PG lipid mixtures with PC
+having different mole fractions from simulations and experiments \cite{borle85,roux90}
+are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
+amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
+The $\beta$-carbon order parameter increases in experiment, but decreases
+in simulations with both tested counterions (Na+ and K+). Larger
+$\alpha$-carbon order parameter decreases with the addition of PC
+in experiment, while the lower remains unchanged. In simulations the
+larger increases and the lower decreases.
+Interestingly, the $\alpha$-carbon order parameters are closer to experiments
+in pure PS system with K+ counterions than with Na+. 
+The changes in PG headgroup order parameters are minor in simulations, which
+is in line with the only available experiment for the $\beta$-carbon.
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPSPGvsPC.eps}
+  \caption{\label{HGorderparametersPSPGvsPC}
+    PS and PG lipid headgroup order parameters from mixtures with PC
+    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
+    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
+    Signs for PS are measures as described in SI.
   }
@@
   pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
-\end{figure}
+\end{figure*}
 

### 2017-11-05 Manuscript/manuscript.tex
Update figures
Add lipids figure
Add dihedral plots by Buslaev into manuscript
Fix error in data due to mistake in taking data from scherer87.

@@
 %\subsection{Molecular dynamics simulations}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
+  \caption{\label{lipids}
+    Chemical structures and labels for the headgroup carbons.
+  }
+\end{figure}
 
@@
 smaller than in experiments.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
+  \caption{\label{HGorderparametersPCvsPEPSPGchol}
+    PC headgroup order parameters from experiments of mixtures with
+    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
+    Signs are determined as discussed in \cite{botan15,ollila16}.
+  }
+\end{figure}
+
 \begin{figure*}[]
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPG.eps}
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
   \caption{\label{changesWITHCaClPG}
@@
 
+\section{Dihedrals}
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
+  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
+  \caption{\label{dihedrals}
+    Experimental results for sign measurement for POPS sample
+  }
+\end{figure*}
+
 % Create the reference section using BibTe

### 2017-11-29 Manuscript/manuscriptPGPE.tex
Add separate manuscript for PS, and for PG and PE.

--- 2017-11-05 Manuscript/manuscript.tex
@@
 
-%Phospholipids containing various polar headgroups and acyl
-%chains are essential building blocks of biological membranes.
-%Atomistic level structural details of lipids and lipid-ion 
-%interactions are considered highly important in several
-%biological processes. The lipid structure and ion binding
-%can be studied in detail with NMR spectroscopy. However,
-%the structural interpretation of NMR data requires usage
-%of models. The combination of classical molecular dynamics
-%simulations with NMR data, especially with C-H bond
-%order parameters, can potentially give atomistic resolution
-%interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.
-
-%Our recent studies concluded that MD models are cabable to
-%give structural interpretation for phosphatidylcholine 
-%hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
-%backbone and cation binding posed a major challenge for current
-%force fields \cite{botan15,ollila16,catte16,ferreira16}. 
-%These conclusions were reached by reviewing extensively 
-%available experimental data from various sources and using
-%NMRlipids Open Collaboration project to collect massive 
-%amount of simulation data \cite{botan15,catte16}.
-
-%Here we apply the same approach in search of MD simulation
-%models which would reproduce glycerol backbone and 
-%headgroup structures of lipids with PE, PG and PS headgroups.  
-%In addition, we attempt to find a MD simulation model
-%that would be able to correctly describe cation binding in
-%bilayer containing negatively charged PG and PS lipids.
-
-%Several MD simulation models have succesfully described 
-%acyl chain structures and their qaulitative changes in different
-%conditions for phosphatidylcholine lipids (for review see \cite{ollila16}).
-%However, the current simulation models were not found to be accurate
-%enough for full structural interpretation of phosphatidylcholine headgroup and
-%glycerol backbone \cite{botan15}. Also Na$ +$  binding affinities
-%were found to be significantly overestimated in several models and 
-%none of the available models was able to Ca$2+$
-
-%\section{Methods}
-
-%\subsection{Molecular dynamics simulations}
-
 \begin{figure}[]
@@
 \end{figure}
-
-\subsection{PS headgroup}
-The headgroup order parameters of DOPS and POPS bilayers 
-from different experiments and simulations are shown in
-Figs. \ref{HGorderParametersPS}.
-None of the tested models gives satisfactory agreement with
-experiments for order parameters in headgroup $\alpha$ and $\beta$
-carbons.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.eps}
-  \caption{\label{HGorderParametersPS}
-    Order parameters for DOPS headgroup and glycerol
-    backbone from simulations with different models and experiments without CaCl$_2$ 
-    Experimental data from \cite{browning80} contains 0.1M of NaCl.
-    Signs are taken from experiments for POPS described in Supplementary Information.
-  }
-  \todo{Check and report all the counterions.}
-  \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
-\end{figure}
-
-Glycerol backbone order parameters seems similar in all models,
-except in Slipids. Even thought glycerol backbone order parameter
-values are not yet experimentally available for PS lipids,
-the comparison with the results for PC lipids suggest that
-Slipid model do not correctly capture the glycerol backbone
-structure \cite{botan15}. The glycerol backbone structures between PC and
-PS lipids simulated with CHARMM36 are compared with the structures
-simulated with CHARMM36 in Fig. \ref{glycerol_buslaev}.
-The differences in sampled conformation leading to the order parameter
-differences are clearly visible in the figure.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
-  \caption{\label{glycerol_buslaev}
-    Snapshots overlayed from different simulations for glycerol backbone region
-    by Pavel Buslaev.
-  }
-\end{figure}
-
-%Comparison between experiments and simulations with CHARMM GUI
-%parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
-%the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
-%and forking in $\alpha$.
 

### 2017-11-29 Manuscript/manuscriptPS.tex
Add separate manuscript for PS, and for PG and PE.

--- 2017-11-05 Manuscript/manuscript.tex
@@
 
-\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper
+\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids} %Title of paper
 
@@
 
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
+  \caption{\label{lipids}
+    Chemical structures and labels for the headgroup carbons.
+  }
+\end{figure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
+  \caption{\label{HGorderParameters}
+    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
+    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
+    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
+    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+  }
+\end{figure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
+  \caption{\label{HGorderparametersPCvsPEPSPGchol}
+    PC headgroup order parameters from experiments of mixtures with
+    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
+    Signs are determined as discussed in \cite{botan15,ollila16}.
+  }
+\end{figure}
+
+Absolute values of experimental order parameters for different lipid headgroups are
+collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
+thus only absolute values are used for now.
+
+Based on superficial reading, the conclusions in the literature are roughly \\ 
+1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
+2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
+3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
+Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
+In contrast to PC lipids (see \cite{botan15} and references therein).
+
 %Phospholipids containing various polar headgroups and acyl
@@
 
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
-  \caption{\label{lipids}
-    Chemical structures and labels for the headgroup carbons.
-  }
-\end{figure}
-
 \section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}
-
-Absolute values of experimental order parameters for different lipid headgroups are
-collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
-thus only absolute values are used for now.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
-  \caption{\label{HGorderParameters}
-    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
-    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
-    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
-    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
-  }
-\end{figure}
-
-Based on superficial reading, the conclusions in the literature are roughly \\ 
-1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
-2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
-3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
-Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
-In contrast to PC lipids (see \cite{botan15} and references therein).
 
@@
 the experiments (based on superficial reading of literature).
-
-\subsection{PE headgroup}
-Order parameters from Slipids simulations 
-and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
-Glycerol backbone order parameters in Slipids are off from experiments,
-as already observed previously for PC lipids \cite{botan15}.
-Order parameter signs for PE are not experimentally measured yet.
-For headgroup the signs are set to give best agreement with simulations
-and for glycerol to be consistent with experimental signs for PC.
-Order parameter for $\beta$ carbon shows apparent agreement
-with experiments. However, the sign of beta order parameter is positive,
-in contrast to PC where negative sign was measured. Thus, the the beta order
-parameter agrees with experiment with the assumption that its sign is opposite
-than for PC. This is yet to be confirmed by experiments.
-Order parameter for $\alpha$ carbon is too close to zero, even if the sign
-would be correct.
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.eps}
-  \caption{\label{HGorderParametersPE}
-    Order parameters for DPPE headgroup and glycerol
-    backbone from simulations with Slipids \cite{??} and experiments
-    (DPPE from \cite{seelig76} and E.coliPE from \cite{gally81}).
-    Absolute values are shown, because signs are not known experimentally.
-  }
-  \todo{Experimental signs of the order parameters would be highly useful.} \\
-\end{figure}
 
@@
 
-\subsection{PG headgroup}
-Comparison between experiments and simulations for PG lipids is shown
-in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
-experimentally. They are set to give the best argeement with experiments.
-This would suggest that the $\beta$ order parameter would be positive,
-in contrast to PC and PS headgroups, were negative signs were measured.
-Even thought the signs turned out to be correct, the tested models would
-not give a very good argeement with the experiments.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
-  \caption{\label{HGorderParametersPOPG}
-    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
-    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
-    E.Coli PG results from \cite{gally81}).
-    Signs are not known for experimental data. They are determined to give best agreement
-    with simulations. This is not reliable and should be corrected when experimental
-    data becomes available.
-  }
-  \todo{More simulation data for lipids with different headgroups to be collected} \\
-  \todo{CHARMM GUI simulation contains only counter ions as potassium.
-    All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.} \\
-  \todo{Experimental signs of the order parameters would highly useful.}
-\end{figure}
-
 \section{Lipid bilayers with mixed compositions}
@@
 smaller than in experiments.
+
 \begin{figure}[]
   \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
-  \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    PC headgroup order parameters from experiments of mixtures with
-    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
-  }
-\end{figure}
-
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPSPG.eps}
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
   \caption{\label{HGorderparametersPCvsPEPSPG}
@@
   \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
-\end{figure*}
+\end{figure}
 
@@
 is in line with the only available experiment for the $\beta$-carbon.
-\begin{figure*}[]
+\begin{figure}[]
   \centering
-  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPSPGvsPC.eps}
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
   \caption{\label{HGorderparametersPSPGvsPC}
@@
   pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
-\end{figure*}
+\end{figure}
 
@@
 negatively charged PS or PG lipids.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
-  \caption{\label{OrderParametersWithCaCl}
-    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
-    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
-    and POPC:POPG mixture data without NaCl from \cite{borle85}.
-  }
-  \todo{Check the NaCl concentrations in the samples.}
-\end{figure}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
-  \caption{\label{OrderParametersWithCaClBELOW1M}
-    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
-  }
-\end{figure}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
-  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    The change of PC headgroup order parameters
-    in the presence of different amount of negatively charged lipids
-    respect to the values without added CaCl$_2$.
-    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
-  }
-\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
+%  \caption{\label{OrderParametersWithCaCl}
+%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
+%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
+%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
+%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
+%  }
+%  \todo{Check the NaCl concentrations in the samples.}
+%\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
+%  \caption{\label{OrderParametersWithCaClBELOW1M}
+%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
+%  }
+%\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
+%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
+%    The change of PC headgroup order parameters
+%    in the presence of different amount of negatively charged lipids
+%    respect to the values without added CaCl$_2$.
+%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
+%  }
+%\end{figure}
 

### 2017-12-01 Manuscript/manuscriptPGPE.tex
Updates related to the separation of manuscripts

@@
 
-\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper
+\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids} %Title of paper
 
@@
 added salt is now collected and presented in this manuscript.
-Simulations for bilayers containing PG and PS with low salt conditions
-are ran with parameters from CHARMM GUI
-and compared to experiments. Some order parameters from
-simulations seems to be off from experimental values.
-Simulation data from other models would be highly useful 
-to see if some of the existing models would reproduce the 
-experimental order parameters and to analyze different 
-conformations predicted by different models respect to 
-experiments.
 
@@
 \end{figure}
-
-\section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}
 
@@
 
-\section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}
-
 Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
@@
 
-\subsection{PE headgroup}
+\section{Results and Discussion}
+
+\subsection{PE headgroup and glycerol backbone}
 Order parameters from Slipids simulations 
@@
 
-\subsection{PG headgroup}
+\subsection{PG headgroup and glycerol backbone}
 Comparison between experiments and simulations for PG lipids is shown
@@
 
-\section{Lipid bilayers with mixed compositions}
-
 \subsection{Effect of PE, PS and PG on PC headgroup}
@@
 smaller than in experiments.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
-  \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    PC headgroup order parameters from experiments of mixtures with
-    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
-  }
-\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
+%  \caption{\label{HGorderparametersPCvsPEPSPGchol}
+%    PC headgroup order parameters from experiments of mixtures with
+%    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
+%    Signs are determined as discussed in \cite{botan15,ollila16}.
+%  }
+%\end{figure}
 
@@
 
-\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}
+\subsection{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}
 
@@
 
-Also the experimental order parameters for PS and PG headgroups
-as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
-\todo{These should be compared to simulations for potential structural interpretation of the changes.}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
-  \caption{\label{PSPGchangesWITHCaCl}
-    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
-  }
-  \todo{Get the small concentration data from the inserts}
-\end{figure}
-
-\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids in simulations}
-
 Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
@@
 POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
-parameter of bound charge \todo{Should we check this against cationic surfactant experiments \cite{scherer89}?}.
+parameter of bound charge
+\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
+  I have already ran the simulations, analysis to be done.}.
 \begin{figure}[]
@@
 \todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
+
+\section{Effect of Ca$2+$ binding on PG and PE headgroup}
+
+Also the experimental order parameters for PS and PG headgroups
+as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
+\todo{These should be compared to simulations for potential structural interpretation of the changes.}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
+  \caption{\label{PSPGchangesWITHCaCl}
+    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
+  }
+  \todo{Get the small concentration data from the inserts}
+\end{figure}
 
@@
 
-\section{Measurements of order parameter sign}
+%\section{Measurements of order parameter sign}
 
-Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
-measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
-In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
-identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
-splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
-of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
-calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
-random collection of slices from noise such that it gets clear what is significant. The S-DROSS
-slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
-the higher order parameter is positive and suggests that the smaller order parameter is negative
-(from the deviation towards negative values in the longer t1 times).
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
-  \caption{\label{PShgSIGNS}
-    Experimental results for sign measurement for POPS sample
-  }
-\end{figure}
+%Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
+%measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
+%In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
+%identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
+%splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
+%of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
+%calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
+%random collection of slices from noise such that it gets clear what is significant. The S-DROSS
+%slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
+%the higher order parameter is positive and suggests that the smaller order parameter is negative
+%(from the deviation towards negative values in the longer t1 times).
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
+%  \caption{\label{PShgSIGNS}
+%    Experimental results for sign measurement for POPS sample
+%  }
+%\end{figure}
 
-The results updated with SIMPSON simulations for the SDROSS profiles
-are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
-alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
-because resolution in 13C NMR experiments was nor high enough to determine
-numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
-the following. The error bars and points are the experimental SDROSS data.
-The thick lines are SIMPSON simulations. The simulations were done by using
-the order parameter for beta equal to -0.12 and for alpha one order parameter
-equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
-Since the black lines agree with experimental data, we conlude that
-the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
-order parameters are 0.09 and -0.02.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
-  \caption{\label{PShgSIGNSsimpson}
-    Experimental results for sign measurement for POPS sample
-  }
-\end{figure}
+%The results updated with SIMPSON simulations for the SDROSS profiles
+%are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
+%alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
+%because resolution in 13C NMR experiments was nor high enough to determine
+%numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
+%the following. The error bars and points are the experimental SDROSS data.
+%The thick lines are SIMPSON simulations. The simulations were done by using
+%the order parameter for beta equal to -0.12 and for alpha one order parameter
+%equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
+%Since the black lines agree with experimental data, we conlude that
+%the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
+%order parameters are 0.09 and -0.02.
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
+%  \caption{\label{PShgSIGNSsimpson}
+%    Experimental results for sign measurement for POPS sample
+%  }
+%\end{figure}
 
-\section{Dihedrals}
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
-  \caption{\label{dihedrals}
-    Experimental results for sign measurement for POPS sample
-  }
-\end{figure*}
+%\section{Dihedrals}
+%\begin{figure*}[]
+%  \centering
+%  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
+%  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
+%  \caption{\label{dihedrals}
+%    Experimental results for sign measurement for POPS sample
+%  }
+%\end{figure*}
 

### 2017-12-01 Manuscript/manuscriptPS.tex
Updates related to the separation of manuscripts

@@
 backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
-Here we extend the same goal for other than PC lipids.
-Currently the focus is on PE, PG and PS bilayers and their
-mixtures with PC. Experimental data with different amounts of 
-added salt is now collected and presented in this manuscript.
-Simulations for bilayers containing PG and PS with low salt conditions
-are ran with parameters from CHARMM GUI
-and compared to experiments. Some order parameters from
-simulations seems to be off from experimental values.
-Simulation data from other models would be highly useful 
-to see if some of the existing models would reproduce the 
-experimental order parameters and to analyze different 
-conformations predicted by different models respect to 
-experiments.
-
+Here we extend the same goal for lipids with negatively charged PS headgroup.
+Chemical structure of PS headgroup together with other common biological
+lipids is shown in Fig. \ref{lipids}.
 \begin{figure}[]
@@
 
+Absolute values of experimental order parameters for different lipid headgroups are
+collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
+signs are known only for PC, only absolute values are shown.
+Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
+1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
+2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
+3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
+Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
+in contrast to PC lipids (see \cite{botan15} and references therein).
 \begin{figure}[]
@@
 
+As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
+headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
+i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
+surfactants. The changes with the addition of neutral lipids is significantly smaller.
 \begin{figure}[]
@@
 \end{figure}
-
-Absolute values of experimental order parameters for different lipid headgroups are
-collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
-thus only absolute values are used for now.
-
-Based on superficial reading, the conclusions in the literature are roughly \\ 
-1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
-2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
-3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
-Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
-In contrast to PC lipids (see \cite{botan15} and references therein).
 
@@
 
-\section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}
+\section{Results and Discussion}
 
-\section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}
-
-Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
-however, glycerol backbone and headgroup order parameters are not compared to
-the experiments (based on superficial reading of literature).
-
-\subsection{PS headgroup}
+\subsection{PS headgroup and glycerol backbone structure in simulations and experiments}
 The headgroup order parameters of DOPS and POPS bilayers 
@@
 
-%Comparison between experiments and simulations with CHARMM GUI
-%parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
-%the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
-%and forking in $\alpha$.
+\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be included in the discussion.}
 
-\section{Lipid bilayers with mixed compositions}
-
-\subsection{Effect of PE, PS and PG on PC headgroup}
-The headgroup order parameters for PC lipids (POPC and DOPC)
-mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
+\subsection{Effect of PS on PC headgroup}
+The headgroup order parameters of POPC 
+mixed with PS lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
 from different simulation model and experiments \cite{scherer87} with different
@@
 This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
-The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
-order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
-in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
-headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
-smaller than in experiments.
-
+%The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
+%order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
+%in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
+%headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
+%smaller than in experiments.
 \begin{figure}[]
@@
 
-\subsection{Effect of PC on PS and PG headgroups}
-The headgroup order parameters for PS and PG lipid mixtures with PC
-having different mole fractions from simulations and experiments \cite{borle85,roux90}
+\todo{More data to be collected before discussion.}
+
+\subsection{Effect of PC on PS headgroup}
+The headgroup order parameters of PS mixed with varying amounts of PC
+from simulations and experiments \cite{borle85,roux90}
 are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
@@
 in pure PS system with K+ counterions than with Na+. 
-The changes in PG headgroup order parameters are minor in simulations, which
-is in line with the only available experiment for the $\beta$-carbon.
+%The changes in PG headgroup order parameters are minor in simulations, which
+%is in line with the only available experiment for the $\beta$-carbon.
 \begin{figure}[]
@@
   \caption{\label{HGorderparametersPSPGvsPC}
-    PS and PG lipid headgroup order parameters from mixtures with PC
+    PS order parameters from mixtures with PC
     lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
-    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
     Signs for PS are measures as described in SI.
@@
 
-\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}
+\subsection{Ca$^{2+}$ binding affinity in bilayers with negatively charged PS lipids}
 
 PC lipid headgroup order parameters can used to measure ion binding
-affinity, because their magnitude is linearly proportional
+affinity, because their magnitude is proportional
 to the amount of bound charge in bilayer \cite{seelig87,catte16}.
-This molecular electrometer concept can be used also
+The molecular electrometer concept can be used also
 for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
-This is demonstrated in Figs \ref{OrderParametersWithCaCl},
-\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
-showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
+This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
+%\ref{OrderParametersWithCaClBELOW1M} and
+\ref{OrderParameterCHANGESWithCaClBELOW1M},
+showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
 as a function of CaCl$_2$ concentration in the presence of different amounts of
@@
 %\end{figure}
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
-%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-%    The change of PC headgroup order parameters
-%    in the presence of different amount of negatively charged lipids
-%    respect to the values without added CaCl$_2$.
-%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
-%  }
-%\end{figure}
-
-PC headgroup order parameters increase when negatively charged
-PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
-as expected based on electrometer concept \cite{seelig87}
-(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
+  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
+    The change of PC headgroup order parameters
+    in the presence of different amount of negatively charged lipids
+    respect to the values without added CaCl$_2$.
+    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
+  }
+\end{figure}
+%PC headgroup order parameters increase when negatively charged
+%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
+%as expected based on electrometer concept \cite{seelig87}
+%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
 %In electrometer concept this is explained by the tilting of
 %headgroup more parallel to membrane normal \cite{??}.
-Further, the order parameters decrease with the addition
-of CaCl$_2$ and the decrease is more pronounced for systems with more
-negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
-the Ca2+ binding presumably fully cancels the charge from negative lipids and
+The decrease of order parameters with CaCl$_2$
+is more pronounced for systems with more
+negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
+At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
 overcharging occurs above these concenterations.
@@
 
-Also the experimental order parameters for PS and PG headgroups
-as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
-\todo{These should be compared to simulations for potential structural interpretation of the changes.}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
-  \caption{\label{PSPGchangesWITHCaCl}
-    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
-  }
-  \todo{Get the small concentration data from the inserts}
-\end{figure}
-
-\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids in simulations}
-
+\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
+to other manuscript.}
 Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
@@
 POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
-parameter of bound charge \todo{Should we check this against cationic surfactant experiments \cite{scherer89}?}.
+parameter of bound charge
+\todo{The response of CHARMM36 against cationic surfactant experiments \cite{scherer89} is to be checked.
+  I have already ran the simulations, analysis yet to be done}.
 \begin{figure}[]
@@
 \todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
+
+\subsection{Effect of Ca$2+$ binding to PS headgroup}
+
+Also the experimental order parameters for PS and PG headgroups
+as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
+\todo{These should be compared to simulations for potential structural interpretation of the changes.}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
+  \caption{\label{PSPGchangesWITHCaCl}
+    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
+  }
+  \todo{Get the small concentration data from the inserts}
+\end{figure}
 

### 2017-12-08 Manuscript/manuscriptPS.tex
Update table with simulated systems

@@
     \hline
-    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-    \hline
-    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
-    \hline
-    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
-    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 110:22	   		& 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+%    \hline
+%    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
+%    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
+%    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
+%    \hline
+%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
+%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
 \end{tabular}

### 2017-12-08 Manuscript/manuscriptPS.tex
Add results from simulations with cationic surfactants

@@
 
+Before using headgroup order parameters to compare ion binding affinity between simulations
+and experiments, the response of the order parameters to bound charge has to be quantified.
+The response of headgroup order parameters to the fixed amount of cationic surfactants in
+POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
+The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
+while CHARMM36 is in better agreement with experiments. This has to be taken into account when
+analysin the binding affinities.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
+  \caption{\label{CHANGESwithCaClPGPS}
+  The response of headgroup order parameters to the fixed amount of cationic surfactants in
+  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+\end{figure}
+
 \todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
@@
 parameter of bound charge
-\todo{The response of CHARMM36 against cationic surfactant experiments \cite{scherer89} is to be checked.
-  I have already ran the simulations, analysis yet to be done}.
 \begin{figure}[]

### 2017-12-15 Manuscript/manuscriptPS.tex
Include the recent contributions by Markus Miettinen to the version containing only PS results

@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.eps}
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
   \caption{\label{HGorderParametersPS}
-    Order parameters for DOPS headgroup and glycerol
+    Order parameters for PS headgroup and glycerol
     backbone from simulations with different models and experiments without CaCl$_2$ 
@@
     Signs are taken from experiments for POPS described in Supplementary Information.
+    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
   }
   \todo{Check and report all the counterions.}
-  \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
+\end{figure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
+  \caption{\label{comparisonTablePS}
+  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
+Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity.
+Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red).
+The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
+For full details of the assessment, see Supplementary Information.
+  }
 \end{figure}
@@
 
+\pagebreak
+\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
+
+The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
+%
+First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
+we looked separately at deviations in magnitude and forking.
+
+{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
+the force-field-produced OPs were.
+%
+For each carbon, the following 5-step scale was used:
+%
+\begin{description}
+\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+%
+\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
+%
+\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
+%
+\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
+%
+\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
+\end{description}
+
+{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
+Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.
+For the $\alpha$ carbon,  for which a considerable forking of 0.105 is experimentally seen, the following 5-step scale was used:
+\begin{description}
+\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
+%
+\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.06 < D < 0.08)$ OR $(0.13 < D < 0.15)$.}
+%
+\item [2  ({\textsf{\small F}}):] \noindent {$(0.04 < D < 0.06)$ OR $(0.15 < D < 0.17)$.}
+%
+\item [3 ({\textsf{\large F}}):] \noindent {$(0.02 < D < 0.04)$ OR $(0.17 < D < 0.19)$.}
+%
+\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.02)$ OR $(0.19<D)$.}
+\end{description}
+%
+For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was used:
+%
+\begin{description}
+\item [0 (~):] \noindent {$ D< 0.02$.}
+%
+\item [1 ({\textsf{\tiny F}}):] \noindent {$0.02 < D < 0.04$.}
+%
+\item [2  ({\textsf{\small F}}):] \noindent {$0.04 < D < 0.06$.}
+%
+\item [3 ({\textsf{\large F}}):] \noindent {$0.06 < D < 0.08$.}
+%
+\item [4 ({\textsf{\Large F}}):] \noindent {$0.08 < D$.}
+\end{description}
+%
+For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was used:
+%
+\begin{description}
+\item [0 (~):] \noindent {$0.11 < D < 0.15$.}
+%
+\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.09 < D < 0.11)$ OR $(0.15 < D < 0.17)$.}
+%
+\item [2  ({\textsf{\small F}}):] \noindent {$(0.07 < D < 0.09)$ OR $(0.17 < D < 0.19)$.}
+%
+\item [3 ({\textsf{\large F}}):] \noindent {$(0.05 < D < 0.07)$ OR $(0.19 < D < 0.21)$.}
+%
+\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.05)$ OR $(0.21<D)$.}
+\end{description}
+
+Based on these assessments of magnitude and forking deviations,
+each carbon was then assigned to one of the following groups:
+"within experimental error"
+(magnitude and forking deviations both on step 0 of the scales described above),
+"almost within experimental error"
+(sum of the magnitude and forking deviation steps 1 or 2),
+"clear deviation from experiments"
+(sum of magnitude and forking deviation steps from 3 to 5), and
+"major deviation from experiments"
+(sum of magnitude and forking deviation steps from 6 to 8).
+These groups are indicated by colors in Fig.~4.
+(Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
+the corresponding group assigment limits were: 0, 1, 2, and 3.)
+
+Finally, the total ability of the force field to describe the headgroup and
+glycerol structure was estimated.
+To this end, the groups were given the following weights:
+0 (within experimental error),
+1 (almost within experimental error),
+2 (clear deviation from experiments),
+4 (major deviation from experiments),
+and the weights of the five carbons were summed up.
+The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
+was then used to (roughly and subjectively, as should be clear from the
+above description) rank the force fields.
+
+\newpage
+
 % Create the reference section using BibTe

### 2017-12-20 Manuscript/manuscriptPS.tex
Update figures and some text

@@
 
-\subsection{PS headgroup and glycerol backbone structure in simulations and experiments}
+\subsection{Headgroup and glycerol backbone structure in simulations and experiments of PS lipid bilayers}
 The headgroup order parameters of DOPS and POPS bilayers 
@@
 Glycerol backbone order parameters seems similar in all models,
-except in Slipids. Even thought glycerol backbone order parameter
-values are not yet experimentally available for PS lipids,
-the comparison with the results for PC lipids suggest that
+except in Slipids. 
 Slipid model do not correctly capture the glycerol backbone
-structure \cite{botan15}. The glycerol backbone structures between PC and
-PS lipids simulated with CHARMM36 are compared with the structures
+structure, as already observed previously for PC lipids \cite{botan15}.
+The glycerol backbone structures between PC and
+PS lipids simulated with Slipids are compared with the structures
 simulated with CHARMM36 in Fig. \ref{glycerol_buslaev}.
@@
 
-\subsection{Effect of PS on PC headgroup}
+\subsection{Headgroup structure in PS and PC mixtures}
 The headgroup order parameters of POPC 
@@
 %smaller than in experiments.
-\begin{figure}[]
+\begin{figure*}[]
   \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
+  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
   \caption{\label{HGorderparametersPCvsPEPSPG}
-    PC headgroup order parameters from mixtures with PE, PS and PG
-    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
+    Headgroup order parameters from PC:PS mixtures from 
+    different simulation models and experiments.
+    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
+    signs are determined as discussed in \cite{botan15,ollila16}).
+    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
+    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
+    Counterions in expriments are sodium, while potassium is used in simulations.
+    Using sodium in simulations do not have a significant effect.
   }
-  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
-\end{figure}
+  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
+  \todo{We need results also from other than CHARMM36 force field.}
+\end{figure*}
 
@@
 
-\subsection{Effect of PC on PS headgroup}
+%\subsection{Effect of PC on PS headgroup}
 The headgroup order parameters of PS mixed with varying amounts of PC
@@
 %is in line with the only available experiment for the $\beta$-carbon.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
-  \caption{\label{HGorderparametersPSPGvsPC}
-    PS order parameters from mixtures with PC
-    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
-    Signs for PS are measures as described in SI.
-  }
-  \todo{Some simulations contain potassium as counterions, while some sodium.
-    All experiments here contain some amount of sodium salt. The best
-    ion concentrations for comparison should be figured out.} \\
-  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
-  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
-\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
+%  \caption{\label{HGorderparametersPSPGvsPC}
+%    PS order parameters from mixtures with PC
+%    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
+%    Signs for PS are measures as described in SI.
+%  }
+%  \todo{Some simulations contain potassium as counterions, while some sodium.
+%    All experiments here contain some amount of sodium salt. The best
+%    ion concentrations for comparison should be figured out.} \\
+%\end{figure}
 

### 2017-12-28 Manuscript/manuscriptPS.tex
Add the data by M. Javanainen for MacRog POPS and update figures and manuscript
Updates and reorganization of the manuscript

@@
 
-Absolute values of experimental order parameters for different lipid headgroups are
-collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
-signs are known only for PC, only absolute values are shown.
-Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
-1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
-2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
-3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
-Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
-in contrast to PC lipids (see \cite{botan15} and references therein).
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
-  \caption{\label{HGorderParameters}
-    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
-    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
-    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
-    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
-  }
-\end{figure}
-
-As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
-headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
-i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
-surfactants. The changes with the addition of neutral lipids is significantly smaller.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
-  \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    PC headgroup order parameters from experiments of mixtures with
-    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
-  }
-\end{figure}
-
 %Phospholipids containing various polar headgroups and acyl
@@
 
-%\section{Methods}
+\section{Methods}
 
-%\subsection{Molecular dynamics simulations}
+\subsection{Solid state NMR experiments}
+The experimental protocol is the same used in Ref. \citenum{ferreira16}.
+
+\subsection{Molecular dynamics simulations}
+Molecular dynamics simulation data was collected with Open Collaboration method.
+The simulated systems are listed in Table \ref{IONsystems}.
+The simulation details are in SI or in references in the table.
+\begin{table*}[!htb]
+%\begin{sidewaystable*}[!p]
+\centering
+\caption{List of MD simulations. The salt concentrations calculated as 
+   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+%   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+%   The lipid force fields named as in our previous work~\cite{botan15}.
+}\label{IONsystems}
+%begin{minipage}[t]{\textwidth}
+\begin{tabular}{l c c r r r r r r c c}
+  %\hline
+  % some footnotes are not visible in typeset-MS (pdf)
+ lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  \hline
+  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
+    \hline
+    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
+We need to decide the switching version or discuss this somehow.}  \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 			   		& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    \hline
+    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
+    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    \hline
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 110:22	   		& 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+%    \hline
+%    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
+%    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
+%    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
+%    \hline
+%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
+%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+\end{tabular}
+%\end{minipage}
+%\end{sidewaystable*} 
+\end{table*}
 
 \section{Results and Discussion}
+
+\section{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}
+
+Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR} summarize the experimental NMR results for POPS
+bilayer sample.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=6.0cm]{../Figs/PShgSIGNSsimpson.pdf}
+  \caption{\label{PShgSIGNSsimpson}
+    (a) The headgroup region of the INEPT spectrum where alpha and beta are
+    identified.
+    (b) The R-PDLF slices for alpha and beta showing one single
+    splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
+    of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot
+    resolved with the available resolution.
+    (c) Points with error bars are the experimental SDROSS data.
+    The thick lines are SIMPSON simulations.
+    The S-DROSS slice for beta clearly shows that the order parameter is negative,
+    which is confirmed by SIMPSON simulations using the order parameter value of -0.12.
+    The S-DROSS slice for alpha suggests that the higher order parameter is positive
+    and the deviation towards negative values in the longer t1 times suggests
+    that the smaller order parameter is negative.
+    This is confirmed by SIMPSON simulation using value of 0.09
+    for the larger alpha order parameter and the value of -0.02 for
+    smaller (black curve). The value for the smaller
+    alpha order parameter for SIMPSON calculation was taken from Fig 3 in Ref. \citenum{roux91},
+    because resolution in 13C NMR experiments was nor high enough to determine
+    numerical value for this. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller order parameter gave did not agree with experiments (dashed grey),
+    confirming the interpretation that the smaller order parameter is negative.
+  }
+  \todo{Maybe we should combine this with \ref{glycerolNMR}}
+\end{figure}
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
+  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
+  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
+  \caption{\label{glycerolNMR}
+    R-PDFL slices for gycerol backbone carbons.
+  }
+  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
+  \todo{What are the top figures actually?}
+\end{figure}
+
+Absolute values of experimental order parameters for different lipid headgroups are
+collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
+signs are known only for PC, only absolute values are shown.
+Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
+1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
+2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
+3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
+Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
+in contrast to PC lipids (see \cite{botan15} and references therein).
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
+  \caption{\label{HGorderParameters}
+    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
+    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
+    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
+    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+  }
+\end{figure}
+
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
+%  \caption{\label{PShgSIGNS}
+%    Experimental results for sign measurement for POPS sample
+%  }
+%\end{figure}
 
@@
 parameter of bound charge
-\begin{figure}[]
+\begin{figure*}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
+  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
   \caption{\label{changesWITHCaClPG}
@@
   }
-\end{figure}
+\end{figure*}
 
@@
 
-\subsection{Effect of Ca$2+$ binding to PS headgroup}
+%\subsection{Effect of Ca$2+$ binding to PS headgroup}
 
@@
 \todo{These should be compared to simulations for potential structural interpretation of the changes.}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
-  \caption{\label{PSPGchangesWITHCaCl}
-    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
-  }
-  \todo{Get the small concentration data from the inserts}
-\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
+%  \caption{\label{PSPGchangesWITHCaCl}
+%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
+%  }
+%  \todo{Get the small concentration data from the inserts}
+%\end{figure}
 
@@
 \section{Simulated systems}
-\begin{table*}[!p]
-%\begin{sidewaystable*}[!p]
-\centering
-\caption{List of MD simulations. The salt concentrations calculated as 
-   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-%   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
-%   The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
-\begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
-  %\hline
-  % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
-  \hline
-  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
-    \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
-We need to decide the switching version or discuss this somehow.}  \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 			   		& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    \hline
-    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    \hline
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 110:22	   		& 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-%    \hline
-%    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-%    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-%    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
-%    \hline
-%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
-%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
-\end{tabular}
-\end{minipage}
-%\end{sidewaystable*} 
-\end{table*}
 
-\section{Measurements of order parameter sign}
-
-Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
-measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
-In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
-identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
-splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
-of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
-calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
-random collection of slices from noise such that it gets clear what is significant. The S-DROSS
-slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
-the higher order parameter is positive and suggests that the smaller order parameter is negative
-(from the deviation towards negative values in the longer t1 times).
+\section{PC lipid headgroup response to different mixtures in experiments}
+As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
+headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
+i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
+surfactants. The changes with the addition of neutral lipids is significantly smaller.
 \begin{figure}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
-  \caption{\label{PShgSIGNS}
-    Experimental results for sign measurement for POPS sample
-  }
-\end{figure}
-
-The results updated with SIMPSON simulations for the SDROSS profiles
-are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
-alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
-because resolution in 13C NMR experiments was nor high enough to determine
-numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
-the following. The error bars and points are the experimental SDROSS data.
-The thick lines are SIMPSON simulations. The simulations were done by using
-the order parameter for beta equal to -0.12 and for alpha one order parameter
-equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
-Since the black lines agree with experimental data, we conlude that
-the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
-order parameters are 0.09 and -0.02.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
-  \caption{\label{PShgSIGNSsimpson}
-    Experimental results for sign measurement for POPS sample
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
+  \caption{\label{HGorderparametersPCvsPEPSPGchol}
+    PC headgroup order parameters from experiments of mixtures with
+    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
+    Signs are determined as discussed in \cite{botan15,ollila16}.
   }

### 2018-01-01 Manuscript/manuscriptPS.tex
Organize and update the manuscript

@@
 \end{figure}
+
+Glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
 
@@
 
-\section{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}
+\subsection{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}
 
@@
 
-Absolute values of experimental order parameters for different lipid headgroups are
-collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
-signs are known only for PC, only absolute values are shown.
-Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
-1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
-2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
-3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
-Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
-in contrast to PC lipids (see \cite{botan15} and references therein).
+The headgroup and glycerol backbone order parameters of 
+POPS measured in this work are compared to the literature values of
+DOPS \cite{browning80} and POPC \cite{ferreira13} in Fig. \ref{HGorderParameters}. 
+Our results for POPS are in good agreement with the previously reported
+values for DOPS measured with $^2$H NMR. Significant differences are
+observed between PC and PS lipids, especially at the headgroup region.  
+Previous discussions in the literature have concluded that  
+glycerol backbone structure is largely similar in
+PC, PE, PG and PS lipids \cite{gally81}. The headgroup region was
+found to be similar in PC, PE and PG lipid (assuming that the signs of PE and PG
+order parameters are the same as in PC), while the PS headgroup was suggested
+to be more rigid \cite{wohlgemuth80,buldt81}. 
+The detailed structural differences between the headgroups is, however, not known.
 \begin{figure}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPS.eps}
   \caption{\label{HGorderParameters}
-    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
-    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
-    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
-    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
+    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
+    as measured in this work and signs for PC as measured in Refs \cite{??,ferreira16}.
+    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
+    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
   }
@@
 
-\subsection{Headgroup and glycerol backbone structure in simulations and experiments of PS lipid bilayers}
-The headgroup order parameters of DOPS and POPS bilayers 
-from different experiments and simulations are shown in
-Figs. \ref{HGorderParametersPS}.
-None of the tested models gives satisfactory agreement with
-experiments for order parameters in headgroup $\alpha$ and $\beta$
-carbons.
+\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers}
 \begin{figure}[]
@@
   }
-  \todo{Check and report all the counterions.}
+  \todo{Check and report all the counterions. There may be some simulation data 
+    sets with both Na+ and K+ ions?}
 \end{figure}
@@
   }
+  \todo{CHARMM36-UA point for alpha in wrong place?}
 \end{figure}
 
-Glycerol backbone order parameters seems similar in all models,
-except in Slipids. 
-Slipid model do not correctly capture the glycerol backbone
-structure, as already observed previously for PC lipids \cite{botan15}.
-The glycerol backbone structures between PC and
-PS lipids simulated with Slipids are compared with the structures
-simulated with CHARMM36 in Fig. \ref{glycerol_buslaev}.
-The differences in sampled conformation leading to the order parameter
-differences are clearly visible in the figure.
+The headgroup order parameters of DOPS and POPS bilayers 
+from different simulation models are compared with the experimental
+data in Fig. \ref{HGorderParametersPS}. In line with the previous
+study for PC lipids \cite{botan15}, the glycerol backbone order 
+parameters in CHARMM36 and Berger force fields roughly agree with
+the experimental data, but significant discrepancies are observed for
+Slipids and CKP models. The glycerol backbone structures from CHARMM36 and
+Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
+missing structures in Slipids model, which probably lead to the incorrect
+order parameters. For the headgroup $\alpha$ and $\beta$ carbon order parameters of PS lipids,
+the tested models perform less well than for PC headgroup in previous
+study \cite{botan15}. None of the tested models give a satisfactory 
+agreement with experiments for the headgroup order paramaters of 
+PS lipids. Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
+the CHARMM36 is the best performing model for both lipids.
+However, the total deviation from the experiments for the PS lipids is significantly
+higher for PS lipids (8) than for PC (3) \cite{botan15}. Therefore the 
+interpretation of structural differences between PC and PS lipids 
+from current MD simulation models is very challenging.
+
+\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be analyzed from different models 
+  and included in the discussion. Maybe also figures similar to \ref{glycerol_buslaev}.}
+
 \begin{figure}[]
@@
 
-\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be included in the discussion.}
+\subsection{Headgroup structure in PS and PC mixtures}
+Lipid mixtures containing PC and PS lipids are present in biological
+systems. Also the Ca2+ binding affinity to bilayers containing
+PS lipids is measured from mixtures with PC, because pure PS
+bilayers exhibit instant transition to the ordered phase with the
+addition of the ions. Therefore, the characterization of mutual interactions
+between PC and PS lipid headgroups is necessary in order to
+understand PS lipids in biological environments.
 
-\subsection{Headgroup structure in PS and PC mixtures}
-The headgroup order parameters of POPC 
-mixed with PS lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
-from different simulation model and experiments \cite{scherer87} with different
-mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
-follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
-parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
-remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
-This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
-%The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
-%order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
-%in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
-%headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
-%smaller than in experiments.
+Fig. \ref{HGorderparametersPCvsPS} shows the headgroup order parameters
+of POPC and POPS lipids in mixtures with various mole fractions from 
+simulations and experiments \cite{scherer87,roux90}.
+The headgroup order parameters of POPC increase with the increasing
+amount of negatively charged POPS, as expected from the 
+the electrometer concept \cite{seelig87}. Notably, the concept
+seems to be generally valid, at least qualitatively, also for 
+other lipid mixed with PC lipids; the PC lipid headgroup order
+parameters increase when mixed with negatively charged 
+lipids (PS, PI, CL, PA and PG) and remains almost unchaged when 
+mixed with neutral lipids (PE and SM) \cite{scherer87} or cholesterol \cite{ferreira13}.
+This is, however, not reproduced in the simulation data 
+in Fig. \ref{HGorderparametersPCvsPS}. The order parameter for
+$\beta$-carbon of POPC decreases and almost negligible increase is 
+seen in $\alpha$ order parameter with the addition of PS lipid in CHARMM36 simulation.
+This may indicate incorrect interactions between PC and PS headgroups
+in simulations. On the other hand, the previously reported overestimated
+ion binding affinity may also play a role, as the amount of counterions
+is increasing with PS.
+\todo{Data from other force fields to be collected before discussion can be finalized.}
+
 \begin{figure*}[]
@@
   \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
-  \caption{\label{HGorderparametersPCvsPEPSPG}
+  \caption{\label{HGorderparametersPCvsPS}
     Headgroup order parameters from PC:PS mixtures from 
@@
 
-\todo{More data to be collected before discussion.}
-
-%\subsection{Effect of PC on PS headgroup}
-The headgroup order parameters of PS mixed with varying amounts of PC
-from simulations and experiments \cite{borle85,roux90}
-are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
-amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
-The $\beta$-carbon order parameter increases in experiment, but decreases
-in simulations with both tested counterions (Na+ and K+). Larger
-$\alpha$-carbon order parameter decreases with the addition of PC
-in experiment, while the lower remains unchanged. In simulations the
-larger increases and the lower decreases.
-Interestingly, the $\alpha$-carbon order parameters are closer to experiments
-in pure PS system with K+ counterions than with Na+. 
-%The changes in PG headgroup order parameters are minor in simulations, which
-%is in line with the only available experiment for the $\beta$-carbon.
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
-%  \caption{\label{HGorderparametersPSPGvsPC}
-%    PS order parameters from mixtures with PC
-%    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
-%    Signs for PS are measures as described in SI.
-%  }
-%  \todo{Some simulations contain potassium as counterions, while some sodium.
-%    All experiments here contain some amount of sodium salt. The best
-%    ion concentrations for comparison should be figured out.} \\
-%\end{figure}
+Also the headgroup order parameters of POPS mixed with varying amounts of POPC
+from simulations and experiments \cite{borle85,roux90} are shown in Fig. \ref{HGorderparametersPCvsPS}.
+The $\beta$-carbon order parameter of POPS slightly decreases and positive 
+order parameter of $\alpha$-carbon slightly increases with increasing
+amount of PS lipids. This may indicate increasing order of the headgroup.
+It should be, however, noted that the experimental data for pure POPS and
+POPC/POPS mixture come from different experimental sets, 13C NMR in this work
+and 2H NMR from Ref. \citenum{roux90}, respectively. Therefore the accuracy
+of the order parameter change is not as high as typically in the measurements
+of order parameter changes, see discussion about qualitative and quantitative 
+accuracy in Ref. \citenum{ollila16}. The changes of PS headgroup order parameters
+are not reproduced by the tested simulation models. The $\beta$-carbon order parameter
+increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
+to the experimental data. The smaller $\alpha$-carbon order parameter increase in both
+simulation models with increasing amount of PS, while it is almost unchanged in experiments.
+The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
+with increasing amount of PS, both model exhibiting a poor agreement with experiments. 
+Signifincant improvement in the MD simulation models are needed to interpret the
+PS headgroup structures and mutual interactions with PC lipids.
 

### 2018-01-04 Manuscript/manuscriptPS.tex
Update table containing simulated systems by adding links to the simulation data delivered by Tom Piggot

@@
    [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+   CKP1 referes to the version with Berger/chiu NH3 charges compatible with Berger
+   and CKP2 to the version with more Gromos compatible version.
    %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
@@
     \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
-We need to decide the switching version or discuss this somehow.}  \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
     DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    &0 & 0        & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
     \hline
-    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
+    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
+    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
+    POPS/Na$^+$  & Berger \cite{??}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & ? \todoi{Equilibration?} & \cite{macrogPOPS298K} \\
+    POPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
     \hline
@@
     sets with both Na+ and K+ ions?}
+  \todo{CHARMM36-UA point for alpha in wrong place?}
+  \todo{MacRog POPS data missing. The equilibration yet to be checked, thought.}
+  \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with 
+    Berger/chiu NH3 charges compatible with Berger
+    and CKP2 to the version with more Gromos compatible version.'' Is this correct
+    also in this figure?}
 \end{figure}
@@
   }
-  \todo{CHARMM36-UA point for alpha in wrong place?}
 \end{figure}

### 2018-01-20 Manuscript/manuscriptPS.tex
Update manuscript

@@
 Glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
+
+PC lipid headgroup order parameters can used to measure ion binding
+affinity, because their magnitude is proportional
+to the amount of bound charge in bilayer \cite{seelig87,catte16}.
+The molecular electrometer concept can be used also
+for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
+This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
+%\ref{OrderParametersWithCaClBELOW1M} and
+\ref{OrderParameterCHANGESWithCaClBELOW1M},
+showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
+as a function of CaCl$_2$ concentration in the presence of different amounts of
+negatively charged PS or PG lipids.
+
+%PC headgroup order parameters increase when negatively charged
+%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
+%as expected based on electrometer concept \cite{seelig87}
+%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
+%In electrometer concept this is explained by the tilting of
+%headgroup more parallel to membrane normal \cite{??}.
+The decrease of order parameters with CaCl$_2$
+is more pronounced for systems with more
+negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
+At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
+overcharging occurs above these concenterations.
+The interpretation of this data and some other results has been that \cite{seelig90}
+\begin{displayquote}
+  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
+    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
+    (ii) the free Ca$^{2+}$
+    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
+    charge, either due to protein or to lipid; \\
+    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
+    of bound Ca$^{2+}$ at neutral and charged lipids; \\
+    (iv) the actual binding step can be described by a Langmuir
+    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
+    used to describe the chemical binding equilibrium.''}
+\end{displayquote}
 
@@
 
-PC lipid headgroup order parameters can used to measure ion binding
-affinity, because their magnitude is proportional
-to the amount of bound charge in bilayer \cite{seelig87,catte16}.
-The molecular electrometer concept can be used also
-for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
-This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
-%\ref{OrderParametersWithCaClBELOW1M} and
-\ref{OrderParameterCHANGESWithCaClBELOW1M},
-showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
-as a function of CaCl$_2$ concentration in the presence of different amounts of
-negatively charged PS or PG lipids.
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
-%  \caption{\label{OrderParametersWithCaCl}
-%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
-%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
-%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
-%  }
-%  \todo{Check the NaCl concentrations in the samples.}
-%\end{figure}
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
-%  \caption{\label{OrderParametersWithCaClBELOW1M}
-%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
-%  }
-%\end{figure}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
-  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    The change of PC headgroup order parameters
-    in the presence of different amount of negatively charged lipids
-    respect to the values without added CaCl$_2$.
-    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
-  }
-\end{figure}
-%PC headgroup order parameters increase when negatively charged
-%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
-%as expected based on electrometer concept \cite{seelig87}
-%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
-%In electrometer concept this is explained by the tilting of
-%headgroup more parallel to membrane normal \cite{??}.
-The decrease of order parameters with CaCl$_2$
-is more pronounced for systems with more
-negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
-At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
-overcharging occurs above these concenterations.
-The interpretation of this data and some other results has been that \cite{seelig90}
-\begin{displayquote}
-  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
-    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
-    (ii) the free Ca$^{2+}$
-    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
-    charge, either due to protein or to lipid; \\
-    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
-    of bound Ca$^{2+}$ at neutral and charged lipids; \\
-    (iv) the actual binding step can be described by a Langmuir
-    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
-    used to describe the chemical binding equilibrium.''}
-\end{displayquote}
-
 Before using headgroup order parameters to compare ion binding affinity between simulations
@@
     Signs are determined as discussed in \cite{botan15,ollila16}.
+  }
+\end{figure}
+
+\section{Cation binding affinity to lipid bilayers with different amount of charge}
+
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
+%  \caption{\label{OrderParametersWithCaCl}
+%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
+%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
+%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
+%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
+%  }
+%  \todo{Check the NaCl concentrations in the samples.}
+%\end{figure}
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
+%  \caption{\label{OrderParametersWithCaClBELOW1M}
+%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
+%  }
+%\end{figure}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
+  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
+    The change of PC headgroup order parameters
+    in the presence of different amount of negatively charged lipids
+    respect to the values without added CaCl$_2$.
+    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
+    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
+    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
+    and POPC:POPG mixture data without NaCl from \cite{borle85}.
   }

### 2018-01-22 Manuscript/manuscriptPS.tex
Update manuscript

@@
 \section{Introduction}
+Cellular membranes are composed of mixtures of zwitterionic
+and negatively charged lipids. Negatively charged lipids are important for
+the signalling and lateral organization of cellular membranes \cite{li4}.
+Direct interactions of cations and proteins with negatively charged
+lipids have been proposed to trigger signaling events \cite{li14,??}.
+On the other hand, it has been proposed that negatively charged lipids 
+only increase the local concentration of cations in the vicinity of
+the membrane and that the binding constant of cations is similar
+to zwitterionic and negatively charged lipids \cite{seelig90,??}.
 
-In NMRlipids I and II project we were looking for a MD model
-which would correctly reproduce headgroup and glycerol
-backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
-Here we extend the same goal for lipids with negatively charged PS headgroup.
-Chemical structure of PS headgroup together with other common biological
-lipids is shown in Fig. \ref{lipids}.
+Details of negatively charged lipids and their interactions
+with ions and proteins have been experimentally analyzed with
+various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
+%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
+which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.
+The structure of negatively charged lipids and their detailed
+interactions with ions or proteins have not, however, been fully resolved
+due to the difficulties to interpret the experimental data \cite{??} and
+the inaccuracies in simulation models at the
+headgroup region \cite{botan15,catte16,ollila16}.
+
+Headgroup and glycerol backbone C-H bond order parameters
+calculated from MD simulations have been
+recently used to interpret the lipid structures in NMR experiments
+and to validate lipid structure and ion binding in simulations of
+PC lipid bilayers \cite{botan15,catte16,ollila16,ferreira16}.
+In this work we apply the same approach to negatively charged
+phosphatidylserine lipids (PS) which compose   
+25-35\% of plasma membrane \cite{li14,??}, regulate
+surface charge and protein localization \cite{yeung08}, and
+induce protein aggregation \cite{zhao04,??}.
+
+%In NMRlipids I and II project we were looking for a MD model
+%which would correctly reproduce headgroup and glycerol
+%backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
+%Here we extend the same goal for lipids with negatively charged PS headgroup.
+%Chemical structure of PS headgroup together with other common biological
+%lipids is shown in Fig. \ref{lipids}.
 \begin{figure}[]
@@
 
-Glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
+Glycerol backbone and headgroup structure and behaviour are similar
+in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
 
@@
 The experimental protocol is the same used in Ref. \citenum{ferreira16}.
+\todo{Some basic details should be given.}
 
@@
  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
-  \hline
-  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 & 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
   \hline
@@
 Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR} summarize the experimental NMR results for POPS
-bilayer sample.
+bilayer sample. \todo{New figure, probably combining Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR}, should
+be done. This discussion to be finished after this.}
 

### 2018-01-24 Manuscript/manuscriptPS.tex
Add results of CaCl binding to PC:PS mixture delivered by Jesper Madsen http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1516642859204#c3440534109745026509

@@
     DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18}          &0 & 0        & ?    & ?   & ?  & ?  & ? & ? & \cite{??} \\
     \hline
@@
     POPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & lipid17 \cite{gould18}  &0 & 0  & ? & ? & ? & ? & ? & ? & ? \\
     \hline
@@
   \todo{Check and report all the counterions. There may be some simulation data 
-    sets with both Na+ and K+ ions?}
-  \todo{CHARMM36-UA point for alpha in wrong place?}
-  \todo{MacRog POPS data missing. The equilibration yet to be checked, thought.}
+    sets with both Na+ and K+ ions?} \\
+  \todo{CHARMM36-UA point for alpha in wrong place?} \\
+  \todo{MacRog POPS data missing. The equilibration yet to be checked, thought.} \\
   \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  
@@
     and CKP2 to the version with more Gromos compatible version.'' Is this correct
-    also in this figure?}
+    also in this figure?} \\
+  \todo{Lipid17 results should be added.} \\
+  \todo{We should also add CHARMM36 POPS ran with K counterions: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1\#issuecomment-308570874}
 \end{figure}

### 2018-01-24 Manuscript/manuscriptPS.tex
Update manuscript

@@
 \section{Introduction}
-Cellular membranes are composed of mixtures of zwitterionic
-and negatively charged lipids. Negatively charged lipids are important for
-the signalling and lateral organization of cellular membranes \cite{li4}.
-Direct interactions of cations and proteins with negatively charged
-lipids have been proposed to trigger signaling events \cite{li14,??}.
-On the other hand, it has been proposed that negatively charged lipids 
-only increase the local concentration of cations in the vicinity of
-the membrane and that the binding constant of cations is similar
-to zwitterionic and negatively charged lipids \cite{seelig90,??}.
+Phosphatidylserine (PS) is the most common negatively
+charged lipid in eykaryotic membranes.
+PS lipids compose 8.5\% of total lipid weight of erythrocytes,
+but the abundance varies between different organelles up to
+25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
+Despite of the relatively low abundance, PS lipids
+are important signaling molecules. They interact with
+signaling proteins \cite{leventis10}, regulate
+surface charge and protein localization \cite{yeung08}, and
+induce protein aggregation \cite{zhao04,??}.
+Some domains spesifically interact PS lipids,
+while others are attracted by general electrostatics and the
+binding can be regulated by calcium \cite{leventis10}.
+Therefore, it is crucial to resolve the structural details
+of PS headgroup and calcium binding in order to fully
+understand PS mediated signaling processes.
 
-Details of negatively charged lipids and their interactions
-with ions and proteins have been experimentally analyzed with
-various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
-%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
-which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.
-The structure of negatively charged lipids and their detailed
-interactions with ions or proteins have not, however, been fully resolved
+The structure of PS lipid headgroups and their 
+interactions with ions have been studied with
+various experimental methods and theoretical techniques \cite{roux90,melcrova16,??}.
+However, the consesus has not been reached 
 due to the difficulties to interpret the experimental data \cite{??} and
@@
 headgroup region \cite{botan15,catte16,ollila16}.
+Some studies propose specific binding of calcium directly to PS lipid
+headgroups \cite{??}, while others propose that only the local concentration
+of cations in the vicinity of negatively charged membranes containing PS lipids
+and that the binding constant of cations is similar
+to zwitterionic and negatively charged lipids \cite{seelig90,??}.
+The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
+headgroups, but more detailed interperetation has not been done.
+
+%Details of negatively charged lipids and their interactions
+%with ions and proteins have been experimentally analyzed with
+%various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
+%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
+%which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.
 
@@
 PC lipid bilayers \cite{botan15,catte16,ollila16,ferreira16}.
-In this work we apply the same approach to negatively charged
-phosphatidylserine lipids (PS) which compose   
-25-35\% of plasma membrane \cite{li14,??}, regulate
-surface charge and protein localization \cite{yeung08}, and
-induce protein aggregation \cite{zhao04,??}.
+In this work we apply this approach to PS lipid headgroup
+in order to elucidate the structural details and ion binding
+to negatively charged lipids. The results are expected to elucidate
+also PS mediated signalling events because 
+glycerol backbone and headgroup structure and behaviour are similar
+in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
 
@@
 \end{figure}
-
-Glycerol backbone and headgroup structure and behaviour are similar
-in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
 
@@
 
-\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
-to other manuscript.}
-Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
-is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
-is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
-The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
-It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
-but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
-Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
-suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
-POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
-parameter of bound charge
+The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
+as a function of Ca$2+$ concentration from different
+simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPG}.
+As shown previously, the changes of PC headgroup order parameters can be
+used to measure the amount of bound ions to the bilayer according to the
+electrometer concept \cite{akutsu81,seelig87,catte16}.
+This would suggest that the Ca2+ binding is overestimated in
+MacRog simulation and underestimated in CHARMM36 simulation.
+It should be however noted, that MacRog simulation with low
+concetrations gives a good agreemet with experiments.
+The result from CHARMM simulation is somewhat surprising because
+previous study suggested that Ca2+ ions overbind to pure PC lipid
+bilayers \cite{catte16}.
+\todo{The role of counterions and equilibration should be somehow checked
+and discussed.}
 \begin{figure*}[]
@@
   \caption{\label{changesWITHCaClPG}
-    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
-    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
+    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
   }
@@
 
-\todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}
+\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}
 
@@
 
-Also the experimental order parameters for PS and PG headgroups
+Also the experimental order parameters of POPS headgroup 
 as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
+The order parameter changes with added CaCl$_2$ are overestimated in
+both simulation models. However, the changes can be considered qualitatively
+correct for both segments in CHARMM36 model and in $\beta$ segment in MacRog model. 
 \todo{These should be compared to simulations for potential structural interpretation of the changes.}
@@
 \end{acknowledgments}
-\newpage
+\pagebreak
 \appendix

### 2018-01-25 Manuscript/manuscriptPS.tex
Update density profiles

@@
 
-\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers}
+\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
 \begin{figure}[]
@@
     by Pavel Buslaev.
+  }
+\end{figure}
+
+In contrast to PC lipids discussed previously \cite{botan15,catte16},
+systems with negatively charged lipids always contains counterions.
+Since the ion binding changes the headgroup order parameters \cite{seelig87,catte16},
+the counterion distribution have to be taken into account also
+when additional ions are not present.
+The Na+ density profiles from different simulation are shown in Fig. \ref{NAdensPOPS}.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
+  \caption{\label{NAdensPOPS}
+    Sodium density of POPS lipid bilayer along the membrane normal from
+    simulations with different force fields.
   }

### 2018-02-01 Manuscript/manuscriptPS.tex
Add Ca density plots and some updates

@@
 It should be however noted, that MacRog simulation with low
-concetrations gives a good agreemet with experiments.
+concentrations gives a good agreemet with experiments.
 The result from CHARMM simulation is somewhat surprising because
@@
   and discussed.}
+The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
+Overall, it seems that calcium clearly overbinds in MacRog simulations
+as expected based on previous results \cite{catte16}.
+However, the calcium binding in CHARMM36 simulations seems to
+surprisingly low. Thi
 \begin{figure*}[]
@@
 \end{figure*}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
+  \caption{\label{CAdensPCPSmixture}
+    Ca2+ density profiles from simulations.
+  }
+  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
+  \todo{These are now mass densities, numbers would be probably better.}
+  \todo{Counterions should be also included.}
+  \todo{Not all the data from MacRog is included.}
+\end{figure}
 

### 2018-02-19 Manuscript/manuscriptPS.tex
Update manuscript

@@
 binding can be regulated by calcium \cite{leventis10}.
-Therefore, it is crucial to resolve the structural details
-of PS headgroup and calcium binding in order to fully
-understand PS mediated signaling processes.
+Therefore, the structural details
+of lipid headgroups and the details of calcium binding
+are crucial for the PS mediated signaling processes.
 
@@
 
-PC lipid headgroup order parameters can used to measure ion binding
-affinity, because their magnitude is proportional
-to the amount of bound charge in bilayer \cite{seelig87,catte16}.
-The molecular electrometer concept can be used also
-for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
-This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
-%\ref{OrderParametersWithCaClBELOW1M} and
-\ref{OrderParameterCHANGESWithCaClBELOW1M},
-showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
-as a function of CaCl$_2$ concentration in the presence of different amounts of
-negatively charged PS or PG lipids.
-
-%PC headgroup order parameters increase when negatively charged
-%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
-%as expected based on electrometer concept \cite{seelig87}
-%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
-%In electrometer concept this is explained by the tilting of
-%headgroup more parallel to membrane normal \cite{??}.
-The decrease of order parameters with CaCl$_2$
-is more pronounced for systems with more
-negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
-At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
-overcharging occurs above these concenterations.
-The interpretation of this data and some other results has been that \cite{seelig90}
-\begin{displayquote}
-  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
-    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
-    (ii) the free Ca$^{2+}$
-    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
-    charge, either due to protein or to lipid; \\
-    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
-    of bound Ca$^{2+}$ at neutral and charged lipids; \\
-    (iv) the actual binding step can be described by a Langmuir
-    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
-    used to describe the chemical binding equilibrium.''}
-\end{displayquote}
+%The interpretation of this data and some other results has been that \cite{seelig90}
+%\begin{displayquote}
+%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
+%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
+%    (ii) the free Ca$^{2+}$
+%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
+%    charge, either due to protein or to lipid; \\
+%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
+%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
+%    (iv) the actual binding step can be described by a Langmuir
+%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
+%    used to describe the chemical binding equilibrium.''}
+%\end{displayquote}
 
@@
 
-Before using headgroup order parameters to compare ion binding affinity between simulations
-and experiments, the response of the order parameters to bound charge has to be quantified.
+The headgroup order parameters of PC lipids can be used to measure ion binding
+affinity because they change proportionally
+to the amount of bound charge in bilayer \cite{seelig87,catte16}.
+This molecular electrometer concept can be applied also to
+lipid bilayers with mixtures of PC and negatively charged
+lipids \cite{borle85,macdonald87,roux90}.
+This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
+%\ref{OrderParametersWithCaClBELOW1M} and
+\ref{OrderParameterCHANGESWithCaClBELOW1M}
+using the previuosly reported experimental data
+for mixtures with varying proportions of
+negatively charged PS or PG lipids.
+%
+%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
+%as a function of CaCl$_2$ concentration are shown
+%
+As expected, the decrease of order parameters with CaCl$_2$
+is more pronounced for systems with more
+negatively charged lipids.
+% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+
+%PC headgroup order parameters increase when negatively charged
+%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
+%as expected based on electrometer concept \cite{seelig87}
+%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
+%In electrometer concept this is explained by the tilting of
+%headgroup more parallel to membrane normal \cite{??}.
+
+%Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
+%At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
+%overcharging occurs above these concenterations.
+
+Before using the headgroup order parameters to compare ion binding affinity between simulations
+and experiments, it is important to quantify the response of the order parameters to the
+bound charge in simulations.
 The response of headgroup order parameters to the fixed amount of cationic surfactants in

### 2018-02-22 Manuscript/manuscriptPS.tex
Updates in the manuscript and density plot

@@
 surface charge and protein localization \cite{yeung08}, and
-induce protein aggregation \cite{zhao04,??}.
+induce protein aggregation \cite{zhao04,gorbenko06}.
 Some domains spesifically interact PS lipids,
@@
 headgroup region \cite{botan15,catte16,ollila16}.
-Some studies propose specific binding of calcium directly to PS lipid
-headgroups \cite{??}, while others propose that only the local concentration
-of cations in the vicinity of negatively charged membranes containing PS lipids
-and that the binding constant of cations is similar
-to zwitterionic and negatively charged lipids \cite{seelig90,??}.
+Some studies propose that the negatively charged lipids
+attract cations only due to the increase of local concentration
+in the vicinity of membranes and that the binding constant of cations is similar
+to zwitterionic and negatively charged lipids \cite{seelig90,sinn06,??}.
+On the other hand, some studies propose specific binding of calcium directly to PS lipid
+headgroups \cite{vernier09,boettcher11,??},
 The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
@@
     POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
+    POPS/K$^+$   & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
     POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
@@
     POPS/Na$^+$  & Berger \cite{??}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & ? \todoi{Equilibration?} & \cite{macrogPOPS298K} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
     POPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
@@
   }
-  \todo{Check and report all the counterions. There may be some simulation data 
-    sets with both Na+ and K+ ions?} \\
   \todo{CHARMM36-UA point for alpha in wrong place?} \\
-  \todo{MacRog POPS data missing. The equilibration yet to be checked, thought.} \\
+  \todo{MacRog POPS data should be added.} \\
   \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  

### 2018-02-22 Manuscript/manuscriptPS.tex
Updates in the manuscript and density plot
Add new dihedrals to manuscript

@@
   }
+  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4}
 \end{figure}
@@
 study for PC lipids \cite{botan15}, the glycerol backbone order 
-parameters in CHARMM36 and Berger force fields roughly agree with
-the experimental data, but significant discrepancies are observed for
-Slipids and CKP models. The glycerol backbone structures from CHARMM36 and
+parameters in CHARMM36 roughly agree with the experimental data,
+while significant discrepancies are observed in other models at
+least for one glycerol backbone carbons.
+\todo{Discussion to be finished once all the results are in the plot.}
+
+The glycerol backbone structures from CHARMM36 and
 Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
-  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
-  \caption{\label{dihedrals}
-    Experimental results for sign measurement for POPS sample
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops1.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops2.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops3.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops4.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops5.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops6.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops7.png}
+  \caption{\label{dihedralsGLY}
+    Dihedral angle distributions of bonds from phosphate to acyl chain carbonyls from different simulation models.
+  }
+\end{figure*}
+\section{Dihedrals}
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops8.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops9.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops10.png}
+  \includegraphics[width=8.0cm]{../Figs/diheds_pops11.png}
+  \caption{\label{dihedralsHG}
+    Dihedral angle distributions of bonds from phosphate to headgroup from different simulation models.
   }

### 2018-02-22 Manuscript/manuscriptPS.tex
Updates to text

@@
 parameters in CHARMM36 roughly agree with the experimental data,
-while significant discrepancies are observed in other models at
-least for one glycerol backbone carbons.
+while significant discrepancies for glycerol backbone carbons
+are observed in other models.
 \todo{Discussion to be finished once all the results are in the plot.}
 
+Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
+the CHARMM36 is the best performing model for both lipids.
+However, none of the tested models give a satisfactory 
+agreement with experiments for the headgroup order paramaters of 
+PS lipids. The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
+for the best performing CHARMM36 model is larger for PS lipids (8)
+than for PC (3) \cite{botan15}. Therefore the 
+interpretation of structural details of PS headgroup or
+differences between PC and PS lipids is very challenging 
+from the current MD simulation models.
+Figures \ref{dihedralsGLY} and \ref{dihedralsHG} show dihedral
+angle distributions calculated from different models. 
 The glycerol backbone structures from CHARMM36 and
@@
 the tested models perform less well than for PC headgroup in previous
-study \cite{botan15}. None of the tested models give a satisfactory 
-agreement with experiments for the headgroup order paramaters of 
-PS lipids. Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
-the CHARMM36 is the best performing model for both lipids.
-However, the total deviation from the experiments for the PS lipids is significantly
-higher for PS lipids (8) than for PC (3) \cite{botan15}. Therefore the 
-interpretation of structural differences between PC and PS lipids 
-from current MD simulation models is very challenging.
-
-\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be analyzed from different models 
-  and included in the discussion. Maybe also figures similar to \ref{glycerol_buslaev}.}
+study \cite{botan15}.
+\todo{Discussion will be finished when we have all the data in
+  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}}
 
@@
 
-In contrast to PC lipids discussed previously \cite{botan15,catte16},
-systems with negatively charged lipids always contains counterions.
-Since the ion binding changes the headgroup order parameters \cite{seelig87,catte16},
-the counterion distribution have to be taken into account also
-when additional ions are not present.
-The Na+ density profiles from different simulation are shown in Fig. \ref{NAdensPOPS}.
+Since systems with negatively charged PS lipids always contain counterions
+and the ion binding presumably affects the PS headgroup
+order parameters, the discussion of ion binding affinity and
+headgroup structure cannot be separated as done previously
+for PC lipids \cite{botan15,catte16}. The density profiles
+of counterions along membrane normal from different simulations
+are shown in Fig. \ref{NAdensPOPS}.
+Significant differences are observed in counterion
+densities between different simulation models. 
+\todo{We should try to figure out which one of these are more
+  realistic. One option could be to use PS headgroup order parameter
+  data measured as a function of NaCl \cite{roux86}. Another
+  approach could be to use the area per molecule, which is suggested to
+  depend on the binding affinity of counterions \cite{pandit02,mukhopadhyay04,pedersen06}. The
+  experimental value for area per molecule is available at \cite{pan14}.}
 \begin{figure}[]
@@
   \caption{\label{NAdensPOPS}
-    Sodium density of POPS lipid bilayer along the membrane normal from
+    Counterion densities of POPS lipid bilayer along the membrane normal from
     simulations with different force fields.

### 2018-03-02 Manuscript/manuscriptPS.tex
Updates to manuscript

@@
 \todo{Discussion will be finished when we have all the data in
-  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}}
+  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}.
+  One possible conclusion could be the following:
+  The main differences between the models in the headgroup region are observed
+  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
+  CHARMM36, CHARMM36UA and Slipids give very similar results to the
+  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
+  The order parameters of $\beta$-carbons for these three models
+  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
+  On the other hand, Gromos-CKP models give better order parameters
+  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
+  In conclusion, the suggestion would be that the single peak for
+  observed at 120 degrees in CHARMMs and Slipids would
+  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
+  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
+}
 
@@
 
-The headgroup order parameters of PC lipids can be used to measure ion binding
-affinity because they change proportionally
-to the amount of bound charge in bilayer \cite{seelig87,catte16}.
-This molecular electrometer concept can be applied also to
-lipid bilayers with mixtures of PC and negatively charged
-lipids \cite{borle85,macdonald87,roux90}.
-This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
+The headgroup order parameters of PC lipids
+decrease proportionally to the bound positive
+charge in to a bilayer \cite{seelig87,catte16} and can be
+therefore used to measure the ion binding
+affinity. This molecular electrometer concept can
+be also applied to lipid bilayers with mixtures
+of PC and negatively charged lipids \cite{borle85,macdonald87,roux90}
+(see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+%This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
 %\ref{OrderParametersWithCaClBELOW1M} and
-\ref{OrderParameterCHANGESWithCaClBELOW1M}
-using the previuosly reported experimental data
-for mixtures with varying proportions of
-negatively charged PS or PG lipids.
+%using the previuosly reported experimental data
+%for mixtures with varying proportions of
+%negatively charged PS or PG lipids.
 %
@@
 %
-As expected, the decrease of order parameters with CaCl$_2$
-is more pronounced for systems with more
-negatively charged lipids.
 % (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
@@
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    The change of PC headgroup order parameters
-    in the presence of different amount of negatively charged lipids
-    respect to the values without added CaCl$_2$.
-    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
-    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
-    and POPC:POPG mixture data without NaCl from \cite{borle85}.
+    The change of PC headgroup order parameters as a function of CaCl$_2$
+    measured from bilayers containing different amount of negatively charged lipids.
+    The values are taken from 2H NMR experiments reported in the
+    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
+    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
+    and POPC:POPG (1:1) without NaCl \cite{borle85}).
+    As expected, the decrease of order parameters with the added CaCl$_2$
+    is more pronounced for systems with larger fraction of
+    negatively charged lipids, indicating larger amount amount
+    of bound cations.
   }

### 2018-03-06 Manuscript/manuscriptOLD.tex

--- 2017-11-05 Manuscript/manuscript.tex
@@
     Signs are taken from experiments for POPS described in Supplementary Information.
+    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
   }
@@
   \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
+\end{figure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
+  \caption{\label{comparisonTablePS}
+  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
+Here “M” indicates a magnitude problem, “F” a forking problem; letter size increases with problem severity.
+Color scheme: “within experimental error” (dark green), “almost within experimental error” (light green), “clear deviation from experiments” (light red), and “major deviation from experiments” (dark red).
+The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
+For full details of the assessment, see Supplementary Information.
+  }
 \end{figure}

### 2018-03-06 Manuscript/manuscriptPS.tex
Added MacRog POPS data to Fig 5.

@@
     Signs are taken from experiments for POPS described in Supplementary Information.
-    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
+    The vertical bars shown for most computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
   }
-  \todo{CHARMM36-UA point for alpha in wrong place?} \\
-  \todo{MacRog POPS data should be added.} \\
   \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  

### 2018-03-06 Manuscript/manuscriptPS.tex
Update manuscript

--- 2018-03-02 Manuscript/manuscriptPS.tex
@@
 \subsection{Headgroup structure in PS and PC mixtures}
-Lipid mixtures containing PC and PS lipids are present in biological
-systems. Also the Ca2+ binding affinity to bilayers containing
-PS lipids is measured from mixtures with PC, because pure PS
-bilayers exhibit instant transition to the ordered phase with the
-addition of the ions. Therefore, the characterization of mutual interactions
-between PC and PS lipid headgroups is necessary in order to
-understand PS lipids in biological environments.
+Cellular membranes often compose of mixtures of zwitterionic and negatively
+charged lipids. The response of PC headgroup to mixtures of differently charged
+lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
+As expected from the electrometer concept \cite{seelig87}, the headgroup 
+order parameters increase with the addition of negatively charged PS and PG lipids,
+decrease when mixed with positively charged surfactants and
+are less affected by the addition of zwitterionic PE lipids or cholesterol.
+In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
+also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
+follow the eletrometer concept \cite{scherer87}.
 
-Fig. \ref{HGorderparametersPCvsPS} shows the headgroup order parameters
-of POPC and POPS lipids in mixtures with various mole fractions from 
+%Also the Ca2+ binding affinity to bilayers containing
+%PS lipids is measured from mixtures with PC, because pure PS
+%bilayers exhibit instant transition to the ordered phase with the
+%addition of the ions.
+%Therefore, the characterization of mutual interactions
+%between PC and PS lipid headgroups is necessary in order to
+%understand PS lipids in biological environments.
+
+Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
+from POPC:POPS mixtures with various mole fractions from 
 simulations and experiments \cite{scherer87,roux90}.
-The headgroup order parameters of POPC increase with the increasing
-amount of negatively charged POPS, as expected from the 
-the electrometer concept \cite{seelig87}. Notably, the concept
-seems to be generally valid, at least qualitatively, also for 
-other lipid mixed with PC lipids; the PC lipid headgroup order
-parameters increase when mixed with negatively charged 
-lipids (PS, PI, CL, PA and PG) and remains almost unchaged when 
-mixed with neutral lipids (PE and SM) \cite{scherer87} or cholesterol \cite{ferreira13}.
-This is, however, not reproduced in the simulation data 
-in Fig. \ref{HGorderparametersPCvsPS}. The order parameter for
-$\beta$-carbon of POPC decreases and almost negligible increase is 
-seen in $\alpha$ order parameter with the addition of PS lipid in CHARMM36 simulation.
-This may indicate incorrect interactions between PC and PS headgroups
-in simulations. On the other hand, the previously reported overestimated
-ion binding affinity may also play a role, as the amount of counterions
-is increasing with PS.
-\todo{Data from other force fields to be collected before discussion can be finalized.}
+The experimentally observed increase of PC headgroup order parameters with the increasing
+amount of negatively charged PS lipid is reproduced in
+MacRog simulations with potassium counterions, but not in CHARMM36 simulations
+with potassium or sodium conterions. The results can be
+understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
+Both counterions are more strongly bound to bilayers in
+CHARMM36 simulations than potassium in MacRog simulation.
+The cation binding probably neutralizes the effect of negatively
+charged lipids in CHARMM36 simulations, while the
+weaker binding of potassium in MacRog simulations enables the
+increase of order parameters with the added amount of PS.
+
+%$\beta$-carbon of POPC decreases and almost negligible increase is 
+%seen in $\alpha$ order parameter with the addition of PS lipid 
+%This may indicate incorrect interactions between PC and PS headgroups
+%in simulations. On the other hand, the previously reported overestimated
+%ion binding affinity may also play a role, as the amount of counterions
+%is increasing with PS.
+%\todo{Data from other force fields to be collected before discussion can be finalized.}
 
@@
 \end{figure*}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
+  \caption{  Counterion density distributions from PC:PS mixtures.
+\label{CIdensPSOCmixt}
+  }
+\end{figure}
 
@@
 
-Before using the headgroup order parameters to compare ion binding affinity between simulations
-and experiments, it is important to quantify the response of the order parameters to the
-bound charge in simulations.
-The response of headgroup order parameters to the fixed amount of cationic surfactants in
-POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
-The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
-while CHARMM36 is in better agreement with experiments. This has to be taken into account when
-analysin the binding affinities.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
-  \caption{\label{CHANGESwithCaClPGPS}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
-\end{figure}
-
 The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
 as a function of Ca$2+$ concentration from different
-simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPG}.
-As shown previously, the changes of PC headgroup order parameters can be
-used to measure the amount of bound ions to the bilayer according to the
-electrometer concept \cite{akutsu81,seelig87,catte16}.
-This would suggest that the Ca2+ binding is overestimated in
-MacRog simulation and underestimated in CHARMM36 simulation.
-It should be however noted, that MacRog simulation with low
-concentrations gives a good agreemet with experiments.
-The result from CHARMM simulation is somewhat surprising because
-previous study suggested that Ca2+ ions overbind to pure PC lipid
-bilayers \cite{catte16}.
-\todo{The role of counterions and equilibration should be somehow checked
-  and discussed.}
+simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPS}.
+The results suggest that Ca2+ ions clearly overbind in simulations with
+MacRog model, as expected from previous results for PC lipid bilayers \cite{catte16}.
+It should be noted, however, that the lowest concentration (100mM) gives
+a good agreemet with experiments \todo{Should be analyze/discuss this further?}.
+Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations
+\todo{This is most likely due to the new interactions added to CHARMM-GUI:
+  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1519913688317\#c5312059274016196300
+}.
 The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
-Overall, it seems that calcium clearly overbinds in MacRog simulations
-as expected based on previous results \cite{catte16}.
-However, the calcium binding in CHARMM36 simulations seems to
-surprisingly low. Thi
+\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}
+
 \begin{figure*}[]
@@
   \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
-  \caption{\label{changesWITHCaClPG}
+  \caption{\label{changesWITHCaClPS}
     POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
@@
 
-Also dependence of $\beta$-carbon of PG on CaCl$_2$ concentration is compared with
-experiments \cite{borle85} in Fig. \ref{PSPGchangesWITHCaCl}. Absolute value of
-the order parameter is too large without ions, but rapid decrease due to addition of
-CaCl$_2$ is observed in agreement with experiments for systems with 1:1 mixture of POPC and POPG.
-In addition, absolute value in systems with CaCl$_2$ is in agreement with experiments.
-However, system with 4:1 mixture of POPC and POPG behaves differently, but experimental
-data is not available for comparison for this mixture.
-
-\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}
-
-%\subsection{Effect of Ca$2+$ binding to PS headgroup}
-
-Also the experimental order parameters of POPS headgroup 
-as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
-The order parameter changes with added CaCl$_2$ are overestimated in
-both simulation models. However, the changes can be considered qualitatively
-correct for both segments in CHARMM36 model and in $\beta$ segment in MacRog model. 
-\todo{These should be compared to simulations for potential structural interpretation of the changes.}
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
-%  \caption{\label{PSPGchangesWITHCaCl}
-%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
-%  }
-%  \todo{Get the small concentration data from the inserts}
-%\end{figure}
+Also the order parameters of PS headgroup as a function of CaCl$_2$ concentration
+%from mixtures with PC  
+are shown in Fig. \ref{changesWITHCaClPS}. The order parameters exhibit
+a strong dependence small CaCl$_2$ concentrations with a rapid saturation. 
+The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
+both simulation models. The response of $\beta$-carbon order parameter could be
+interpreted qualitatively correct in simulations, but the general agreement with
+experiments is not sufficient to use simulations to interpret the response of
+PS headgroup to the added CaCl$_2$.
 
@@
   }
+\end{figure}
+
+Before using the headgroup order parameters to compare ion binding affinity between simulations
+and experiments, it is important to quantify the response of the order parameters to the
+bound charge in simulations.
+The response of headgroup order parameters to the fixed amount of cationic surfactants in
+POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
+The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
+while CHARMM36 is in better agreement with experiments. This has to be taken into account when
+analysin the binding affinities.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
+  \caption{\label{CHANGESwithCaClPGPS}
+  The response of headgroup order parameters to the fixed amount of cationic surfactants in
+  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
 \end{figure}

### 2018-03-14 scripts/calcAREAperLIPID.sh
Add Lipid17 density profiles by Kav and area per molecule values

@@ -1,0 +1,10 @@
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_CHARMM36/area.dat | awk '{print "CHARMM36  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_CHARMM36_POTASSIUMcounterion/area.dat | awk '{print "CHARMM36 potassium  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_CHARMM36UA/area.dat | awk '{print "CHARMM36ua  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_SLIPIDS/area.dat | awk '{print "Slipids  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_Berger/area.dat | awk '{print "Berger  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_MACROG/area.dat | awk '{print "MacRog  "$1*2/128}'
+echo 'Lipid17 ff99 ions   0.573274'  #ff99 ions from http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1520519296669#c1058693839316306625
+echo 'Lipid17 JC-ions   0.527930'  #JC-ions from http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1520519296669#c1058693839316306625
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_GROMOS-CKP1/area.dat | awk '{print "GROMOS-CKP1  "$1*2/128}'
+cat ../../MATCH/Data/Lipid_Bilayers/POPS/T298K/MODEL_GROMOS-CKP2/area.dat | awk '{print "CROMOS-CKP2  "$1*2/128}'

### 2018-03-15 Manuscript/manuscriptPS.tex
Update table with simulations

@@
     DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18}          &0 & 0        & ?    & ?   & ?  & ?  & ? & ? & \cite{??} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
     \hline
@@
     POPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
-    POPS/Na$^+$  & lipid17 \cite{gould18}  &0 & 0  & ? & ? & ? & ? & ? & ? & ? \\
+    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
+    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
     \hline
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}  &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
 %    \hline

### 2018-03-20 Manuscript/manuscriptPS.tex
Update the table listing simulated systems

@@
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}  &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-%    \hline
-%    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-%    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-%    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
-%    \hline
-%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
-%    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS\todoi{This is also probably OPPS?
+These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    \hline
+    POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref.
+\citenum{kim16}}          &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?
+& ?
+& 298  & 200 & ?
+& \cite{??} \todoi{Trajectories and further details to be added by J.
+Madsen}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref.
+\citenum{kim16}}          &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?
+& ?
+& 298  & 200 & ?
+& \cite{??} \todoi{Trajectories and further details to be added by J.
+Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
 \end{tabular}

### 2018-03-21 Manuscript/manuscriptPS.tex
Add sodium binding results to DMPS/DMPC bilayers

@@
 
+\subsection{Counterion binding to lipid bilayers containing PS lipids}
 Since systems with negatively charged PS lipids always contain counterions
@@
 of counterions along membrane normal from different simulations
-are shown in Fig. \ref{NAdensPOPS}.
-Significant differences are observed in counterion
-densities between different simulation models. 
-\todo{We should try to figure out which one of these are more
-  realistic. One option could be to use PS headgroup order parameter
-  data measured as a function of NaCl \cite{roux86}. Another
-  approach could be to use the area per molecule, which is suggested to
-  depend on the binding affinity of counterions \cite{pandit02,mukhopadhyay04,pedersen06}. The
-  experimental value for area per molecule is available at \cite{pan14}.}
+show significant binding of counterions in bilayers in all
+models in Fig. \ref{NAdensPOPS}.
 \begin{figure}[]
@@
 \end{figure}
+Differences are, however,
+observed in the total binding affinity and in the shape of density profiles.
+Comparison of bulk concentrations suggests that strongest binding
+is observed in MacRog, Berger and Lipid17 models, respectively.
+Other models gives roughly similar total binding affinity
+based on rough comparison of bulk concentrations.
+In CHARMM36, CHARMM36ua and Gromos-CKP models two maxima are observed
+in sodium density plots, possible corresponding binding to phoshate
+and carboxylic grops in the PS headgroup. In other models only
+a single maxima is observed. \todo{More quantitative analysis of
+binding positions and affinity may be useful.}
+
+Based on previous simulation results it has been suggested that
+the counterion binding and concomitant screening of headgroup repulsion
+leads to smaller area per lipid in PS lipid bilayers than in PC
+bilayers \cite{pandit02,mukhopadhyay04,pedersen06}. Therefore we also show the area per lipid from different models
+in Fig. \ref{NAdensPOPS}. The proposed trend is observed for Lipid17 model,
+for which the Joung-Cheatham ions give higher affinity and smaller area
+per molecule. However, the trend is not observed when comparing accross different
+force fields. For example, CHARMM36ua and MacRog give similar area per molecule
+but binding affinity is significantly higher in MacRog. On the other hand,
+Gromos-CKP models gives a larger area per lipid than other models, being only
+in agreement with experiments \cite{pan14}, but the binding affinity is comparable
+to the CHARMM36, CHARMM36us and Slipids, which give smaller are per molecule.
+
+In order to compare the sodium interaction with PS containing membranes between
+experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the order parameters of the
+$\alpha$ carbon of PS headgroup measured from the mixture with PC as a
+function of added NaCl \cite{roux86} together with the resuls from CHARMM36 model. 
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
+  \caption{\label{PSresponseTONaCl}
+    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
+  }
+\end{figure}
+The experimental results show essentially no changes in the order parameters as a function of
+added NaCl, while significant changes are observed in simulations. However,
+the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
+Therefore, we cannot exclude the possibility that the NaCl induced changes were already
+saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
+\todo{It would be probably better to use datasets for NaCl and KCl from Ref. \citenum{roux90}
+to compare monovalent ion binding to PS containing bilayer.}
 

### 2018-03-23 Manuscript/manuscriptPS.tex
Add results delivered by R. Nencini and conclusion that currect CHARMM-GUI parameters give too weak Ca2+ binding to the manuscript

@@
 a good agreemet with experiments \todo{Should be analyze/discuss this further?}.
-Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations
-\todo{This is most likely due to the new interactions added to CHARMM-GUI:
-  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1519913688317\#c5312059274016196300
-}.
+Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
+This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, which are incorporated
+to CHARMM-GUI. These parameters give also too weak binding of calcium to pure POPC bilayer,
+as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
 The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
@@
 
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
+  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
+  The response of headgroup order parameters to the fixed amount of cationic surfactants in
+  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+\end{figure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
+  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
+  The response of headgroup order parameters to the fixed amount of cationic surfactants in
+  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+\end{figure}
+
 \pagebreak

### 2018-03-27 Manuscript/manuscriptPS.tex
Add Berger results as a function of monovalent ions plotted with experimental data delivered in previous commit https://github.com/NMRLipids/NMRlipidsIVotherHGs/commit/b93473d40004729f0b8d0346a362ba9594fc8008

@@
 function of added NaCl \cite{roux86} together with the resuls from CHARMM36 model. 
-\begin{figure}[]
+\begin{figure*}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
+  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
   \caption{\label{PSresponseTONaCl}
-    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
+    Order parameters of PC (left) and PS (right) headgroup as a function of added NaCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
   }
-\end{figure}
+\end{figure*}
 The experimental results show essentially no changes in the order parameters as a function of
@@
   \centering
+  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
+  \caption{\label{PSresponseTONaCl}
+    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
+  }
+\end{figure}
+
+\begin{figure}[]
+  \centering
   \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}

### 2018-04-13 Manuscript/manuscriptPS.tex
Add Figure 5 with new data incorporated into the manuscript
Renamed the two CKP models in Table I and clarified their difference in the caption of Table I.

@@
    [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   CKP1 referes to the version with Berger/chiu NH3 charges compatible with Berger
-   and CKP2 to the version with more Gromos compatible version.
+   CKP-TMA$^+$ refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids)
+   and CKP-Lys refers to the version with more Gromos compatible version
+   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).  
    %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
@@
     POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
-    POPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
-    POPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP-TMA$^+$ \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP-Lys \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
     POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
@@
     and CKP2 to the version with more Gromos compatible version.'' Is this correct
-    also in this figure?} \\
-  \todo{Lipid17 results should be added.} \\
-  \todo{We should also add CHARMM36 POPS ran with K counterions: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1\#issuecomment-308570874}
+    also in this figure?
+    A: YES. Renamed CKPs to clarify this.} \\
 \end{figure}

### 2018-04-16 Manuscript/manuscriptPS.tex
Named the two CKP models following the naming scheme of Piggot.

@@
    [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   CKP-TMA$^+$ refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids)
-   and CKP-Lys refers to the version with more Gromos compatible version
+   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges)
+   and CKP refers to the version with more Gromos compatible version
    (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).  
@@
     POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
-    POPS/Na$^+$  & GROMOS-CKP-TMA$^+$ \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
-    POPS/Na$^+$  & GROMOS-CKP-Lys \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
     POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
@@
     Order parameters for PS headgroup and glycerol
-    backbone from simulations with different models and experiments without CaCl$_2$ 
-    Experimental data from \cite{browning80} contains 0.1M of NaCl.
+    backbone from simulations with different models and experiments without CaCl$_2$.
+    Experimental data from \cite{browning80} contain 0.1~M of NaCl.
     Signs are taken from experiments for POPS described in Supplementary Information.
@@
   }
-  \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  
-    Berger/chiu NH3 charges compatible with Berger
-    and CKP2 to the version with more Gromos compatible version.'' Is this correct
-    also in this figure?
-    A: YES. Renamed CKPs to clarify this.} \\
 \end{figure}

### 2018-04-20 Manuscript/manuscriptPS.tex
Update manuscript

@@
   }
-  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4}
+  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4} \\
+  \todo{Lipid17 and MacRog results should be added into this plot.}
 \end{figure}
@@
 are observed in other models.
-\todo{Discussion to be finished once all the results are in the plot.}
+\todo{Discussion to be finished once all the results are in the plot.} \\
+\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
 
@@
 In order to compare the sodium interaction with PS containing membranes between
-experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the order parameters of the
-$\alpha$ carbon of PS headgroup measured from the mixture with PC as a
-function of added NaCl \cite{roux86} together with the resuls from CHARMM36 model. 
+experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
+order parameters of POPC and POPS measured from 5:1 mixture as a
+function of different monovalent ions added to the buffer.
+The experimental experimental results \cite{roux86} are compared with different simulation
+models.
 \begin{figure*}[]
@@
   \caption{\label{PSresponseTONaCl}
-    Order parameters of PC (left) and PS (right) headgroup as a function of added NaCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
+    Order parameters of PC (left) and PS (right) headgroup as a function of
+    added NaCl, KCl and LiCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
   }
+  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
+    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
+  \todo{Maybe we should get the CHARMM results as well?}
 \end{figure*}
-The experimental results show essentially no changes in the order parameters as a function of
-added NaCl, while significant changes are observed in simulations. However,
-the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
-Therefore, we cannot exclude the possibility that the NaCl induced changes were already
-saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
-\todo{It would be probably better to use datasets for NaCl and KCl from Ref. \citenum{roux90}
-to compare monovalent ion binding to PS containing bilayer.}
+In line with the trend observed for bilayers in the absence of negatively
+charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
+significantly larger than to potassium, indicating stronger binding of lithium
+ions to lipid bilayers. The unique nature of lithium is also evident in the
+response of PS headgroup order parameters to bounds ions, which decrease
+with the addition of lithium or divalent cations but increase with the
+addition of sodium or potassium. The headgroup order parameter response
+to the added NaCl in the Berger model is similar to the experiments of LiCl,
+except for the $\alpha$ carbon of POPS for which the change is overestimated
+with the order of magnitude. The results suggests that the binding affinity
+of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
+and that the response of $\alpha$ carbon of POPS to the added sodium
+is significantly overestimated. The response of POPC to the added potassium
+in MacRog model is only slightly overestimated, in contrast to the response
+of POPS where order of magnitude overestimations of reposonse are observed. 
+In conclusion, it seems that the response of PS headgroup to bound cations
+is not correctly captured in the tested models.
+\todo{Finish the discussion when more models are available}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
+  \caption{  Counterion density distributions from PC:PS mixtures.
+\label{CIdensPSOCmixt}
+  }
+  \todo{Add Berger results.}
+\end{figure}
 
@@
 increase of order parameters with the added amount of PS.
+\todo{This discussion will be finalized when we have the
+results for monovalent ions also from other simulations.}
 
@@
   \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
-  \todo{We need results also from other than CHARMM36 force field.}
+  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
+    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
+  \todo{Maybe we should show the changes also in this plot?}
 \end{figure*}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
-  \caption{  Counterion density distributions from PC:PS mixtures.
-\label{CIdensPSOCmixt}
-  }
-\end{figure}
 
@@
 amount of PS lipids. This may indicate increasing order of the headgroup.
+\todo{The existing literature about this should be mentioned/discussed.}
 It should be, however, noted that the experimental data for pure POPS and
@@
 PS headgroup structures and mutual interactions with PC lipids. 
+\todo{This discussion should be updated.}
 
@@
   }
+  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
+  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
+  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
+    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
 \end{figure*}
@@
   \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
-  \caption{\label{PSresponseTONaCl}
+  \caption{\label{PSresponseTONaClDMPC}
     Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
@@
 \end{figure}
+The experimental results show essentially no changes in the order parameters as a function of
+added NaCl, while significant changes are observed in simulations. However,
+the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
+Therefore, we cannot exclude the possibility that the NaCl induced changes were already
+saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
 

### 2018-04-20 Manuscript/manuscriptPS.tex

--- 2018-04-16 Manuscript/manuscriptPS.tex
@@
   % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+ lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
     backbone from simulations with different models and experiments without CaCl$_2$.
+    All DOPS data at 303~K, POPS at 298~K.
     Experimental data from \cite{browning80} contain 0.1~M of NaCl.
     Signs are taken from experiments for POPS described in Supplementary Information.
-    The vertical bars shown for most computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
+    The vertical bars shown %for most computational values
+    are not error bars, but demonstrate that %for these systems
+    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
   }

### 2018-04-20 Manuscript/manuscriptPS.tex
Update density figures

@@
   }
-  \todo{Add Berger results.}
 \end{figure}
@@
   }
-  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
-  \todo{These are now mass densities, numbers would be probably better.}
-  \todo{Counterions should be also included.}
+  \todo{The CHARMM results are mass densities, numbers should be used.} \\
+  \todo{Should we include also counterions into the plot?} \\
   \todo{Not all the data from MacRog is included.} 

### 2018-04-20 Manuscript/manuscriptPS.tex
update manuscript

@@
 \centering
-\caption{List of MD simulations. The salt concentrations calculated as 
-   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+\caption{List of MD simulations without additional salt.
    CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
@@
     POPS/Na$^+$  & Berger \cite{??}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & ? & ?? & 0  & ?  & ? & ?  & \cite{??} \todoi{Data to be added by Piggot} \\
+    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
     POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
@@
     POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
+    \end{tabular}
+%\end{minipage}
+%\end{sidewaystable*} 
+\end{table*}
+
+\begin{table*}[!htb]
+%\begin{sidewaystable*}[!p]
+\centering
+\caption{  List of MD simulations. The salt concentrations calculated as 
+   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{??})
+   and CKP refers to the version with more Gromos compatible version
+   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
+   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+   %   The lipid force fields named as in our previous work~\cite{botan15}.
+}\label{IONsystems}
+%begin{minipage}[t]{\textwidth}
+\begin{tabular}{l c c r r r r r r c c}
+  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
     POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
-    POPC:POPS\todoi{This is also probably OPPS?
-These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    \hline
      POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref.
@@
 Madsen}  \\
+    \hline
+    POPC:POPS\todoi{This is also probably OPPS?
+These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
@@
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    \todo{MacRog simulations with KCl to be added}\\
+    \hline
+    \todo{Berger simulations with NaCl and CaCl to be added}
 \end{tabular}
@@
 %\end{sidewaystable*}
+
 \end{table*}
@@
 are observed in other models.
-\todo{Discussion to be finished once all the results are in the plot.} \\
+\todo{Discussion is to be finished.} \\
 \todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
@@
 study \cite{botan15}.
-\todo{Discussion will be finished when we have all the data in
-  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}.
+\todo{Discussion is to be finished.
   One possible conclusion could be the following:
@@
   }
+  \todo{Fix CKP1 and CKP2 captions.}
 \end{figure}
@@
 \section{Simulated systems}
+
+\subsection{CHARMM36}
+\todo{To be written by Piggot, Madsen and Ollila}
+
+\subsection{CHARMM36ua}
+\todo{To be written by Piggot}
+
+\subsection{Slipids}
+\todo{To be written by Piggot and Favela}
+
+\subsection{Berger}
+\todo{To be wiritten by Piggot and Ollila}
+Simulations with sodium were taken directly from Ref. \citenum{??} and
+simulations with calcium directly from \citenum{melcrova16}.
+
+\subsection{GROMOS-CKP}
+\todo{To be written by Piggot}
+
+\subsection{Lipid17}
+\todo{To be written by Kav and Miettinen}
+
+\subsection{MacRog}
+\todo{To be written by Javanainen and Piggot}
 

### 2018-04-26 Manuscript/manuscriptPS.tex
Update manuscript and figures. Add Berger data for PC change as a function of PS.

@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
+
+\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
+\affiliation{nmrlipids.blogspot.fi}
 
@@
   \caption{\label{PSresponseTONaCl}
-    Order parameters of PC (left) and PS (right) headgroup as a function of
-    added NaCl, KCl and LiCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
+    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
+    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of y-axis.
+    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
+    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
+    parameter value is at zero.
   }
@@
   \caption{\label{HGorderparametersPCvsPS}
-    Headgroup order parameters from PC:PS mixtures from 
-    different simulation models and experiments.
-    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
-    signs are determined as discussed in \cite{botan15,ollila16}).
-    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
-    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
-    Counterions in expriments are sodium, while potassium is used in simulations.
-    Using sodium in simulations do not have a significant effect.
+    Changes of PC (left panel) and PS (right panel) headgroup order parameters
+    from POPC:POPS mixtures with increasing amount of POPS.
+    Experimental results of POPC are taken from Ref. \citenum{scherer87}
+    (signs are determined as discussed in \cite{botan15,ollila16}).
+    Experimental values for POPS in pure bilayer and in mixture are  
+    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
+    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
+    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
+    set to the value from POPC:POPS (5:1) mixture.
+    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
+    with the same value for both order parameters such that the lower order
+    parameter value from POPC:POPS (5:1) mixture is at zero
+    to correctly illustrate the significant forking.
   }
@@
     http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
-  \todo{Maybe we should show the changes also in this plot?}
 \end{figure*}
@@
 
-The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
-as a function of Ca$2+$ concentration from different
-simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPS}.
-The results suggest that Ca2+ ions clearly overbind in simulations with
-MacRog model, as expected from previous results for PC lipid bilayers \cite{catte16}.
+The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) mixtures
+are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
+simulations and experiments \cite{roux90}. 
+The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
+The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
+MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
 It should be noted, however, that the lowest concentration (100mM) gives
-a good agreemet with experiments \todo{Should be analyze/discuss this further?}.
+a good agreemet with experiments
+\todo{Should be analyze/discuss this further?
+  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
+It should be noted that Berger simulation do not have counterions.}.
 Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
-This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, which are incorporated
-to CHARMM-GUI. These parameters give also too weak binding of calcium to pure POPC bilayer,
+This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
+in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
+The binding of calcium to pure POPC bilayer is also too weak with these parameters
 as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
-The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
-\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}
+\todo{The discussion is to be finished when we have all the data in the plot.}
 
@@
   \caption{\label{changesWITHCaClPS}
-    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
-  }
+    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of y-axis.
+    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
+    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
+    parameter value is at zero. 
+  }
+  \todo{Information about the cuonterions in different simulations should be added} \\
   \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
@@
 
-Also the order parameters of PS headgroup as a function of CaCl$_2$ concentration
-%from mixtures with PC  
-are shown in Fig. \ref{changesWITHCaClPS}. The order parameters exhibit
-a strong dependence small CaCl$_2$ concentrations with a rapid saturation. 
+Also the order parameters of PS headgroup from POPC:POPS (5:1) mixture
+are shown in Fig. \ref{changesWITHCaClPS} as a function of CaCl$_2$ concentration.
+In experiments the order parameters exhibit a strong dependence of CaCl$_2$ with
+small concentrations with a rapid saturation around 50 mM. 
 The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
-both simulation models. The response of $\beta$-carbon order parameter could be
-interpreted qualitatively correct in simulations, but the general agreement with
-experiments is not sufficient to use simulations to interpret the response of
-PS headgroup to the added CaCl$_2$.
+all tested simulation models. Furthermore, the changes of the headgroup order parameters
+do not qualitatively agree with experiments. This is in contrast to previous results
+for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
+observed despite of significant discrepancies in the headgroup structure without additional ions.
 
@@
 simulations with calcium directly from \citenum{melcrova16}.
+Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.
 

### 2018-05-17 Manuscript/manuscriptPS.tex
Add plot comparing POPS and OPPS from MacRog simulations.
Update preliminary figure summarising experimental results

@@
 
-Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR} summarize the experimental NMR results for POPS
-bilayer sample. \todo{New figure, probably combining Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR}, should
-be done. This discussion to be finished after this.}
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=6.0cm]{../Figs/PShgSIGNSsimpson.pdf}
-  \caption{\label{PShgSIGNSsimpson}
-    (a) The headgroup region of the INEPT spectrum where alpha and beta are
-    identified.
-    (b) The R-PDLF slices for alpha and beta showing one single
+Figure \ref{PShgSIGNSsimpson} summarizes the experimental NMR results for POPS bilayer sample.
+showing one single
     splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
@@
     resolved with the available resolution.
-    (c) Points with error bars are the experimental SDROSS data.
-    The thick lines are SIMPSON simulations.
-    The S-DROSS slice for beta clearly shows that the order parameter is negative,
-    which is confirmed by SIMPSON simulations using the order parameter value of -0.12.
+    S-DROSS slice for beta clearly shows that the order parameter is negative,
+    which is confirmed by SIMPSON simulations using the 
     The S-DROSS slice for alpha suggests that the higher order parameter is positive
@@
     numerical value for this. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller order parameter gave did not agree with experiments (dashed grey),
+    for the smaller order parameter (dashed grey), did not agree with experiments 
     confirming the interpretation that the smaller order parameter is negative.
-  }
-  \todo{Maybe we should combine this with \ref{glycerolNMR}}
-\end{figure}
-
 \begin{figure}[!htb]
   \centering
-  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
-  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
-  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
-  \caption{\label{glycerolNMR}
-    R-PDFL slices for gycerol backbone carbons.
-  }
-  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
-  \todo{What are the top figures actually?}
-\end{figure}
+  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
+  \caption{\label{PShgSIGNSsimpson}
+    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
+    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
+    (c) Slices for $\alpha$ and $\beta$ barbons.
+    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
+    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
+    for the larger and smaller $\alpha$-carbon slittings were used in the
+    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller order parameter (dashed grey).
+  }
+  \todo{This is preliminary figure, should be polished.}
+  \todo{Should we show slices for all the analyzed carbons in (c)?}
+\end{figure}
+
+%\begin{figure}[!htb]
+%  \centering
+%  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
+%  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
+%  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
+%  \caption{\label{glycerolNMR}
+%    R-PDFL slices for gycerol backbone carbons.
+%  }
+%  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
+%  \todo{What are the top figures actually?}
+%\end{figure}
 
@@
 
+\section{Difference between POPC and OPPS in MacRog model}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
+  \caption{\label{CHANGESwithCaClPGPS}
+    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
+\end{figure}
+
 \section{Dihedrals}

### 2018-05-22 Manuscript/manuscriptPS.tex
Update manuscript and some cleaning

@@
 %lipids is shown in Fig. \ref{lipids}.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
-  \caption{\label{lipids}
-    Chemical structures and labels for the headgroup carbons.
-  }
-\end{figure}
+%\begin{figure}[]
+
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
+%  \caption{\label{lipids}
+%    Chemical structures and labels for the headgroup carbons.
+%  }
+%\end{figure}
 
@@
 \subsection{Solid state NMR experiments}
-The experimental protocol is the same used in Ref. \citenum{ferreira16}.
-\todo{Some basic details should be given.}
+The magnitude and signs of the C-H bond order parameters in
+headgroup and glycerol backbone were measured
+using natural abundance $^{13}$C solid state NMR spectroscopy as
+described previously \cite{ferreira13,ferreira16}.
+Shortly, the absolute values of the order parameters were determined from the dipolar splittings
+given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
+the signs were measured using S-DROSS experiments \cite{gross97}.
+
+\todo{Details of the used spectrometer and maybe some other details should be given.}
+\todo{Sample preparation should be described.}
 
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPS.eps}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
   \caption{\label{HGorderParameters}

### 2018-05-31 Manuscript/manuscriptPS.tex
Update Lipid17 results

@@
 \subsection{Molecular dynamics simulations}
-Molecular dynamics simulation data was collected with Open Collaboration method.
-The simulated systems are listed in Table \ref{IONsystems}.
-The simulation details are in SI or in references in the table.
+Molecular dynamics simulation data was collected using
+the Open Collaboration method \cite{botan15}.
+The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
+the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
+were used as the communication platforms.
+The simulated systems are listed in Table \ref{IONsystems} and
+simulation details are given in the SI.
+The simulation data is also indexed in the
+searchable database (\url{nmrlipids.fi}),
+and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).  
 \begin{table*}[!htb]
@@
 
+The C-H bond order parameters were calculated directly
+from the definition
+\begin{equation}
+S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
+\end{equation}
+where $\theta$ is the angle between the C-H bond and the membrane normal.
+Angular brackets point to the average over all sampled configurations.
+\todo{Error estimation should be discussed.}\\
+The number density profiles were calculated using {\it gmx density} tool
+from Gromacs sofware package \cite{gromacsMANUAL}.
+
 \section{Results and Discussion}

### 2018-05-31 scripts/grepOPdata.sh

@@ -1,0 +1,7 @@
+grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep beta1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep beta2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep beta ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"   "}' 

### 2018-06-01 scripts/grepOPdata.sh
Update and correct Lipid17 plots

@@ -1,7 +1,7 @@
-grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMKCl/T298/MODEL_LIPID17/OrdParsPOPS.dat | awk '{print $2"  "$3"   "}' 
+grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
+grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
[…]

### 2018-06-06 scripts/grepOPdata.sh
Add results from Lipid17 simulation with monovalent ions by B. Kav http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1524581650542#c4627966291656120911

@@ -1,7 +1,8 @@
-grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPC.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha1 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep alpha2 ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPS.dat | awk '{print $2"  "$3"  "}' | tr -d "\n"
-grep beta ../../MATCH/Data/Lipid_Bilayers/POPS+83%popc/T298K/MODEL_LIPID17_POTASSIUMcounterion/OrdParsPOPS.dat | awk '{print $2"  "$3"   "}' 
+directory=../../MATCH/Data/Lipid_Bilayers/POPS+83%popc+4000mMNaCl/T298K/MODEL_LIPID17
+grep alpha1 $directory/OrdParsPOPC.dat | awk '{print $5"  "$7"  "}' | tr -d "\n"
+grep alpha2 $directory/OrdParsPOPC.dat  | awk '{print $5"  "$7"  "}' | tr -d "\n"
+grep beta1 $directory/OrdParsPOPC.dat | awk '{print $5"  "$7"  "}' | tr -d "\n"
[…]

### 2018-06-07 Manuscript/manuscriptPS.tex
Update text

@@
 from different simulation models are compared with the experimental
-data in Fig. \ref{HGorderParametersPS}. In line with the previous
-study for PC lipids \cite{botan15}, the glycerol backbone order 
-parameters in CHARMM36 roughly agree with the experimental data,
-while significant discrepancies for glycerol backbone carbons
-are observed in other models.
-\todo{Discussion is to be finished.} \\
+data in Fig. \ref{HGorderParametersPS}. Subjective ranking
+of the quality of the models is shown in Fig. \ref{comparisonTablePS}.
+The tested models perform generally less well than in the previous
+study for PC headgroup~\cite{botan15} and none of the models 
+reproduce the experimental order parameters within
+experimental error bars.
+Therefore, the models cannot be straightforwardly used to interpret the
+structure of PS headgroup. However, the differences between PC and PS headgroups
+are partially reproduced by some of the models.
+
+The two best performing models for the $\alpha$ and $\beta$-carbons of
+PS, Slipids and CHARMM36, reproduce the large forking in the $\alpha$-carbon
+and the Slipids model gives also a good agreement with experiments for
+the $\beta$-carbon order parameter in both PC and PS headgroups
+(Fig. \ref{HGorderParametersPS} and Ref.~\citenum{botan15}).
+Interestingly, the dihedral angle distributions of CHARMM36
+and Slipids in Fig.~\ref{dihedralsHG} share significant similarities
+in the headgroup region. However, the experimental order parameters in the
+glycerol backbone region are not well reproduced by the Slipids model,
+which was also the case in for PC lipids \cite{botan15}.
+This difference probably arises from the differences in
+the dihedral angle distributions of C1-C2-C3-O31 and C2-C3-O31-C31
+in Figure \ref{dihedralsGLY}, which are also illustrated in
+Figure \ref{glycerol_buslaev}.
+
+%\todo{Discussion is to be finished.
+%  One possible conclusion could be the following:
+%  The main differences between the models in the headgroup region are observed
+%  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
+%  CHARMM36, CHARMM36UA and Slipids give very similar results to the
+%  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
+%  The order parameters of $\beta$-carbons for these three models
+%  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
+%  On the other hand, Gromos-CKP models give better order parameters
+%  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
+%  In conclusion, the suggestion would be that the single peak for
+%  observed at 120 degrees in CHARMMs and Slipids would
+%  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
+%  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
+%}
+
 \todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
-
-Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
-the CHARMM36 is the best performing model for both lipids.
-However, none of the tested models give a satisfactory 
-agreement with experiments for the headgroup order paramaters of 
-PS lipids. The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
-for the best performing CHARMM36 model is larger for PS lipids (8)
-than for PC (3) \cite{botan15}. Therefore the 
-interpretation of structural details of PS headgroup or
-differences between PC and PS lipids is very challenging 
-from the current MD simulation models.
-Figures \ref{dihedralsGLY} and \ref{dihedralsHG} show dihedral
-angle distributions calculated from different models. 
-The glycerol backbone structures from CHARMM36 and
-Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
-missing structures in Slipids model, which probably lead to the incorrect
-order parameters. For the headgroup $\alpha$ and $\beta$ carbon order parameters of PS lipids,
-the tested models perform less well than for PC headgroup in previous
-study \cite{botan15}.
-\todo{Discussion is to be finished.
-  One possible conclusion could be the following:
-  The main differences between the models in the headgroup region are observed
-  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
-  CHARMM36, CHARMM36UA and Slipids give very similar results to the
-  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
-  The order parameters of $\beta$-carbons for these three models
-  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
-  On the other hand, Gromos-CKP models give better order parameters
-  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
-  In conclusion, the suggestion would be that the single peak for
-  observed at 120 degrees in CHARMMs and Slipids would
-  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
-  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
-}
 
@@
 \end{figure}
+
+%
+%The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
+%for the best performing CHARMM36 model is larger for PS lipids (8)
+%than for PC (3) \cite{botan15}.
+%Therefore the interpretation of structural details of PS headgroup or
+%differences between PC and PS lipids is very challenging 
+%from the current MD simulation models.
 

### 2018-06-07 Manuscript/manuscriptPS.tex
Update text

@@
 
-\todo{Details of the used spectrometer and maybe some other details should be given.}
-\todo{Sample preparation should be described.}
+\todo{Details of the used spectrometer and maybe some other details should be given.} \\
+\todo{Sample preparation should be described.} \\
+\todo{How is the peak assignment done?}
 
@@
 
-Figure \ref{PShgSIGNSsimpson} summarizes the experimental NMR results for POPS bilayer sample.
-showing one single
-    splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
-    of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot
-    resolved with the available resolution.
-    S-DROSS slice for beta clearly shows that the order parameter is negative,
-    which is confirmed by SIMPSON simulations using the 
-    The S-DROSS slice for alpha suggests that the higher order parameter is positive
-    and the deviation towards negative values in the longer t1 times suggests
-    that the smaller order parameter is negative.
-    This is confirmed by SIMPSON simulation using value of 0.09
-    for the larger alpha order parameter and the value of -0.02 for
-    smaller (black curve). The value for the smaller
-    alpha order parameter for SIMPSON calculation was taken from Fig 3 in Ref. \citenum{roux91},
-    because resolution in 13C NMR experiments was nor high enough to determine
-    numerical value for this. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller order parameter (dashed grey), did not agree with experiments 
+The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
+carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
+was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
+Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
+show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
+and a superposition of a large and a very small splitting for the $\alpha$-carbon.
+The larger splitting gives a order parameter value of 0.09, while the numerical value
+from the small splitting cannot resolved with the available resolution.
+Since only the absolute values of the PS headgroup order parameters were measured
+previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{??} to determine the signs of
+the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
+order parameter is negative, which is confirmed by SIMPSON simulations.
+The beginning of the S-DROSS slice suggests that the higher order parameter of
+the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
+that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
+where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
+The literature value was used because the
+resolution of our experiment was not sufficient to determine the
+small value of the order parameter. The S-DROSS curve from
+SIMPSON simulation with a positive value for the smaller order parameter
+(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
 confirming the interpretation that the smaller order parameter is negative.
@@
 The headgroup and glycerol backbone order parameters of 
-POPS measured in this work are compared to the literature values of
-DOPS \cite{browning80} and POPC \cite{ferreira13} in Fig. \ref{HGorderParameters}. 
-Our results for POPS are in good agreement with the previously reported
-values for DOPS measured with $^2$H NMR. Significant differences are
-observed between PC and PS lipids, especially at the headgroup region.  
-Previous discussions in the literature have concluded that  
-glycerol backbone structure is largely similar in
-PC, PE, PG and PS lipids \cite{gally81}. The headgroup region was
-found to be similar in PC, PE and PG lipid (assuming that the signs of PE and PG
-order parameters are the same as in PC), while the PS headgroup was suggested
-to be more rigid \cite{wohlgemuth80,buldt81}. 
-The detailed structural differences between the headgroups is, however, not known.
+POPS measured in this work are in good agreement with the previously reported
+values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
+The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
+experiences a significant forking in PS headgroup when compared with
+the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
+These features have been intepreted to arise from a rigid PS headgroup
+conformation, stabilized by hydrogen bonds or electrostatic
+interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
+available. 
+%The detailed structural differences between the headgroups is, however, not known.
 \begin{figure}[]
@@
   }
+  \todo{There should be values in \cite{roux90} which should be added.}
 \end{figure}

### 2018-06-12 Manuscript/manuscriptPS.tex

@@
 Therefore, the structural details
-of lipid headgroups and the details of calcium binding
+of lipid headgroups and the details of cation binding
 are crucial for the PS mediated signaling processes.
+
+Previous experimental studies have concluded that
+PS headgroups have more rigid conformation than
+phophocholines (PC) due to hydrogen bonding or
+electrostatic interactions \cite{browning80,buldt81}.
+Multivalent cations or Li$^+$ are able to form strong
+dehydrated molecular complexes with PS lipids,
+while monovalent ions interact more weakly with PS
+containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
+Dilution of PS rich bilayers with PC lipids
+reduces the rigidity and propensity of the
+strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
 

### 2018-06-19 Manuscript/manuscriptPS.tex

@@
 electrostatic interactions \cite{browning80,buldt81}.
-Multivalent cations or Li$^+$ are able to form strong
+Multivalent cations and Li$^+$ are able to form strong
 dehydrated molecular complexes with PS lipids,
@@
 containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
-Dilution of PS rich bilayers with PC lipids
-reduces the rigidity and propensity of the
+Dilution of bilayers with PC lipids makes PS headgroups
+less rigid and reduces propensity for the formation of
 strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
@@
   \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
+    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
     DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
@@
     \hline
-    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
-    POPS/K$^+$   & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
+    POPS/Na$^+$  & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
+    POPS/K$^+$   & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
     POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
@@
   \hline
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}  &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
-    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
-     POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref.
-\citenum{kim16}}          &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?
+& 0  & 298  & 200 & ?
+& \cite{??} \todoi{Trajectories and further details to be added by J.
+Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?
+& 0  & 298  & 200 & ?
+& \cite{??} \todoi{Trajectories and further details to be added by J.
+Madsen}  \\
+     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?
 & ?
@@
 Madsen}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref.
-\citenum{kim16}}          &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?
 & ?
@@
     http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
-  \todo{Maybe we should get the CHARMM results as well?}
+  \todo{CHARMM36 results for this plot would be highly useful.}
 \end{figure*}
@@
   }
-  \todo{Fix CKP1 and CKP2 captions.}
+  \todo{Lipid 17 is to be added.}
 \end{figure}
@@
   \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
-  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
-    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
+  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
+  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.}
 \end{figure*}

### 2018-06-19 Manuscript/manuscriptPS.tex
Update text

@@
 \subsection{Counterion binding to lipid bilayers containing PS lipids}
-Since systems with negatively charged PS lipids always contain counterions
-and the ion binding presumably affects the PS headgroup
-order parameters, the discussion of ion binding affinity and
-headgroup structure cannot be separated as done previously
-for PC lipids \cite{botan15,catte16}. The density profiles
-of counterions along membrane normal from different simulations
-show significant binding of counterions in bilayers in all
-models in Fig. \ref{NAdensPOPS}.
+Membranes containing PS lipids are always accomppanied with counterions, which
+modulate electrostatic interactions between lipids and other biomolecules. 
+Counterions are also suggested screen the repulsion between charged lipid headgroups 
+in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
+bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
+The counterion density profiles along membrane normal 
+show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
+The strongest counterion binding, i.e., the lowest concentrations in bulk water,
+are observed in MacRog, Berger and Lipid17/JC simulations.
+CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
+density, while a single maxima is observed in the other models. 
+\todo{More detailed discussion may be possible after comparing monovalent ion binding
+to bilayers between CHARMM simulations and experiments.}
+Area per lipid is in agreement with experiments \cite{pan14} only
+in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
+The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
+counterion binding because CHARMM36, CHARMM36ua and Slipid models give
+smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
+%The proposed trend is observed for Lipid17 model,
+%for which the Joung-Cheatham ions give higher affinity and smaller area
+%per molecule. However, the trend is not observed when comparing accross different
+%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
+%but binding affinity is significantly higher in MacRog. 
 \begin{figure}[]
@@
 \end{figure}
-Differences are, however,
-observed in the total binding affinity and in the shape of density profiles.
-Comparison of bulk concentrations suggests that strongest binding
-is observed in MacRog, Berger and Lipid17 models, respectively.
-Other models gives roughly similar total binding affinity
-based on rough comparison of bulk concentrations.
-In CHARMM36, CHARMM36ua and Gromos-CKP models two maxima are observed
-in sodium density plots, possible corresponding binding to phoshate
-and carboxylic grops in the PS headgroup. In other models only
-a single maxima is observed. \todo{More quantitative analysis of
-binding positions and affinity may be useful.}
-
-Based on previous simulation results it has been suggested that
-the counterion binding and concomitant screening of headgroup repulsion
-leads to smaller area per lipid in PS lipid bilayers than in PC
-bilayers \cite{pandit02,mukhopadhyay04,pedersen06}. Therefore we also show the area per lipid from different models
-in Fig. \ref{NAdensPOPS}. The proposed trend is observed for Lipid17 model,
-for which the Joung-Cheatham ions give higher affinity and smaller area
-per molecule. However, the trend is not observed when comparing accross different
-force fields. For example, CHARMM36ua and MacRog give similar area per molecule
-but binding affinity is significantly higher in MacRog. On the other hand,
-Gromos-CKP models gives a larger area per lipid than other models, being only
-in agreement with experiments \cite{pan14}, but the binding affinity is comparable
-to the CHARMM36, CHARMM36us and Slipids, which give smaller are per molecule.
-
-In order to compare the sodium interaction with PS containing membranes between
-experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
-order parameters of POPC and POPS measured from 5:1 mixture as a
-function of different monovalent ions added to the buffer.
-The experimental experimental results \cite{roux86} are compared with different simulation
-models.
+
+To evaluate counterion binding in different simulation against experimental data~\cite{roux90},
+we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
+as a function of function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
 \begin{figure*}[]

### 2018-06-21 Manuscript/manuscriptPS.tex
Update text

@@
 
-To evaluate counterion binding in different simulation against experimental data~\cite{roux90},
+To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
 we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
-as a function of function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
+as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
+Experimental order parameter data for POPC headgroup in the mixture is available as a function
+of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
+as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
+ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
+This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
+but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
+POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
+but only modest changes as a function of KCl concentration, indicating singificant 
+Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
+electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
+Berger model, the headgroup order parameter response of POPC
+to the added NaCl is similar to the experiments of LiCl,
+indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
+Indeed, the sodium density profile shows a significant binding peak in the 
+Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
+is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
+changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
+\todo{Discussion about Lipid17 to be written when we have the density profiles.}
+All the tested models overestimate the changes of POPS headgroup order parameters as
+a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
+model development is necessary to interpret the PS headgroup-ion interactions 
+from MD simulations.
 \begin{figure*}[]
@@
   }
-  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
-    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
   \todo{CHARMM36 results for this plot would be highly useful.}
 \end{figure*}
-In line with the trend observed for bilayers in the absence of negatively
-charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
-significantly larger than to potassium, indicating stronger binding of lithium
-ions to lipid bilayers. The unique nature of lithium is also evident in the
-response of PS headgroup order parameters to bounds ions, which decrease
-with the addition of lithium or divalent cations but increase with the
-addition of sodium or potassium. The headgroup order parameter response
-to the added NaCl in the Berger model is similar to the experiments of LiCl,
-except for the $\alpha$ carbon of POPS for which the change is overestimated
-with the order of magnitude. The results suggests that the binding affinity
-of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
-and that the response of $\alpha$ carbon of POPS to the added sodium
-is significantly overestimated. The response of POPC to the added potassium
-in MacRog model is only slightly overestimated, in contrast to the response
-of POPS where order of magnitude overestimations of reposonse are observed. 
-In conclusion, it seems that the response of PS headgroup to bound cations
-is not correctly captured in the tested models.
-\todo{Finish the discussion when more models are available}
 
@@
 \subsection{Headgroup structure in PS and PC mixtures}
-Cellular membranes often compose of mixtures of zwitterionic and negatively
-charged lipids. The response of PC headgroup to mixtures of differently charged
-lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
-As expected from the electrometer concept \cite{seelig87}, the headgroup 
-order parameters increase with the addition of negatively charged PS and PG lipids,
-decrease when mixed with positively charged surfactants and
-are less affected by the addition of zwitterionic PE lipids or cholesterol.
-In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
-also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
-follow the eletrometer concept \cite{scherer87}.
-
-%Also the Ca2+ binding affinity to bilayers containing
-%PS lipids is measured from mixtures with PC, because pure PS
-%bilayers exhibit instant transition to the ordered phase with the
-%addition of the ions.
-%Therefore, the characterization of mutual interactions
-%between PC and PS lipid headgroups is necessary in order to
-%understand PS lipids in biological environments.
-
-Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
-from POPC:POPS mixtures with various mole fractions from 
-simulations and experiments \cite{scherer87,roux90}.
-The experimentally observed increase of PC headgroup order parameters with the increasing
-amount of negatively charged PS lipid is reproduced in
-MacRog simulations with potassium counterions, but not in CHARMM36 simulations
-with potassium or sodium conterions. The results can be
-understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
-Both counterions are more strongly bound to bilayers in
-CHARMM36 simulations than potassium in MacRog simulation.
-The cation binding probably neutralizes the effect of negatively
-charged lipids in CHARMM36 simulations, while the
-weaker binding of potassium in MacRog simulations enables the
-increase of order parameters with the added amount of PS.
-\todo{This discussion will be finalized when we have the
-results for monovalent ions also from other simulations.}
-
-%$\beta$-carbon of POPC decreases and almost negligible increase is 
-%seen in $\alpha$ order parameter with the addition of PS lipid 
-%This may indicate incorrect interactions between PC and PS headgroups
-%in simulations. On the other hand, the previously reported overestimated
-%ion binding affinity may also play a role, as the amount of counterions
-%is increasing with PS.
-%\todo{Data from other force fields to be collected before discussion can be finalized.}
+Dilution of PS lipid bilayers with PC lipids reduces propensity for the formation of
+strong complexes with multivalent ions and is proposed to make PS headgroups 
+less rigid~\cite{browning80,buldt81,roux90,roux91}. Therefore, the intermolecular
+interactions at the headgroup region seems to determine important physical properties 
+of mixed lipid bilayers. The headgroup order parameters of POPC increase in experiments 
+with increasing amount of POPS in the mixed lipid bilayer (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
+This behaviour is generally observed when negatively charged lipids or surfactants 
+are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
+tilting of lipid headgroup more parallel to the membrane plane according to
+the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
+shift closer to zero when bilayer is diluted with PC lipids in 
+experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
+which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
+
+The increase of POPC headgroup order parameters with the increasing
+amount of negatively charged POPS lipid is reproduced in
+MacRog simulations with potassium counterions, but not in 
+Berger simulations with sodium or in CHARMM36 simulations
+with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
+The observations can be explained using the electrometer concept.
+The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
+which surpasses the effect of negatively charged lipids as the amount of
+counterions increase with increasing amount of PS.
+In CHARMM36 simulations, the counterion binding neutralizes the effect
+of PS and the headgroup order parameters are not changed with increasing amount
+of PS. Finally, the weak binding of potassium in the MacRog simulations
+enables the increase of order parameters with the added amount of PS 
+(Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).
 
@@
 
+%Also the Ca2+ binding affinity to bilayers containing
+%PS lipids is measured from mixtures with PC, because pure PS
+%bilayers exhibit instant transition to the ordered phase with the
+%addition of the ions.
+%Therefore, the characterization of mutual interactions
+%between PC and PS lipid headgroups is necessary in order to
+%understand PS lipids in biological environments.
+
 The headgroup order parameters of PC lipids
@@
 \todo{To be written by Javanainen and Piggot}
-
-\section{PC lipid headgroup response to different mixtures in experiments}
-As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
-headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
-i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
-surfactants. The changes with the addition of neutral lipids is significantly smaller.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
-  \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    PC headgroup order parameters from experiments of mixtures with
-    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
-  }
-\end{figure}
 

### 2018-06-21 Manuscript/manuscriptPS.tex
Update text

@@
     measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
+    Since the experimental data of POPS in pure and diluted mixture come from
+    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
+    the experimental change of the order parameter is less accurate than in typical measurements
+    where same technique is used in all conditions, see discussion about qualitative and quantitative 
+    accuracy in Ref. \citenum{ollila16}.
     For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
@@
 
-Also the headgroup order parameters of POPS mixed with varying amounts of POPC
-from simulations and experiments \cite{borle85,roux90} are shown in Fig. \ref{HGorderparametersPCvsPS}.
-The $\beta$-carbon order parameter of POPS slightly decreases and positive 
-order parameter of $\alpha$-carbon slightly increases with increasing
-amount of PS lipids. This may indicate increasing order of the headgroup.
-\todo{The existing literature about this should be mentioned/discussed.}
-It should be, however, noted that the experimental data for pure POPS and
-POPC/POPS mixture come from different experimental sets, 13C NMR in this work
-and 2H NMR from Ref. \citenum{roux90}, respectively. Therefore the accuracy
-of the order parameter change is not as high as typically in the measurements
-of order parameter changes, see discussion about qualitative and quantitative 
-accuracy in Ref. \citenum{ollila16}. The changes of PS headgroup order parameters
-are not reproduced by the tested simulation models. The $\beta$-carbon order parameter
-increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
-to the experimental data. The smaller $\alpha$-carbon order parameter increase in both
-simulation models with increasing amount of PS, while it is almost unchanged in experiments.
-The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
-with increasing amount of PS, both model exhibiting a poor agreement with experiments. 
-Signifincant improvement in the MD simulation models are needed to interpret the
-PS headgroup structures and mutual interactions with PC lipids. 
-\todo{This discussion should be updated.}
+Oppositely to experiments, the headgroup order parameter of POPS shift away
+from zero in CHARM36 simulations when bilayer is diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
+In lipid14/17 simulations, the POPS order parameter shift closer to zero when
+bilayer is diluted with POPC, but the numerical values of order parameters
+are too far from experiments to enable interpretation of the experimental data.
+Therefore, we conlcude that the force field development is necessary before
+MD simulations can be used to interpret the interactions between PC and PS headgroups.
+%The $\beta$-carbon order parameter
+%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
+%to the experimental data.
+%The smaller $\alpha$-carbon order parameter increase in both
+%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
+%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
+%with increasing amount of PS, both model exhibiting a poor agreement with experiments.
 

### 2018-06-22 Manuscript/manuscriptPS.tex
Update texrt

@@
 
+\subsection{Comparison of ion binding between simulations and experiments using the electrometer concept}
+
+The headgroup order parameters of PC lipids
+decrease proportionally to the bound positive
+charge in to a bilayer \cite{seelig87,catte16} and can be
+therefore used to measure the ion binding
+affinity. This molecular electrometer concept can
+be also applied to lipid bilayers with mixtures
+of PC and negatively charged lipids \cite{borle85,macdonald87,roux90}
+(see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+%This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
+%\ref{OrderParametersWithCaClBELOW1M} and
+%using the previuosly reported experimental data
+%for mixtures with varying proportions of
+%negatively charged PS or PG lipids.
+%
+%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
+%as a function of CaCl$_2$ concentration are shown
+%
+% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+
+%PC headgroup order parameters increase when negatively charged
+%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
+%as expected based on electrometer concept \cite{seelig87}
+%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
+%In electrometer concept this is explained by the tilting of
+%headgroup more parallel to membrane normal \cite{??}.
+
+%Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
+%At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
+%overcharging occurs above these concenterations.
+
 \section{Results and Discussion}
@@
 \subsection{Headgroup structure in PS and PC mixtures}
-Dilution of PS lipid bilayers with PC lipids reduces propensity for the formation of
-strong complexes with multivalent ions and is proposed to make PS headgroups 
-less rigid~\cite{browning80,buldt81,roux90,roux91}. Therefore, the intermolecular
-interactions at the headgroup region seems to determine important physical properties 
-of mixed lipid bilayers. The headgroup order parameters of POPC increase in experiments 
-with increasing amount of POPS in the mixed lipid bilayer (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
+Dilution of PS lipid bilayers with PC lipids reduces the propensity
+of PS headgroup-multivalent ion complexes and is proposed to make PS
+headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
+Therefore, the intermolecular interactions at the headgroup region seems
+to be important for the physical properties of mixed lipid bilayers.
+These interactions can be indirectly monitored by measuring the
+headgroup order parameters from PS:PC mixtures with different molar
+ratios. The headgroup order parameters of POPC increase in such experiments 
+with increasing amount of POPS (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
 This behaviour is generally observed when negatively charged lipids or surfactants 
@@
 The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
-which surpasses the effect of negatively charged lipids as the amount of
+which surpasses the effect of negatively charged lipids as also the amount of
 counterions increase with increasing amount of PS.
@@
 of PS. Finally, the weak binding of potassium in the MacRog simulations
-enables the increase of order parameters with the added amount of PS 
-(Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).
+enables the increase of order parameters with the increasing amount of
+negatively charged PS lipids (Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).
 
@@
 
-%Also the Ca2+ binding affinity to bilayers containing
-%PS lipids is measured from mixtures with PC, because pure PS
-%bilayers exhibit instant transition to the ordered phase with the
-%addition of the ions.
-%Therefore, the characterization of mutual interactions
-%between PC and PS lipid headgroups is necessary in order to
-%understand PS lipids in biological environments.
-
-The headgroup order parameters of PC lipids
-decrease proportionally to the bound positive
-charge in to a bilayer \cite{seelig87,catte16} and can be
-therefore used to measure the ion binding
-affinity. This molecular electrometer concept can
-be also applied to lipid bilayers with mixtures
-of PC and negatively charged lipids \cite{borle85,macdonald87,roux90}
-(see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-%This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
-%\ref{OrderParametersWithCaClBELOW1M} and
-%using the previuosly reported experimental data
-%for mixtures with varying proportions of
-%negatively charged PS or PG lipids.
-%
-%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
-%as a function of CaCl$_2$ concentration are shown
-%
-% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-
-%PC headgroup order parameters increase when negatively charged
-%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
-%as expected based on electrometer concept \cite{seelig87}
-%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
-%In electrometer concept this is explained by the tilting of
-%headgroup more parallel to membrane normal \cite{??}.
-
-%Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
-%At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
-%overcharging occurs above these concenterations.
-
-The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) mixtures
+Calcium ions form strong dehydrated molecular complexes with PS lipids,
+which can also lead to phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
+Such complexes and phase separation are not observed in
+bilayers that are sufficiently diluted with PC
+lipids \cite{feigenson86,mattai89,roux90,roux91}.
+Therefore, the Ca2+ binding affinity to bilayers if often measured
+from POPC:POPS (5:1) mixtures \cite{roux90,roux91}.
+
+The headgroup order parameter changes of POPC and POPS from 
 are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different

### 2018-06-24 Manuscript/manuscriptPS.tex
Update text

@@
 
-Calcium ions form strong dehydrated molecular complexes with PS lipids,
-which can also lead to phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-Such complexes and phase separation are not observed in
-bilayers that are sufficiently diluted with PC
-lipids \cite{feigenson86,mattai89,roux90,roux91}.
-Therefore, the Ca2+ binding affinity to bilayers if often measured
-from POPC:POPS (5:1) mixtures \cite{roux90,roux91}.
-
-The headgroup order parameter changes of POPC and POPS from 
-are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
-simulations and experiments \cite{roux90}. 
-The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
-The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
-MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
-It should be noted, however, that the lowest concentration (100mM) gives
-a good agreemet with experiments
-\todo{Should be analyze/discuss this further?
-  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
-It should be noted that Berger simulation do not have counterions.}.
-Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
-This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
-in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
-The binding of calcium to pure POPC bilayer is also too weak with these parameters
-as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
-\todo{The discussion is to be finished when we have all the data in the plot.}
+The dehydrated complexes of PS headgroup and calcium ions can also lead to the
+phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
+Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
+to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
+the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
+The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
+POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
+when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
+special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
+According to the electrometer concept, this means that the calcium binding
+seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
+is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
+The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
+PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
+CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
+binding to PC bilayers. The difference can be explained by the NBfix interaction
+parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
+the CHARMM-GUI at the time of running the simulations (January 2018).
+These parameters underestimate also the binding of calcium to pure POPC bilayers
+(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).
+
+%It should be noted, however, that the lowest concentration (100mM) gives
+%a good agreemet with experiments
+%\todo{Should be analyze/discuss this further?
+%  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
+%It should be noted that Berger simulation do not have counterions.}.
 
@@
 \end{figure*}
+
 \begin{figure}[]
@@
 
-Also the order parameters of PS headgroup from POPC:POPS (5:1) mixture
-are shown in Fig. \ref{changesWITHCaClPS} as a function of CaCl$_2$ concentration.
-In experiments the order parameters exhibit a strong dependence of CaCl$_2$ with
-small concentrations with a rapid saturation around 50 mM. 
-The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
-all tested simulation models. Furthermore, the changes of the headgroup order parameters
-do not qualitatively agree with experiments. This is in contrast to previous results
-for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
+The headgroup order parameters of POPS headgroup measured from POPC:POPS (5:1) mixture
+exhibit a strong dependence of CaCl$_2$ with small concentrations with a rapid saturation
+below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order parameter of
+POPS increase with the added CaCl$_2$ in the experiment and in all the tested simulation models,
+but simulations significantly overestimated the change. The larger $\alpha$-carbon order
+parameter of POPS decrease and the smaller one slightly increase with the added
+CaCl$_2$ in the experiment. The changes are again significantly overestimated in
+the simulations, however, in this case all simulations predict qualitatively different
+behaviour. Notably, the changes of POPS headgroup order parameters are overestimated
+also in the CHARMM36/NBfix model where the calcium binding affinity was too low. 
+We conclude that the effect of bound ions to the headgroup order parameters of POPS
+is not qualitatively reproduced by the tested simulations models. This is in contrast
+to previous results for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
 observed despite of significant discrepancies in the headgroup structure without additional ions. 
+The response of POPS headgroup order parameters to the bound charge is systematic but
+less well understood than the responce of PC headgroups used in the electrometer
+concept \cite{seelig87,roux90}. The force field development is necessary to
+generate MD simulations that could be used to explain the interactions between
+PS headgroup and calcium ions.
 

### 2018-06-25 Manuscript/manuscriptPS.tex
Update text and make separate file for SI

@@
 \usepackage{csquotes}
+\usepackage{xr}
+\externaldocument{manuscriptPSsuppl}
 
@@
 
-\subsection{Comparison of ion binding between simulations and experiments using the electrometer concept}
+\subsection{Comparison of ion binding to negatively charged lipid bilayers 
+between simulations and experiments using the electrometer concept}
 
-The headgroup order parameters of PC lipids
-decrease proportionally to the bound positive
-charge in to a bilayer \cite{seelig87,catte16} and can be
-therefore used to measure the ion binding
-affinity. This molecular electrometer concept can
-be also applied to lipid bilayers with mixtures
-of PC and negatively charged lipids \cite{borle85,macdonald87,roux90}
-(see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
-%This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
-%\ref{OrderParametersWithCaClBELOW1M} and
-%using the previuosly reported experimental data
-%for mixtures with varying proportions of
-%negatively charged PS or PG lipids.
-%
-%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
-%as a function of CaCl$_2$ concentration are shown
-%
-% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
+The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
+can be used to measure the ion binding affinity because they 
+decrease proportionally to the amound of bound positive
+charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
+This molecular electrometer concept is especially useful for 
+the comparison between simulations and experiments because
+the headgroup order parameters can be directly calculated from 
+simulations~\cite{catte16}. Also the headgroup order parameters
+of negatively charged PS and PG lipids exhibit systemic, but less
+characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
+Therefore, the ion binding affinity to negatively charged bilayers
+can be better characterized measuring the PC headgroup order parameters from 
+mixed bilayers, see section \ref{electrometerFORmixtures} in the supplementary information.
 
-%PC headgroup order parameters increase when negatively charged
-%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
-%as expected based on electrometer concept \cite{seelig87}
-%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
-%In electrometer concept this is explained by the tilting of
-%headgroup more parallel to membrane normal \cite{??}.
-
-%Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
-%At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
-%overcharging occurs above these concenterations.
+Before using the PC headgroup order parameters to quantify the ion binding
+affinity, it is important to quantify the response of the headgroup order
+parameters to the known amount of bound charge~\cite{catte16,melcr18}.
+This can be done using the experimental data from the mixtures of
+monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
+see section \ref{electrometerCALIBRATION} in the supplementary information.
+In this work, we also quantify the response of PC headgroup order parameters
+to the negatively charged PS headgroups, which also follows the electrometer
+concept in the experiments~\cite{scherer87},
+see section \ref{electrometerFORmixtures} in the supplementary information.
 
@@
 \end{acknowledgments}
-\pagebreak
-\appendix
-\begin{center}
-{\bf SUPPLEMENTARY INFORMATION}
-\end{center}
-
-\section{Simulated systems}
-
-\subsection{CHARMM36}
-\todo{To be written by Piggot, Madsen and Ollila}
-
-\subsection{CHARMM36ua}
-\todo{To be written by Piggot}
-
-\subsection{Slipids}
-\todo{To be written by Piggot and Favela}
-
-\subsection{Berger}
-\todo{To be wiritten by Piggot and Ollila}
-Simulations with sodium were taken directly from Ref. \citenum{??} and
-simulations with calcium directly from \citenum{melcrova16}.
-Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.
-
-\subsection{GROMOS-CKP}
-\todo{To be written by Piggot}
-
-\subsection{Lipid17}
-\todo{To be written by Kav and Miettinen}
-
-\subsection{MacRog}
-\todo{To be written by Javanainen and Piggot}
-
-\section{Cation binding affinity to lipid bilayers with different amount of charge}
-
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
-%  \caption{\label{OrderParametersWithCaCl}
-%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
-%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
-%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
-%  }
-%  \todo{Check the NaCl concentrations in the samples.}
-%\end{figure}
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
-%  \caption{\label{OrderParametersWithCaClBELOW1M}
-%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
-%  }
-%\end{figure}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
-  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    The change of PC headgroup order parameters as a function of CaCl$_2$
-    measured from bilayers containing different amount of negatively charged lipids.
-    The values are taken from 2H NMR experiments reported in the
-    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
-    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
-    and POPC:POPG (1:1) without NaCl \cite{borle85}).
-    As expected, the decrease of order parameters with the added CaCl$_2$
-    is more pronounced for systems with larger fraction of
-    negatively charged lipids, indicating larger amount amount
-    of bound cations.
-  }
-\end{figure}
-
-Before using the headgroup order parameters to compare ion binding affinity between simulations
-and experiments, it is important to quantify the response of the order parameters to the
-bound charge in simulations.
-The response of headgroup order parameters to the fixed amount of cationic surfactants in
-POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
-The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
-while CHARMM36 is in better agreement with experiments. This has to be taken into account when
-analysin the binding affinities.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
-  \caption{\label{CHANGESwithCaClPGPS}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
-\end{figure}
-
-\section{Difference between POPC and OPPS in MacRog model}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
-  \caption{\label{CHANGESwithCaClPGPS}
-    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
-\end{figure}
-
-\section{Dihedrals}
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops1.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops2.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops3.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops4.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops5.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops6.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops7.png}
-  \caption{\label{dihedralsGLY}
-    Dihedral angle distributions of bonds from phosphate to acyl chain carbonyls from different simulation models.
-  }
-\end{figure*}
-\section{Dihedrals}
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops8.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops9.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops10.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops11.png}
-  \caption{\label{dihedralsHG}
-    Dihedral angle distributions of bonds from phosphate to headgroup from different simulation models.
-  }
-\end{figure*}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
-  \caption{\label{PSresponseTONaClDMPC}
-    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
-  }
-\end{figure}
-The experimental results show essentially no changes in the order parameters as a function of
-added NaCl, while significant changes are observed in simulations. However,
-the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
-Therefore, we cannot exclude the possibility that the NaCl induced changes were already
-saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
-  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
-\end{figure}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
-  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
-\end{figure}
-
-\pagebreak
-\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
-
-The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
-%
-First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
-we looked separately at deviations in magnitude and forking.
-
-{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
-the force-field-produced OPs were.
-%
-For each carbon, the following 5-step scale was used:
-%
-\begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
-%
-\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
-%
-\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
-%
-\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
-%
-\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
-\end{description}
-
-{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
-Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.
-For the $\alpha$ carbon,  for which a considerable forking of 0.105 is experimentally seen, the following 5-step scale was used:
-\begin{description}
-\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
-%
-\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.06 < D < 0.08)$ OR $(0.13 < D < 0.15)$.}
-%
-\item [2  ({\textsf{\small F}}):] \noindent {$(0.04 < D < 0.06)$ OR $(0.15 < D < 0.17)$.}
-%
-\item [3 ({\textsf{\large F}}):] \noindent {$(0.02 < D < 0.04)$ OR $(0.17 < D < 0.19)$.}
-%
-\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.02)$ OR $(0.19<D)$.}
-\end{description}
-%
-For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was used:
-%
-\begin{description}
-\item [0 (~):] \noindent {$ D< 0.02$.}
-%
-\item [1 ({\textsf{\tiny F}}):] \noindent {$0.02 < D < 0.04$.}
-%
-\item [2  ({\textsf{\small F}}):] \noindent {$0.04 < D < 0.06$.}
-%
-\item [3 ({\textsf{\large F}}):] \noindent {$0.06 < D < 0.08$.}
-%
-\item [4 ({\textsf{\Large F}}):] \noindent {$0.08 < D$.}
-\end{description}
-%
-For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was used:
-%
-\begin{description}
-\item [0 (~):] \noindent {$0.11 < D < 0.15$.}
-%
-\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.09 < D < 0.11)$ OR $(0.15 < D < 0.17)$.}
-%
-\item [2  ({\textsf{\small F}}):] \noindent {$(0.07 < D < 0.09)$ OR $(0.17 < D < 0.19)$.}
-%
-\item [3 ({\textsf{\large F}}):] \noindent {$(0.05 < D < 0.07)$ OR $(0.19 < D < 0.21)$.}
-%
-\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.05)$ OR $(0.21<D)$.}
-\end{description}
-
-Based on these assessments of magnitude and forking deviations,
-each carbon was then assigned to one of the following groups:
-"within experimental error"
-(magnitude and forking deviations both on step 0 of the scales described above),
-"almost within experimental error"
-(sum of the magnitude and forking deviation steps 1 or 2),
-"clear deviation from experiments"
-(sum of magnitude and forking deviation steps from 3 to 5), and
-"major deviation from experiments"
-(sum of magnitude and forking deviation steps from 6 to 8).
-These groups are indicated by colors in Fig.~4.
-(Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
-the corresponding group assigment limits were: 0, 1, 2, and 3.)
-
-Finally, the total ability of the force field to describe the headgroup and
-glycerol structure was estimated.
-To this end, the groups were given the following weights:
-0 (within experimental error),
-1 (almost within experimental error),
-2 (clear deviation from experiments),
-4 (major deviation from experiments),
-and the weights of the five carbons were summed up.
-The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
-was then used to (roughly and subjectively, as should be clear from the
-above description) rank the force fields.
 

### 2018-06-25 Manuscript/manuscriptPSsuppl.tex
Update text and make separate file for SI

\documentclass[journal=jpcbfk]{achemso}

\usepackage[version=3]{mhchem}
\usepackage[T1]{fontenc}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

\usepackage{upgreek}				
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{lmodern}
\usepackage{microtype}

\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
\affiliation{nmrlipids.blogspot.fi}

\SectionNumbersOn

\renewcommand{\thetable}{S\arabic{table}}%
\renewcommand{\thefigure}{S\arabic{figure}}%
\renewcommand{\thesection}{S\arabic{section}}%
\renewcommand{\thepage}{S\arabic{page}}%

\title{ Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids}

\begin{document}

\newpage
%\tableofcontents

\section{Simulated systems}

\subsection{CHARMM36}
\todo{To be written by Piggot, Madsen and Ollila}

\subsection{CHARMM36ua}
\todo{To be written by Piggot}

\subsection{Slipids}
\todo{To be written by Piggot and Favela}

\subsection{Berger}
\todo{To be wiritten by Piggot and Ollila}
Simulations with sodium were taken directly from Ref. \citenum{??} and
simulations with calcium directly from \citenum{melcrova16}.
Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.

\subsection{GROMOS-CKP}
\todo{To be written by Piggot}

\subsection{Lipid17}
\todo{To be written by Kav and Miettinen}

\subsection{MacRog}
\todo{To be written by Javanainen and Piggot}

\section{Electrometer concept in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89}. 
The concept is also valid for lipid mixtures where PC order parameters 
increase when mixed with anionic lipids, in contrast to the mixtures with neutral or
cationic surfactants~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

Based on the electrometer concept, the headgroup order parameters of
PC lipids can be used to measure the ion binding affinity to lipid 
bilayers \cite{akutsu81,altenbach84,seelig87}. Changes of the headgroup order 
parameters of negatively charged PS and PG lipids with bound charge are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers. The PC headgroup order parameters in mixtures with anionic
lipids are larger than in pure PC lipid because addition of
negative charges increase the order parameters (Figs. \ref{HGorderparametersPCvsPEPSPGchol} 
and \ref{OrderParametersWithCaCl}). Upon addition of CaCl$_2$, the PC headgroup order 
parameters in the mixed bilayers decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}). At this point the charge of
bound Ca$^{2+}$ probably cancels the charge of anionic negative lipids and the bilayer 
has neutral charge. The overcharging then occurs with larger concenterations.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments 
    different mole fractions of negatively charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
    The right column shows the results zoomed to concentrations below 1M.
  }
\end{figure}
As expected, the more pronounced order parameter decrease is observed in systems with more
negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}) indicating 
higher binding affitiny of cations into the negatively charged membranes. In conclusion,
the consistency of the experimental results suggests that the electrometer concept can
be used to determine the cation binding also to the lipid bilayers containing negatively
charged lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters as a function of CaCl$_2$
    measured from bilayers containing different amount of negatively charged lipids.
    The values are taken from 2H NMR experiments reported in the
    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
    and POPC:POPG (1:1) without NaCl \cite{borle85}).
    As expected, the decrease of order parameters with the added CaCl$_2$
    is more pronounced for systems with larger fraction of
    negatively charged lipids, indicating larger amount amount
    of bound cations.
  }
\end{figure}

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, it is important to quantify the response of the order parameters to the
bound charge in simulations~\cite{catte16,melcr18}. Comparison of the headgroup order parameter
response to the cationic surfactants in POPC bilayer to the experiments \cite{scherer89} shows
that the both order parameters are too sensitive to the bound charge in the Lipid14 model,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
The ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ from Lipid14 model was in good agreement 
with experiments in the previous study~\cite{catte16} because both order parameters are equally too 
sensitive to the bound cations (Fig. \ref{CHANGESwithDHMDMAB}). The ratio was, however, overestimated
for the CHARMM36 model because the $\beta$ order parameter relatively more sensitive than the $\alpha$
order parameter. These have to be taken into account when analysing the ion binding affinities
using headgroup order parameters in simulations.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

\pagebreak
\section{Difference between POPC and OPPS in MacRog model}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
  \caption{\label{CHANGESwithCaClPGPS}
    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
\end{figure}

\pagebreak
\section{Dihedrals}
\begin{figure*}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/diheds_pops1.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops2.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops3.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops4.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops5.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops6.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops7.png}
  \caption{\label{dihedralsGLY}
    Dihedral angle distributions of bonds from phosphate to acyl chain carbonyls from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/diheds_pops8.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops9.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops10.png}
  \includegraphics[width=8.0cm]{../Figs/diheds_pops11.png}
  \caption{\label{dihedralsHG}
    Dihedral angle distributions of bonds from phosphate to headgroup from different simulation models.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaClDMPC}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

\pagebreak
\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
%
First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
we looked separately at deviations in magnitude and forking.

{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
the force-field-produced OPs were.
%
For each carbon, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
%
\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
%
\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
%
\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
%
\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
\end{description}

{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally. Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen. For the $\alpha$ carbon,  for which a considerable forking of 0.105 is experimentally seen, the following 5-step scale was used:
\begin{description}
\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.06 < D < 0.08)$ OR $(0.13 < D < 0.15)$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$(0.04 < D < 0.06)$ OR $(0.15 < D < 0.17)$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$(0.02 < D < 0.04)$ OR $(0.17 < D < 0.19)$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.02)$ OR $(0.19<D)$.}
\end{description}
%
For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {$ D< 0.02$.}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$0.02 < D < 0.04$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$0.04 < D < 0.06$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$0.06 < D < 0.08$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$0.08 < D$.}
\end{description}
%
For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {$0.11 < D < 0.15$.}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.09 < D < 0.11)$ OR $(0.15 < D < 0.17)$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$(0.07 < D < 0.09)$ OR $(0.17 < D < 0.19)$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$(0.05 < D < 0.07)$ OR $(0.19 < D < 0.21)$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.05)$ OR $(0.21<D)$.}
\end{description}

Based on these assessments of magnitude and forking deviations,
each carbon was then assigned to one of the following groups:
"within experimental error"
(magnitude and forking deviations both on step 0 of the scales described above),
"almost within experimental error"
(sum of the magnitude and forking deviation steps 1 or 2),
"clear deviation from experiments"
(sum of magnitude and forking deviation steps from 3 to 5), and
"major deviation from experiments"
(sum of magnitude and forking deviation steps from 6 to 8).
These groups are indicated by colors in Fig.~4.
(Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
the corresponding group assigment limits were: 0, 1, 2, and 3.)

Finally, the total ability of the force field to describe the headgroup and
glycerol structure was estimated.
To this end, the groups were given the following weights:
0 (within experimental error),
1 (almost within experimental error),
2 (clear deviation from experiments),
4 (major deviation from experiments),
and the weights of the five carbons were summed up.
The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
was then used to (roughly and subjectively, as should be clear from the
above description) rank the force fields.

\newpage
\bibliography{refs.bib}

\end{document}

### 2018-06-25 Manuscript/manuscriptPS.tex
Update text

@@
 Previous experimental studies have concluded that
-PS headgroups have more rigid conformation than
-phophocholines (PC) due to hydrogen bonding or
+PS headgroups are more rigid than phophocholines (PC)
+due to the hydrogen bonding network or
 electrostatic interactions \cite{browning80,buldt81}.
@@
 while monovalent ions interact more weakly with PS
-containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
+containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+On the other hand, some studies propose that the specific binding
+affinity is similar to the negatively charged and zwitterionic lipids and that
+the increased cation binding to negatively charged lipid bilayer arise only due
+to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
 Dilution of bilayers with PC lipids makes PS headgroups
@@
 strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
+The molecular level interpretation of these observations is,
+however, not available.
 
-The structure of PS lipid headgroups and their 
-interactions with ions have been studied with
-various experimental methods and theoretical techniques \cite{roux90,melcrova16,??}.
+Several classical molecular dynamics (MD) simulation studies are done
+to understand PS headgroups, their influence on lipid bilayer properties and
+interactions with ions \cite{vernier09,boettcher11,melcrova16,??}.
 However, the consesus has not been reached 
@@
 headgroup region \cite{botan15,catte16,ollila16}.
-Some studies propose that the negatively charged lipids
-attract cations only due to the increase of local concentration
-in the vicinity of membranes and that the binding constant of cations is similar
-to zwitterionic and negatively charged lipids \cite{seelig90,sinn06,??}.
-On the other hand, some studies propose specific binding of calcium directly to PS lipid
-headgroups \cite{vernier09,boettcher11,??},
-The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
-headgroups, but more detailed interperetation has not been done.
 

### 2018-06-25 Manuscript/manuscriptPS.tex
Update text

@@
 to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with ions \cite{vernier09,boettcher11,melcrova16,??}.
-However, the consesus has not been reached 
-due to the difficulties to interpret the experimental data \cite{??} and
-the inaccuracies in simulation models at the
-headgroup region \cite{botan15,catte16,ollila16}.
-
-%Details of negatively charged lipids and their interactions
-%with ions and proteins have been experimentally analyzed with
-%various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
-%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
-%which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.
-
-Headgroup and glycerol backbone C-H bond order parameters
-calculated from MD simulations have been
-recently used to interpret the lipid structures in NMR experiments
-and to validate lipid structure and ion binding in simulations of
-PC lipid bilayers \cite{botan15,catte16,ollila16,ferreira16}.
-In this work we apply this approach to PS lipid headgroup
-in order to elucidate the structural details and ion binding
-to negatively charged lipids. The results are expected to elucidate
-also PS mediated signalling events because 
-glycerol backbone and headgroup structure and behaviour are similar
-in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.
+interactions with ions \cite{vernier09,boettcher11,pan14,melcrova16,??}.
+However, the recent comparisons of PC lipid headgroup and glycerol backbone
+C-H bond order parameters calculated from different simulation models
+revealed that improvements in the current force fields are needed
+to correctly reproduce the headgroup structure and ion binding to
+lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
+to POPC bilayer was then improved by implicitly including the electronic
+polarizability using the electronic continuum correction \cite{melcr18}.
+Here, we collect the set of experimentally measured lipid headgroup and
+glycerol backbone C-H bond order parameters, which can be used to
+evaluate the quality of headgroup structure and the ion binding affinity
+in MD simulations of lipid bilayers containing PS lipids. 
+The available MD simulation models of PS are then compared against
+the collected experimental data. The results pave the way for the
+development of MD simulation force fields that correctly describe
+PS lipid headgroup structure and its interactions with ions.
+Such models are expected to be useful in elucidating the biological
+role of PS and other lipid headgroups because glycerol backbone and
+lipid headgroups behave similarly in model membranes and in
+bacteria \cite{gally81,scherer87,seelig90}.
 

### 2018-06-27 Manuscript/manuscriptPS.tex
Update text

@@
 while monovalent ions interact more weakly with PS
-containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
 On the other hand, some studies propose that the specific binding
@@
 to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with ions \cite{vernier09,boettcher11,pan14,melcrova16,??}.
+interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
 However, the recent comparisons of PC lipid headgroup and glycerol backbone
@@
 
-\subsection{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}
+\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}
 
@@
 Since only the absolute values of the PS headgroup order parameters were measured
-previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{??} to determine the signs of
+previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
 the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
@@
 
-The headgroup order parameters of DOPS and POPS bilayers 
-from different simulation models are compared with the experimental
-data in Fig. \ref{HGorderParametersPS}. Subjective ranking
-of the quality of the models is shown in Fig. \ref{comparisonTablePS}.
-The tested models perform generally less well than in the previous
-study for PC headgroup~\cite{botan15} and none of the models 
-reproduce the experimental order parameters within
-experimental error bars.
+None of the tested models reproduce the experimental headgroup and 
+glycerol backbone order parameters of DOPS and POPS within the
+experimental error bars (Fig. \ref{HGorderParametersPS}). 
+The tested models perform generally less well than the models tested
+for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
+which is also evindent from the comparison between subjective rankings
+of the model quality for the headgroup and glycerol backbone 
+(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
 Therefore, the models cannot be straightforwardly used to interpret the
-structure of PS headgroup. However, the differences between PC and PS headgroups
-are partially reproduced by some of the models.
-
-The two best performing models for the $\alpha$ and $\beta$-carbons of
-PS, Slipids and CHARMM36, reproduce the large forking in the $\alpha$-carbon
-and the Slipids model gives also a good agreement with experiments for
-the $\beta$-carbon order parameter in both PC and PS headgroups
-(Fig. \ref{HGorderParametersPS} and Ref.~\citenum{botan15}).
-Interestingly, the dihedral angle distributions of CHARMM36
-and Slipids in Fig.~\ref{dihedralsHG} share significant similarities
-in the headgroup region. However, the experimental order parameters in the
-glycerol backbone region are not well reproduced by the Slipids model,
-which was also the case in for PC lipids \cite{botan15}.
-This difference probably arises from the differences in
-the dihedral angle distributions of C1-C2-C3-O31 and C2-C3-O31-C31
-in Figure \ref{dihedralsGLY}, which are also illustrated in
-Figure \ref{glycerol_buslaev}.
+structural differences between PC and PS headgroups. 
+However, the differences are partially reproduced by the two best performing 
+models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
+Both reproduce the larger forking of the $\alpha$-carbon 
+and the Slipids model reproduces also the lower of the $\beta$-carbon order 
+in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
+Interestingly, the dihedral angle distributions 
+%of CHARMM36 and Slipids in 
+in these two models share significant similarities in the headgroup region (Fig.~\ref{dihedralsHG}). 
+\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
+chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
+However, the glycerol backbone order parameters in Slipids model
+significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
+which can be related to the differences in the dihedral angle distributions 
+of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}). 
+Similar difference was previously observed for PC lipids \cite{botan15}
+and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}.
 
@@
 \todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
-  \caption{\label{glycerol_buslaev}
-    Snapshots overlayed from different simulations for glycerol backbone region
-    by Pavel Buslaev.
-  }
-\end{figure}
 

### 2018-06-28 Manuscript/manuscriptPS.tex
Add abstract

@@
   % insert abstract here
-  Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine
-  (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between 
-  different lipid molecules in bilayers.
+Phosphatidylserine (PS) is the most common negatively
+charged lipid in eykaryotic membranes.
+PS lipids interact with signaling and other proteins via
+electrostatic interactions and direct binding, and induce
+membrane fusion and phase separation together with calcium ions.
+Molecular details of these phenomena are not well understood
+because accurate models to interpret the experimental data has not
+been available. Here, we collect a set of experimental NMR data which
+could be used together with molecular dynamics (MD) simulations
+to interpret the lipid headgroup structures and details of ion binding
+in pure and mixed PS and PS:PC lipid bilayers. Aiming to interpret the
+data, we use the open collaboration method to go through the available MD simulation models for PS lipids.
+However, none of the models reproduce the experimental data with sufficient
+accuracy to interpet the structural details of lipid headgroups or
+ion binding details in lipid bilayers containing PS lipids.
+In contrast to PC lipids, the tested MD simulation models do not
+correctly reproduce the qualitative response of PS lipid headgroups
+to the bound ions or changes in the lipid composition. 
+Our results pave the way for the model improvement to correctly
+describe negatively charged membranes and their interactions with ions.
 \end{abstract}
@@
 
-\section{Methods}
-
-\subsection{Solid state NMR experiments}
-The magnitude and signs of the C-H bond order parameters in
-headgroup and glycerol backbone were measured
-using natural abundance $^{13}$C solid state NMR spectroscopy as
-described previously \cite{ferreira13,ferreira16}.
-Shortly, the absolute values of the order parameters were determined from the dipolar splittings
-given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
-the signs were measured using S-DROSS experiments \cite{gross97}.
-
-\todo{Details of the used spectrometer and maybe some other details should be given.} \\
-\todo{Sample preparation should be described.} \\
-\todo{How is the peak assignment done?}
-
-\subsection{Molecular dynamics simulations}
-Molecular dynamics simulation data was collected using
-the Open Collaboration method \cite{botan15}.
-The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
-the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
-were used as the communication platforms.
-The simulated systems are listed in Table \ref{IONsystems} and
-simulation details are given in the SI.
-The simulation data is also indexed in the
-searchable database (\url{nmrlipids.fi}),
-and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).  
 \begin{table*}[!htb]
@@
 \end{table*}
+
+\section{Methods}
+
+\subsection{Solid state NMR experiments}
+The magnitude and signs of the C-H bond order parameters in
+headgroup and glycerol backbone were measured
+using natural abundance $^{13}$C solid state NMR spectroscopy as
+described previously \cite{ferreira13,ferreira16}.
+Shortly, the absolute values of the order parameters were determined from the dipolar splittings
+given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
+the signs were measured using S-DROSS experiments \cite{gross97}.
+
+\todo{Details of the used spectrometer and maybe some other details should be given.} \\
+\todo{Sample preparation should be described.} \\
+\todo{How is the peak assignment done?}
+
+\subsection{Molecular dynamics simulations}
+Molecular dynamics simulation data was collected using
+the Open Collaboration method \cite{botan15}.
+The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
+the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
+were used as the communication platforms.
+The simulated systems are listed in Table \ref{IONsystems} and
+simulation details are given in the SI.
+The simulation data is also indexed in the
+searchable database (\url{nmrlipids.fi}),
+and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).
 

### 2018-06-28 Manuscript/manuscriptPS.tex
Add preliminary version of conclusions section

@@
 }\label{IONsystems}
+    \todo{Berger simulations with NaCl and CaCl to be added}
 %begin{minipage}[t]{\textwidth}
@@
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    \todo{MacRog simulations with KCl to be added}\\
     \hline
-    \todo{Berger simulations with NaCl and CaCl to be added}
 \end{tabular}
@@
 \section{Conclusions}
+We have collected a set of experimental NMR order parameter data,
+which could be combined with MD simulations to interpret the
+headgroup structure and cation binding details to negatively charged
+membranes containing PS lipids. Using open collaboration method, we
+tried to find a MD simulation model which would be sufficiently accurate
+to interpret the experimental data. However, none of the tested models
+was accurate enough. In line with the previous study for PC lipids \cite{catte16},
+MD simulation models seems to generally overestimate cation binding also
+to negatively charged bilayers containing PS lipids, with some exceptions.
+The response of PS lipid headgroup order parameters to the bound cations
+does not agree with experiments, even in the cases where binding affinity is
+not overestimated. This is in contrast to the previous results with PC lipids,
+where the qualitative response of the headgroup order parameters was in agreement
+with experiments even in the cases where the headgroup structure without ions was not
+correct and the cation binding affinity was overestimated. In addition, the inaccurate
+responses of PS headgroup order parameters to the dilution with PC lipids suggests
+that the PC-PS interactions are not accurately described by the tested models.
+
+Our results pave the way for improving the PS lipid parameters for MD simulations
+by offering the set of experimental data for the quality measurement, by
+pinpointing problems areas in the models and suggesting directions for the corrections.
+Improvements using the electronic continuum correction is already in progress \url{https://github.com/jmelcr/ecc_lipids},
+following the recent work for PC lipids \cite{melcr18}.
 

### 2018-06-29 Manuscript/manuscriptPGPE.tex

@@
 \subsection{Effect of PE, PS and PG on PC headgroup}
+, the headgroup 
+order parameters increase with the addition of negatively charged PS and PG lipids,
+decrease when mixed with positively charged surfactants and
+are less affected by the addition of zwitterionic PE lipids or cholesterol.
+In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
+also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
+follow the eletrometer concept \cite{scherer87}.
+to mixtures of differently charged
+lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
+
+%\section{PC lipid headgroup response to different mixtures in experiments}
+As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
+headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
+i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
+surfactants. The changes with the addition of neutral lipids is significantly smaller.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
+  \caption{\label{HGorderparametersPCvsPEPSPGchol}
+    PC headgroup order parameters from experiments of mixtures with
+    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
+    Signs are determined as discussed in \cite{botan15,ollila16}.
+  }
+\end{figure}
+
 The headgroup order parameters for PC lipids (POPC and DOPC)

### 2018-07-03 Manuscript/manuscriptPSsuppl.tex
Update text

@@
 \usepackage{microtype}
+\usepackage{xr}
+\externaldocument{manuscriptPS}
 
@@
 \subsection{Berger}
-\todo{To be wiritten by Piggot and Ollila}
-Simulations with sodium were taken directly from Ref. \citenum{??} and
+\todo{To be written by Piggot and Ollila}
+Simulations with excess sodium were taken directly from Ref. \citenum{jurkiewicz12} and
 simulations with calcium directly from \citenum{melcrova16}.
@@
 \subsection{Lipid17}
-\todo{To be written by Kav and Miettinen}
+\todo{To be written by Kav, Miettinen and Meclr.}
 
@@
   \centering
+  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
+  \caption{\label{glycerol_buslaev}
+    Snapshots overlayed from different simulations for glycerol backbone region
+    by Pavel Buslaev.
+  }
+\end{figure}
+\pagebreak
+
+\section{Sodium binding to DMPC:DOPS mixture}
+
+\begin{figure}[]
+  \centering
   \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
@@
 saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
+\pagebreak
+
+\section{Calcium bindinf to POPC in CHARMM36 simulation with NBfix}
 

### 2018-07-04 Manuscript/manuscriptPS.tex
Update authors and small thinbgs

@@
 
-\author{O. H. Samuli Ollila}
+\author{Tiago M. Ferreira}
+\affiliation{Halle, Germany}
+\author{Josef Melcr}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{O. H. Samuli Ollila  \todo{Authorlist is not yet complete}}
 \email[]{samuli.ollila@helsinki.fi}
@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
-
-\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
-\affiliation{nmrlipids.blogspot.fi}
+%\author{NMRlipids collaboration}
+%\affiliation{nmrlipids.blogspot.fi}
 
@@
 to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
+interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16}.
 However, the recent comparisons of PC lipid headgroup and glycerol backbone
@@
     POPS/Na$^+$  & Berger \cite{??}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & ? & ?? & 0  & ?  & ? & ?  & \cite{??} \todoi{Data to be added by Piggot} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
     OPPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
@@
   \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
-  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.}
+  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
+  \todo{The data for Gromos-CKP by Tom Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1528213129976\#c8319083196182661431
+    should be added into the plot.}
 \end{figure*}

### 2018-07-04 Manuscript/manuscriptPS.tex
Update authorlist and other details

@@
 \affiliation{Halle, Germany}
+\author{Matti Javanainen}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{Markus Miettinen}
+\affiliation{Potsdam, Germany}
 \author{Josef Melcr}
@@
 \centering
-\caption{List of MD simulations without additional salt.
+\caption{List of MD simulations of pure PS bilayers without additional salt.
    CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
    (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges)
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
    and CKP refers to the version with more Gromos compatible version
@@
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
+\begin{tabular}{l c r r r r r c c}
   %\hline
   % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+ lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    &0 & 0        & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
+    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
     \hline
-    POPS/Na$^+$  & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
-    POPS/K$^+$   & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
-    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
-    POPS/Na$^+$  & Berger \cite{??}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
-    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
-    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
-    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
-    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
-    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
+    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
+    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
+    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
+    POPS/Na$^+$  & Berger \cite{??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
+    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
+    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
+    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
     \end{tabular}
@@
 \centering
-\caption{  List of MD simulations. The salt concentrations calculated as 
-   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{??})
-   and CKP refers to the version with more Gromos compatible version
-   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
+\caption{List of POPC:POPS mixture simulations with different amounts of added ions. 
+   % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
@@
 }\label{IONsystems}
-    \todo{Berger simulations with NaCl and CaCl to be added}
+    \todo{Berger simulations with NaCl and CaCl to be added} \\
+    \todo{The used definition of ion concentrations needs to be specified and standardized.} 
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
-  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+\begin{tabular}{l c c c c c c c c c c}
+  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) & CaCl$_2$\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?
-& 0  & 298  & 200 & ?
-& \cite{??} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
     POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?
-& 0  & 298  & 200 & ?
-& \cite{??} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
-     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked} & 250:50 & ?
 & ?
@@
 Madsen}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked} & 250:50 & ?
 & ?
@@
     \hline
-    POPC:POPS\todoi{This is also probably OPPS?
-These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.1 & 120:24 & 5760 & 10   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.3 & 120:24 & 5760 & 31   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1   & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3   & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.5  & 0   & 120:24 & 5760 & 52   & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1    & 0   & 120:24 & 5760 & 104  & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2    & 0   & 120:24 & 5760 & 208  & 298  & 200 & 145 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3    & 0   & 120:24 & 5760 & 311  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4    & 0   & 120:24 & 5760 & 415  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
     \hline
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5  & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    \hline
+    POPC:POPS (4:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (4:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
 \end{tabular}
 %\end{minipage}
-%\end{sidewaystable*}
+%end{sidewaystable*}
 
@@
 see section \ref{electrometerFORmixtures} in the supplementary information.
+
+\todo{How close are the hydration levels in experiments and simulations?} \\
+\todo{We need to decide how to report the ion concentrations, see discussion in Ref.~\citenum{melcr18}.}
 

### 2018-07-04 Manuscript/manuscriptPS.tex
Update
Update text

@@
 
-\begin{table*}[!htb]
+\begin{table*}[!p]
 %\begin{sidewaystable*}[!p]
@@
 }\label{IONsystems}
-    \todo{Berger simulations with NaCl and CaCl to be added} \\
     \todo{The used definition of ion concentrations needs to be specified and standardized.} 
@@
     \hline
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0  & 0     & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
     POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5  & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
@@
     POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0   & 0   & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5  & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.03 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.17 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.36 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
     \hline
-    POPC:POPS (4:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (4:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & ? & ? & \cite{??} \todoi{To be added by Ollila}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.12   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
 \end{tabular}

### 2018-07-17 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
ffavela section added.

@@
 \todo{To be written by Piggot and Favela}
+{\it DOPC with 288 lipids.}
+The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions.
+The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
+Simulation was performed for 200~ns, and the last 100~ns were used for the analysis.
+Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}.
+Timestep of 2~fs was used with the leapfrog integrator.
+The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath.
+Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
+barostat~\cite{parrinello81} with a time constant of 10.0~ps.
+Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}.
+A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space.
+Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure.
+All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}.
+Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.
 

### 2018-08-07 Manuscript/manuscriptPS.tex
Update authors
Add authors
Update GROMOS-CKP results into the manuscript

@@
 \affiliation{Halle, Germany}
+\author{Ivan Gushchin}
+\affiliation{Moscow Institute of Physics and Technology}
 \author{Matti Javanainen}
@@
 Prague 6, Czech Republic}
+\author{Batuhan Kav}
+\affiliation{Potsdam, Germany}
+\author{Jesper J. Madsen}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
 \author{Markus Miettinen}
@@
 Prague 6, Czech Republic}
-\author{O. H. Samuli Ollila  \todo{Authorlist is not yet complete}}
+\author{O. H. Samuli Ollila}
 \email[]{samuli.ollila@helsinki.fi}
@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
+\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
+\affiliation{Southampton, United Kingdom}
 %\author{NMRlipids collaboration}
@@
     POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
+    \hline
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKP \cite{??}             &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPwithNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKPM \cite{??}            &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPMwithNa}  \\
 \end{tabular}
@@
   \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
-  \todo{The data for Gromos-CKP by Tom Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1528213129976\#c8319083196182661431
-    should be added into the plot.}
+  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
 \end{figure*}

### 2018-08-09 Manuscript/manuscriptPS.tex
Updates in the text

@@
 containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+The dehydrated complexes of PS headgroup and calcium ions can also lead to the
+phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
 On the other hand, some studies propose that the specific binding
@@
    %   The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
+}\label{PSsystems}
 %begin{minipage}[t]{\textwidth}
@@
    % The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
+}\label{mixedIONsystems}
     \todo{The used definition of ion concentrations needs to be specified and standardized.} 
@@
 were used as the communication platforms.
-The simulated systems are listed in Table \ref{IONsystems} and
-simulation details are given in the SI.
+The simulated systems are listed in 
+Tables \ref{PSsystems} (pure PS systems without additional ions) 
+and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
+Simulation details are given in the SI.
 The simulation data is also indexed in the
@@
 Therefore, the ion binding affinity to negatively charged bilayers
-can be better characterized measuring the PC headgroup order parameters from 
-mixed bilayers, see section \ref{electrometerFORmixtures} in the supplementary information.
+can be better characterized by measuring the PC headgroup order parameters from 
+mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.
 
@@
 
-\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}
+\subsection{Headgroup and glycerol backbone order parameters of POPS from $^{13}$C NMR}
 
@@
 from the small splitting cannot resolved with the available resolution.
-Since only the absolute values of the PS headgroup order parameters were measured
-previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
-the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
+Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
+only the absolute values of order parameters, we determined the signs of PS headgroup
+and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
+The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
 order parameter is negative, which is confirmed by SIMPSON simulations.
@@
   }
-  \todo{This is preliminary figure, should be polished.}
+  \todo{This is preliminary figure, should be polished.} \\
   \todo{Should we show slices for all the analyzed carbons in (c)?}
@@
     Headgroup and glycerol backbone order parameters of POPS measured in this work compared
-    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
-    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
-    as measured in this work and signs for PC as measured in Refs \cite{??,ferreira16}.
-    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
-    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
+    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
   }
-  \todo{There should be values in \cite{roux90} which should be added.}
 \end{figure}
@@
 
-\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
+\todo{POPS/OPPS issue with MacRog model is in progress: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
 
@@
 \todo{More detailed discussion may be possible after comparing monovalent ion binding
-to bilayers between CHARMM simulations and experiments.}
+to bilayers between CHARMM simulations and experiments.
+Also, section S6 should be finished.}
 Area per lipid is in agreement with experiments \cite{pan14} only
@@
 
-The dehydrated complexes of PS headgroup and calcium ions can also lead to the
-phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
-to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
-the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
-The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
-POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
-when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
-special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
-According to the electrometer concept, this means that the calcium binding
-seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
-is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
-The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
-PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
-CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
-binding to PC bilayers. The difference can be explained by the NBfix interaction
-parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
-the CHARMM-GUI at the time of running the simulations (January 2018).
-These parameters underestimate also the binding of calcium to pure POPC bilayers
-(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).
+Ion binding affinity to PS containing membranes can be most conveniently measured 
+from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
+not observed \cite{feigenson86,mattai89,roux90,roux91}. Also, the headgroup order 
+parameters of PC can be used to detect ion binding affinity to the mixed lipid 
+bilayers, see section~\ref{electrometerFORmixtures}. As expected from the previous study of PC lipid bilayers \cite{catte16}, 
+the decrease of POPC headgroup order parameters in POPC:POPS mixtures as a function of Ca$^{2+}$ concentration 
+is overestimated with respect to experiments \cite{roux90} in almost all the 
+tested simulation models (Fig. \ref{changesWITHCaClPS}), indicating overestimated calcium binding
+binding affinity. Only exception is the CHARMM36 model with the special NBfix~\cite{kim16} interaction, 
+incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
+which underestimates the calcium binding affinity. With this interaction, the 
+calcium and sodium binding affinities to POPC bilayer are equally weak (see section S7), in contrast to experimental 
+data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the ion binding peaks in
+density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}) are underestimated 
+in the CHARMM36/NBfix model, but overestimated in the other tested models.
 

### 2018-08-09 Manuscript/manuscriptPSsuppl.tex
Updates in the text

--- 2018-07-03 Manuscript/manuscriptPSsuppl.tex
@@
 decrease (increase) proportionally to the bound positive (negative) 
-charge~\cite{akutsu81,altenbach84,seelig87,scherer89}. 
-The concept is also valid for lipid mixtures where PC order parameters 
-increase when mixed with anionic lipids, in contrast to the mixtures with neutral or
-cationic surfactants~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
+charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
+Therefore, the headgroup order parameters can be used to measure the
+ion binding affinity to lipid bilayers containing PC
+lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
+Changes of the headgroup order parameters of other lipids with bound charge,
+like negatively charged PS and PG lipids,  are also systematic,
+but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
+Therefore, the ion binding affinity to negatively charged bilayers
+can be better characterized measuring the PC headgroup order parameters from 
+mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
 \begin{figure}[]
@@
   \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    PC headgroup order parameters from experiments of mixtures with
-    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
-    Signs are determined as discussed in \cite{botan15,ollila16}.
+    Headgroup order parameters of POPC measured from mixtures with
+    PS (bovine brain), POPG, POPE, cholesterol and cationic surfactant~\cite{scherer87,scherer89,ferreira13}.
+    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
   }
@@
 
-Based on the electrometer concept, the headgroup order parameters of
-PC lipids can be used to measure the ion binding affinity to lipid 
-bilayers \cite{akutsu81,altenbach84,seelig87}. Changes of the headgroup order 
-parameters of negatively charged PS and PG lipids with bound charge are also systematic,
-but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
-Therefore, the ion binding affinity to negatively charged bilayers
-can be better characterized measuring the PC headgroup order parameters from 
-mixed bilayers. The PC headgroup order parameters in mixtures with anionic
-lipids are larger than in pure PC lipid because addition of
-negative charges increase the order parameters (Figs. \ref{HGorderparametersPCvsPEPSPGchol} 
-and \ref{OrderParametersWithCaCl}). Upon addition of CaCl$_2$, the PC headgroup order 
-parameters in the mixed bilayers decrease and reach the values of pure PC bilayer close 
+When comparing ion binding measured from the PC headgroup order parameters between
+neutral and anionic membranes, it is important to note that the order parameters
+increase due to the addition of negative charged lipids according to the electrometer
+concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
+Therefore, the headgroup order parameters of POPC in mixtures with anionic
+lipids are larger than in pure POPC lipid bilayer, which is also seen in the ion binding
+experiments from different mixtures in the absence of added calcium (Fig. \ref{OrderParametersWithCaCl}).
+Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
 to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
-lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}). At this point the charge of
-bound Ca$^{2+}$ probably cancels the charge of anionic negative lipids and the bilayer 
-has neutral charge. The overcharging then occurs with larger concenterations.
+lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
+Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
+the negative charge lipids, resulting to a neutral membrane. 
+Above such concentrations, the specific binding of calcium leads to
+the overcharging of the membrane.
 \begin{figure}[]
@@
   \caption{\label{OrderParametersWithCaCl}
-    PC headgroup order parameters as a function of CaCl concentration from experiments 
-    different mole fractions of negatively charged lipids.
-    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
-    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
-    and POPC:POPG mixture data without NaCl from \cite{borle85}.
-    The right column shows the results zoomed to concentrations below 1M.
+    Headgroup order parameters of POPC as a function of CaCl concentration from experiments 
+    different mole fractions of negatively charged lipids (left column).
+    Right column shows the same data zoomed to the concentrations below 1M.
+    Data for Pure DPPC from~\cite{akutsu81},
+    for pure POPC from~\cite{altenbach84}, 
+    for POPC:POPS (5:1) mixture from~\cite{roux90},
+    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from~\cite{macdonald87}
+    and for POPC:POPG (1:1) mixture data without NaCl from \cite{borle85}.
   }
 \end{figure}
-As expected, the more pronounced order parameter decrease is observed in systems with more
-negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}) indicating 
-higher binding affitiny of cations into the negatively charged membranes. In conclusion,
-the consistency of the experimental results suggests that the electrometer concept can
-be used to determine the cation binding also to the lipid bilayers containing negatively
-charged lipids.  
+
+Because the POPC headgroup order parameters in mixtures with different amounts of anionic lipids
+but without additional salt are not equal, the binding affinity of calcium can be better compared
+by plotting the changes of order parameters as a function of added calcium.
+As expected, such a plot reveals more pronounced order parameter decrease in systems
+with larger fractions of negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}),
+indicating an increase in the calcium binding affinity with the increasing amount of negatively charged
+lipids in membranes. In conclusion, the presented empirical comparison of headgroup order parameter changes
+from various mixtures of POPC and anionic lipids with added calcium gives physically
+consistent results, suggesting that the electrometer concept can be used to determine
+the cation binding affinity also to the lipid bilayers containing mixtures of PC and anionic lipids.  
 \begin{figure}[]
@@
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    The change of PC headgroup order parameters as a function of CaCl$_2$
-    measured from bilayers containing different amount of negatively charged lipids.
-    The values are taken from 2H NMR experiments reported in the
-    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
-    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
-    and POPC:POPG (1:1) without NaCl \cite{borle85}).
-    As expected, the decrease of order parameters with the added CaCl$_2$
-    is more pronounced for systems with larger fraction of
-    negatively charged lipids, indicating larger amount amount
-    of bound cations.
+    Changes of POPC headgroup order parameters as a function of CaCl$_2$
+    measured from mixed bilayers containing different amounts of anionic lipids.
+    The original data is the same as in figure~\ref{OrderParametersWithCaCl}.
+%    The values are taken from 2H NMR experiments reported in the
+%    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
+%    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
+%    and POPC:POPG (1:1) without NaCl \cite{borle85}).
   }
@@
 Before using the headgroup order parameters to compare ion binding affinity between simulations
-and experiments, it is important to quantify the response of the order parameters to the
-bound charge in simulations~\cite{catte16,melcr18}. Comparison of the headgroup order parameter
-response to the cationic surfactants in POPC bilayer to the experiments \cite{scherer89} shows
-that the both order parameters are too sensitive to the bound charge in the Lipid14 model,
+and experiments, the response of the order parameters to the bound charge in simulations should be quantified
+against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
+the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
+in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
+the headgroup order parameter responses were compared more carefully with the experiments of cationic
+surfactants in POPC bilayer~\cite{scherer89}. The comparison reveals that the both headgroup
+order parameters in the Lipid14 model are too sensitive to the bound charge,
 while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
-The ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ from Lipid14 model was in good agreement 
-with experiments in the previous study~\cite{catte16} because both order parameters are equally too 
-sensitive to the bound cations (Fig. \ref{CHANGESwithDHMDMAB}). The ratio was, however, overestimated
-for the CHARMM36 model because the $\beta$ order parameter relatively more sensitive than the $\alpha$
-order parameter. These have to be taken into account when analysing the ion binding affinities
-using headgroup order parameters in simulations.
+Therefore, the headgroup order parameter response to the bound charge is actually more realistic
+in CHARMM36 model than in the Lipid14, where both order parameters are equally too 
+sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
+The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
+is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
+into account when analysing the ion binding affinities using headgroup order parameters in simulations.
+However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
+bound charge are typically smaller than then discrepancies arising from ion binding affinity.
 \begin{figure}[]
@@
   \caption{\label{CHANGESwithDHMDMAB}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
+  POPC bilayer from simulations and experiments \cite{scherer89}.
+  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
+  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
+}
 \end{figure}
@@
 \section{Difference between POPC and OPPS in MacRog model}
+
+\todo{This is to be removed once the new data is available:
+https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
 

### 2018-08-10 Manuscript/manuscriptPS.tex
Add figures delivered by Pavel Buslaev: http://nmrlipids.blogspot.com/2018/07/nmrlipids-iv-first-draft-of-manuscript.html?showComment=1533838323396#c5055037960206662973

@@
 
+\author{Pavel Buslaev}
+\affiliation{Moscow Institute of Physics and Technology}
 \author{Tiago M. Ferreira}
@@
 
+\todo{New figures are delivered by Pavel Buslaev. This should be rewritten.}
 None of the tested models reproduce the experimental headgroup and 
@@
 in these two models share significant similarities in the headgroup region (Fig.~\ref{dihedralsHG}). 
-\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
-chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
 However, the glycerol backbone order parameters in Slipids model
@@
 and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}. 
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=18.0cm]{../Figs/figS8.png}
+  \caption{\label{HGstructuresPS}
+    Overlayed snapshots from glycerol backbone and headgroup region from 
+    different simulations of PS lipids.
+  }
+\end{figure*}
 

### 2018-08-10 Manuscript/manuscriptPSsuppl.tex
Add figures delivered by Pavel Buslaev: http://nmrlipids.blogspot.com/2018/07/nmrlipids-iv-first-draft-of-manuscript.html?showComment=1533838323396#c5055037960206662973

@@
 
+\author{Pavel Buslaev}
+\affiliation{Moscow Institute of Physics and Technology}
+\author{Tiago M. Ferreira}
+\affiliation{Halle, Germany}
+\author{Ivan Gushchin}
+\affiliation{Moscow Institute of Physics and Technology}
+\author{Matti Javanainen}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{Batuhan Kav}
+\affiliation{Potsdam, Germany}
+\author{Jesper J. Madsen}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
+\author{Markus Miettinen}
+\affiliation{Potsdam, Germany}
+\author{Josef Melcr}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
 \author{O. H. Samuli Ollila}
@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
-
-\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
-\affiliation{nmrlipids.blogspot.fi}
+\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
+\affiliation{Southampton, United Kingdom}
 
@@
 
-\todo{This is to be removed once the new data is available:
+\todo{This is to be removed once the new data is available: \\
 https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
@@
 \pagebreak
-\section{Dihedrals}
+\section{Dihedral angle distributions of the headgroup and glycerol backbone
+  regions of PS lipids from different simulation models}
+
 \begin{figure*}[]
   \centering
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops1.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops2.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops3.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops4.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops5.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops6.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops7.png}
-  \caption{\label{dihedralsGLY}
-    Dihedral angle distributions of bonds from phosphate to acyl chain carbonyls from different simulation models.
+  \includegraphics[width=8.0cm]{../Figs/PS_labels.png}
+  \includegraphics[width=16.0cm]{../Figs/figS7.png}
+  \caption{\label{dihedralsHG}
+    Dihedral angle distributions of headgroup region of PS lipids from different simulation models.
   }
@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops8.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops9.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops10.png}
-  \includegraphics[width=8.0cm]{../Figs/diheds_pops11.png}
-  \caption{\label{dihedralsHG}
-    Dihedral angle distributions of bonds from phosphate to headgroup from different simulation models.
+  \includegraphics[width=15.0cm]{../Figs/figS6.png}
+  \caption{\label{dihedralsGLY}
+    Dihedral angle distributions of glycerol backbone region of PS lipids from different simulation models.
   }
@@
 
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
-  \caption{\label{glycerol_buslaev}
-    Snapshots overlayed from different simulations for glycerol backbone region
-    by Pavel Buslaev.
-  }
-\end{figure}
 \pagebreak

### 2018-08-14 Manuscript/manuscriptPS.tex
Updates in figures and text

@@
   \caption{\label{changesWITHCaClPS}
-    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of y-axis.
-    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
-    parameter value is at zero. 
+    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
+    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
+    The order parameter values from systems without calcium are set as the zero point of y-axis,
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
+    such that the lower order parameter is zero without additional ions to correctly illustrate
+    the forking with different concentrations of calcium.
+    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
+    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
   }
-  \todo{Information about the cuonterions in different simulations should be added} \\
   \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:

### 2018-08-14 Manuscript/manuscriptPSsuppl.tex
Update SI

@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/PS_labels.png}
+  \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
   \includegraphics[width=16.0cm]{../Figs/figS7.png}
@@
 
-\section{Calcium bindinf to POPC in CHARMM36 simulation with NBfix}
+\section{Calcium binding to POPC in CHARMM36 simulation with NBfix}
+
+The response of headgroup order parameters to the CaCl$_2$ concentration
+are underestimated in simulations with CHARMM36 with the NBfix term emloyed
+to the interactions between calicum and lipid oxygens \cite{kim16}
+(Fig. \ref{changesWITHCaClPS} in the main text),
+indicating that the calcium binding to the bilayer is too weak with these
+parameters. Without the NBfix term, the calcium binding affinity to PC
+lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
+However, after employing the NBfix term, the response of headgroup order
+parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
+(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to
+POPC bilayer are underestimated. Notably, the CHARMM36 simulations with NBfix
+terms predict similar binding affinity to sodium and calcium.
+In conclusion, the special NBfix~\cite{kim16} for calcium, 
+incorporated in the parameters give by the CHARMM-GUI at the time
+of running the simulations (January 2018), leads to the underestimation
+calcium binding affinity to PC and PS lipid bilayers.
 
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
+  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+    Simulation files for CHARMM36 with the NBfix term in calcium are available from Ref. \citenum{CHARMM36withCaClNBfix}.
+  }
 \end{figure}

### 2018-08-15 Manuscript/manuscriptPS.tex
Update text

@@
 
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=18.0cm]{../Figs/figS8.png}
+  \caption{\label{HGstructuresPS}
+    Overlayed snapshots from glycerol backbone and headgroup region from 
+    different simulations of PS lipids.
+  }
+\end{figure*}
+
 \begin{figure}[]
@@
 
-\todo{New figures are delivered by Pavel Buslaev. This should be rewritten.}
-None of the tested models reproduce the experimental headgroup and 
-glycerol backbone order parameters of DOPS and POPS within the
-experimental error bars (Fig. \ref{HGorderParametersPS}). 
-The tested models perform generally less well than the models tested
-for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
-which is also evindent from the comparison between subjective rankings
-of the model quality for the headgroup and glycerol backbone 
-(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
-Therefore, the models cannot be straightforwardly used to interpret the
-structural differences between PC and PS headgroups. 
-However, the differences are partially reproduced by the two best performing 
-models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
-Both reproduce the larger forking of the $\alpha$-carbon 
-and the Slipids model reproduces also the lower of the $\beta$-carbon order 
-in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
+As previously observed for PC lipids, the order parameters and structures
+of headgroup and glycerol backbone show wide variety between different simulation
+models for PS lipids (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}).
+The models for PS lipids perform generally less well than for
+PC lipids in the previous study
+(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
+Therefore, interpretation of structural differences between PC and PS headgroups
+from simulations is not straightforward. The two best performing models,
+Slipids and CHARMM36, reproduce the larger forking of the $\alpha$-carbon 
+and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
+when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
 Interestingly, the dihedral angle distributions 
@@
 and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}.
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=18.0cm]{../Figs/figS8.png}
-  \caption{\label{HGstructuresPS}
-    Overlayed snapshots from glycerol backbone and headgroup region from 
-    different simulations of PS lipids.
-  }
-\end{figure*}
 
@@
 from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
-not observed \cite{feigenson86,mattai89,roux90,roux91}. Also, the headgroup order 
-parameters of PC can be used to detect ion binding affinity to the mixed lipid 
-bilayers, see section~\ref{electrometerFORmixtures}. As expected from the previous study of PC lipid bilayers \cite{catte16}, 
-the decrease of POPC headgroup order parameters in POPC:POPS mixtures as a function of Ca$^{2+}$ concentration 
-is overestimated with respect to experiments \cite{roux90} in almost all the 
-tested simulation models (Fig. \ref{changesWITHCaClPS}), indicating overestimated calcium binding
-binding affinity. Only exception is the CHARMM36 model with the special NBfix~\cite{kim16} interaction, 
-incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
-which underestimates the calcium binding affinity. With this interaction, the 
-calcium and sodium binding affinities to POPC bilayer are equally weak (see section S7), in contrast to experimental 
-data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the ion binding peaks in
-density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}) are underestimated 
-in the CHARMM36/NBfix model, but overestimated in the other tested models.
+not observed~\cite{feigenson86,mattai89,roux90,roux91}. In addition, the ion
+binding affinity to such mixtures can be detected using the PC lipid headgroup order parameters,
+see section~\ref{electrometerFORmixtures}. As expected from the previous study of pure PC lipid
+bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
+POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
+with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
+calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
+indicating weaker binding affinity than experiments.
+%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
+Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
+data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
+manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
+is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
+in all the other tested models.
 

### 2018-08-15 Manuscript/manuscriptPSsuppl.tex
Update text

--- 2018-08-10 Manuscript/manuscriptPSsuppl.tex
@@
 
-\section{Calcium bindinf to POPC in CHARMM36 simulation with NBfix}
+\section{Calcium binding to POPC in CHARMM36 simulation with NBfix}\label{CHARMMcalciumNBfix}
+
+The response of POPC headgroup order parameters to the CaCl$_2$ concentration
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 containing
+the NBfix for interactions between calicum and lipid oxygens \cite{kim16}
+(Fig. \ref{changesWITHCaClPS} in the main text),
+indicating that the calcium binding to the bilayer is too weak with these
+parameters. Without the NBfix term, the calcium binding affinity to pure POPC
+lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
+However, after employing the NBfix term, the response of headgroup order
+parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
+(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
+POPC bilayer are underestimated. Notably, CHARMM36 simulations with the NBfix
+terms \cite{venable13,kim16} employed predict similar binding affinity for sodium and calcium.
+In conclusion, the special NBfix~\cite{kim16} for calcium, 
+incorporated in the parameters give by the CHARMM-GUI at the time
+of running the simulations (January 2018), leads to the underestimation of
+calcium binding affinity to pure PC and mixed PC:PS bilayers.
 
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
+  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having the NBfix term employed for sodium \cite{venable13} ({\it left})
+    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
+    Simulation files for CHARMM36 with the NBfix term in calcium are available from Ref. \citenum{CHARMM36withCaClNBfix}.
+  }
+  \todo{Sources for other simulations by Ricky Nencini.}
 \end{figure}
@@
   \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
-  The response of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
+    Density profiles along membrane normal from CHARMM36 simulations with (middle)
+    and without (top) the NBfix term for calcium \cite{kim16} compared to the simulation
+    with the NBfix term for sodium \cite{venable13} (bottom). The simulation data are the same as in figure \ref{OP_CHARMM_CaCl_POPC_NBFix}.
+  }
 \end{figure}

### 2018-08-16 Manuscript/manuscriptPS.tex
Update the text

@@
     DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP1 \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP2 \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
     DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
@@
 Before using the PC headgroup order parameters to quantify the ion binding
-affinity, it is important to quantify the response of the headgroup order
-parameters to the known amount of bound charge~\cite{catte16,melcr18}.
+affinity, it is important to quantify their response to the known amount of bound charge~\cite{catte16,melcr18}.
 This can be done using the experimental data from the mixtures of
@@
 values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
-The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking in PS headgroup when compared with
-the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
+When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
+the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
+experiences a significant forking in PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
 
-As previously observed for PC lipids, the order parameters and structures
-of headgroup and glycerol backbone show wide variety between different simulation
-models for PS lipids (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}).
-The models for PS lipids perform generally less well than for
+The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
+models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
+as previously observed also for PC lipids. The models for PS lipids perform generally less well than for
 PC lipids in the previous study
@@
 Therefore, interpretation of structural differences between PC and PS headgroups
-from simulations is not straightforward. The two best performing models,
-Slipids and CHARMM36, reproduce the larger forking of the $\alpha$-carbon 
+from simulations is not straightforward.
+
+The best performing models,
+Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
 and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
 when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
-Interestingly, the dihedral angle distributions 
-%of CHARMM36 and Slipids in 
-in these two models share significant similarities in the headgroup region (Fig.~\ref{dihedralsHG}). 
-However, the glycerol backbone order parameters in Slipids model
-significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
-which can be related to the differences in the dihedral angle distributions 
-of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}). 
-Similar difference was previously observed for PC lipids \cite{botan15}
-and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}.
+Interestingly, the headgroup structures of these models share significant similarities
+and are distinct from other models (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}). 
+The distribution of the dihedral angle for C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ shows
+a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
+in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
+visible in the sampled conformations (Fig.~\ref{HGstructuresPS}) and might
+manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
+Another distinct feature in the three best performing models is the distribution
+of N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral, which gives a large maxima close to 300$^{\circ}$
+and a smaller maxima close to 60$^{\circ}$. This is also visible in the sampled
+structures (Fig.~\ref{HGstructuresPS}).
+\todo{There seems to be discrepancy between structures and dihedral distributions for CHARMM36ua}.
+The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
+differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
+as observed previously also for PC lipids \cite{botan15}.
+The origin of this difference is more difficult to track without more elaborate analysis,
+because different models show very complicated patterns of distinct structures
+in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
 
-%\todo{Discussion is to be finished.
+%which can be related to the differences in the dihedral angle distributions 
+%of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}).
+
 %  One possible conclusion could be the following:
@@
 %than for PC (3) \cite{botan15}.
-%Therefore the interpretation of structural details of PS headgroup or
-%differences between PC and PS lipids is very challenging 
-%from the current MD simulation models.
 

### 2018-08-17 Manuscript/manuscriptPS.tex
Update text and figures

@@
 \caption{List of POPC:POPS mixture simulations with different amounts of added ions. 
-   % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+  The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  these correspond the concentrations reported in the experiments by Roux et al.~\cite{roux90}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
 }\label{mixedIONsystems}
-    \todo{The used definition of ion concentrations needs to be specified and standardized.} 
 %begin{minipage}[t]{\textwidth}
 \begin{tabular}{l c c c c c c c c c c}
-  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) & CaCl$_2$\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) & [CaCl$_2$]\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
     POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
 & ?
@@
 Madsen}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
 & ?
@@
     POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.1 & 120:24 & 5760 & 10   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.3 & 120:24 & 5760 & 31   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1   & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3   & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.5  & 0   & 120:24 & 5760 & 52   & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1    & 0   & 120:24 & 5760 & 104  & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2    & 0   & 120:24 & 5760 & 208  & 298  & 200 & 145 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3    & 0   & 120:24 & 5760 & 311  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4    & 0   & 120:24 & 5760 & 415  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.30 & 120:24 & 5760 & 31   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1.00   & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 200 & 145 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
     \hline
     POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0  & 0     & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5  & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
     POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0   & 0   & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5  & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4    & 0   & 120:24 & 5760 & ?   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
     POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.03 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.17 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.36 & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.03 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.17 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.36 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
     \hline
     POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & ? & ? & \cite{??} \todoi{To be added by Ollila}  \\
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.12   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.12\footnote{Calculation of concetration complicated due the scaled ions.
+Concentration taken as reported in the delivered data.}   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715\footnote{Calculation of concetration complicated due the scaled ions.
+Concentration taken as reported in the delivered data.}  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
     \hline
@@
 
-\todo{How close are the hydration levels in experiments and simulations?} \\
-\todo{We need to decide how to report the ion concentrations, see discussion in Ref.~\citenum{melcr18}.}
+Two different definitions for the salt concentrations have been used when
+electrometer concept is applied to study ion binding affinity.
+The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
+or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
+In this work, we use the former definition to be consistent with the reference
+experimental data \cite{roux90}. The used definition has only a marginal effect
+to the results is simulations with realistic ion binding affinity.
+
+\todo{Should we discuss about hydration levels in experiments and simulations?}
 
@@
     Changes of the PC (left) and PS (right) headgroup order parameters as a function of
-    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
+    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
+    (except Berger simulations are (4:1) mixture at 310~K).
+    The experimental data is from Ref. \citenum{roux90}.
     The values from counterion-only systems are set as a zero point of y-axis.

### 2018-08-17 Manuscript/manuscriptPSsuppl.tex
Update text and figures

@@
 Therefore, the headgroup order parameter response to the bound charge is actually more realistic
-in CHARMM36 model than in the Lipid14, where both order parameters are equally too 
+in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
 sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.

### 2018-08-31 Manuscript/manuscriptPS.tex
Add discussion about hydration levels in simulations and experiments

@@
 
+In the experiment from the literature used in this work~\cite{browning80,roux90},
+the lipids were first soluted to the buffer and then centrifuged to a pellet which
+was used in the measurements. In such experiments the samples contain approximately
+10~wt~\% of lipids~\cite{browning80,roux88,roux90}, which is less than in
+gravimetric experiments (60~wt~\%) and simulations (approx. 50-60~wt~\%) in this work.
+Since multilamellar structures of pure PS lipids swell due to the electrostatic repulsion,
+the larger repeat distances are expected in samples with lower lipid concentrations~\cite{millman82}.
+However, the PS headgroup order parameters from gravimetric sample are in good agreement
+with centrifuged sample (Fig. \ref{HGorderParameters}) and the equlibrium repeat distance decreases
+already above 500~mM concentration of monovalent ions close to the simulation box sizes \cite{millman82,rand89}.
+Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
+in simulations and the reference experiments.
+
 Two different definitions for the salt concentrations have been used when
@@
 to the results is simulations with realistic ion binding affinity.
-
-\todo{Should we discuss about hydration levels in experiments and simulations?}
 

### 2018-09-03 Manuscript/manuscriptPSsuppl.tex
Add author and update information about CHARMM36 simulations with different NBfix terms

@@
 \author{Josef Melcr}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{Ricky Nencini}
 \affiliation{Institute of Organic Chemistry and Biochemistry,
@@
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having the NBfix term employed for sodium \cite{venable13} ({\it left})
+    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
+    the NBfix term employed for sodium \cite{venable13} ({\it left})
     and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
-    Simulation files for CHARMM36 with the NBfix term in calcium are available from Ref. \citenum{CHARMM36withCaClNBfix}.
+    Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
+    with the NBfix term in sodium at Ref. \citenum{CHARMM36withCaClNBfix}, 
+    with the NBfix term in calcium at Ref. \citenum{CHARMM36withCaClNBfix} and
+    without the NBfix in calcium at Ref. \citenum{charmmPOPC450mMCaClfiles}.
   }
-  \todo{Sources for other simulations by Ricky Nencini.}
 \end{figure}

### 2018-09-03 Manuscript/manuscriptPS.tex
Add author and update information about CHARMM36 simulations with different NBfix terms
Add new figures and text by Tiago Ferreira

@@
 \subsection{Solid state NMR experiments}
-The magnitude and signs of the C-H bond order parameters in
-headgroup and glycerol backbone were measured
-using natural abundance $^{13}$C solid state NMR spectroscopy as
-described previously \cite{ferreira13,ferreira16}.
-Shortly, the absolute values of the order parameters were determined from the dipolar splittings
-given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
-the signs were measured using S-DROSS experiments \cite{gross97}.
 
-\todo{Details of the used spectrometer and maybe some other details should be given.} \\
+Headgroup and glycerol backbone C-H bond order parameters of POPS
+were determined from chemical-shift resolved dipolar splittings
+measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}
+and corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
+using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
+The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
+Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
+The following experimental setups were used.
+
+{\emph{R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
+for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
+Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
+50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
+dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.
+
+\emph{S-DROSS experiment}. The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
+experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
+blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
+frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
+The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
+measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.
+
+\emph{NMR numerical simulations}
+The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package using the $^{13}$C--$^1$H
+dipolar couplings determined by the R-PDLF experiments or calculated from known $^2$H quadrupolar couplings as input.
+Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
+distribution of bilayer orientations in the samples studied.
+
+\emph{Sample preparation}
 \todo{Sample preparation should be described.} \\
-\todo{How is the peak assignment done?}
 
@@
 
+\todo{How is the peak assignment done?}
+
 The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
@@
 confirming the interpretation that the smaller order parameter is negative.
-\begin{figure}[!htb]
+\begin{figure}[!tb]
   \centering
-  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
+  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
   \caption{\label{PShgSIGNSsimpson}
-    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
-    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
-    (c) Slices for $\alpha$ and $\beta$ barbons.
-    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
-    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the larger and smaller $\alpha$-carbon slittings were used in the
-    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller order parameter (dashed grey).
+    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
+    (B) 2D R-PDLF spectra.
+    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
+    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
+    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller $\alpha$-carbon order parameter (dashed grey).
   }
-  \todo{This is preliminary figure, should be polished.} \\
-  \todo{Should we show slices for all the analyzed carbons in (c)?}
 \end{figure}

### 2018-09-03 Manuscript/manuscriptPSsuppl.tex
Add new figures and text by Tiago Ferreira

@@
 \pagebreak
-\section{Difference between POPC and OPPS in MacRog model}
-
-\todo{This is to be removed once the new data is available: \\
-https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
-
+\section{Dipolar slices of the R-PDFL experiment}
 \begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
-  \caption{\label{CHANGESwithCaClPGPS}
-    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
+%  \centering
+  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
+  \caption{\label{DPslices}
+    Dipolar coupling slices from the R-PDFL experiment.
+  }
+  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
+  Also, the size of the figure in the file may not be optimal.}
 \end{figure}
@@
 
+\pagebreak
+\section{Difference between POPC and OPPS in MacRog model}
+
+\todo{This is to be removed once the new data is available: \\
+https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
+  \caption{\label{CHANGESwithCaClPGPS}
+    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
+\end{figure}
+
 \newpage

### 2018-09-07 Manuscript/manuscriptPS.tex
Add corrected MacRog simulations by M. Javanainen: http://nmrlipids.blogspot.com/2018/07/nmrlipids-iv-first-draft-of-manuscript.html?showComment=1535364446749#c4563235956514366812

@@
     \hline
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.30 & 120:24 & 5760 & 31   & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1.00   & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 200 & 190 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 200 & 145 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 200 & 125 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.30 & 120:24 & 5760 & 31   & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1.00   & 120:24 & 5760 & 104  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+%    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
     \hline
@@
   http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
-  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
-    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
 \end{figure*}
@@
   }
-  \todo{The CHARMM results are mass densities, numbers should be used.} \\
+  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
   \todo{Should we include also counterions into the plot?} \\
-  \todo{Not all the data from MacRog is included.} 
 \end{figure}

### 2018-09-11 Manuscript/manuscriptPS.tex
Updates in the manuscript

@@
 \emph{Sample preparation}
-\todo{Sample preparation should be described.} \\
+The sample was prepared simply by mixing the POPS with water (lipid:water 60:40 wt\%) in an eppendorf
+tube mixing and centrifuging the sample repeatedly until an homogeneous viscous fluid was obtained.
+20 mg of sample was then transferred to an NMR insert suitable for 4 mm NMR rotors.  
+\todo{Maybe we need little bit more information about the mixing procedure?}
 
@@
 Angular brackets point to the average over all sampled configurations.
-\todo{Error estimation should be discussed.}\\
+The order parameters were first calculated separately for each lipid molecule
+in the system averaging over time. The average and
+the standard error of the mean were then calculated over different lipids.
 The number density profiles were calculated using {\it gmx density} tool
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
   \caption{\label{HGorderParameters}
@@
 
-\todo{POPS/OPPS issue with MacRog model is in progress: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
-
 %
@@
   }
-  \todo{Lipid 17 is to be added.}
+  \todo{Lipid 17 is to be added when we have the density profiles.
+  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
 \end{figure}

### 2018-09-14 Manuscript/manuscriptPS.tex
Add CHARMM36 data with additional potassium

@@
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
     POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
@@
 Madsen}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     \hline

### 2018-09-17 Manuscript/manuscriptPS.tex
Update discussion about PS headgroup structure based on new figures by P. Buslaev

@@
 models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
-as previously observed also for PC lipids. The models for PS lipids perform generally less well than for
+as previously observed also for PC lipids \cite{botan15}. The models for PS lipids perform generally less well than for
 PC lipids in the previous study
@@
 when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
-Interestingly, the headgroup structures of these models share significant similarities
-and are distinct from other models (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}). 
-The distribution of the dihedral angle for C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ shows
+Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
 a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
@@
 manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
-Another distinct feature in the three best performing models is the distribution
-of N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral, which gives a large maxima close to 300$^{\circ}$
-and a smaller maxima close to 60$^{\circ}$. This is also visible in the sampled
-structures (Fig.~\ref{HGstructuresPS}).
-\todo{There seems to be discrepancy between structures and dihedral distributions for CHARMM36ua}.
+\todo{We should consider making the conformation figures from CHARMM36 and Slipids POPC, and compare these
+  to the results from POPS.}
+
+\todo{Maybe we should move the glycerol backbone results to SI and focus here on the headgroup,
+and especially to the difference between PC and PS.}
 The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations

### 2018-09-18 Manuscript/manuscriptPS.tex
Update text

@@
 \emph{NMR numerical simulations}
-The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package using the $^{13}$C--$^1$H
-dipolar couplings determined by the R-PDLF experiments or calculated from known $^2$H quadrupolar couplings as input.
+The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
+dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
 Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
@@
 Angular brackets point to the average over all sampled configurations.
-The order parameters were first calculated separately for each lipid molecule
-in the system averaging over time. The average and
+The order parameters were first calculated averaging over time separately
+for each lipid molecule in the system. The average and
 the standard error of the mean were then calculated over different lipids.
@@
 
-In the experiment from the literature used in this work~\cite{browning80,roux90},
+In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
 the lipids were first soluted to the buffer and then centrifuged to a pellet which
-was used in the measurements. In such experiments the samples contain approximately
-10~wt~\% of lipids~\cite{browning80,roux88,roux90}, which is less than in
-gravimetric experiments (60~wt~\%) and simulations (approx. 50-60~wt~\%) in this work.
-Since multilamellar structures of pure PS lipids swell due to the electrostatic repulsion,
-the larger repeat distances are expected in samples with lower lipid concentrations~\cite{millman82}.
-However, the PS headgroup order parameters from gravimetric sample are in good agreement
-with centrifuged sample (Fig. \ref{HGorderParameters}) and the equlibrium repeat distance decreases
-already above 500~mM concentration of monovalent ions close to the simulation box sizes \cite{millman82,rand89}.
+was used in the measurements. Such samples have lower lipid concentration
+(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
+gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
+Larger multilamellar repeat distances are expected in the samples with lower lipid
+concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
+However, the PS headgroup order parameters measured from gravimetric sample in this work
+are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
+Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
+salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
 Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
-in simulations and the reference experiments.
+in the used simulations and reference experiments.
 
@@
 to the results is simulations with realistic ion binding affinity.
+\todo{Maybe we should add a figure in the SI.}
 
@@
 
-\todo{How is the peak assignment done?}
-
 The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
-carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
-was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
-Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
-show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
-and a superposition of a large and a very small splitting for the $\alpha$-carbon.
-The larger splitting gives a order parameter value of 0.09, while the numerical value
-from the small splitting cannot resolved with the available resolution.
+carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
+The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
+known order parameters from the $^2$H NMR experiments~\cite{browning80}.
+Slices of the R-PDFL spectra and the resulting order parameters values
+are shown in the supplementary information (Fig. \ref{DPslices}). 
 Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
 only the absolute values of order parameters, we determined the signs of PS headgroup
-and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
-The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
-order parameter is negative, which is confirmed by SIMPSON simulations.
-The beginning of the S-DROSS slice suggests that the higher order parameter of
-the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
+order parameters using the S-DROSS experiment \cite{gross97}.
+The S-DROSS slice clearly shows that the order parameter of
+the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
+which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
+suggests that the larger order parameter of the $\alpha$-carbon 
+is positive and the deviation towards negative values with longer T$_1$ times suggests
 that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
-where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
+using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
 The literature value was used because the
 resolution of our experiment was not sufficient to determine the
-small value of the order parameter. The S-DROSS curve from
-SIMPSON simulation with a positive value for the smaller order parameter
-(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
+small value of the order parameter.
+The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
+(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
 confirming the interpretation that the smaller order parameter is negative.
@@
 \end{figure}
-
-%\begin{figure}[!htb]
-%  \centering
-%  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
-%  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
-%  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
-%  \caption{\label{glycerolNMR}
-%    R-PDFL slices for gycerol backbone carbons.
-%  }
-%  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
-%  \todo{What are the top figures actually?}
-%\end{figure}
 
@@
   }
-  \todo{CHARMM36 results for this plot would be highly useful.}
 \end{figure*}

### 2018-09-18 Manuscript/manuscriptPSsuppl.tex
Remove POPS vs OPPS discussion from the SI
Update text

@@
 \section{Dipolar slices of the R-PDFL experiment}
+
+Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
+show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
+and a superposition of a large and a very small splitting for the $\alpha$-carbon.
+The larger splitting gives a order parameter value of 0.09, while the numerical value
+from the small splitting cannot resolved with the available resolution.
 \begin{figure}[]
@@
 
-\pagebreak
-\section{Difference between POPC and OPPS in MacRog model}
-
-\todo{This is to be removed once the new data is available: \\
-https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
-  \caption{\label{CHANGESwithCaClPGPS}
-    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
-\end{figure}
-
 \newpage

### 2018-09-20 Manuscript/manuscriptPGPE.tex
Add todos to PGPE manuscript

@@
 
+\section{Methods}
+\section{Molecular dynamics simulations}
+Molecular dynamics simulation data was collected using open collaboration \cite{botan15}.
+The simulations are listed in table \ref{systems}.
+\begin{table*}[htb]
+  %\begin{sidewaystable*}[!p]
+  \centering
+  \caption{List of MD simulations. The salt concentrations calculated as 
+    [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+    %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+    %   The lipid force fields named as in our previous work~\cite{botan15}.
+  }\label{systems}
+  \begin{minipage}[t]{\textwidth}
+    \begin{tabular}{l c c r r r r r r c c}
+      %\hline
+      % some footnotes are not visible in typeset-MS (pdf)
+      lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+      \hline
+      DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
+      \hline
+      POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
+      \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
+      POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
+      \hline
+      DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
+      DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+    \end{tabular}
+  \end{minipage}
+  %\end{sidewaystable*} 
+\end{table*}
+
 \section{Results and Discussion}
@@
 \subsection{PE headgroup and glycerol backbone}
+\todo{Add results from this discussion: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1502290977005#c4964787551006026647}
 Order parameters from Slipids simulations 
@@
 
+\todo{There are mass density profiles available from CHARMM simulations, preliminary figures should be made:
+http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1501446779291#c7311317114567929503}
+
 \section{Conclusions}
@@
 \end{center}
-
-\section{Simulated systems}
-\begin{table*}[!p]
-%\begin{sidewaystable*}[!p]
-\centering
-\caption{List of MD simulations. The salt concentrations calculated as 
-   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-%   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
-%   The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
-\begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
-  %\hline
-  % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
-  \hline
-  DPPE  & Slipids \cite{jambeck12b} &0 & 0        & 288 			   		& 9386 & 0  & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
-    \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{By Piggot: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1491425687561\#c4932902612512697301.
-We need to decide the switching version or discuss this somehow.}  \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 			   		& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}        &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. Data to be uploaded in Zenodo?}  \\
-    DOPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}       &0 & 0        & 128 			   		& 4480 & 0  & 303  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide between RF and PME or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    \hline
-    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the switching version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 			   		& 4480 & 0  & 298  & 500 & 100 & \cite{??}
-    \todoi{Delivered by Piggot. We need to decide the cut-off version or discuss this somehow. Data to be uploaded in Zenodo?}  \\
-    \hline
-    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-    \hline
-    POPG/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}        &0 & 0        & ? 			   		& ? & 0  & ?  & ? & ? & \cite{??}
-    \todoi{Details to be filled and data to be uploaded in Zenodo by Ollila.}  \\
-    POPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 10664 & 0  & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
-    \hline
-    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
-    DPPG/Na$^+$  & Slipids \cite{jambeck13}     &0 & 0        & 288 			   		& 11232 & 0  & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
-\end{tabular}
-\end{minipage}
-%\end{sidewaystable*} 
-\end{table*}
 

### 2018-09-21 Manuscript/manuscriptPGPE.tex
Update PGPE manuscript

@@
 \subsection{PE headgroup and glycerol backbone}
-\todo{Add results from this discussion: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1502290977005#c4964787551006026647}
+\todo{Add results from this discussion: \url{http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1502290977005#c4964787551006026647}}
 Order parameters from Slipids simulations 

### 2018-09-30 Manuscript/manuscriptPS.tex
Update manuscript

@@
 
-\subsection{Counterion binding to lipid bilayers containing PS lipids}
-Membranes containing PS lipids are always accomppanied with counterions, which
+\subsection{Counterion binding and interactions between PC and PS headgroups}
+Membranes containing PS lipids are always accomppanied with counterions which
 modulate electrostatic interactions between lipids and other biomolecules. 
-Counterions are also suggested screen the repulsion between charged lipid headgroups 
+Counterions are also suggested to screen the repulsion between charged lipid headgroups 
 in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
@@
 The counterion density profiles along membrane normal 
-show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
-The strongest counterion binding, i.e., the lowest concentrations in bulk water,
-are observed in MacRog, Berger and Lipid17/JC simulations.
-CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
-density, while a single maxima is observed in the other models. 
-\todo{More detailed discussion may be possible after comparing monovalent ion binding
-to bilayers between CHARMM simulations and experiments.
-Also, section S6 should be finished.}
+show significant differences between simulation models in both binding affinity
+and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
+The strongest sodium counterion binding, i.e., the lowest concentrations in bulk water,
+is observed in MacRog, Berger and Lipid17/JC simulations.
+Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
+counterion density, but slightly at different distances from the bilayer
+center. Other models exhibit more complicated distribution, often including
+two local maxima in the density profile.
+%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
+%density, while a single maxima is observed in the other models.
+
 Area per lipid is in agreement with experiments \cite{pan14} only
-in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
-The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
+in the MacRog simulation with potassium counterions and in Gromos-CKP models,
+while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
+The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
 counterion binding because CHARMM36, CHARMM36ua and Slipid models give
 smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
-%The proposed trend is observed for Lipid17 model,
-%for which the Joung-Cheatham ions give higher affinity and smaller area
-%per molecule. However, the trend is not observed when comparing accross different
-%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
-%but binding affinity is significantly higher in MacRog. 
+On the other hand, increasing counterion binding affinity reduces the area per molecule
+when comparing between sodium and potassium in MacRog simulations or between
+ff99 and jc ions in Lipid 17 simulations.
+This suggests that the low area per molecule in PS lipid bilayers cannot be
+explained only by counterion binding but also the hydrogen bonding network has to
+be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
 \begin{figure}[]
@@
 
-To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
-we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
-as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
-Experimental order parameter data for POPC headgroup in the mixture is available as a function
-of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
-as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
-ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
-but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
-POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
-but only modest changes as a function of KCl concentration, indicating singificant 
-Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
-electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
-Berger model, the headgroup order parameter response of POPC
-to the added NaCl is similar to the experiments of LiCl,
-indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
-Indeed, the sodium density profile shows a significant binding peak in the 
-Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
-is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
-changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
-\todo{Discussion about Lipid17 to be written when we have the density profiles.}
-All the tested models overestimate the changes of POPS headgroup order parameters as
-a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
-model development is necessary to interpret the PS headgroup-ion interactions 
-from MD simulations.
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
-  \caption{\label{PSresponseTONaCl}
-    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
-    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
-    (except Berger simulations are (4:1) mixture at 310~K).
-    The experimental data is from Ref. \citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of y-axis.
-    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
-    parameter value is at zero.
-  }
-\end{figure*}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
-  \caption{  Counterion density distributions from PC:PS mixtures.
-\label{CIdensPSOCmixt}
-  }
-  \todo{Lipid 17 is to be added when we have the density profiles.
-  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
-\end{figure}
-
-\subsection{Headgroup structure in PS and PC mixtures}
-Dilution of PS lipid bilayers with PC lipids reduces the propensity
-of PS headgroup-multivalent ion complexes and is proposed to make PS
-headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
-Therefore, the intermolecular interactions at the headgroup region seems
-to be important for the physical properties of mixed lipid bilayers.
-These interactions can be indirectly monitored by measuring the
-headgroup order parameters from PS:PC mixtures with different molar
-ratios. The headgroup order parameters of POPC increase in such experiments 
-with increasing amount of POPS (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
-This behaviour is generally observed when negatively charged lipids or surfactants 
-are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
-tilting of lipid headgroup more parallel to the membrane plane according to
-the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
-shift closer to zero when bilayer is diluted with PC lipids in 
-experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
-which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
-
-The increase of POPC headgroup order parameters with the increasing
-amount of negatively charged POPS lipid is reproduced in
-MacRog simulations with potassium counterions, but not in 
-Berger simulations with sodium or in CHARMM36 simulations
-with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
-The observations can be explained using the electrometer concept.
-The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
-which surpasses the effect of negatively charged lipids as also the amount of
-counterions increase with increasing amount of PS.
+Previously used method to evaluate counterion binding affinity to lipid bilayers
+using the changes of hedgroup order parameters as a function
+additional ions \cite{catte16} is difficult to apply for counterions because
+reference state without ions in solution does not exist, see section \ref{mixtureTOadditionalCIs}
+in supplementary information. However, the counterion binding affinity is
+reflected to the increase of POPC headgroup order parameters due to the
+increasing amount of negatively charged POPS lipids in the bilayer.
+%The headgroup order parameters of POPC increase
+%with increasing amount of negatively charged POPS
+The experimentally measured order parameter increase~\cite{scherer87}
+is reproduced in simulations with MacRog and potassium counterions, but not in 
+other simulations (Fig. \ref{HGorderparametersPCvsPS}). 
+The increase of POPC headgroup order parameters with increasing amount of POPS 
+can be explained using the electrometer concept \cite{seelig87,scherer89}:
+The PC headgroup tilts more parallel to the membrane plane with increasing
+amount of negative charge in the membrane. In Berger simulations,
+the strong sodium binding surpasses the effect of negatively charged PS
+lipids and the order parameters actually decrease because also the amount of
+positive counterions increase with increasing amount of PS.
 In CHARMM36 simulations, the counterion binding neutralizes the effect
 of PS and the headgroup order parameters are not changed with increasing amount
-of PS. Finally, the weak binding of potassium in the MacRog simulations
-enables the increase of order parameters with the increasing amount of
-negatively charged PS lipids (Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).
-
+of PS. Therefore, we conclude that the weak binding of potassium in the MacRog
+simulations (Fig. \ref{NAdensPOPS}) leads to the correct surface charge
+density and the response of POPC headgroup to the increasing amount of negatively
+charges PS lipids in the lipid mixtures. 
 \begin{figure*}[]
@@
 \end{figure*}
+
+The headgroup order parameters of PS lipids
+shift closer to zero when bilayer is diluted with PC lipids in 
+experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
+which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
+
+Dilution of PS lipid bilayers with PC lipids reduces the propensity
+of PS headgroup-multivalent ion complexes and is proposed to make PS
+headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
+Therefore, the intermolecular interactions at the headgroup region seems
+to be important for the physical properties of mixed lipid bilayers.
+These interactions can be indirectly monitored by measuring the
+headgroup order parameters from PS:PC mixtures with different molar
+ratios.
 

### 2018-09-30 Manuscript/manuscriptPSsuppl.tex
Update manuscript

@@
 
-\section{Sodium binding to DMPC:DOPS mixture}
+\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
+To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
+we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
+as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
+Experimental order parameter data for POPC headgroup in the mixture is available as a function
+of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
+as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
+ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
+This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
+but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
+POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
+but only modest changes as a function of KCl concentration, indicating singificant 
+Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
+electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
+Berger model, the headgroup order parameter response of POPC
+to the added NaCl is similar to the experiments of LiCl,
+indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
+Indeed, the sodium density profile shows a significant binding peak in the 
+Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
+is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
+changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
+\todo{Discussion about Lipid17 to be written when we have the density profiles.}
+All the tested models overestimate the changes of POPS headgroup order parameters as
+a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
+model development is necessary to interpret the PS headgroup-ion interactions 
+from MD simulations.
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
+  \caption{\label{PSresponseTONaCl}
+    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
+    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
+    (except Berger simulations are (4:1) mixture at 310~K).
+    The experimental data is from Ref. \citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of y-axis.
+    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
+    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
+    parameter value is at zero.
+  }
+\end{figure*}
 
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
-  \caption{\label{PSresponseTONaClDMPC}
-    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
+  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
+  \caption{  Counterion density distributions from PC:PS mixtures.
+\label{CIdensPSOCmixt}
   }
+  \todo{Lipid 17 is to be added when we have the density profiles.
+  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
 \end{figure}
-The experimental results show essentially no changes in the order parameters as a function of
-added NaCl, while significant changes are observed in simulations. However,
-the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
-Therefore, we cannot exclude the possibility that the NaCl induced changes were already
-saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
+
+%%
+%% I think that this section is not very useful, I think that we should remove it.
+%%
+%\section{Sodium binding to DMPC:DOPS mixture}
+%
+%\begin{figure}[]
+%  \centering
+%  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
+%  \caption{\label{PSresponseTONaClDMPC}
+%    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
+%  }
+%\end{figure}
+%The experimental results show essentially no changes in the order parameters as a function of
+%added NaCl, while significant changes are observed in simulations. However,
+%the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
+%Therefore, we cannot exclude the possibility that the NaCl induced changes were already
+%saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
 \pagebreak

### 2018-10-01 Manuscript/manuscriptPS.tex
Update text

@@
 ff99 and jc ions in Lipid 17 simulations.
-This suggests that the low area per molecule in PS lipid bilayers cannot be
-explained only by counterion binding but also the hydrogen bonding network has to
+This suggests that the counterion binding alone cannot explain the low area per molecule
+in PS lipid bilayers but also the hydrogen bonding network has to
 be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
@@
 
-Previously used method to evaluate counterion binding affinity to lipid bilayers
-using the changes of hedgroup order parameters as a function
-additional ions \cite{catte16} is difficult to apply for counterions because
-reference state without ions in solution does not exist, see section \ref{mixtureTOadditionalCIs}
-in supplementary information. However, the counterion binding affinity is
+Evaluation of counterion binding affinity to charged lipid bilayers
+using the changes of headgroup order parameters is difficult because
+reference state without ions does not exist, in contrast to previous
+studies with zwitterionic membranes and coions \cite{catte16},
+see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
+However, the counterion binding affinity is
 reflected to the increase of POPC headgroup order parameters due to the
 increasing amount of negatively charged POPS lipids in the bilayer.
-%The headgroup order parameters of POPC increase
-%with increasing amount of negatively charged POPS
-The experimentally measured order parameter increase~\cite{scherer87}
-is reproduced in simulations with MacRog and potassium counterions, but not in 
-other simulations (Fig. \ref{HGorderparametersPCvsPS}). 
-The increase of POPC headgroup order parameters with increasing amount of POPS 
-can be explained using the electrometer concept \cite{seelig87,scherer89}:
-The PC headgroup tilts more parallel to the membrane plane with increasing
-amount of negative charge in the membrane. In Berger simulations,
-the strong sodium binding surpasses the effect of negatively charged PS
-lipids and the order parameters actually decrease because also the amount of
-positive counterions increase with increasing amount of PS.
-In CHARMM36 simulations, the counterion binding neutralizes the effect
-of PS and the headgroup order parameters are not changed with increasing amount
-of PS. Therefore, we conclude that the weak binding of potassium in the MacRog
-simulations (Fig. \ref{NAdensPOPS}) leads to the correct surface charge
-density and the response of POPC headgroup to the increasing amount of negatively
-charges PS lipids in the lipid mixtures. 
+The experimentally measured increase in the order parameters~\cite{scherer87}
+is reproduced only in MacRog simulations, while other simulations
+predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
+The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
+In experiments and MacRog simulations, the increasing amount of negative charge
+in the membrane tilts the PC headgroup more parallel to the membrane plane
+leading to the larger order parameters. In Berger simulations,
+the strong sodium binding surpasses the effect of negatively charged
+lipids and the order parameters decrease with the increasing amount of
+positive counterions of PS lipid.
+In CHARMM36 simulations, the counterion binding neutralizes the negative charge
+of lipids at the interface and the headgroup order parameters are not changed
+with increasing amount of PS. Therefore, the relatively weak binding of potassium
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
+response of POPC headgroup to the increasing amount of negatively charged lipids,
+indicating that these simulations have realistic surface charge density in membranes
+containing PS lipids.
 \begin{figure*}[]

### 2018-10-03 Manuscript/manuscriptPSsuppl.tex
Add figure showing the effect of two different concentration definitions to the changes of headgroup order parameters Figure and data contibuted by Ricky Nencini by email.

@@
   The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
+}
+\end{figure}
+
+\pagebreak
+\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFfigure}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
+  \caption{\label{concentrationDEFfigure}
+    Difference between order parameter changes as a function of CaCl$_2$ concentration demonstrated
+    for the two possible definitions of ion concentration, concentration in buffer before solvating the
+    lipids (buffer concentration) and concenrtration in bulk after solvating the lipids (bulk concentration) \cite{melcr18}. 
 }

### 2018-10-03 Manuscript/manuscriptPS.tex
Update text

@@
 experimental data \cite{roux90}. The used definition has only a marginal effect
-to the results is simulations with realistic ion binding affinity.
-\todo{Maybe we should add a figure in the SI.}
+to the results is simulations with realistic ion binding affinity
+(section \ref{concentrationDEFsection} in the supplementary information).
 
@@
   \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
-  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
-  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
   \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
 \end{figure*}
-
-The headgroup order parameters of PS lipids
-shift closer to zero when bilayer is diluted with PC lipids in 
-experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
-which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
 
@@
 headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
-Therefore, the intermolecular interactions at the headgroup region seems
-to be important for the physical properties of mixed lipid bilayers.
-These interactions can be indirectly monitored by measuring the
-headgroup order parameters from PS:PC mixtures with different molar
-ratios.
-
-Oppositely to experiments, the headgroup order parameter of POPS shift away
-from zero in CHARM36 simulations when bilayer is diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
-In lipid14/17 simulations, the POPS order parameter shift closer to zero when
-bilayer is diluted with POPC, but the numerical values of order parameters
-are too far from experiments to enable interpretation of the experimental data.
-Therefore, we conlcude that the force field development is necessary before
-MD simulations can be used to interpret the interactions between PC and PS headgroups.
+This is related to the shift of the PS headgroup order parameters 
+closer to zero when bilayer is diluted with PC
+lipids (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90}.
+%which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
+This shift is not seen in CHARM36 and Gromos-CKP simulations when bilayer is
+diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
+In lipid14/17 simulations, the headgroup order parameters of POPS shift closer
+to zero when the bilayer is diluted with POPC, but the numerical values of
+order parameters are too far from experiments, having also different signs,
+for the proper interpretation of the experimental data.
+Therefore, the imporvements in force fields are necessary
+for accurate studies of interactions between PC and PS headgroups
+with MD simulations.
 %The $\beta$-carbon order parameter
@@
 %with increasing amount of PS, both model exhibiting a poor agreement with experiments.
+
+%Therefore, the intermolecular interactions at the headgroup region seems
+%to be important for the physical properties of mixed lipid bilayers.
+%These interactions can be indirectly monitored by measuring the
+%headgroup order parameters from PS:PC mixtures with different molar
+%ratios.
 

### 2018-10-03 Manuscript/manuscriptPSsuppl.tex
Update text

@@
 \pagebreak
-\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFfigure}
+\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
 

### 2018-10-09 Manuscript/manuscriptPS.tex
Update text

@@
 and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
-The strongest sodium counterion binding, i.e., the lowest concentrations in bulk water,
-is observed in MacRog, Berger and Lipid17/JC simulations.
-Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
-counterion density, but slightly at different distances from the bilayer
-center. Other models exhibit more complicated distribution, often including
-two local maxima in the density profile.
+The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
+is reproduced only in Gromos-CKP simulations and in the MacRog simulation
+with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
+The counterion binding and concomintant electrostatic screening of the headgroup
+repulsion does not fully explain the low area per molecule values
+because the MacRog simulation with strongest sodium binding
+(the lowest concentrations in bulk water) gives the same area per molecule
+as CHARMM36ua simulation with significantly weaker counterion binding
+affinity.
+%CHARMM36, CHARMM36ua and Slipid models give
+%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
+On the other hand, changing counterions from sodium to potassium, having weaker binding
+affinity, increase the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
+%with potassium counterions and weaker
+%binding affinity give larger area per molecule 
+%increasing counterion binding affinity reduces the area per molecule
+%when comparing between sodium and potassium in 
+%or between
+%ff99 and jc ions in Lipid 17 simulations.
+In conclusion, the results are in line with the previous study suggesting that the
+low area per molecule in PS lipid bilayers originate from the combination
+of both counterion binding and hydrogen bonding network between lipid
+headgroups \cite{petrache04}.
+%, Berger and Lipid17/JC simulations.
+%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
+%counterion density, but slightly at different distances from the bilayer
+%center. Other models exhibit more complicated distribution, often including
+%two local maxima in the density profile.
 %CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
 %density, while a single maxima is observed in the other models. 
-
-Area per lipid is in agreement with experiments \cite{pan14} only
-in the MacRog simulation with potassium counterions and in Gromos-CKP models,
-while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
-The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
-counterion binding because CHARMM36, CHARMM36ua and Slipid models give
-smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
-On the other hand, increasing counterion binding affinity reduces the area per molecule
-when comparing between sodium and potassium in MacRog simulations or between
-ff99 and jc ions in Lipid 17 simulations.
-This suggests that the counterion binding alone cannot explain the low area per molecule
-in PS lipid bilayers but also the hydrogen bonding network has to
-be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
 \begin{figure}[]
@@
 
-Evaluation of counterion binding affinity to charged lipid bilayers
-using the changes of headgroup order parameters is difficult because
-reference state without ions does not exist, in contrast to previous
-studies with zwitterionic membranes and coions \cite{catte16},
-see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
+Binding of coions to zwitterionic PC lipid bilayers has been previously
+evaluated against experiments using the changes of headgroup order parameters
+as a function of ion concentration \cite{catte16}. This is less straighforward
+for charged lipid bilayers because counterions are always present and the
+ion free reference state does not exist
+(section \ref{mixtureTOadditionalCIs} in the supplementary information).
 However, the counterion binding affinity is

### 2018-10-10 Manuscript/manuscriptPS.tex
Update text

@@
 (section \ref{mixtureTOadditionalCIs} in the supplementary information).
-However, the counterion binding affinity is
-reflected to the increase of POPC headgroup order parameters due to the
-increasing amount of negatively charged POPS lipids in the bilayer.
-The experimentally measured increase in the order parameters~\cite{scherer87}
-is reproduced only in MacRog simulations, while other simulations
-predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
-The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
-In experiments and MacRog simulations, the increasing amount of negative charge
-in the membrane tilts the PC headgroup more parallel to the membrane plane
-leading to the larger order parameters. In Berger simulations,
+Here we evaluate the amount of bound charge using the changes of
+headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
+According to the electrometer concept, the headgroup order parameters of POPC
+increase when negativaly charged POPS lipids are incorporated in lipid bilayer
+(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
+This is reproduced in the MacRog simulations with potassium counterions having
+the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}),
+while other simulations predict no change or decrease in the order parameters
+(Fig. \ref{HGorderparametersPCvsPS}).
+%In experiments and MacRog simulations, the increasing amount of negative charge
+%in the membrane tilts the PC headgroup more parallel to the membrane plane
+%leading to the larger order parameters.
+In Berger simulations,
 the strong sodium binding surpasses the effect of negatively charged
@@
 of lipids at the interface and the headgroup order parameters are not changed
-with increasing amount of PS. Therefore, the relatively weak binding of potassium
+with increasing amount of PS.
+Therefore, the relatively weak binding of potassium
 in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct

### 2018-10-11 Manuscript/manuscriptPS.tex
Update text

@@
 %leading to the larger order parameters.
-In Berger simulations,
-the strong sodium binding surpasses the effect of negatively charged
-lipids and the order parameters decrease with the increasing amount of
-positive counterions of PS lipid.
-In CHARMM36 simulations, the counterion binding neutralizes the negative charge
-of lipids at the interface and the headgroup order parameters are not changed
-with increasing amount of PS.
-Therefore, the relatively weak binding of potassium
-in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
-response of POPC headgroup to the increasing amount of negatively charged lipids,
-indicating that these simulations have realistic surface charge density in membranes
-containing PS lipids.
+In Berger and CHARMM36 simulations, the stronger counterion binding cancel
+the effect of negatively charged headgroups and the headgroup order parameters
+do not increase with increasing amount of PS lipids.
+%the order parameters decrease with the increasing amount of
+%positive counterions of PS lipid.
+%In CHARMM36 simulations, the counterion binding neutralizes the negative charge
+%of lipids at the interface and 
+Therefore, we suggest that the relatively weak binding of potassium
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
+realistic surface charge density in membranes containing PS lipids,
+while the other tested simulation models overestimate the counterion
+binding affinity.
+%reproduces the correct
+%response of POPC headgroup to the increasing amount of negatively charged lipids,
+%indicating that these simulations have 
 \begin{figure*}[]

### 2018-10-11 Manuscript/manuscriptPS.tex
Update text

@@
 
-Dilution of PS lipid bilayers with PC lipids reduces the propensity
-of PS headgroup-multivalent ion complexes and is proposed to make PS
-headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
-This is related to the shift of the PS headgroup order parameters 
-closer to zero when bilayer is diluted with PC
-lipids (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90}.
-%which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
-This shift is not seen in CHARM36 and Gromos-CKP simulations when bilayer is
+The headgroup order parameteres of POPS shift closer to zero when bilayer
+is diluted with POPC (Fig.~\ref{HGorderparametersPCvsPS}),
+which is intereted to indicate less rigid structure of PS headgroups in
+the mixture~\cite{browning80,buldt81,roux90,roux91,scherer87}.
+This shift is observed only in lipid14/17 simulations
+%, the headgroup order parameters of POPS shift closer
+%to zero when the bilayer is diluted with POPC,
+but the numerical values of order parameters are too far from experiments, having also different signs,
+for the proper interpretation of the experimental data.
+In CHARM36 and Gromos-CKP simulations, the shift of headgroup order
+parameters toward zero are not observed when bilayer
 diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
-In lipid14/17 simulations, the headgroup order parameters of POPS shift closer
-to zero when the bilayer is diluted with POPC, but the numerical values of
-order parameters are too far from experiments, having also different signs,
-for the proper interpretation of the experimental data.
-Therefore, the imporvements in force fields are necessary
-for accurate studies of interactions between PC and PS headgroups
-with MD simulations.
+Therefore, we conclude that more accurate force fields are necessary
+for MD simulation studies of PC-PS headgroup interactions.  
 %The $\beta$-carbon order parameter

### 2018-10-12 Manuscript/manuscriptPS.tex
Update text and figures. Figure updates related to https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21

@@
 
-\subsection{Ca$^{2+}$ binding affinity in bilayers with negatively charged PS lipids}
+\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
 
-Ion binding affinity to PS containing membranes can be most conveniently measured 
-from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
-not observed~\cite{feigenson86,mattai89,roux90,roux91}. In addition, the ion
-binding affinity to such mixtures can be detected using the PC lipid headgroup order parameters,
-see section~\ref{electrometerFORmixtures}. As expected from the previous study of pure PC lipid
+Ion binding affinity to membranes containing negative charged PS lipids can be
+measured by detecteding the PC lipid headgroup order parameters
+from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
+where the dehydrated lipid-ion complexes and phase separation are 
+not observed~\cite{feigenson86,mattai89,roux90,roux91}.
+As expected from the previous study of pure PC lipid
 bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
@@
 
-The headgroup order parameters of POPS headgroup measured from POPC:POPS (5:1) mixture
-exhibit a strong dependence of CaCl$_2$ with small concentrations with a rapid saturation
-below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order parameter of
-POPS increase with the added CaCl$_2$ in the experiment and in all the tested simulation models,
-but simulations significantly overestimated the change. The larger $\alpha$-carbon order
-parameter of POPS decrease and the smaller one slightly increase with the added
-CaCl$_2$ in the experiment. The changes are again significantly overestimated in
-the simulations, however, in this case all simulations predict qualitatively different
-behaviour. Notably, the changes of POPS headgroup order parameters are overestimated
-also in the CHARMM36/NBfix model where the calcium binding affinity was too low. 
-We conclude that the effect of bound ions to the headgroup order parameters of POPS
-is not qualitatively reproduced by the tested simulations models. This is in contrast
-to previous results for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
-observed despite of significant discrepancies in the headgroup structure without additional ions. 
-The response of POPS headgroup order parameters to the bound charge is systematic but
-less well understood than the responce of PC headgroups used in the electrometer
-concept \cite{seelig87,roux90}. The force field development is necessary to
-generate MD simulations that could be used to explain the interactions between
-PS headgroup and calcium ions.
+The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
+exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
+below 100 mM (Fig. \ref{changesWITHCaClPS}).
+The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
+decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
+the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
+tested simulation models, including the CHARMM36 with the NBfix for calcium \cite{kim16},
+where the binding affinity was underestimated. 
+In addition, different simulation models predict qualitatively different behaviour
+for the POPS $\alpha$-carbon order parameters with the added calcium.
+For example, both order parameters decrease in Berger simulations but increase
+in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
+more complicated. This is in contrast to the PC headgroup, where
+qualitatively correct reponse to the bound ions is observed
+in all simulation models despite of the significant discrepancies in the headgroup
+structure without additional ions~\cite{catte16}. Therefore, we conclude that
+the improvement of force fields is necessary to correctly describe interactions between
+PS headgroup and calcium ions using MD simulations.
+%The response of POPS headgroup order parameters to the bound charge is systematic but
+%less well understood than the responce of PC headgroups used in the electrometer
+%concept \cite{seelig87,roux90}.
 

### 2018-10-25 Manuscript/manuscriptPSsuppl.tex
Updates in the SI

@@
 changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
-\todo{Discussion about Lipid17 to be written when we have the density profiles.}
+\todo{Discussion about Lipid17 have be written based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21}
 All the tested models overestimate the changes of POPS headgroup order parameters as
@@
   }
-  \todo{Lipid 17 is to be added when we have the density profiles.
-  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
 \end{figure}
@@
     the NBfix term employed for sodium \cite{venable13} ({\it left})
-    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
+    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}
+    and simulations without NBfix for the calcium.
     Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},

### 2018-10-31 Manuscript/manuscriptPS.tex
Updates in text and figures
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21 and some other updates

@@
     POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
-    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
+    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
     POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
@@
     \hline
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & ? & ? & \cite{??} \todoi{To be added by Ollila}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & 80 & 120 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
     POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
@@
   }
-  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4} \\
-  \todo{Lipid17 and MacRog results should be added into this plot.}
 \end{figure}
@@
 models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
-as previously observed also for PC lipids \cite{botan15}. The models for PS lipids perform generally less well than for
-PC lipids in the previous study
+as previously observed also for PC lipids \cite{botan15}.
+The models perform generally less well for PS lipids than for PC lipids in the previous study
 (Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
@@
 
-\subsection{Counterion binding and interactions between PC and PS headgroups}
+\subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
 Membranes containing PS lipids are always accomppanied with counterions which
@@
 On the other hand, changing counterions from sodium to potassium, having weaker binding
-affinity, increase the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
+affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
 %with potassium counterions and weaker
@@
 for charged lipid bilayers because counterions are always present and the
-ion free reference state does not exist
-(section \ref{mixtureTOadditionalCIs} in the supplementary information).
-Here we evaluate the amount of bound charge using the changes of
+ion free reference state does not exist.
+In addition, the analysis is complicated by the artificial aggregation of counterions
+in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
+Here, we evaluate the amount of bound charge using the changes of
 headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
@@
 while the other tested simulation models overestimate the counterion
-binding affinity.
+binding affinity. The results are in line with the changes of headgroup order
+parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
+in the supplementary information.
 %reproduces the correct
@@
 is diluted with POPC (Fig.~\ref{HGorderparametersPCvsPS}),
-which is intereted to indicate less rigid structure of PS headgroups in
+which is interpreted to indicate less rigid structure of PS headgroups in
 the mixture~\cite{browning80,buldt81,roux90,roux91,scherer87}.
@@
   \todo{Should we include also counterions into the plot?} \\
+  \todo{Figure needs general improvement.}
 \end{figure}

### 2018-10-31 Manuscript/manuscriptPSsuppl.tex
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21 and some other updates

@@
     Headgroup order parameters of POPC measured from mixtures with
-    PS (bovine brain), POPG, POPE, cholesterol and cationic surfactant~\cite{scherer87,scherer89,ferreira13}.
+    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
     Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
@@
 in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
-the headgroup order parameter responses were compared more carefully with the experiments of cationic
-surfactants in POPC bilayer~\cite{scherer89}. The comparison reveals that the both headgroup
-order parameters in the Lipid14 model are too sensitive to the bound charge,
+the headgroup order parameter responses were compared more carefully with the experiments of
+cationic dihexadecyldimethylammonium bromide (DHAB) surfactants in POPC bilayer~\cite{scherer89}.
+The comparison reveals that the both headgroup order parameters in the Lipid14 model are too sensitive to the bound charge,
 while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
@@
 \section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
-
+Previous studies using the electrometer concept to assess the ion binding affinity to lipid
+bilayers report ion concentrations either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
+or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
+In this work, we use the former definition to be consistent with the reference
+experimental data \cite{roux90}. The difference between these two concentrations increases
+with the increasing ion binding affinity. However, the difference is essentially negligible
+in the recent model \cite{melcr18} with realistic ion binding affinity (Fig. \ref{concentrationDEFfigure}).
 \begin{figure}[]
@@
     for the two possible definitions of ion concentration, concentration in buffer before solvating the
-    lipids (buffer concentration) and concenrtration in bulk after solvating the lipids (bulk concentration) \cite{melcr18}. 
+    lipids (buffer concentration) and concentration in bulk after solvating the lipids (bulk concentration),
+    from the recent model with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}. 
 }
@@
 as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
-Experimental order parameter data for POPC headgroup in the mixture is available as a function
+Experimental order parameters of POPC headgroup in the mixture are available as a function
 of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
@@
 Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
-electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
-Berger model, the headgroup order parameter response of POPC
-to the added NaCl is similar to the experiments of LiCl,
-indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
-Indeed, the sodium density profile shows a significant binding peak in the 
-Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
-is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
-changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
-\todo{Discussion about Lipid17 have be written based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21}
-All the tested models overestimate the changes of POPS headgroup order parameters as
-a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
-model development is necessary to interpret the PS headgroup-ion interactions 
-from MD simulations.
+electrometer concept~\cite{akutsu81,altenbach84,seelig87}.
+
+In simulations with Berger and CHARMM36 models, the responses of both headgroup order parameter
+of POPC and the $\beta$ carbon of POPS to the added sodium or potassium are similar to the experiments
+of LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of the counterions 
+in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
+(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
+less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
+Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
+with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
+for which the data is not shown due to the formation of
+ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
+Such clusters appering also in MagRog simulations with 4 M concentration of KCl
+could explain the unsystematic changes of the order parameters in this model with the added KCl.
+In conclusion, the results are in line with the section \ref{ciBINDINGsection}
+in the main text, suggesting that the MacRog simulations with KCl give the most
+realistic surface charge at the lipid bilayer interface.
 \begin{figure*}[]
   \centering
-  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
+  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
   \caption{\label{PSresponseTONaCl}
@@
 \end{figure*}
-
 \begin{figure}[]
@@
 \end{figure}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
+  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
+  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
+    and MacRog simulations with 4 M of KCl (right).
+\label{ionCLUSTERS}
+  }
+\end{figure}
+\newpage
 
@@
 %saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
-\pagebreak
 

### 2018-11-20 Manuscript/manuscriptPS.tex

@@
 & 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.03 & 60:12 & ?
-& ?
-& 298  & ?
-& ?
-& \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.17 & 60:12 & ?
-& ?
-& 298  & ?
-& ?
-& \cite{??} \todoi{Data to be delivered by Melcr}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.36 & 60:12 & ?
-& ?
-& 298  & ?
-& ?
-& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0      & 60:12 & 3600  & 0   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.08   & 60:12 & 3561  & 5   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.13   & 60:12 & 3561  & 8   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.20   & 60:12 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.41   & 60:12 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.62   & 60:12 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
     \hline

### 2018-11-20 Manuscript/manuscriptPS.tex
Update text

@@
 
-\begin{figure*}[]
+\begin{figure}[]
   \centering
-  \includegraphics[width=18.0cm]{../Figs/figS8.png}
-  \caption{\label{HGstructuresPS}
+  \includegraphics[width=9.0cm]{../Figs/structures.eps}
+  \caption{\label{HGstructuresPSandPC}
     Overlayed snapshots from glycerol backbone and headgroup region from 
@@
   }
-\end{figure*}
+\end{figure}
 
@@
 The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
-models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
+models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
 as previously observed also for PC lipids \cite{botan15}.
@@
 in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
-visible in the sampled conformations (Fig.~\ref{HGstructuresPS}) and might
-manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
-\todo{We should consider making the conformation figures from CHARMM36 and Slipids POPC, and compare these
-  to the results from POPS.}
-
-\todo{Maybe we should move the glycerol backbone results to SI and focus here on the headgroup,
-and especially to the difference between PC and PS.}
-The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
-differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
-as observed previously also for PC lipids \cite{botan15}.
-The origin of this difference is more difficult to track without more elaborate analysis,
-because different models show very complicated patterns of distinct structures
-in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
+visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} a) and \ref{HGandGLYstructuresPS})
+suggesting that the rotation of carboxyl group is limited in the serine headgroup.
+In addition, the comparison of N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral between PC and PS headgroups
+from CHARMM36 simulations, which are in best agreement with experiments for both lipids,
+shows a more asymmetric and restricted
+rotations around this bond (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
+These results might manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
+Also the dihedral distributions and sampled conformations of glycerol backbone region
+from different models are shown in the supplementary information
+(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
+but further analysis is beyond the scope of this work which the PS headgroup.
 

### 2018-11-20 Manuscript/manuscriptPSsuppl.tex
Update text

@@
 
+The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
+differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
+as observed previously also for PC lipids \cite{botan15}.
+The origin of this difference is more difficult to track without more elaborate analysis,
+because different models show very complicated patterns of distinct structures
+in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
+
 \begin{figure*}[]
@@
 \end{figure*}
+
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=18.0cm]{../Figs/figS8.png}
+  \caption{\label{HGandGLYstructuresPS}
+    Overlayed snapshots from glycerol backbone and headgroup region from 
+    different simulations of PS lipids.
+  }
+\end{figure*}
+
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=16.0cm]{../Figs/figS7PC.png}
+  \caption{\label{dihedralsHGpc}
+    Dihedral angle distributions of headgroup region of PS lipids from different simulation models.
+  }
+\end{figure*}
+
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=15.0cm]{../Figs/figS6PC.png}
+  \caption{\label{dihedralsGLYpc}
+    Dihedral angle distributions of glycerol backbone region of PS lipids from different simulation models.
+  }
+\end{figure*}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/figS8_POPC.png}
+  \caption{\label{HGandGLYstructuresPSPC}
+    Overlayed snapshots from glycerol backbone and headgroup region from 
+    different simulations of PS lipids.
+  }
+\end{figure}
 

### 2018-11-21 Manuscript/manuscriptPS.tex
Update text

@@
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith450mMK}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMK}  \\
     POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith440mMNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMNa}  \\
     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from glycerol backbone and headgroup region from 
-    different simulations of PS lipids.
+    Overlayed snapshots from CHARMM36 simulations in best agreement with experiments
+    to demonstrate the conformational fluctuations around
+    a) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
+    of PS headgroup and b) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    dihedrals of PS and PC headgroups.
   }
+  \todo{We need atom labeling for b). Similar to the one in a) would be good} \\
 \end{figure}
@@
 The models perform generally less well for PS lipids than for PC lipids in the previous study
-(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
-Therefore, interpretation of structural differences between PC and PS headgroups
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
+Therefore, the interpretation of structural differences between PC and PS headgroups
 from simulations is not straightforward.
@@
 and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
-when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
+when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
 Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
@@
 suggesting that the rotation of carboxyl group is limited in the serine headgroup.
-In addition, the comparison of N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral between PC and PS headgroups
-from CHARMM36 simulations, which are in best agreement with experiments for both lipids,
-shows a more asymmetric and restricted
-rotations around this bond (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
-These results might manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
+In addition, the CHARMM36 simulations, in best agreement with experiments for both lipids,
+show more asymmetric and restricted N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral distribution 
+in PS headgroup than in PC (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
+These results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
 Also the dihedral distributions and sampled conformations of glycerol backbone region
-from different models are shown in the supplementary information
+significantly vary between different simulations models 
 (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
@@
 
-%which can be related to the differences in the dihedral angle distributions 
-%of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}).
-
-%  One possible conclusion could be the following:
-%  The main differences between the models in the headgroup region are observed
-%  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
-%  CHARMM36, CHARMM36UA and Slipids give very similar results to the
-%  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
-%  The order parameters of $\beta$-carbons for these three models
-%  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
-%  On the other hand, Gromos-CKP models give better order parameters
-%  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
-%  In conclusion, the suggestion would be that the single peak for
-%  observed at 120 degrees in CHARMMs and Slipids would
-%  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
-%  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
-%}
+The observed conformations of PS headgroup may be useful when interpreting experiments and
+can guide the further force field development. However, more accurate simulations are
+required to confirm the results because the simulated PS headgroup order parameters
+are not within the experimental error bars in any of the tested models.
 
@@
 \subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
+  \caption{\label{NAdensPOPS}
+    Counterion densities of POPS lipid bilayer along the membrane normal from
+    simulations with different force fields.
+  }
+\end{figure}
+
+\begin{figure*}[ht]
+  \centering
+  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
+  \caption{\label{changesWITHCaClPS}
+    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
+    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
+    The order parameter values from systems without calcium are set as the zero point of y-axis,
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
+    such that the lower order parameter is zero without additional ions to correctly illustrate
+    the forking with different concentrations of calcium.
+    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
+    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
+  }
+  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
+  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
+\end{figure*}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
+  \caption{\label{CAdensPCPSmixture}
+    Ca2+ density profiles from simulations.
+  }
+  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
+  \todo{Should we include also counterions into the plot?} \\
+  \todo{Figure needs general improvement.}
+\end{figure}
+
 Membranes containing PS lipids are always accomppanied with counterions which
@@
 %density, while a single maxima is observed in the other models.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
-  \caption{\label{NAdensPOPS}
-    Counterion densities of POPS lipid bilayer along the membrane normal from
-    simulations with different force fields.
-  }
-\end{figure}
 
@@
 
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
-  \caption{\label{changesWITHCaClPS}
-    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
-    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
-    The order parameter values from systems without calcium are set as the zero point of y-axis,
-    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
-    such that the lower order parameter is zero without additional ions to correctly illustrate
-    the forking with different concentrations of calcium.
-    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
-    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
-  }
-  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
-  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
-\end{figure*}
-
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
-  \caption{\label{CAdensPCPSmixture}
-    Ca2+ density profiles from simulations.
-  }
-  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
-  \todo{Should we include also counterions into the plot?} \\
-  \todo{Figure needs general improvement.}
-\end{figure}
-
 The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture

### 2018-12-02 Manuscript/manuscriptPS.tex
Update text

@@
 been available. Here, we collect a set of experimental NMR data which
-could be used together with molecular dynamics (MD) simulations
+can be used together with molecular dynamics (MD) simulations
 to interpret the lipid headgroup structures and details of ion binding
-in pure and mixed PS and PS:PC lipid bilayers. Aiming to interpret the
-data, we use the open collaboration method to go through the available MD simulation models for PS lipids.
-However, none of the models reproduce the experimental data with sufficient
-accuracy to interpet the structural details of lipid headgroups or
-ion binding details in lipid bilayers containing PS lipids.
-In contrast to PC lipids, the tested MD simulation models do not
-correctly reproduce the qualitative response of PS lipid headgroups
-to the bound ions or changes in the lipid composition. 
-Our results pave the way for the model improvement to correctly
+in pure PS and mixed PS:PC lipid bilayers. We use the open collaboration
+method to collect data from available MD simulation models of PS lipids.
+None of the models reproduce the NMR data with experimental accuracy,
+but the best models suggest that the carboxyl group in the serine headgroup
+does not rotate freely. In line with the previous results for PC lipids,
+all the tested force fields overestimate the cation binding affinity to lipid bilayers containing
+PS lipids. In contrast to PC lipids, the response of PS headgroups
+to the bound ions qualitatively differs from experiments in the tested MD
+simulation models. The collected experimental dataset and simulation
+results pave the way for improvement of lipid force fields to correctly
 describe negatively charged membranes and their interactions with ions.
+The work is performed in the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
 \end{abstract}
@@
 to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16}.
-However, the recent comparisons of PC lipid headgroup and glycerol backbone
-C-H bond order parameters calculated from different simulation models
-revealed that improvements in the current force fields are needed
-to correctly reproduce the headgroup structure and ion binding to
-lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
-to POPC bilayer was then improved by implicitly including the electronic
-polarizability using the electronic continuum correction \cite{melcr18}.
-Here, we collect the set of experimentally measured lipid headgroup and
-glycerol backbone C-H bond order parameters, which can be used to
-evaluate the quality of headgroup structure and the ion binding affinity
-in MD simulations of lipid bilayers containing PS lipids. 
-The available MD simulation models of PS are then compared against
-the collected experimental data. The results pave the way for the
-development of MD simulation force fields that correctly describe
-PS lipid headgroup structure and its interactions with ions.
-Such models are expected to be useful in elucidating the biological
-role of PS and other lipid headgroups because glycerol backbone and
-lipid headgroups behave similarly in model membranes and in
-bacteria \cite{gally81,scherer87,seelig90}.
+interactions with
+ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18}.
+However, the headgroup structures and ion binding affinity 
+predicted by the MD simulation force fields for bilayers containing PS lipids are not
+evaluated against experiments. Recent studies comparing the
+C-H bond order parameters between simulations and experiments show that the
+structure of PC lipid headgroup and glycerol backbone are not well
+captured by most simulation models \cite{botan15} and that the cation binding to PC
+lipid bilayers is overestimated \cite{catte16}. Based on this data,
+the cation binding affinity to POPC bilayer was then improved by implicitly
+including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
 
-%In NMRlipids I and II project we were looking for a MD model
-%which would correctly reproduce headgroup and glycerol
-%backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
-%Here we extend the same goal for lipids with negatively charged PS headgroup.
-%Chemical structure of PS headgroup together with other common biological
-%lipids is shown in Fig. \ref{lipids}.
-%\begin{figure}[]
-
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
-%  \caption{\label{lipids}
-%    Chemical structures and labels for the headgroup carbons.
-%  }
-%\end{figure}
-
-%The interpretation of this data and some other results has been that \cite{seelig90}
-%\begin{displayquote}
-%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
-%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
-%    (ii) the free Ca$^{2+}$
-%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
-%    charge, either due to protein or to lipid; \\
-%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
-%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
-%    (iv) the actual binding step can be described by a Langmuir
-%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
-%    used to describe the chemical binding equilibrium.''}
-%\end{displayquote}
-
-%Phospholipids containing various polar headgroups and acyl
-%chains are essential building blocks of biological membranes.
-%Atomistic level structural details of lipids and lipid-ion 
-%interactions are considered highly important in several
-%biological processes. The lipid structure and ion binding
-%can be studied in detail with NMR spectroscopy. However,
-%the structural interpretation of NMR data requires usage
-%of models. The combination of classical molecular dynamics
-%simulations with NMR data, especially with C-H bond
-%order parameters, can potentially give atomistic resolution
-%interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.
-
-%Our recent studies concluded that MD models are cabable to
-%give structural interpretation for phosphatidylcholine 
-%hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
-%backbone and cation binding posed a major challenge for current
-%force fields \cite{botan15,ollila16,catte16,ferreira16}. 
-%These conclusions were reached by reviewing extensively 
-%available experimental data from various sources and using
-%NMRlipids Open Collaboration project to collect massive 
-%amount of simulation data \cite{botan15,catte16}.
-
-%Here we apply the same approach in search of MD simulation
-%models which would reproduce glycerol backbone and 
-%headgroup structures of lipids with PE, PG and PS headgroups.  
-%In addition, we attempt to find a MD simulation model
-%that would be able to correctly describe cation binding in
-%bilayer containing negatively charged PG and PS lipids.
-
-%Several MD simulation models have succesfully described 
-%acyl chain structures and their qaulitative changes in different
-%conditions for phosphatidylcholine lipids (for review see \cite{ollila16}).
-%However, the current simulation models were not found to be accurate
-%enough for full structural interpretation of phosphatidylcholine headgroup and
-%glycerol backbone \cite{botan15}. Also Na$ +$  binding affinities
-%were found to be significantly overestimated in several models and 
-%none of the available models was able to Ca$2+$
-
-\begin{table*}[!htb]
+\begin{table*}[htb]
 %\begin{sidewaystable*}[!p]
 \centering
-\caption{List of MD simulations of pure PS bilayers without additional salt.
-   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
-   and CKP refers to the version with more Gromos compatible version
-   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
-   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+\caption{The list of MD simulations of pure PS bilayers without additional salt.
+  Simulation details are given in the supplementary information
     %   The lipid force fields named as in our previous work~\cite{botan15}.
@@
 
+Here, we collect the set of experimentally measured lipid headgroup and
+glycerol backbone C-H bond order parameters, which can be used to
+evaluate the quality of headgroup structure and the ion binding affinity
+in MD simulations of lipid bilayers containing PS lipids. 
+The available MD simulation models of PS are then compared with
+the collected experimental data using the NMRlipids open collaboration
+project (\url{www.nmrlipids.blogspot.fi}). The results pave the way
+for the development of lipid force fields with realistic description of
+the headgroup region of negatively charged lipids in physiological salt
+conditions. Such models are expected to be useful in understanding
+biological function of lipid headgroups and glycerol backbone because
+they behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.
+
 \begin{table*}[!p]
@@
 \centering
-\caption{List of POPC:POPS mixture simulations with different amounts of added ions. 
-  The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-  these correspond the concentrations reported in the experiments by Roux et al.~\cite{roux90}.
+\caption{The list of POPC:POPS mixtures simulated with different amounts of added ions. 
+  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  This corresponds the concentration in buffer before solvating lipids, which are
+  reported in the experiments by Roux et al.~\cite{roux90}.
+  The simulation details are given in the supplementary information.
    % The lipid force fields named as in our previous work~\cite{botan15}.
@@
 Headgroup and glycerol backbone C-H bond order parameters of POPS
-were determined from chemical-shift resolved dipolar splittings
-measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}
-and corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
+were determined from the chemical-shift resolved dipolar splittings
+measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}.
+The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
 using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
@@
 dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
-Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
+The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
 distribution of bilayer orientations in the samples studied.
@@
 the standard error of the mean were then calculated over different lipids.
+The python program ({\it scripts/calcOrderParameters.py}) utilizing
+MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
 The number density profiles were calculated using {\it gmx density} tool
@@
   \caption{\label{HGorderParameters}
-    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
+    (A) Chemical structures and labels for the headgroup carbons.
+    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
     with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 

### 2018-12-02 Manuscript/manuscriptPSsuppl.tex
Update text

@@
 \todo{To be written by Piggot}
+  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
+   and CKP refers to the version with more Gromos compatible version
+   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
 

### 2018-12-05 Manuscript/manuscriptPS.tex
Update text and figures

@@
 the headgroup order parameters can be directly calculated from 
-simulations~\cite{catte16}. Also the headgroup order parameters
+simulations and compared to the experimental data with varying cation
+concentrations~\cite{catte16}. The headgroup order parameters
 of negatively charged PS and PG lipids exhibit systemic, but less
@@
 
+\begin{figure}[!tb]
+  \centering
+  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
+  \caption{\label{PShgSIGNSsimpson}
+    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
+    (B) 2D R-PDLF spectra.
+    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
+    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
+    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller $\alpha$-carbon order parameter (dashed grey).
+  }
+  \todo{I think that the peak labeling would be good to show also in (A).}
+\end{figure}
+
 The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
@@
 confirming the interpretation that the smaller order parameter is negative.
-\begin{figure}[!tb]
-  \centering
-  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
-  \caption{\label{PShgSIGNSsimpson}
-    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
-    (B) 2D R-PDLF spectra.
-    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
-    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller $\alpha$-carbon order parameter (dashed grey).
-  }
-\end{figure}
 
@@
 the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking in PS headgroup.
+experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
 available. 
-%The detailed structural differences between the headgroups is, however, not known.
 \begin{figure}[]
@@
   \caption{\label{HGorderParameters}
-    (A) Chemical structures and labels for the headgroup carbons.
+    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
     (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
-    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
     POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
-    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
+    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
   }
 \end{figure}
-
-%\begin{figure}[]
-%  \centering
-%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
-%  \caption{\label{PShgSIGNS}
-%    Experimental results for sign measurement for POPS sample
-%  }
-%\end{figure}
 
@@
   \caption{\label{HGorderParametersPS}
-    Order parameters for PS headgroup and glycerol
-    backbone from simulations with different models and experiments without CaCl$_2$.
-    All DOPS data at 303~K, POPS at 298~K.
-    Experimental data from \cite{browning80} contain 0.1~M of NaCl.
-    Signs are taken from experiments for POPS described in Supplementary Information.
-    The vertical bars shown %for most computational values
-    are not error bars, but demonstrate that %for these systems
-    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
+    Order parameters of PS headgroup and glycerol backbone from experiments
+    and simulations with different force fields.
+    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
+    while all the other data is without additional ions.
+    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
+    Light blue areas spanning 0.04 units around the average of the extremal experimental values,
+    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
+    The vertical bars shown for some simulation values
+    are not error bars, but demonstrate that for these systems
+    we had at least two data sets; the ends of the bars mark the extreme values
+    from the sets, and the dot marks their measurement-time-weighted average.
   }
@@
     to demonstrate the conformational fluctuations around
-    a) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
-    of PS headgroup and b) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
+    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
     dihedrals of PS and PC headgroups.
   }
-  \todo{We need atom labeling for b). Similar to the one in a) would be good} \\
+  \todo{We need atom labeling for (B). Similar to the one in (A) would be good} \\
 \end{figure}
@@
 
-The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
-models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
+The headgroup and glycerol backbone of PS lipids show wide variety
+in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS})
+between different simulation models,
 as previously observed also for PC lipids \cite{botan15}.
@@
 
-The best performing models,
+However, the best performing models,
 Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
@@
 when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
-Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
-a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
-in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
-visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} a) and \ref{HGandGLYstructuresPS})
+Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
+a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
+in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
+The restricted motion is also
+visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
 suggesting that the rotation of carboxyl group is limited in the serine headgroup.
-In addition, the CHARMM36 simulations, in best agreement with experiments for both lipids,
-show more asymmetric and restricted N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral distribution 
-in PS headgroup than in PC (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
-These results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
-Also the dihedral distributions and sampled conformations of glycerol backbone region
-significantly vary between different simulations models 
+In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
+and restricted angle distribution for PS than for PC headgroup in
+CHARMM36 simulations in best agreement with experiments
+(Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
+The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
+Also, the sampled conformations of glycerol backbone
+significantly vary between different simulation models 
 (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
@@
 \subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
-\begin{figure}[]
+
+\begin{figure}[!htb]
   \centering
@@
 \end{figure}
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
+  \caption{\label{HGorderparametersPCvsPS}
+    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
+    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\cite{ferreira16}.
+  }
+  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
+  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
+\end{figure}
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
+  \caption{\label{HGorderparametersPSvsPC}
+    Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
+    Experimental values and order parameter signs of pure POPS and values in mixture are  
+    measured at 298~K in this work and in Ref. \citenum{roux90}, respectively.
+    Since the experimental data of POPS in pure and diluted mixture come from
+    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
+    the experimental change of the order parameter is less accurate than in typical measurements
+    where same technique is used in all conditions, see discussion about qualitative and quantitative 
+    accuracy in Ref. \citenum{ollila16}.
+    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
+    with the same value for both order parameters such that the lower order
+    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
+  }
+\end{figure}
+
+Membranes containing PS lipids are always accomppanied with counterions which
+modulate electrostatic interactions between lipids and other biomolecules. 
+Counterions are also suggested to screen the repulsion between charged lipid headgroups 
+in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
+bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
+The counterion density profiles along membrane normal 
+show significant differences between simulation models in both binding affinity
+and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
+The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
+is reproduced only in Gromos-CKP simulations and in the MacRog simulation
+with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
+The counterion binding and concomintant electrostatic screening of the headgroup
+repulsion does not fully explain the low area per molecule values
+because the MacRog simulation with strongest sodium binding
+(the lowest concentrations in bulk water) gives the same area per molecule
+as CHARMM36ua simulation with significantly weaker counterion binding
+affinity. On the other hand, changing counterions from sodium to potassium, having weaker binding
+affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
+In conclusion, the results are in line with the previous study suggesting that the
+low area per molecule in PS lipid bilayers originate from the combination
+of both counterion binding and hydrogen bonding network between lipid
+headgroups \cite{petrache04}.
 
@@
 
+Binding of cations to zwitterionic PC lipid bilayers has been previously
+evaluated against experiments using the changes of headgroup order parameters
+as a function of ion concentration \cite{catte16}. This is less straighforward
+for charged lipid bilayers because counterions are always present and the
+ion free reference state does not exist.
+In addition, the analysis is complicated by the artificial aggregation of counterions
+in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
+Therefore, we evaluate the amount of bound charge also using the changes of
+headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
+According to the electrometer concept, the headgroup order parameters of POPC
+increase when negativaly charged POPS lipids are incorporated in lipid bilayer
+(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
+This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS})
+having the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
+The CHARMM36 and Berger simulations predict no change or decrease
+in the POPC headgroup order parameters as a function of increased amount of POPS (Fig. \ref{HGorderparametersPCvsPS}).
+This can be explained by the stronger counterion binding affinity which cancels
+the effect of negatively charged headgroups and prevents the experimentally observed
+increase of headgroup order parameters with increasing amount of PS lipids.
+Therefore, we suggest that the relatively weak binding of potassium
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
+realistic surface charge density in membranes containing PS lipids,
+while the other tested simulation models overestimate the counterion
+binding affinity. The results are in line with the changes of headgroup order
+parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
+in the supplementary information.
+
+The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
+and other experimental results suggest less rigid structure of PS headgroups when diluted with
+POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
+None of the tested models reproduce the changes of POPS headgroup order
+parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
+Therefore, we conclude that more accurate force fields are necessary
+to correctly describe the PC-PS headgroup interactions in MD simulations.
+
+\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
 \begin{figure}[ht]
@@
 
-Membranes containing PS lipids are always accomppanied with counterions which
-modulate electrostatic interactions between lipids and other biomolecules. 
-Counterions are also suggested to screen the repulsion between charged lipid headgroups 
-in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
-bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
-The counterion density profiles along membrane normal 
-show significant differences between simulation models in both binding affinity
-and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
-The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
-is reproduced only in Gromos-CKP simulations and in the MacRog simulation
-with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
-The counterion binding and concomintant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values
-because the MacRog simulation with strongest sodium binding
-(the lowest concentrations in bulk water) gives the same area per molecule
-as CHARMM36ua simulation with significantly weaker counterion binding
-affinity.
-%CHARMM36, CHARMM36ua and Slipid models give
-%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
-On the other hand, changing counterions from sodium to potassium, having weaker binding
-affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
-%with potassium counterions and weaker
-%binding affinity give larger area per molecule 
-%increasing counterion binding affinity reduces the area per molecule
-%when comparing between sodium and potassium in 
-%or between
-%ff99 and jc ions in Lipid 17 simulations.
-In conclusion, the results are in line with the previous study suggesting that the
-low area per molecule in PS lipid bilayers originate from the combination
-of both counterion binding and hydrogen bonding network between lipid
-headgroups \cite{petrache04}.
-%, Berger and Lipid17/JC simulations.
-%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
-%counterion density, but slightly at different distances from the bilayer
-%center. Other models exhibit more complicated distribution, often including
-%two local maxima in the density profile.
-%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
-%density, while a single maxima is observed in the other models.
-
-Binding of coions to zwitterionic PC lipid bilayers has been previously
-evaluated against experiments using the changes of headgroup order parameters
-as a function of ion concentration \cite{catte16}. This is less straighforward
-for charged lipid bilayers because counterions are always present and the
-ion free reference state does not exist.
-In addition, the analysis is complicated by the artificial aggregation of counterions
-in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
-Here, we evaluate the amount of bound charge using the changes of
-headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
-According to the electrometer concept, the headgroup order parameters of POPC
-increase when negativaly charged POPS lipids are incorporated in lipid bilayer
-(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
-This is reproduced in the MacRog simulations with potassium counterions having
-the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}),
-while other simulations predict no change or decrease in the order parameters
-(Fig. \ref{HGorderparametersPCvsPS}).
-%In experiments and MacRog simulations, the increasing amount of negative charge
-%in the membrane tilts the PC headgroup more parallel to the membrane plane
-%leading to the larger order parameters.
-In Berger and CHARMM36 simulations, the stronger counterion binding cancel
-the effect of negatively charged headgroups and the headgroup order parameters
-do not increase with increasing amount of PS lipids.
-%the order parameters decrease with the increasing amount of
-%positive counterions of PS lipid.
-%In CHARMM36 simulations, the counterion binding neutralizes the negative charge
-%of lipids at the interface and 
-Therefore, we suggest that the relatively weak binding of potassium
-in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
-realistic surface charge density in membranes containing PS lipids,
-while the other tested simulation models overestimate the counterion
-binding affinity. The results are in line with the changes of headgroup order
-parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
-in the supplementary information.
-%reproduces the correct
-%response of POPC headgroup to the increasing amount of negatively charged lipids,
-%indicating that these simulations have 
-\begin{figure*}[]
-  \centering
-  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
-  \caption{\label{HGorderparametersPCvsPS}
-    Changes of PC (left panel) and PS (right panel) headgroup order parameters
-    from POPC:POPS mixtures with increasing amount of POPS.
-    Experimental results of POPC are taken from Ref. \citenum{scherer87}
-    (signs are determined as discussed in \cite{botan15,ollila16}).
-    Experimental values for POPS in pure bilayer and in mixture are  
-    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
-    Since the experimental data of POPS in pure and diluted mixture come from
-    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
-    the experimental change of the order parameter is less accurate than in typical measurements
-    where same technique is used in all conditions, see discussion about qualitative and quantitative 
-    accuracy in Ref. \citenum{ollila16}.
-    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
-    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
-    set to the value from POPC:POPS (5:1) mixture.
-    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
-    with the same value for both order parameters such that the lower order
-    parameter value from POPC:POPS (5:1) mixture is at zero
-    to correctly illustrate the significant forking.
-  }
-  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
-  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
-\end{figure*}
-
-The headgroup order parameteres of POPS shift closer to zero when bilayer
-is diluted with POPC (Fig.~\ref{HGorderparametersPCvsPS}),
-which is interpreted to indicate less rigid structure of PS headgroups in
-the mixture~\cite{browning80,buldt81,roux90,roux91,scherer87}.
-This shift is observed only in lipid14/17 simulations
-%, the headgroup order parameters of POPS shift closer
-%to zero when the bilayer is diluted with POPC,
-but the numerical values of order parameters are too far from experiments, having also different signs,
-for the proper interpretation of the experimental data.
-In CHARM36 and Gromos-CKP simulations, the shift of headgroup order
-parameters toward zero are not observed when bilayer
-diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
-Therefore, we conclude that more accurate force fields are necessary
-for MD simulation studies of PC-PS headgroup interactions.  
-%The $\beta$-carbon order parameter
-%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
-%to the experimental data.
-%The smaller $\alpha$-carbon order parameter increase in both
-%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
-%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
-%with increasing amount of PS, both model exhibiting a poor agreement with experiments.
-
-%Therefore, the intermolecular interactions at the headgroup region seems
-%to be important for the physical properties of mixed lipid bilayers.
-%These interactions can be indirectly monitored by measuring the
-%headgroup order parameters from PS:PC mixtures with different molar
-%ratios.
-
-\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
-
-Ion binding affinity to membranes containing negative charged PS lipids can be
+Calcium binding affinity to membranes containing negative charged PS lipids can be
 measured by detecteding the PC lipid headgroup order parameters
@@
 not observed~\cite{feigenson86,mattai89,roux90,roux91}.
+Despite the lack of ion free reference state
+in the presence of negatively charged lipids, the simulations give
+coherent results for POPC headgroup order parameters as a function of
+CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
 As expected from the previous study of pure PC lipid
-bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
-POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
-with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
-calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
+bilayers \cite{catte16}, almost all the tested simulation models overestimate the
+experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
+in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
+indicating overestimated calcium binding binding affinity.
+Only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameter 
 indicating weaker binding affinity than experiments.
 %incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
-Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
+Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
 calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
@@
 in all the other tested models.
-
-%It should be noted, however, that the lowest concentration (100mM) gives
-%a good agreemet with experiments
-%\todo{Should be analyze/discuss this further?
-%  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
-%It should be noted that Berger simulation do not have counterions.}.
 

### 2018-12-06 Manuscript/manuscriptPS.tex
Update calcium density plot

@@
     POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.12\footnote{Calculation of concetration complicated due the scaled ions.
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.102\footnote{Calculation of concetration complicated due the scaled ions.
 Concentration taken as reported in the delivered data.}   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
@@
   \centering
-  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWcons.eps}
   \caption{\label{CAdensPCPSmixture}
@@
   \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
+  \todo{y-axis scale of Lipid17 should be checked.} \\
   \todo{Should we include also counterions into the plot?} \\
-  \todo{Figure needs general improvement.}
 \end{figure}

### 2018-12-06 Manuscript/manuscriptPSsuppl.tex
Update calcium density plot

@@
 \pagebreak
+\section{Calcium density profiles from simulations with POPC:POPS(5:1) mixture}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
+  \caption{\label{CAdensPCPSmixtureALL}
+    Ca2+ density profiles from simulations.
+  }
+  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
+  \todo{Should we include also counterions into the plot?} \\
+\end{figure}
+
+\pagebreak
 \section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

### 2018-12-06 Manuscript/manuscriptPS.tex
Update text

@@
     dihedrals of PS and PC headgroups.
+    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
   }
@@
 calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
-data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
-manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
+data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
 is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
-in all the other tested models.
+in all the other tested models. This is evident in the calcium density distributions
+along membrane normal where almost all Ca$^{2+}$ ions bind to the membrane interface in
+all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
 
@@
 the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
-tested simulation models, including the CHARMM36 with the NBfix for calcium \cite{kim16},
-where the binding affinity was underestimated. 
+tested simulation models, including the CHARMM36 with underestimated binding affinity.
 In addition, different simulation models predict qualitatively different behaviour
@@
 the improvement of force fields is necessary to correctly describe interactions between
-PS headgroup and calcium ions using MD simulations.
-%The response of POPS headgroup order parameters to the bound charge is systematic but
-%less well understood than the responce of PC headgroups used in the electrometer
-%concept \cite{seelig87,roux90}.
+PS headgroup and calcium ions in MD simulations.
 

### 2018-12-06 Manuscript/manuscriptPSsuppl.tex
Update text

@@
 Previous studies using the electrometer concept to assess the ion binding affinity to lipid
-bilayers report ion concentrations either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
-or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
+bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
+or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
 In this work, we use the former definition to be consistent with the reference
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
+  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
   \caption{\label{concentrationDEFfigure}
-    Difference between order parameter changes as a function of CaCl$_2$ concentration demonstrated
-    for the two possible definitions of ion concentration, concentration in buffer before solvating the
-    lipids (buffer concentration) and concentration in bulk after solvating the lipids (bulk concentration),
-    from the recent model with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}. 
+    Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
+    two possible definitions of ion concentration from the recent force field
+    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
+    together with the experimental data \cite{akutsu81,altenbach84}. 
   }
@@
 Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
-show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
+show a single splitting for the $\beta$-carbon with the order parameter value of 0.11,
 and a superposition of a large and a very small splitting for the $\alpha$-carbon.
-The larger splitting gives a order parameter value of 0.09, while the numerical value
+The larger splitting from the $\alpha$-carbon gives a order parameter value of 0.09, while the numerical value
 from the small splitting cannot resolved with the available resolution.
@@
   \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
+    The splitting corresponding the larger $\alpha$-carbon order parameter should be also shown.
   Also, the size of the figure in the file may not be optimal.}
@@
 
-The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
-differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
-as observed previously also for PC lipids \cite{botan15}.
-The origin of this difference is more difficult to track without more elaborate analysis,
-because different models show very complicated patterns of distinct structures
-in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
+The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
+wide variety between different simulation models (Figs. \ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
+Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
+headgroup and glycerol backbone order parameters. However, some structural characteristics
+of PS headgroup can be suggested based on the best available models
+(Figs. \ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
+The glycerol backbone structures are not discussed in this works because our focus is on PS headgroup.
+However, the data presented here can be useful for future investigations.
+%backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
+%differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
+%as observed previously also for PC lipids \cite{botan15}.
+%The origin of this difference is more difficult to track without more elaborate analysis,
+%because different models show very complicated patterns of distinct structures
+%in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
 
@@
   \caption{\label{dihedralsHG}
-    Dihedral angle distributions of headgroup region of PS lipids from different simulation models.
+    Dihedral angle distributions of the headgroup region of POPS from different simulation models.
   }
@@
   \centering
-  \includegraphics[width=15.0cm]{../Figs/figS6.png}
+    \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
+  \includegraphics[width=13.0cm]{../Figs/figS6.png}
   \caption{\label{dihedralsGLY}
-    Dihedral angle distributions of glycerol backbone region of PS lipids from different simulation models.
+    Dihedral angle distributions of the glycerol backbone region of POPS lipids from different simulation models.
   }
@@
   \caption{\label{HGandGLYstructuresPS}
-    Overlayed snapshots from glycerol backbone and headgroup region from 
-    different simulations of PS lipids.
+    Overlayed snapshots of the glycerol backbone and headgroup regions from different POPS simulations.
   }
@@
   \caption{\label{dihedralsHGpc}
-    Dihedral angle distributions of headgroup region of PS lipids from different simulation models.
+    Dihedral angle distributions of the headgroup regions from CHARMM36 and Slipids simulations
+    compared between the POPC and POPS lipids. 
+    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
   }
@@
   \caption{\label{dihedralsGLYpc}
-    Dihedral angle distributions of glycerol backbone region of PS lipids from different simulation models.
+    Dihedral angle distributions of the glycerol backbone regions from CHARMM36 and Slipids simulations
+    compared between the POPC and POPS lipids.
+    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
   }
@@
   \caption{\label{HGandGLYstructuresPSPC}
-    Overlayed snapshots from glycerol backbone and headgroup region from 
-    different simulations of PS lipids.
+    Overlayed snapshots of the headgroup and glycerol backbone regions 
+    from CHARMM36 and Slipids simulations compared between the POPC and POPS lipids.
+    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
   }
@@
 ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
+The different binding behaviour is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
 but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
@@
 but only modest changes as a function of KCl concentration, indicating singificant 
-Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
+Li$^+$ binding but only weak K$^+$ binding to the mixture when interpreted using the
 electrometer concept~\cite{akutsu81,altenbach84,seelig87}.
 
-In simulations with Berger and CHARMM36 models, the responses of both headgroup order parameter
-of POPC and the $\beta$ carbon of POPS to the added sodium or potassium are similar to the experiments
-of LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of the counterions 
+In simulations with Berger and CHARMM36 models, the responses of headgroup order parameter
+of both POPC and POPS to the added sodium or potassium are more similar to the experiments
+with LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
 in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
@@
 ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
-Such clusters appering also in MagRog simulations with 4 M concentration of KCl
+Appearance of such clusters also in the MagRog simulations with 4 M concentration of KCl
 could explain the unsystematic changes of the order parameters in this model with the added KCl.
@@
 in the main text, suggesting that the MacRog simulations with KCl give the most
-realistic surface charge at the lipid bilayer interface.
+realistic surface charge at the lipid bilayer interface among the tested simulation models.
 \begin{figure*}[]

### 2018-12-07 Manuscript/manuscriptPS.tex
New version of conclusions

@@
 \section{Conclusions}
-We have collected a set of experimental NMR order parameter data,
-which could be combined with MD simulations to interpret the
-headgroup structure and cation binding details to negatively charged
-membranes containing PS lipids. Using open collaboration method, we
-tried to find a MD simulation model which would be sufficiently accurate
-to interpret the experimental data. However, none of the tested models
-was accurate enough. In line with the previous study for PC lipids \cite{catte16},
-MD simulation models seems to generally overestimate cation binding also
-to negatively charged bilayers containing PS lipids, with some exceptions.
-The response of PS lipid headgroup order parameters to the bound cations
-does not agree with experiments, even in the cases where binding affinity is
-not overestimated. This is in contrast to the previous results with PC lipids,
-where the qualitative response of the headgroup order parameters was in agreement
-with experiments even in the cases where the headgroup structure without ions was not
-correct and the cation binding affinity was overestimated. In addition, the inaccurate
-responses of PS headgroup order parameters to the dilution with PC lipids suggests
-that the PC-PS interactions are not accurately described by the tested models.
+Lipids with PS headgroups and their interactions with ions
+play an important role in lipid mediated signaling
+processes \cite{leventis10,yeung08}. Recent studies using molecular
+dynamics simulations to interpret the various spectroscopic data
+give contradictory results for the calcium binding details to PS
+headgroups \cite{melcrova16,valentine18,hallock18}.  
+Here, we used the headgroup C-H bond order parameters and open collaboration method to evaluate the quality
+of headgroup structure and ion binding affinity to PS lipids
+in available MD simulation force fields,
+as previously done for PC lipids \cite{botan15,catte16}.
 
-Our results pave the way for improving the PS lipid parameters for MD simulations
-by offering the set of experimental data for the quality measurement, by
-pinpointing problems areas in the models and suggesting directions for the corrections.
-Improvements using the electronic continuum correction is already in progress \url{https://github.com/jmelcr/ecc_lipids},
-following the recent work for PC lipids \cite{melcr18}.
+First, we complemented the available experimental data of PS
+lipid headgroup order parameters \cite{browning80,roux90} by measuring the signs of the order parameters.
+None of the available force fields tested using the NMRlipids open
+collaboration was accurate enough to reproduce the PS
+headgroup order parameters with the experimental accuracy, but
+the best models suggested a characteristic rigid conformation for the
+carboxyl group in the serine headgroup. Comparison to the previously
+measured headgroup order parameters from POPC:POPS (5:1) bilayers with different ion
+concentrations \cite{roux90} showed that the tested MD simulation force fields
+overestimate the cation binding affinity to the negatively charged bilayers
+containing PS lipids with two exceptions. The apparently most realisitic monovalent ion binding
+affinity to PS containing lipid bilayers was observed in the MacRog
+simulations with potassium counterions and the CHARMM36 force field with recently introduced
+NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
+However, the experimentally measured trends of the PS headgroup order parameter response
+to the bound cations and to the dilution of bilayer with zwitterionic PC lipids were not
+qualitatively reproduced in any of the tested force fields, indicating that improvements in
+the MD simulation force fields are necessary to study interactions between PS lipids and other biomolecules.
+This is different to the previous results with PC lipids,
+where the experimentally measured headgroup order parameter responses to the bound charge
+were qualitative reproduced even thought the headgroup structures without ions were not
+correct and the cation binding affinities were overestimated \cite{catte16}.
+
+Our results pave the way for the development of better MD simulations force fields
+for PS lipids. Using the headgroup order parameters, we were able to evaluate the
+quality of various conformational ensembles in different force fields.
+This can guide the development of force fields that would correctly reproduce the
+conformations sampled by PS headgroups. The experimental dataset of headgroup order
+parameters from POPC:POPS (5:1) mixture with different cation concentrations can
+be used to improve cation binding details in the force fields, as recently demonstrated
+for POPC using the electronic continuum correction \cite{melcr18}.
+Similar study for POPS is progressed separately \cite{ECCpops}.
 

### 2018-12-10 Manuscript/manuscriptPS.tex
Update simulation details

@@
     \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
-    \hline
     POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
@@
     POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
-    POPS/Na$^+$  & Berger \cite{??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
     POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
     POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
+    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
+    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
+    POPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
     POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
     POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
-    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
-    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
+  \hline
+    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
     \end{tabular}
@@
   \hline
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
@@
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
@@
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
 %    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\

### 2018-12-10 Manuscript/manuscriptPSsuppl.tex
Update simulation details

@@
 \subsection{CHARMM36}
-\todo{To be written by Piggot, Madsen and Ollila}
+
+\noindent
+{\it POPC bilayer}
+\todo{By Fernando Favela at http://dx.doi.org/10.5281/zenodo.16830} \\
+% It should be noted that pure POPC are ran with Gromacs 4, but the results with
+% PS are ran with Gromacs 5. This has significant effect on acyl chains:
+% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
+% The effect for headgroups can be estimated by comparing these two data sets:
+% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
+% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}
+
+\noindent
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
+\todo{T. Piggot, please write simulation details} \\
+
+\noindent
+{\it POPC:POPS mixtures without additional ions}\\
+Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
+with potassium and sodium countertions were ran using CHARMM36 force field.
+The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
+and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.
+
+Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids
+with potassium countertions were ran using CHARMM36 force field.
+The data are available from Refs. \citenum{??} \todo{J. Madsen, please write simulation details}. \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional monovalent ions} \\
+POPC:POPS (5:1) mixtures containing total 132 (110:22) lipids
+were simulated with the additional potassium and sodium countertion concentrations
+of $\sim$450 mM and 890 mM using CHARMM36 force field at 298 K. The data, force field and
+input files generated using parameters given by CHARMM-GUI \cite{lee16,jo18} are
+available from Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional calcium} \\
+POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
+were simulated with the additional calcium concentrations
+\todo{J. Madsen, please write simulation details}. \\
 
 \subsection{CHARMM36ua}
-\todo{To be written by Piggot}
+\noindent
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
+\todo{T. Piggot, please write simulation details} \\
+
+\subsection{MacRog}
+\noindent
+{\it POPC bilayers.} \\
+
+\noindent
+{\it POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
+\todo{T. Piggot and M. Javanainen, please write simulation details} \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
+The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
+\todo{M. Javanainen, please write simulation details} \\
+
+\subsection{Lipid17}
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs \citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions}
+\todo{B. Kav and M. Miettinen, please write simulation details} \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} \\
+The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}
+\todo{B. Kav and M. Miettinen, please write simulation details} \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional calcium} \\
+The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series}
+\todo{J. Melcr, please write simulation details} \\
+
+\subsection{Berger}
+\noindent
+{\it POPC bilayers.} \\
+Previously published simulation of POPC at 310 K \cite{ollila07a}, available from Ref. \cite{bergerPOPC310K},
+was used.
+
+\noindent
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
+\todo{T. Piggot, please write simulation details} \\
+
+\noindent
+{\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions} \\
+Simulations with excess sodium were taken directly from Ref. \citenum{jurkiewicz12} and
+simulations with calcium directly from \citenum{melcrova16}.
 
@@
 \todo{To be written by Piggot and Favela}
-
-\subsection{Berger}
-\todo{To be written by Piggot and Ollila}
-Simulations with excess sodium were taken directly from Ref. \citenum{jurkiewicz12} and
-simulations with calcium directly from \citenum{melcrova16}.
-Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.
 
@@
    (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
-
-\subsection{Lipid17}
-\todo{To be written by Kav, Miettinen and Meclr.}
-
-\subsection{MacRog}
-\todo{To be written by Javanainen and Piggot}
 

### 2018-12-10 Manuscript/manuscriptPS.tex
Update simulation detilas

@@
     \hline
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & 80 & 120 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & 120 & 80 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
     POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.102\footnote{Calculation of concetration complicated due the scaled ions.

### 2018-12-10 Manuscript/manuscriptPSsuppl.tex
Update simulation detilas

@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional monovalent ions} \\
-POPC:POPS (5:1) mixtures containing total 132 (110:22) lipids
+{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
+Systems containing total 132 (110:22) lipids
 were simulated with the additional potassium and sodium countertion concentrations
@@
 \noindent
-{\it POPC bilayers.} \\
+    {\it POPC bilayers.} Simulation data available from Ref. \citenum{macrogPOPC298K}.
+    \todo{For example this data set has pretty good description of simulation details in the
+      reference. I am not sure how much we should repeat these in here.}
 
@@
 \noindent
-{\it POPC bilayers.} \\
-Previously published simulation of POPC at 310 K \cite{ollila07a}, available from Ref. \cite{bergerPOPC310K},
-was used.
+{\it POPC bilayers.} 
+Previously published simulation \cite{ollila07a} available from Ref. \citenum{bergerPOPC310K}
+was used for POPC at 310 K.
 
@@
 \noindent
-{\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions} \\
-Simulations with excess sodium were taken directly from Ref. \citenum{jurkiewicz12} and
-simulations with calcium directly from \citenum{melcrova16}.
+{\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions.} 
+Previously published simulations with the additional amount of sodium \cite{jurkiewicz12}
+available from Ref. \citenum{POPCpopsBERGERwith1000mMNa} and with the additional amounts
+of calcium \cite{melcrova16} available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa}
+were used. The additional sodium ions were removed from the structure file of the simulation
+with excess amount of sodium \cite{POPCpopsBERGERwith1000mMNa} to generate the initial structure for the reference simulation
+with sodium counterions only. The data, input and force field parameter files of the reference simulation
+are available from Ref. \citenum{bergerPOPSPOPC4:1mixtureT310K}. \todo{The simulation details of this are
+  not clearly described in Zenodo repository, but can be read from the mdp file available there. Should we
+write the detailes here or not?}
 
-\subsection{Slipids}
-\todo{To be written by Piggot and Favela}
-
-\subsection{GROMOS-CKP}
-\todo{To be written by Piggot}
+\subsection{GROMOS-CKP and GROMOS-CKPM}
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
@@
    (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
+\todo{T. Piggot, please write the description of force field.}
+
+\noindent
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs. \citenum{ckp1POPS303K,ckp2POPS303K,ckp1DOPS303K,ckp2DOPS303K}
+\todo{T. Piggot, please write simulation details} \\
+
+\noindent
+    {\it POPC:POPS (5:1) mixtures without additional ions}\\
+    The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
+\todo{T. Piggot, please write simulation details} \\
+
+\subsection{Slipids}
+
+\noindent
+{\it DOPS and POPS bilayers.} \\
+The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.
+\todo{T. Piggot and F. Favela, please write simulation details} \\
 

### 2018-12-11 Manuscript/manuscriptPS.tex
Update manuscript

@@
 does not rotate freely. In line with the previous results for PC lipids,
-all the tested force fields overestimate the cation binding affinity to lipid bilayers containing
+none of the tested force fields correctly captures the cation binding affinity to lipid bilayers containing
 PS lipids. In contrast to PC lipids, the response of PS headgroups
@@
 induce protein aggregation \cite{zhao04,gorbenko06}.
-Some domains spesifically interact PS lipids,
+Some protein domains spesifically interact PS lipids,
 while others are attracted by general electrostatics and the
@@
 
-Several classical molecular dynamics (MD) simulation studies are done
-to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with
-ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18}.
-However, the headgroup structures and ion binding affinity 
-predicted by the MD simulation force fields for bilayers containing PS lipids are not
-evaluated against experiments. Recent studies comparing the
-C-H bond order parameters between simulations and experiments show that the
-structure of PC lipid headgroup and glycerol backbone are not well
-captured by most simulation models \cite{botan15} and that the cation binding to PC
-lipid bilayers is overestimated \cite{catte16}. Based on this data,
-the cation binding affinity to POPC bilayer was then improved by implicitly
-including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
-
 \begin{table*}[htb]
@@
     POPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
     POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
     POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
   \hline
@@
     DOPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
     DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
     DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
     \end{tabular}
@@
 \end{table*}
+
+Several classical molecular dynamics (MD) simulation studies are done
+to understand PS headgroups, their influence on lipid bilayer properties and
+interactions with
+ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
+However, the results strongly depend on the used force field parameters.
+For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
+CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
+suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
+the same force field without the NBfix parameters together with the NMR chemical shifts and
+REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
+On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
+combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
+calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}. 
+We have recently demonstrated that such controversies can be resolved by
+comparing C-H bond order parameters of lipid headgroups between simulations
+and experiments \cite{botan15,catte16}, which can be directly
+measured from NMR experiments with high accuracy and compared to simulations
+in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
+Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
+captured by most simulation models \cite{botan15} and that the cation binding to PC
+lipid bilayers is overestimated \cite{catte16}. Based on this data, the cation binding
+affinity to POPC bilayer was then improved by implicitly including the electronic polarizability
+using the electronic continuum correction \cite{melcr18}.
 
@@
 
-\subsection{Solid state NMR experiments}
+\subsection{C-H bond order parameters from the natural abundance $^{13}$C NMR}
 
@@
 
-{\emph{R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
+{\emph{C-H bond order parameters from the R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
 for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
@@
 
-\emph{S-DROSS experiment}. The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
+\emph{Order parameter signs from the S-DROSS experiment}.
+The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
 experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
@@
 
-\emph{NMR numerical simulations}
+\emph{Numerical simulations of S-DROSS curves.}
 The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
@@
 the Open Collaboration method \cite{botan15}.
-The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
-the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
+The NMRlipids project blog (\url{www.nmrlipids.blogspot.fi}) and
+the GitHub repository (\url{www.github.com/NMRLipids/NMRlipidsIVotherHGs})
 were used as the communication platforms.
@@
 and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
-Simulation details are given in the SI.
+Further simulation details are given in the SI.
 The simulation data is also indexed in the
-searchable database (\url{nmrlipids.fi}),
+searchable database (\url{www.nmrlipids.fi}),
 and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).
@@
 \end{figure}
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
+  \caption{\label{HGorderParameters}
+    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
+    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
+    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
+    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
+  }
+\end{figure}
 
@@
 
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
+  \caption{\label{HGorderParametersPS}
+    Order parameters of PS headgroup and glycerol backbone from experiments
+    and simulations with different force fields.
+    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
+    while all the other data is without additional ions.
+    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
+    Light blue areas span 0.04 units around the average of the extremal experimental values,
+    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
+    The vertical bars shown for some simulation values
+    are not error bars, but demonstrate that for these systems
+    we had at least two data sets; the ends of the bars mark the extreme values
+    from the sets, and the dot marks their measurement-time-weighted average.
+  }
+\end{figure}
+
 The headgroup and glycerol backbone order parameters of 
@@
 available.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
-  \caption{\label{HGorderParameters}
-    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
-    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
-    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
-    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
-    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
-  }
-\end{figure}
 
 \subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
-  \caption{\label{HGorderParametersPS}
-    Order parameters of PS headgroup and glycerol backbone from experiments
-    and simulations with different force fields.
-    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
-    while all the other data is without additional ions.
-    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
-    Light blue areas spanning 0.04 units around the average of the extremal experimental values,
-    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
-    The vertical bars shown for some simulation values
-    are not error bars, but demonstrate that for these systems
-    we had at least two data sets; the ends of the bars mark the extreme values
-    from the sets, and the dot marks their measurement-time-weighted average.
-  }
-\end{figure}
 
@@
 
-The headgroup and glycerol backbone of PS lipids show wide variety
-in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS})
-between different simulation models,
+The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
+between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
 as previously observed also for PC lipids \cite{botan15}.
-The models perform generally less well for PS lipids than for PC lipids in the previous study
+The models perform generally less well for PS lipids than for PC lipids 
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
@@
 from simulations is not straightforward.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
+  \caption{\label{NAdensPOPS}
+    Counterion densities of POPS lipid bilayer along the membrane normal from
+    simulations with different force fields.
+  }
+\end{figure}
 
@@
 CHARMM36 simulations in best agreement with experiments
-(Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
+(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
 The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
@@
 
-The observed conformations of PS headgroup may be useful when interpreting experiments and
+The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
 can guide the further force field development. However, more accurate simulations are
-required to confirm the results because the simulated PS headgroup order parameters
+required to confirm the suggested conformations because the simulated PS headgroup order parameters
 are not within the experimental error bars in any of the tested models.
@@
 
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
-  \caption{\label{NAdensPOPS}
-    Counterion densities of POPS lipid bilayer along the membrane normal from
-    simulations with different force fields.
-  }
-\end{figure}
-
-\begin{figure}[!htb]
+\begin{figure}[!tb]
   \centering
@@
     Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
-    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\cite{ferreira16}.
+    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
   }
-  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
-  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
+  \todo{After we know which force field is used for POPC in Gromos-CKP simulations, we might be able to
+    add Gromos-CKP data into this plot.}
 \end{figure}
 
-\begin{figure}[!htb]
+\begin{figure}[!tb]
   \centering
@@
     Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
-    Experimental values and order parameter signs of pure POPS and values in mixture are  
-    measured at 298~K in this work and in Ref. \citenum{roux90}, respectively.
-    Since the experimental data of POPS in pure and diluted mixture come from
-    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
-    the experimental change of the order parameter is less accurate than in typical measurements
-    where same technique is used in all conditions, see discussion about qualitative and quantitative 
-    accuracy in Ref. \citenum{ollila16}.
+    Experimental values with the signs are measured for pure POPS system in this work.
+    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
+%    Since the experimental data of POPS in pure and diluted mixture come from
+%    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
+%    the experimental change of the order parameter is less accurate than in typical measurements
+%    where same technique is used in all conditions, see discussion about qualitative and quantitative 
+%    accuracy in Ref. \citenum{ollila16}.
     The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
@@
   \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
-  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
+    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
+  these are not necessary, but could be added here if delivered.} \\
 \end{figure*}
@@
 The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
-and other experimental results suggest less rigid structure of PS headgroups when diluted with
+together with other experimental results suggest less rigid structure of PS headgroups when diluted with
 POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
@@
 \subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
-\begin{figure}[ht]
+\begin{figure}[tb]
   \centering
@@
   \caption{\label{CAdensPCPSmixture}
-    Ca2+ density profiles from simulations.
+    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
+    The smallest simulated CaCl$_2$ concentrations are shown.
+    For the density profiles from all the simulated concentrations see figure \ref{CAdensPCPSmixtureALL}
+    in the supplementary information.
   }
@@
 as previously done for PC lipids \cite{botan15,catte16}.
+The main advantage of this approach is the direct connection
+between accurately measured experimental order parameters and simulations
+which reduces the ambiguity in the interpretation of experiments.
 
@@
 However, the experimentally measured trends of the PS headgroup order parameter response
-to the bound cations and to the dilution of bilayer with zwitterionic PC lipids were not
+to the bound calcium and to the dilution of bilayer with zwitterionic PC lipids were not
 qualitatively reproduced in any of the tested force fields, indicating that improvements in
@@
 % Put your acknowledgments here.
+  OHSO acknowledges financial support from Academy of Finland (315596),
+  Integrated Structural Biology Research Infrastructure of
+  Helsinki Institute of Life Science (Instruct-HiLIFE), and
+  CSC-IT center for science for computational resources.
 \end{acknowledgments}

### 2018-12-11 Manuscript/manuscriptPSsuppl.tex
Update manuscript

@@
 \noindent
-{\it POPC bilayer}
-\todo{By Fernando Favela at http://dx.doi.org/10.5281/zenodo.16830} \\
+    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
+    available from Ref. \citenum{charmmPOPC512files} are used.
 % It should be noted that pure POPC are ran with Gromacs 4, but the results with
@@
 % https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
-% The effect for headgroups can be estimated by comparing these two data sets:
+% but the effect to headgroups is small
 % https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
@@
 lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
-Changes of the headgroup order parameters of other lipids with bound charge,
-like negatively charged PS and PG lipids,  are also systematic,
-but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
+Changes of the headgroup order parameters of negatively charged PS and PG lipids
+are also systematic, but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
 Therefore, the ion binding affinity to negatively charged bilayers
@@
 
-When comparing ion binding measured from the PC headgroup order parameters between
-neutral and anionic membranes, it is important to note that the order parameters
+When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
+it is important to note that the order parameters
 increase due to the addition of negative charged lipids according to the electrometer
 concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
-Therefore, the headgroup order parameters of POPC in mixtures with anionic
-lipids are larger than in pure POPC lipid bilayer, which is also seen in the ion binding
-experiments from different mixtures in the absence of added calcium (Fig. \ref{OrderParametersWithCaCl}).
+Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
+than in pure lipid bilayers. This is evident also in the headgroup order parameters
+of POPC in bilayers with different amounts anionic lipids without added calcium (Fig. \ref{OrderParametersWithCaCl}).
 Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
@@
   \caption{\label{OrderParametersWithCaCl}
-    Headgroup order parameters of POPC as a function of CaCl concentration from experiments 
-    different mole fractions of negatively charged lipids (left column).
+    Headgroup order parameters of POPC as a function of CaCl$_2$ concentration from experiments 
+    with different mole fractions of negatively charged lipids (left column).
     Right column shows the same data zoomed to the concentrations below 1M.
-    Data for Pure DPPC from~\cite{akutsu81},
-    for pure POPC from~\cite{altenbach84}, 
-    for POPC:POPS (5:1) mixture from~\cite{roux90},
-    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from~\cite{macdonald87}
-    and for POPC:POPG (1:1) mixture data without NaCl from \cite{borle85}.
+    Data for Pure DPPC from Ref.~\citenum{akutsu81},
+    for pure POPC from Ref.~\citenum{altenbach84}, 
+    for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
+    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from Ref.~\citenum{macdonald87}
+    and for POPC:POPG (1:1) mixture data without NaCl from Ref.\citenum{borle85}.
   }
@@
 (Figs. \ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
-The glycerol backbone structures are not discussed in this works because our focus is on PS headgroup.
+The glycerol backbone structures are not discussed in this work because our focus is on PS headgroup.
 However, the data presented here can be useful for future investigations.

### 2018-12-19 Manuscript/manuscriptPSsuppl.tex

@@
 \externaldocument{manuscriptPS}
+\usepackage{soul} % for highlights with \hl{}
 
@@
 
+The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
+from Ref.~\cite{10.1016/j.dib.2015.09.013}. The bilayer was hydrated by a total of 5120 water 
+molecules, corresponding to 40 water molecules per lipid.
+
+The MacRog lipid parameters \cite{10.1016/j.chemphyslip.2015.07.002}, obtained from 
+\cite{10.1016/j.dib.2015.09.013}, were used, along with the TIP3P water model 
+\cite{10.1063/1.445869}. The simulation parameters were identical to those used for 
+POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns, 
+out of which \hl{??~ns: (this is not tabulated in the MS or the SI)} were used for analyses.
+
 \noindent
@@
 \todo{T. Piggot and M. Javanainen, please write simulation details} \\
+\hl{Piggot: please write about system with Na$^+$ counterions}.
+
+This membrane was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
+were described by the \AA{}qvist parameters \cite{10.1021/j100384a009}. Other topologies
+were unchanged. The simulation parameters were the same as used for the POPC/POPS mixtures with 
+the MacRog force field, see below. The simulation was run for 200~ns, out of which 150~ns was used 
+in the analyses.
 
@@
 The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
-\todo{M. Javanainen, please write simulation details} \\
+
+The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
+two leaflets were set up using CHARMM-GUI \cite{10.1002/jcc.20945}.
+The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
+Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
+
+These bilayers were simulated in the presence of counterions only, as well as together with different
+concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
+were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
+of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
+were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.
+
+The lipids were described using the MacRog model \cite{10.1021/jp5016627,10.1016/j.chemphyslip.2015.07.002,10.1016/j.dib.2016.03.067}. The POPC topologies were obtained from 
+Ref.~\citenum{10.1016/j.dib.2015.09.013}, whereas the POPS topology was adapted from 
+Ref.~\citenum{10.1016/j.dib.2016.03.067}. Notably, the chains were reversed in order to modify 
+OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{10.1016/j.dib.2016.03.067}, the 
+POPS head group conformations generated by CHARMM-GUI are in the naturally occurring L stereoisomer.
+TIP3P model \cite{10.1063/1.445869} was used for water, and the \AA{}qvist parameters 
+\cite{10.1021/j100384a009}, standard for the OPLS force field, were employed for the ions.
+
+The simulation parameters were taken from Ref.~\citenum{10.1016/j.dib.2016.03.067}. 
+Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step 
+\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
+The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
+\cite{10.1063/1.470117} was used for electrostatics. Dispersion correction \cite{10.1021/jp0735987} 
+was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
+Nos\'{e}--Hoover thermostat \cite{10.1063/1.447334,10.1103/PhysRevA.31.1695}. The lipids and 
+the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
+maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
+\cite{10.1063/1.328693}. All non-water bonds were constrained with the LINCS algorithm 
+\cite{10.1002/(SICI)1096-987X(199709)18:12<1463::AID-JCC4>3.0.CO;2-H,10.1021/ct700200b},
+whereas the bonds in water molecules were constrained using SETTLE \cite{10.1002/jcc.540130805}.
+The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
+(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
+250, 200, or 300~ns of the simulations were included in the analyses, respectively.
 

### 2018-12-20 Manuscript/manuscriptPSsuppl.tex
Update references added in recent commit https://github.com/NMRLipids/NMRlipidsIVotherHGs/commit/a3978044eb65f6c1160abefce17e3a8ed7e1d7b1

@@
 The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
-from Ref.~\cite{10.1016/j.dib.2015.09.013}. The bilayer was hydrated by a total of 5120 water 
+from Ref.~\cite{kulig15c}. The bilayer was hydrated by a total of 5120 water 
 molecules, corresponding to 40 water molecules per lipid.
 
-The MacRog lipid parameters \cite{10.1016/j.chemphyslip.2015.07.002}, obtained from 
-\cite{10.1016/j.dib.2015.09.013}, were used, along with the TIP3P water model 
-\cite{10.1063/1.445869}. The simulation parameters were identical to those used for 
+The MacRog lipid parameters \cite{kulig15b}, obtained from 
+\cite{kulig15c}, were used, along with the TIP3P water model 
+\cite{jorgensen83}. The simulation parameters were identical to those used for 
 POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns, 
@@
 This membrane was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
-were described by the \AA{}qvist parameters \cite{10.1021/j100384a009}. Other topologies
+were described by the \AA{}qvist parameters \cite{aqvist90}. Other topologies
 were unchanged. The simulation parameters were the same as used for the POPC/POPS mixtures with 
@@
 The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
-two leaflets were set up using CHARMM-GUI \cite{10.1002/jcc.20945}.
+two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
 The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
@@
 
-The lipids were described using the MacRog model \cite{10.1021/jp5016627,10.1016/j.chemphyslip.2015.07.002,10.1016/j.dib.2016.03.067}. The POPC topologies were obtained from 
-Ref.~\citenum{10.1016/j.dib.2015.09.013}, whereas the POPS topology was adapted from 
-Ref.~\citenum{10.1016/j.dib.2016.03.067}. Notably, the chains were reversed in order to modify 
-OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{10.1016/j.dib.2016.03.067}, the 
+The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
+Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from 
+Ref.~\citenum{rog16}. Notably, the chains were reversed in order to modify 
+OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{rog16}, the 
 POPS head group conformations generated by CHARMM-GUI are in the naturally occurring L stereoisomer.
-TIP3P model \cite{10.1063/1.445869} was used for water, and the \AA{}qvist parameters 
-\cite{10.1021/j100384a009}, standard for the OPLS force field, were employed for the ions.
+TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
+\cite{aqvist90}, standard for the OPLS force field, were employed for the ions.
 
-The simulation parameters were taken from Ref.~\citenum{10.1016/j.dib.2016.03.067}. 
+The simulation parameters were taken from Ref.~\citenum{rog16}. 
 Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step 
@@
 The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
-\cite{10.1063/1.470117} was used for electrostatics. Dispersion correction \cite{10.1021/jp0735987} 
+\cite{essman95} was used for electrostatics. Dispersion correction \cite{shirts07} 
 was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
-Nos\'{e}--Hoover thermostat \cite{10.1063/1.447334,10.1103/PhysRevA.31.1695}. The lipids and 
+Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
 the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
 maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
-\cite{10.1063/1.328693}. All non-water bonds were constrained with the LINCS algorithm 
-\cite{10.1002/(SICI)1096-987X(199709)18:12<1463::AID-JCC4>3.0.CO;2-H,10.1021/ct700200b},
-whereas the bonds in water molecules were constrained using SETTLE \cite{10.1002/jcc.540130805}.
+\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
+\cite{hess97,hess07},
+whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
 The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 

### 2019-01-07 Manuscript/manuscriptPSsuppl.tex

@@
 {\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs \citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions}
-\todo{B. Kav and M. Miettinen, please write simulation details} \\
+
+The structure of the bilayers with 128 DOPS or 128 POPS lipids were
+obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
+4480 water molecules resulting in 31.1 water molecules per lipid. For
+each system, 128 NaCl molecules are added to neutralize the system.\\
+
+Each bilayer structure was first minimized for 2500 steps with steepest
+descent and an additional 2500 step with the conjugent gradient
+algorithms while restrainig the solvent and ion molecules with a force
+constant 500~kcal~mol$^{-1}$~\AA{}$^{-1}$. An additional minimization
+step was used with the same parameters after removing the constraints on
+the solvent and ion molecules. During the minimization procedure,
+non-periodic boundary conditions were applied, no pressure scaling was
+used and none of the bonds were restrained. A three step heating
+procedure was applied: in the first heating step the system temperature
+is increased from 0.0~K to 100.0~K using Langevin thermostat with
+5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
+with harmonic springs of force constant
+20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
+boundary conditions. In the second heating step, the system temperature
+was increased from 100.0~K to 200.0~K using Langevin thermostat with
+5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
+the lipid positions with harmonic springs of force constant
+10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
+heating step, the system temperature is increased from 200.0~K to
+303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
+collision frequency while restraining the lipid positions with harmonic
+springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
+constant pressure using Berendsen barostat with 3.0~ps relaxation time.
+From the second heating step onward, periodic boundary conditions in all
+dimensions were applied. The semi-isotropic pressure coupling along
+\textit{xy}-plane was employed and the surface tension was set to
+0.0~dyn explicitely. After the third heating step, each structure is
+equilibriated for 100~ns at 1~bar and 303~K constant pressure using
+Berendsen barostat and temperature using Langevin thermostat,
+respectively. For the equilibriation step, semi-isotropic pressure
+coupling along \textit{xy}-plane was employed and the surface tension
+was set to 0.0~dyn explicitely. For all heating and equilibration steps,
+the lengths of the hydrogen containing bonds were restrained using SHAKE
+algorithm and a time step of 2~fs was used. In all steps with periodic
+boundary conditions, the non-bonded interactions are calculated with
+particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.\\
+
+For the production runs, the temperature and pressure were set to 303~K
+and 1~bar, using Langevin thermostat with 1.0~ps$^{-1}$ collision
+frequency and Berendsen barostat with 1.0~ps relaxation time,
+respectively. Periodic boundary conditions were applied in all
+dimensions and the non-bonded interactions are calculated with particle
+mesh Ewald~\cite{essman95} using 1.0~nm cutoff. The lengths of the
+hydrogen containing bonds were constrained using SHAKE algorithm and a
+time step of 2~fs was used. Trajectories are saved in every 10~ps
+intervals. Two independent trajectories of 500~ns long was generated and
+the last 200~ns portions were used for the analysis.\\
+
+In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
+to constrain the bonds in TIP3P water molecules. Lipid parameters were obtained from 
+Amber Lipid17 force field. For the ions, one set of simulations were run with Joung-Cheatham
+ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB~\cite{aqvist90} ion parameters.
+The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain the details of the simulations and the raw trajectories, respectively.
 
@@
 {\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} \\
-The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}
-\todo{B. Kav and M. Miettinen, please write simulation details} \\
+
+The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
+Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}.
+For the simulations of the lipids and ions, the Amber Lipid 17 and ff99SB force fields were used, respectively.
+The simulation steps are the same as the the simulations of DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns was used for the analysis.
+The number of ions in each specific system is given in Table~\ref{mixedIONsystems}.
+The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.\\
 

### 2019-01-15 Manuscript/manuscriptPS.tex
Add new figure by Buslaev contributed in http://nmrlipids.blogspot.com/2018/07/nmrlipids-iv-first-draft-of-manuscript.html?showComment=1545837139263#c8389351474310868144
Update information by J. Madsen

@@
     POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMNa}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
-& ?
-& 298  & 200 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & 11190  & 53  & 298  & 200 & ?
 & \cite{??} \todoi{Trajectories and further details to be added by J.
@@
   }
-  \todo{We need atom labeling for (B). Similar to the one in (A) would be good} \\
+  \todo{We need larger labels in (B).} \\
 \end{figure}
@@
   \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
-  \todo{y-axis scale of Lipid17 should be checked.} \\
   \todo{Should we include also counterions into the plot?} \\

### 2019-01-16 Manuscript/manuscriptPS.tex
Update data by J. Madsen

@@
     POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMNa}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & 11190  & 53  & 298  & 200 & ?
-& \cite{??} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.26 & 250:50 & 11190  & 53  & 298  & 200 & ?
+\todoi{Equilibration time to be checked}  & \cite{POPC5POPS1withCaClCHARMM}   \\
     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
@@
   }
-  \todo{We need larger labels in (B).} \\
 \end{figure}
@@
   }
-  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
   \todo{Should we include also counterions into the plot?} \\

### 2019-01-17 Manuscript/manuscriptPS.tex
Separate systems with additional counterions to other table and move it to SI. The previous table II became too large

@@
 
-\begin{table*}[!p]
+\begin{table*}[tb]
 %\begin{sidewaystable*}[!p]
 \centering
-\caption{The list of POPC:POPS mixtures simulated with different amounts of added ions. 
+\caption{The list of POPC:POPS mixtures simulated with different molar fractions and different amounts of added calcium. 
   The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
@@
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c c c c c c c c c}
-  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) & [CaCl$_2$]\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+\begin{tabular}{l c c c c c c c c c }
+  lipid/counter-ions & force field for lipids / ions & [CaCl$_2$]\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
-%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith450mMK}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMK}  \\
-    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith440mMNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMNa}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.26 & 250:50 & 11190  & 53  & 298  & 200 & ?
-\todoi{Equilibration time to be checked}  & \cite{POPC5POPS1withCaClCHARMM}   \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
+%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26 & 250:50 & 11190  & 53  & 298  & 200 & ? \todoi{Equilibration time to be checked}  & \cite{POPC5POPS1withCaClCHARMM}   \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
 & ?
@@
 Madsen}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     \hline
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.30 & 120:24 & 5760 & 31   & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1.00   & 120:24 & 5760 & 104  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 0.30 & 120:24 & 5760 & 31   & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 1.00   & 120:24 & 5760 & 104  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
 %    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
     \hline
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0  & 0     & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0   & 0   & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
-& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0      & 60:12 & 3600  & 0   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.08   & 60:12 & 3561  & 5   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.13   & 60:12 & 3561  & 8   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.20   & 60:12 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.41   & 60:12 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0.62   & 60:12 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18}  & 0      & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} & 0      & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
+    POPC:POPS (5:1)         & Lipid14/17 \cite{dickson14,gould18} &0.50    & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
+    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &1.00    & 120:24 & 5760 & 104   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
+    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &2.00    & 120:24 & 5760 & 208   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
+    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &3.00    & 120:24 & 5760 & 311   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
+    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &4.00    & 120:24 & 5760 & 415   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18}  & 0      & 60:12 & 3600  & 0   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.08   & 60:12 & 3561  & 5   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.13   & 60:12 & 3561  & 8   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.20   & 60:12 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.41   & 60:12 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.62   & 60:12 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
     \hline
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & 120 & 80 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.102\footnote{Calculation of concetration complicated due the scaled ions.
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0   & 102:26 & 4290 & 0   & 310  & 120 & 80 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}         & 0.102\footnote{Calculation of concetration complicated due the scaled ions.
 Concentration taken as reported in the delivered data.}   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715\footnote{Calculation of concetration complicated due the scaled ions.
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}         & 0.715\footnote{Calculation of concetration complicated due the scaled ions.
 Concentration taken as reported in the delivered data.}  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
     \hline
-    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKP \cite{??}             &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPwithNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKPM \cite{??}            &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPMwithNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKP \cite{??}             & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPwithNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKPM \cite{??}            & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPMwithNa}  \\
 \end{tabular}

### 2019-01-17 Manuscript/manuscriptPSsuppl.tex
Separate systems with additional counterions to other table and move it to SI. The previous table II became too large

@@
 
+\usepackage{rotating}
 \usepackage{upgreek}				
@@
 \section{Simulated systems}
+
+%\begin{table*}[!p]
+\begin{sidewaystable*}[!p]
+\centering
+\caption{The list of POPC:POPS mixtures simulated with different amounts of added monovalent ions. 
+  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  This corresponds the concentration in buffer before solvating lipids, which are
+  reported in the experiments by Roux et al.~\cite{roux90}.
+  The simulation details are given in the supplementary information.
+   % The lipid force fields named as in our previous work~\cite{botan15}.
+}\label{mixedIONsystemsMONOVALENT}
+%begin{minipage}[t]{\textwidth}
+\begin{tabular}{l c c c c c c c c c}
+  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  \hline
+%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todo{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0     & 110:22 & 4620  & 0  & 298  & 500 & 100 & \citenum{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45  & 110:22 & 4926  & 40 & 298  & 200 & 150 & \citenum{charmm36pops+83popcT298Kwith450mMK}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89  & 110:22 & 4946  & 79 & 298  & 200 & 150 & \citenum{charmm36pops+83popcT298Kwith890mMK}  \\
+    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0     & 110:22 & 4620  & 0  & 298  & 500 & 100 & \citenum{charmm36pops+83popcT298KpiggotSODIUM}  \\
+    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \citenum{charmm36pops+83popcT298Kwith440mMNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \citenum{charmm36pops+83popcT298Kwith890mMNa}  \\
+    \hline
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 120:24 & 5760 & 0    & 298  & 400 & 250 & \citenum{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50 & 120:24 & 5760 & 52   & 298  & 300 & 200 & \citenum{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00 & 120:24 & 5760 & 104  & 298  & 300 & 200 & \citenum{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00 & 120:24 & 5760 & 208  & 298  & 300 & 200 & \citenum{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00 & 120:24 & 5760 & 311  & 298  & 300 & 200 & \citenum{POPCpopsMACROGwithK}  \\
+    \hline
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0     & 120:24 & 5760 & 0   & 298  & 500 & 200 & \citenum{POPCpopsLIPID17withKCI}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.50  & 120:24 & 5760 & 54   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1.00  & 120:24 & 5760 & 104  & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2.00  & 120:24 & 5760 & 208  & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3.00  & 120:24 & 5760 & 311  & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4.00  & 120:24 & 5760 & 415  & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0    & 120:24 & 5760 & 0   & 298  & 500 & 200 & \citenum{POPCpopsLIPID17withNaCI}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.50 & 120:24 & 5760 & 54   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1.00 & 120:24 & 5760 & 104   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2.00 & 120:24 & 5760 & 208   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3.00 & 120:24 & 5760 & 311   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4.00 & 120:24 & 5760 & 415   & 298  & 300 & 200 & \citenum{POPCpopsLIPID17withNa}  \\
+    \hline
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 102:26 & 4290 & 0   & 310  & 120 & 80 & \citenum{bergerPOPSPOPC4:1mixtureT310K}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &1.03 & 102:26 & 4290 & 80  & 310  & 200 & 50 & \citenum{POPCpopsBERGERwith1000mMNa}  \\
+\end{tabular}
+%\end{minipage}
+\end{sidewaystable*}
+
+% \end{table*}
 
@@
 The simulation steps are the same as the the simulations of DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns was used for the analysis.
-The number of ions in each specific system is given in Table~\ref{mixedIONsystems}.
+The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
 The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.\\
@@
 {\it POPC:POPS (5:1) mixtures with additional calcium} \\
-The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series}
-\todo{J. Melcr, please write simulation details} \\
+The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
+\todo{B. Kav and J. Melcr, please write simulation details} \\
 
@@
   }
+  \todo{Colours with more contrast would be better. The y-axis scale could be zoomed.}
 \end{figure*}

### 2019-01-22 Manuscript/manuscriptPS.tex
Update authorlist
Add data by J. Madsen

@@
 \affiliation{Moscow Institute of Physics and Technology}
+\author{Fernando Favela}
+\affiliation{Mexico}
 \author{Tiago M. Ferreira}
@@
     POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26 & 250:50 & 11190  & 53  & 298  & 200 & ? \todoi{Equilibration time to be checked}  & \cite{POPC5POPS1withCaClCHARMM}   \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
-& ?
-& 298  & 200 & ?
-& \cite{??} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26 & 250:50 & 11190  & 53  & 298  & 200 & 180 & \cite{POPC5POPS1withCaClCHARMM}   \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50 & 11174  & 214  & 298  & 200 & ?  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
     POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\

### 2019-02-01 Manuscript/manuscriptPS.tex
Stylistic changes to Results on pure PS simulations with no salt

@@
 \author{Batuhan Kav}
-\affiliation{Potsdam, Germany}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Jesper J. Madsen}
@@
 \author{Markus Miettinen}
-\affiliation{Potsdam, Germany}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Josef Melcr}
@@
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & 11207 & 0  & 298  & 200 & 180   & \cite{POPC5POPS1noCaClCHARMM}  \\
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
@@
     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26 & 250:50  & 11190  & 53  & 298  & 200 & 180 & \cite{POPC5POPS1withCaClCHARMM}   \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50 & 11174  & 214  & 298  & 200 & ?  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50  & 11174  & 214  & 298  & 200 & 180  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}        & 0    & 150:150 & 10785    & 0  & 298  & 200 & 180   & \cite{POPC1POPS1noCaClCHARMM}  \\
     \hline
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp-eps-converted-to.pdf}
   \caption{\label{HGorderParameters}
@@
 
-The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
-between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
+The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
 as previously observed also for PC lipids \cite{botan15}.
-The models perform generally less well for PS lipids than for PC lipids 
+The models perform generally less well for PS than for PC
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
@@
 
-However, the best performing models,
-Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
-and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
-when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
+However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
+the best performing models,
+Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
+Slipids reproduces also the feature that the $\beta$-carbon order parameter
+in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
 Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
@@
 visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
-suggesting that the rotation of carboxyl group is limited in the serine headgroup.
-In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
+suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
+In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
 and restricted angle distribution for PS than for PC headgroup in
@@
 
-The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
-can guide the further force field development. However, more accurate simulations are
+The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
+can guide further force field development. However, more accurate MD force fields are
 required to confirm the suggested conformations because the simulated PS headgroup order parameters
-are not within the experimental error bars in any of the tested models.
+are not within the experimental error in any of the tested models.
 

### 2019-02-01 Manuscript/manuscriptPSsuppl.tex
Update authorlist
Stylistic changes to Results on pure PS simulations with no salt

@@
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todo{Trajectories and further details to be added by J. Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & 11207 & 0  & 298  & 200 & 180   & \citenum{POPC5POPS1noCaClCHARMM}  \\
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0     & 110:22 & 4620  & 0  & 298  & 500 & 100 & \citenum{charmm36pops+83popcT298Kpiggot}  \\
@@
 with potassium countertions were ran using CHARMM36 force field.
-The data are available from Refs. \citenum{??} \todo{J. Madsen, please write simulation details}. \\
+The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and  \citenum{POPC1POPS1noCaClCHARMM}
+\todo{J. Madsen, please write simulation details}. \\
 
@@
 POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
-were simulated with the additional calcium concentrations
+were simulated with the additional calcium concentrations.
+The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
 \todo{J. Madsen, please write simulation details}. \\

### 2019-02-04 Manuscript/manuscriptPS.tex
Stylistic changes and clarifications to Results on Ion Binding in PS vs PC

@@
 
-Membranes containing PS lipids are always accomppanied with counterions which
+Membranes containing PS lipids are always accompanied with counterions that
 modulate electrostatic interactions between lipids and other biomolecules. 
 Counterions are also suggested to screen the repulsion between charged lipid headgroups 
-in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
+in MD simulations and thus to reduce the area per lipid of PS bilayers to be smaller than in PC
 bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
-The counterion density profiles along membrane normal 
-show significant differences between simulation models in both binding affinity
+Counterion density profiles along membrane normal 
+indicate significant differences between force fields in both binding affinity
 and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
-The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
-is reproduced only in Gromos-CKP simulations and in the MacRog simulation
-with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
-The counterion binding and concomintant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values
-because the MacRog simulation with strongest sodium binding
-(the lowest concentrations in bulk water) gives the same area per molecule
-as CHARMM36ua simulation with significantly weaker counterion binding
+The experimental area per lipid (62.7~\AA$^2$) \cite{pan14} 
+is reproduced only in Gromos-CKP and in the MacRog simulation
+with potassium counterions, while other models give significantly lower areas (Fig. \ref{NAdensPOPS}).
+The counterion binding and the concomitant electrostatic screening of the headgroup
+repulsion does not fully explain the low area per molecule values,
+because the MacRog simulation, which has the strongest sodium binding
+(the lowest concentrations in bulk water), gives the same area per molecule
+as the CHARMM36ua simulation, which has significantly weaker counterion binding
 affinity. On the other hand, changing counterions from sodium to potassium, having weaker binding
-affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
+affinity, increases the area per molecule from 53~\AA$^2$ to 63~\AA$^2$ in MacRog simulations.
 In conclusion, the results are in line with the previous study suggesting that the
-low area per molecule in PS lipid bilayers originate from the combination
+low areas per molecule in PS lipid bilayers originate from the combination
 of both counterion binding and hydrogen bonding network between lipid
@@
 evaluated against experiments using the changes of headgroup order parameters
-as a function of ion concentration \cite{catte16}. This is less straighforward
-for charged lipid bilayers because counterions are always present and the
-ion free reference state does not exist.
+as a function of salt concentration \cite{catte16}.
+Studying binding of cations to negatively charged lipid bilayers is less straighforward,
+because the cationic counterions are always present and the ion-free reference state does thus not exist.
 In addition, the analysis is complicated by the artificial aggregation of counterions
 in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
-Therefore, we evaluate the amount of bound charge also using the changes of
-headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
-According to the electrometer concept, the headgroup order parameters of POPC
-increase when negativaly charged POPS lipids are incorporated in lipid bilayer
+Therefore, we evaluate here the amount of bound charge not by adding salt
+(although also this is discussed in the section \ref{mixtureTOadditionalCIs} in the supplementary information),
+but by studying the changes of the headgroup order parameters with increasing amount of
+negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
+According to the molecular electrometer concept, the headgroup order parameters of POPC
+increase when negatively charged POPS lipids are incorporated in lipid bilayer
 (section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
-This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS})
-having the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
-The CHARMM36 and Berger simulations predict no change or decrease
+This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
+which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
+The CHARMM36 and Berger simulations predict no change, or a decrease,
 in the POPC headgroup order parameters as a function of increased amount of POPS (Fig. \ref{HGorderparametersPCvsPS}).
-This can be explained by the stronger counterion binding affinity which cancels
+This can be explained by the stronger counterion binding affinity, which cancels
 the effect of negatively charged headgroups and prevents the experimentally observed

### 2019-02-05 Manuscript/manuscriptPS.tex

--- 2019-02-01 Manuscript/manuscriptPS.tex
@@
   \caption{\label{HGorderParametersPS}
-    Order parameters of PS headgroup and glycerol backbone from experiments
-    and simulations with different force fields.
-    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
-    while all the other data is without additional ions.
-    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
+    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
+    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
+    and MD simulations with different force fields (dots).
+    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
+    while all the other data are without additional salt.
+    All the data for DOPS at 303~K and all the data for POPS at 298~K.
     Light blue areas span 0.04 units around the average of the extremal experimental values,
-    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
-    The vertical bars shown for some simulation values
+    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
+    The vertical bars shown for all simulation values except MacRog~K$^+$
     are not error bars, but demonstrate that for these systems
@@
 (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
-but further analysis is beyond the scope of this work which the PS headgroup.
+but further analysis is beyond the scope of this work which focuses on the PS headgroup.
 

### 2019-02-05 Manuscript/manuscriptPS.tex

--- 2019-02-04 Manuscript/manuscriptPS.tex
@@
 (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
-but further analysis is beyond the scope of this work which the PS headgroup.
+but further analysis is beyond the scope of this work which focuses on the PS headgroup.
 
-The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
-can guide the further force field development. However, more accurate simulations are
+The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
+can guide further force field development. However, more accurate MD force fields are
 required to confirm the suggested conformations because the simulated PS headgroup order parameters
-are not within the experimental error bars in any of the tested models.
+are not within the experimental error in any of the tested models.
 

### 2019-02-05 Manuscript/manuscriptPS.tex

@@
 but the abundance varies between different organelles up to
-25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
+25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
 Despite of the relatively low abundance, PS lipids

### 2019-02-05 Manuscript/manuscriptPS.tex
Stylistics and trying to improve clarity.

--- 2019-02-05 Manuscript/manuscriptPS.tex
@@
 \author{Batuhan Kav}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation{Potsdam, Germany}
 \author{Jesper J. Madsen}
@@
 \author{Markus Miettinen}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation{Potsdam, Germany}
 \author{Josef Melcr}
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp-eps-converted-to.pdf}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
   \caption{\label{HGorderParameters}
@@
   \caption{\label{HGorderParametersPS}
-    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
-    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
-    and MD simulations with different force fields (dots).
-    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
-    while all the other data are without additional salt.
-    All the data for DOPS at 303~K and all the data for POPS at 298~K.
+    Order parameters of PS headgroup and glycerol backbone from experiments
+    and simulations with different force fields.
+    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
+    while all the other data is without additional ions.
+    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
     Light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
-    The vertical bars shown for all simulation values except MacRog~K$^+$
+    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
+    The vertical bars shown for some simulation values
     are not error bars, but demonstrate that for these systems
@@
 
-The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
+The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
+between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
 as previously observed also for PC lipids \cite{botan15}.
-The models perform generally less well for PS than for PC
+The models perform generally less well for PS lipids than for PC lipids 
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
@@
 
-However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
-the best performing models,
-Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
-Slipids reproduces also the feature that the $\beta$-carbon order parameter
-in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
+However, the best performing models,
+Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
+and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
+when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
 Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
@@
 visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
-suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
-In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
+suggesting that the rotation of carboxyl group is limited in the serine headgroup.
+In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
 and restricted angle distribution for PS than for PC headgroup in
@@
 (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
-but further analysis is beyond the scope of this work which focuses on the PS headgroup.
+but further analysis is beyond the scope of this work which the PS headgroup.
 
-The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
-can guide further force field development. However, more accurate MD force fields are
+The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
+can guide the further force field development. However, more accurate simulations are
 required to confirm the suggested conformations because the simulated PS headgroup order parameters
-are not within the experimental error in any of the tested models.
+are not within the experimental error bars in any of the tested models.
 
@@
 
-Calcium binding affinity to membranes containing negative charged PS lipids can be
-measured by detecteding the PC lipid headgroup order parameters
+Calcium binding affinity to membranes containing the negative charged PS lipids can be
+experimentally measured by detecting the PC lipid headgroup order parameters
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
@@
 not observed~\cite{feigenson86,mattai89,roux90,roux91}.
-Despite the lack of ion free reference state
-in the presence of negatively charged lipids, the simulations give
+Despite the lack of an ion-free reference state
+in the presence of negatively charged lipids, our simulations give
 coherent results for POPC headgroup order parameters as a function of
@@
 in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
-indicating overestimated calcium binding binding affinity.
-Only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameter 
+indicating overestimated calcium binding affinity.
+The only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
 indicating weaker binding affinity than experiments.
@@
 data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
-is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
+is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
 in all the other tested models. This is evident in the calcium density distributions
-along membrane normal where almost all Ca$^{2+}$ ions bind to the membrane interface in
+along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
 all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
 
-The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
+The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
 exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
 below 100 mM (Fig. \ref{changesWITHCaClPS}).
-The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
-decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
-the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
-tested simulation models, including the CHARMM36 with underestimated binding affinity.
+In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
+whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
+the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
+tested simulation models, including CHARMM36 with underestimated binding affinity.
 In addition, different simulation models predict qualitatively different behaviour
-for the POPS $\alpha$-carbon order parameters with the added calcium.
-For example, both order parameters decrease in Berger simulations but increase
-in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
-more complicated. This is in contrast to the PC headgroup, where
-qualitatively correct reponse to the bound ions is observed
-in all simulation models despite of the significant discrepancies in the headgroup
+for the POPS $\alpha$-carbon order parameters with added calcium.
+For example, both order parameters decrease in Berger, but increase
+in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
+This is in contrast to the PC headgroup, where
+qualitatively correct reponse to bound ions is observed
+in all simulation models, despite significant discrepancies in the headgroup
 structure without additional ions~\cite{catte16}. Therefore, we conclude that
-the improvement of force fields is necessary to correctly describe interactions between
+improvement of force fields is necessary to correctly describe interactions between the
 PS headgroup and calcium ions in MD simulations.

### 2019-02-05 Manuscript/manuscriptPSsuppl.tex
update manuscriptPSsupp.tex
Simulation details with lipid17 force field are added.

@@
 The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl}
+
+One set of simulations of POPC:POPS (5:1) mixtures with additional calcium ions has been performed using Amber18 simulation package.
+Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
+The simulation details, minimization, heating, and equilibration runs are the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield.
+For all simulations, ff99SB~\cite{aqvist90} ion parameters and Amber lipid 17 force field were used.
+For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
+The trajectory files can be found in Ref~\citenum{POPCpopsLIPID17withCaCl}
 \todo{B. Kav and J. Melcr, please write simulation details} \\

### 2019-02-05 Manuscript/manuscriptPS.tex
Fixed few typos; asked for a few clarifications.

@@
     are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
+  }\todo{
+  1) Use diamonds for DOPS, spheres for POPS.
+  2) Error bars?
+  3) Change the y-label to $S_\mathrm{CH}$, and invert the y-axis, as in Fig. 3.
+  4) Make lower g1 POPC visible, e.g., by slightly larger point.
   }
@@
 
-The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
-carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
+The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
+carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
 The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
 The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
-known order parameters from the $^2$H NMR experiments~\cite{browning80}.
-Slices of the R-PDFL spectra and the resulting order parameters values
+known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
+Slices of the R-PDLF spectra and the resulting order parameter values
 are shown in the supplementary information (Fig. \ref{DPslices}). 
-Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
-only the absolute values of order parameters, we determined the signs of PS headgroup
+Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
+only the absolute values of order parameters, we determined the signs of the PS headgroup
 order parameters using the S-DROSS experiment \cite{gross97}.
 The S-DROSS slice clearly shows that the order parameter of
-the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
+the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
+Could we explain in a few words how this is clearly seen?}
 which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
@@
 that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
-using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
+using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
 The literature value was used because the
@@
 (dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
-confirming the interpretation that the smaller order parameter is negative.
+confirming \todo{Or rather 'corroborating'?}
+the interpretation that the smaller order parameter is negative.
 
@@
 POPS measured in this work are in good agreement with the previously reported
-values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
+values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
 When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
 the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
+experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
 
-Calcium binding affinity to membranes containing the negative charged PS lipids can be
-experimentally measured by detecting the PC lipid headgroup order parameters
+Calcium binding affinity to membranes containing negative charged PS lipids can be
+measured by detecteding the PC lipid headgroup order parameters
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
@@
 not observed~\cite{feigenson86,mattai89,roux90,roux91}.
-Despite the lack of an ion-free reference state
-in the presence of negatively charged lipids, our simulations give
+Despite the lack of ion free reference state
+in the presence of negatively charged lipids, the simulations give
 coherent results for POPC headgroup order parameters as a function of
@@
 in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
-indicating overestimated calcium binding affinity.
-The only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
+indicating overestimated calcium binding binding affinity.
+Only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameter 
 indicating weaker binding affinity than experiments.
@@
 data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
-is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
+is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
 in all the other tested models. This is evident in the calcium density distributions
-along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
+along membrane normal where almost all Ca$^{2+}$ ions bind to the membrane interface in
 all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
 
-The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
+The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
 exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
 below 100 mM (Fig. \ref{changesWITHCaClPS}).
-In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
-whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
-the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
-tested simulation models, including CHARMM36 with underestimated binding affinity.
+The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
+decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
+the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
+tested simulation models, including the CHARMM36 with underestimated binding affinity.
 In addition, different simulation models predict qualitatively different behaviour
-for the POPS $\alpha$-carbon order parameters with added calcium.
-For example, both order parameters decrease in Berger, but increase
-in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
-This is in contrast to the PC headgroup, where
-qualitatively correct reponse to bound ions is observed
-in all simulation models, despite significant discrepancies in the headgroup
+for the POPS $\alpha$-carbon order parameters with the added calcium.
+For example, both order parameters decrease in Berger simulations but increase
+in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
+more complicated. This is in contrast to the PC headgroup, where
+qualitatively correct reponse to the bound ions is observed
+in all simulation models despite of the significant discrepancies in the headgroup
 structure without additional ions~\cite{catte16}. Therefore, we conclude that
-improvement of force fields is necessary to correctly describe interactions between the
+the improvement of force fields is necessary to correctly describe interactions between
 PS headgroup and calcium ions in MD simulations.

### 2019-02-05 Manuscript/manuscriptPS.tex
Caption of Fig 1.

--- 2019-02-05 Manuscript/manuscriptPS.tex
@@
     (B) 2D R-PDLF spectra.
-    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
+    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
     the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller $\alpha$-carbon order parameter (dashed grey).
+    for the $\alpha$-carbon splittings.
+    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
+    for the smaller $\alpha$-carbon order parameter.
   }
-  \todo{I think that the peak labeling would be good to show also in (A).}
+  \todo{I think that the peak labeling would be good to show also in (A).}\\
+  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
+  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
 \end{figure}

### 2019-02-05 Manuscript/manuscriptPS.tex
Update Fig 2 related to the issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34

@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp-eps-converted-to.pdf}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
   \caption{\label{HGorderParameters}
     (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
-    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
-    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
-    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
+    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
+    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
     are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.

### 2019-02-05 Manuscript/manuscriptPS.tex
Style edits in experimental methods; asked for a few clarifications.

--- 2019-02-05 Manuscript/manuscriptPS.tex
@@
 
-\subsection{C-H bond order parameters from the natural abundance $^{13}$C NMR}
+\subsection{C--H bond order parameters from the natural abundance $^{13}$C\,NMR}
 
-Headgroup and glycerol backbone C-H bond order parameters of POPS
+Headgroup and glycerol backbone C--H bond order parameters of POPS
 were determined from the chemical-shift resolved dipolar splittings
-measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}.
+measured with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
 The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
@@
 
-{\emph{C-H bond order parameters from the R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
-for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
+{\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
+for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
 Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
-50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
-dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.
+50 kHz (SPINAL64 $^1$H decoupling pulses).
+\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
+The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
+dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?} of 149.5 ppm.
 
-\emph{Order parameter signs from the S-DROSS experiment}.
+\emph{Order parameter signs from the S-DROSS experiment.}
 The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
 experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
-blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
-frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
-The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
-measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.
+blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. Radio frequency pulses with the nutation
+frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
+\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
+The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
+measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?}  of 149.5 ppm.
 
 \emph{Numerical simulations of S-DROSS curves.}
-The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
-dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
-The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
+The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00}
+using as input the $^{13}$C--$^1$H
+dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
+The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate a random
 distribution of bilayer orientations in the samples studied.
 
-\emph{Sample preparation}
-The sample was prepared simply by mixing the POPS with water (lipid:water 60:40 wt\%) in an eppendorf
-tube mixing and centrifuging the sample repeatedly until an homogeneous viscous fluid was obtained.
-20 mg of sample was then transferred to an NMR insert suitable for 4 mm NMR rotors.  
+\emph{Sample preparation.}
+The sample was prepared simply by mixing the POPS \todo{From some company?} with water (lipid:water 60:40\,wt-\%) in an Eppendorf
+tube by mixing and centrifuging the sample repeatedly until a \todo{visually(?)} homogeneous viscous fluid was obtained.
+Then 20 mg of the sample was  transferred to an NMR insert suitable for 4 mm NMR rotors.  
 \todo{Maybe we need little bit more information about the mixing procedure?}
@@
     (B) 2D R-PDLF spectra.
-    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
+    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
     the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the $\alpha$-carbon splittings.
-    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
-    for the smaller $\alpha$-carbon order parameter.
+    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller $\alpha$-carbon order parameter (dashed grey).
   }
-  \todo{I think that the peak labeling would be good to show also in (A).}\\
-  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
-  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
+  \todo{I think that the peak labeling would be good to show also in (A).}
 \end{figure}
@@
     are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
-  }\todo{
-  1) Use diamonds for DOPS, spheres for POPS.
-  2) Error bars?
-  3) Change the y-label to $S_\mathrm{CH}$, and invert the y-axis, as in Fig. 3.
-  4) Make lower g1 POPC visible, e.g., by slightly larger point.
   }
@@
 
-The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
-carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
+The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
+carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
 The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
 The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
-known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
-Slices of the R-PDLF spectra and the resulting order parameter values
+known order parameters from the $^2$H NMR experiments~\cite{browning80}.
+Slices of the R-PDFL spectra and the resulting order parameters values
 are shown in the supplementary information (Fig. \ref{DPslices}). 
-Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
-only the absolute values of order parameters, we determined the signs of the PS headgroup
+Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
+only the absolute values of order parameters, we determined the signs of PS headgroup
 order parameters using the S-DROSS experiment \cite{gross97}.
 The S-DROSS slice clearly shows that the order parameter of
-the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
-Could we explain in a few words how this is clearly seen?}
+the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
 which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
@@
 that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
-using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
+using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
 The literature value was used because the
@@
 (dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
-confirming \todo{Or rather 'corroborating'?}
-the interpretation that the smaller order parameter is negative.
+confirming the interpretation that the smaller order parameter is negative.
 
@@
 POPS measured in this work are in good agreement with the previously reported
-values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
+values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
 When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
 the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
+experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup

### 2019-02-05 Manuscript/manuscriptPS.tex
Style edits to MD methods.

--- 2019-02-05 Manuscript/manuscriptPS.tex
@@
 
-\subsection{C--H bond order parameters from the natural abundance $^{13}$C\,NMR}
+\subsection{C-H bond order parameters from the natural abundance $^{13}$C NMR}
 
-Headgroup and glycerol backbone C--H bond order parameters of POPS
+Headgroup and glycerol backbone C-H bond order parameters of POPS
 were determined from the chemical-shift resolved dipolar splittings
-measured with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
+measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}.
 The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
@@
 
-{\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
-for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
+{\emph{C-H bond order parameters from the R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
+for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
 Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
-50 kHz (SPINAL64 $^1$H decoupling pulses).
-\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
-The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
-dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?} of 149.5 ppm.
+50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
+dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.
 
-\emph{Order parameter signs from the S-DROSS experiment.}
+\emph{Order parameter signs from the S-DROSS experiment}.
 The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
 experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
-blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. Radio frequency pulses with the nutation
-frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
-\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
-The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
-measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?}  of 149.5 ppm.
+blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
+frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
+The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
+measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.
 
 \emph{Numerical simulations of S-DROSS curves.}
-The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00}
-using as input the $^{13}$C--$^1$H
-dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
-The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate a random
+The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
+dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
+The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
 distribution of bilayer orientations in the samples studied.
 
-\emph{Sample preparation.}
-The sample was prepared simply by mixing the POPS \todo{From some company?} with water (lipid:water 60:40\,wt-\%) in an Eppendorf
-tube by mixing and centrifuging the sample repeatedly until a \todo{visually(?)} homogeneous viscous fluid was obtained.
-Then 20 mg of the sample was  transferred to an NMR insert suitable for 4 mm NMR rotors.  
+\emph{Sample preparation}
+The sample was prepared simply by mixing the POPS with water (lipid:water 60:40 wt\%) in an eppendorf
+tube mixing and centrifuging the sample repeatedly until an homogeneous viscous fluid was obtained.
+20 mg of sample was then transferred to an NMR insert suitable for 4 mm NMR rotors.  
 \todo{Maybe we need little bit more information about the mixing procedure?}
@@
 \subsection{Molecular dynamics simulations}
-Molecular dynamics simulation data was collected using
-the Open Collaboration method \cite{botan15}.
-The NMRlipids project blog (\url{www.nmrlipids.blogspot.fi}) and
-the GitHub repository (\url{www.github.com/NMRLipids/NMRlipidsIVotherHGs})
-were used as the communication platforms.
+Molecular dynamics simulation data were collected using
+the Open Collaboration method \cite{botan15}, with
+the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
+GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
+as the communication platforms.
 The simulated systems are listed in 
-Tables \ref{PSsystems} (pure PS systems without additional ions) 
-and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
+Tables \ref{PSsystems} (pure PS bilayers without additional ions) 
+and \ref{mixedIONsystems} (mixed PC:PS bilayers at various salt concentrations).
 Further simulation details are given in the SI.
-The simulation data is also indexed in the
-searchable database (\url{www.nmrlipids.fi}),
-and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).
+The simulation data are also indexed in a
+searchable database available at \url{www.nmrlipids.fi},
+and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).
 
-The C-H bond order parameters were calculated directly
-from the definition
+The C--H bond order parameters were calculated directly
+from the carbon and hydrogen positions using the definition
 \begin{equation}
@@
 \end{equation}
-where $\theta$ is the angle between the C-H bond and the membrane normal.
-Angular brackets point to the average over all sampled configurations.
+where $\theta$ is the angle between the C--H bond and the membrane normal
+(taken to align with $z$, with bilayer periodic in the $xy$-plane).
+Angular brackets denote average over all sampled configurations.
 The order parameters were first calculated averaging over time separately
-for each lipid molecule in the system. The average and
+for each lipid in the system. The average and
 the standard error of the mean were then calculated over different lipids.
-The python program ({\it scripts/calcOrderParameters.py}) utilizing
+Python program ({\tt scripts/calcOrderParameters.py}) that uses the
 MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
-The number density profiles were calculated using {\it gmx density} tool
-from Gromacs sofware package \cite{gromacsMANUAL}.
+The number density profiles were calculated using the {\tt gmx density} tool
+of the Gromacs sofware package \cite{gromacsMANUAL}.
 

### 2019-02-05 Manuscript/manuscriptPS.tex
Update according to comments by M. Javanainen http://nmrlipids.blogspot.com/2018/12/nmrlipids-iv-toward-submission-of.html?showComment=1545138569759#c1540577296270662466

@@
   }
+  \todo{Commented by M. Javanainen in blog:
+    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready}
 \end{figure}
@@
     Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
+    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
     at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
@@
   CSC-IT center for science for computational resources.
+  MJ acknowledges financial support from the Emil Aaltonen foundation and
+  and CSC-IT center for science for computational resources.
 \end{acknowledgments}

### 2019-02-05 Manuscript/manuscriptPSsuppl.tex
Update according to comments by M. Javanainen http://nmrlipids.blogspot.com/2018/12/nmrlipids-iv-toward-submission-of.html?showComment=1545138569759#c1540577296270662466

@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
+  \includegraphics[width=17.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
   \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
@@
 \pagebreak
-\section{Calcium density profiles from simulations with POPC:POPS(5:1) mixture}
+\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
 \begin{figure}[ht]
   \centering
-  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
+  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
   \caption{\label{CAdensPCPSmixtureALL}

### 2019-02-06 Manuscript/manuscriptPS.tex
Style edits to methods, electrometer.

--- 2019-02-05 Manuscript/manuscriptPS.tex
@@
 \subsection{Comparison of ion binding to negatively charged lipid bilayers 
-between simulations and experiments using the electrometer concept}
+between simulations and experiments using the molecular electrometer}
 
-The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
-can be used to measure the ion binding affinity because they 
-decrease proportionally to the amound of bound positive
-charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
-This molecular electrometer concept is especially useful for 
-the comparison between simulations and experiments because
-the headgroup order parameters can be directly calculated from 
-simulations and compared to the experimental data with varying cation
-concentrations~\cite{catte16}. The headgroup order parameters
-of negatively charged PS and PG lipids exhibit systemic, but less
-characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
+As the order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
+decrease proportionally to the amount of positive
+charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
+they can be used to measure the ion binding affinity.
+This concept, known as the molecular electrometer, is especially useful for 
+comparison between simulations and experiments, as
+the headgroup order parameters at varying cation
+concentrations can be directly calculated from 
+simulations and compared to the corresponding experimental data~\cite{catte16}. The headgroup order parameters
+of negatively charged PS and PG lipids also exhibit systematic, but less
+understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
 Therefore, the ion binding affinity to negatively charged bilayers
@@
 Before using the PC headgroup order parameters to quantify the ion binding
-affinity, it is important to quantify their response to the known amount of bound charge~\cite{catte16,melcr18}.
-This can be done using the experimental data from the mixtures of
+affinity, it is important to quantify their response to a known amount of bound charge~\cite{catte16,melcr18}.
+This can be done using experimental data from mixtures of
 monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
@@
 In this work, we also quantify the response of PC headgroup order parameters
-to the negatively charged PS headgroups, which also follows the electrometer
-concept in the experiments~\cite{scherer87},
+to the negatively charged PS headgroups, which also follows the molecular electrometer
+concept in experiments~\cite{scherer87},
 see section \ref{electrometerFORmixtures} in the supplementary information.
 
-In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
-the lipids were first soluted to the buffer and then centrifuged to a pellet which
-was used in the measurements. Such samples have lower lipid concentration
+In the experimental $^2$H~NMR literature data used in this work~\cite{browning80,roux90},
+the lipids were first solvated to the buffer and then centrifuged to a pellet that
+was used in the measurements. Such samples have a lower lipid concentration
 (approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
@@
 concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
-However, the PS headgroup order parameters measured from gravimetric sample in this work
-are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
-Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
-salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
+However, the PS headgroup order parameters measured from gravimetric samples in this work
+are in good agreement with the results from centrifuged samples in the literature~\cite{browning80} (Fig.~\ref{HGorderParameters}).
+Furthermore, the equilibrium repeat distance rapidly decreases with the addition of monovalent
+salts, and is close to the typical simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
 Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
@@
 
-Two different definitions for the salt concentrations have been used when
-electrometer concept is applied to study ion binding affinity.
-The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
+Two different definitions for the salt concentrations have been used when the molecular
+electrometer concept is applied to study ion binding affinities.
+The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
 or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
 In this work, we use the former definition to be consistent with the reference
-experimental data \cite{roux90}. The used definition has only a marginal effect
-to the results is simulations with realistic ion binding affinity
+experimental data \cite{roux90}. The choice of definition has only a marginal effect
+to the results in simulations with realistic ion binding affinity
 (section \ref{concentrationDEFsection} in the supplementary information).

### 2019-02-06 Manuscript/manuscriptPSsuppl.tex

@@
 {\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
-Systems containing total 132 (110:22) lipids
-were simulated with the additional potassium and sodium countertion concentrations
-of $\sim$450 mM and 890 mM using CHARMM36 force field at 298 K. The data, force field and
-input files generated using parameters given by CHARMM-GUI \cite{lee16,jo18} are
-available from Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\
+POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
+potassium or sodium ions corresponding concentrations of $\sim$450 mM and 890 mM
+were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
+Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
+parameters given by the CHARMM-GUI at 298 K. For further details, simulation files
+and data, see table \ref{mixedIONsystemsMONOVALENT} and
+Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\
 
@@
 \noindent
-    {\it POPC bilayers.} Simulation data available from Ref. \citenum{macrogPOPC298K}.
-    \todo{For example this data set has pretty good description of simulation details in the
-      reference. I am not sure how much we should repeat these in here.}
-
+{\it POPC bilayers.} 
+% \todo{For example this data set has pretty good description of simulation details in the
+% reference. I am not sure how much we should repeat these in here.}
 The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
-from Ref.~\cite{kulig15c}. The bilayer was hydrated by a total of 5120 water 
+from Ref.~\citenum{kulig15c}. The bilayer was hydrated by a total of 5120 water 
 molecules, corresponding to 40 water molecules per lipid.
-
-The MacRog lipid parameters \cite{kulig15b}, obtained from 
-\cite{kulig15c}, were used, along with the TIP3P water model 
-\cite{jorgensen83}. The simulation parameters were identical to those used for 
-POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns, 
-out of which \hl{??~ns: (this is not tabulated in the MS or the SI)} were used for analyses.
+The MacRog lipid parameters \cite{kulig15b}, obtained from Ref.~\citenum{kulig15c}, were used,
+along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
+POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham15}, 
+out of which 150~ns were used for analyses. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.
 
@@
 Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
-
-These bilayers were simulated in the presence of counterions only, as well as together with different
+These bilayers were simulated using Gromacs ?? \cite{check the version} in the presence of counterions only, as well as together with different
 concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
@@
 In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Lipid parameters were obtained from 
-Amber Lipid17 force field. For the ions, one set of simulations were run with Joung-Cheatham
+to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
+and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
 ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB~\cite{aqvist90} ion parameters.
@@
 The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
-Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}.
-For the simulations of the lipids and ions, the Amber Lipid 17 and ff99SB force fields were used, respectively.
-The simulation steps are the same as the the simulations of DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns was used for the analysis.
-The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
+Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}. For the simulations of the lipids and ions,
+the Amber Lipid 17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
+Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
+The simulation steps are the same as the the simulations of
+DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns
+was used for the analysis. The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
 The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.\\
@@
 The trajectory files can be found in Ref~\citenum{POPCpopsLIPID17withCaCl}
-\todo{B. Kav and J. Melcr, please write simulation details} \\
+
+\todo{J. Melcr, please write simulation details} \\
 

### 2019-02-06 Manuscript/manuscriptPS.tex
Update simulation details
Stylistics tuning of discussion.

@@
 Phosphatidylserine (PS) is the most common negatively
-charged lipid in eykaryotic membranes.
-PS lipids compose 8.5\% of total lipid weight of erythrocytes,
+charged lipid in eukaryotic membranes.
+PS lipids compose 8.5\% of total lipid weight of red blood cells,
 but the abundance varies between different organelles up to
 25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
-Despite of the relatively low abundance, PS lipids
+Despite of the relatively low abundance \todo{Relatively low compared to what?}, PS lipids
 are important signaling molecules. They interact with
@@
 induce protein aggregation \cite{zhao04,gorbenko06}.
-Some protein domains spesifically interact PS lipids,
-while others are attracted by general electrostatics and the
+Some protein domains interact specifically with PS lipids,
+while others are attracted by nonspecific electrostatics and the
 binding can be regulated by calcium \cite{leventis10}.
@@
 of lipid headgroups and the details of cation binding
-are crucial for the PS mediated signaling processes.
+are crucial for the PS-mediated signaling processes.
 
-Previous experimental studies have concluded that
-PS headgroups are more rigid than phophocholines (PC)
+Previous experimental studies have concluded that the
+PS headgroup is more rigid than the phosphocholine (PC)
 due to the hydrogen bonding network or
@@
 dehydrated molecular complexes with PS lipids,
-while monovalent ions interact more weakly with PS
-containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
-The dehydrated complexes of PS headgroup and calcium ions can also lead to the
+while monovalent ions interact more weakly with
+PS-containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+The dehydrated complexes of PS headgroup and calcium ions can also lead to
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
 On the other hand, some studies propose that the specific binding
-affinity is similar to the negatively charged and zwitterionic lipids and that
-the increased cation binding to negatively charged lipid bilayer arise only due
-to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
+affinity is similar for the negatively charged and zwitterionic lipids, and
+the increased cation binding to negatively charged lipid bilayers arises only due
+to the increased local cation concentration in the membranes vicinity~\cite{seelig90,sinn06}.
 Dilution of bilayers with PC lipids makes PS headgroups
-less rigid and reduces propensity for the formation of
-strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
+less rigid and reduces their propensity to form
+strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.
 The molecular level interpretation of these observations is,
-however, not available.
+however, lacking.
 
@@
 
-Several classical molecular dynamics (MD) simulation studies are done
-to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with
+Several classical molecular dynamics (MD) simulation studies have been done
+to understand PS headgroups, their influence on lipid bilayer properties, and their
+interaction with
 ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
-However, the results strongly depend on the used force field parameters.
+However, the results depend strongly on the force field used.
 For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
-CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
-suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
-the same force field without the NBfix parameters together with the NMR chemical shifts and
-REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
-On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
-combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
+CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
+suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
+the same force field without the NBfix parameters, combined with NMR chemical shifts and
+REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
+On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
+combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
 calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
 We have recently demonstrated that such controversies can be resolved by
-comparing C-H bond order parameters of lipid headgroups between simulations
-and experiments \cite{botan15,catte16}, which can be directly
+comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
+and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
 measured from NMR experiments with high accuracy and compared to simulations
@@
 Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
-captured by most simulation models \cite{botan15} and that the cation binding to PC
-lipid bilayers is overestimated \cite{catte16}. Based on this data, the cation binding
-affinity to POPC bilayer was then improved by implicitly including the electronic polarizability
+captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
+lipid bilayers is overestimated \cite{catte16}. Based on these data, the cation binding
+affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability
 using the electronic continuum correction \cite{melcr18}.
 
-Here, we collect the set of experimentally measured lipid headgroup and
+Here, we collect a set of experimentally measured lipid headgroup and
 glycerol backbone C-H bond order parameters, which can be used to
@@
 The available MD simulation models of PS are then compared with
-the collected experimental data using the NMRlipids open collaboration
-project (\url{www.nmrlipids.blogspot.fi}). The results pave the way
+the collected experimental data.
+%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
+The results pave the way
 for the development of lipid force fields with realistic description of
@@
   \hline
+    POPC:POPS (1:0)        & MacRog \cite{maciejewski14} &0    & 120:0  & 5120 & 0    & 298  & 200 & 150 & \cite{macrogPOPC298K}  \\
     POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
@@
   }
-  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
-    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
-  these are not necessary, but could be added here if delivered.} \\
+%  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
+%    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
+%  these are not necessary, but could be added here if delivered.} \\
 \end{figure*}

### 2019-02-06 Manuscript/manuscriptPS.tex
Stylistics tuning of introduction.

@@
 Previous experimental studies have concluded that the
-PS headgroup is more rigid than the phosphocholine (PC)
-due to the hydrogen bonding network or
+PS headgroup is more rigid than the phosphatidylcholine (PC)
+due to hydrogen bonding network or
 electrostatic interactions \cite{browning80,buldt81}.
@@
 the increased cation binding to negatively charged lipid bilayers arises only due
-to the increased local cation concentration in the membranes vicinity~\cite{seelig90,sinn06}.
+to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.
 Dilution of bilayers with PC lipids makes PS headgroups
@@
 Several classical molecular dynamics (MD) simulation studies have been done
-to understand PS headgroups, their influence on lipid bilayer properties, and their
+to understand the PS headgroups, their influence on lipid bilayer properties, and their
 interaction with
 ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
-However, the results depend strongly on the force field used.
-For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
+However, the results have depended strongly on the force field used.
+For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
 CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
@@
 captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
-lipid bilayers is overestimated \cite{catte16}. Based on these data, the cation binding
-affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability
-using the electronic continuum correction \cite{melcr18}.
+lipid bilayers is overestimated \cite{catte16}.
+%
+Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
+\todo{Should we leave the mention of ECC out from the Introduction
+(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}
 
 Here, we collect a set of experimentally measured lipid headgroup and
-glycerol backbone C-H bond order parameters, which can be used to
-evaluate the quality of headgroup structure and the ion binding affinity
+glycerol backbone C--H bond order parameters, which can be used to
+evaluate the quality of headgroup structure and ion binding affinity
 in MD simulations of lipid bilayers containing PS lipids. 
@@
 The results pave the way
-for the development of lipid force fields with realistic description of
+for development of lipid force fields with realistic description of
 the headgroup region of negatively charged lipids in physiological salt
 conditions. Such models are expected to be useful in understanding
-biological function of lipid headgroups and glycerol backbone because
-they behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.
+biological function of lipid headgroups and glycerol backbone, as
+these behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.
 

### 2019-02-06 Manuscript/manuscriptPS.tex
Stylistics tuning of introduction.

@@
 \section{Conclusions}
-Lipids with PS headgroups and their interactions with ions
-play an important role in lipid mediated signaling
-processes \cite{leventis10,yeung08}. Recent studies using molecular
-dynamics simulations to interpret the various spectroscopic data
+Lipids with PS headgroups, and their interactions with ions,
+play an important role in lipid-mediated signaling
+processes \cite{leventis10,yeung08}. Recent studies using MD
+simulations to interpret the various spectroscopic data
 give contradictory results for the calcium binding details to PS
 headgroups \cite{melcrova16,valentine18,hallock18}.  
-Here, we used the headgroup C-H bond order parameters and open collaboration method to evaluate the quality
-of headgroup structure and ion binding affinity to PS lipids
-in available MD simulation force fields,
-as previously done for PC lipids \cite{botan15,catte16}.
+Here, as was previously done for PC lipids \cite{botan15,catte16},
+we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
+of the headgroup structure and ion binding affinity to PS lipids
+in available MD force fields.
 The main advantage of this approach is the direct connection
-between accurately measured experimental order parameters and simulations
+between the accurately measured experimental order parameters and the simulations,
 which reduces the ambiguity in the interpretation of experiments.
@@
 lipid headgroup order parameters \cite{browning80,roux90} by measuring the signs of the order parameters.
-None of the available force fields tested using the NMRlipids open
-collaboration was accurate enough to reproduce the PS
-headgroup order parameters with the experimental accuracy, but
-the best models suggested a characteristic rigid conformation for the
-carboxyl group in the serine headgroup. Comparison to the previously
+Comparison to these data revealed that none of the available force fields
+%tested using the NMRlipids open collaboration
+was accurate enough to reproduce the PS
+headgroup order parameters within the experimental accuracy. However,
+the best models for the serine headgroup suggested a characteristic rigid conformation for its
+carboxyl group. Comparison to the previously
 measured headgroup order parameters from POPC:POPS (5:1) bilayers with different ion
-concentrations \cite{roux90} showed that the tested MD simulation force fields
+concentrations \cite{roux90} then showed that the tested MD force fields
 overestimate the cation binding affinity to the negatively charged bilayers
-containing PS lipids with two exceptions. The apparently most realisitic monovalent ion binding
-affinity to PS containing lipid bilayers was observed in the MacRog
-simulations with potassium counterions and the CHARMM36 force field with recently introduced
+containing PS lipids with two exceptions. 1) The apparently most realistic monovalent ion binding
+affinity to PS-containing lipid bilayers was observed in the MacRog
+simulations with potassium counterions. 2) The CHARMM36 force field with the recently introduced
 NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
-However, the experimentally measured trends of the PS headgroup order parameter response
-to the bound calcium and to the dilution of bilayer with zwitterionic PC lipids were not
-qualitatively reproduced in any of the tested force fields, indicating that improvements in
-the MD simulation force fields are necessary to study interactions between PS lipids and other biomolecules.
+The experimentally measured trends of the PS headgroup order parameter response
+to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields, which indicates that improvements in
+the MD force fields are necessary to study interactions between PS lipids and other biomolecules.
 This is different to the previous results with PC lipids,
 where the experimentally measured headgroup order parameter responses to the bound charge
-were qualitative reproduced even thought the headgroup structures without ions were not
-correct and the cation binding affinities were overestimated \cite{catte16}.
+were qualitative reproduced even though the headgroup structures themselves were
+incorrect and the cation binding affinities overestimated \cite{catte16}.
 
-Our results pave the way for the development of better MD simulations force fields
+Our results pave the way for the development of better MD force fields
 for PS lipids. Using the headgroup order parameters, we were able to evaluate the
@@
 for POPC using the electronic continuum correction \cite{melcr18}.
-Similar study for POPS is progressed separately \cite{ECCpops}.
+Similar study for POPS is being progressed separately \cite{ECCpops}.
 

### 2019-02-06 Manuscript/manuscriptPS.tex

@@
 Phosphatidylserine (PS) is the most common negatively
-charged lipid in eykaryotic membranes.
+charged lipid in eukaryotic membranes.
 PS lipids interact with signaling and other proteins via
@@
 membrane fusion and phase separation together with calcium ions.
-Molecular details of these phenomena are not well understood
-because accurate models to interpret the experimental data has not
-been available. Here, we collect a set of experimental NMR data which
+Molecular details of these phenomena are not well understood,
+because accurate models to interpret the experimental data have not
+been available. Here, we collect a set of experimental NMR data that
 can be used together with molecular dynamics (MD) simulations
 to interpret the lipid headgroup structures and details of ion binding
-in pure PS and mixed PS:PC lipid bilayers. We use the open collaboration
-method to collect data from available MD simulation models of PS lipids.
-None of the models reproduce the NMR data with experimental accuracy,
-but the best models suggest that the carboxyl group in the serine headgroup
+in pure PS, and mixed PS:PC, lipid bilayers. We use the open collaboration
+method to collect data from a wide range of available MD models of PS lipids.
+None of the models reproduce the NMR data within experimental accuracy,
+but the best headgroup models suggest that the carboxyl group in the serine headgroup
 does not rotate freely. In line with the previous results for PC lipids,
-none of the tested force fields correctly captures the cation binding affinity to lipid bilayers containing
-PS lipids. In contrast to PC lipids, the response of PS headgroups
-to the bound ions qualitatively differs from experiments in the tested MD
-simulation models. The collected experimental dataset and simulation
+none of the force fields correctly captures the cation binding affinity.
+In contrast to PC lipids, the response of PS headgroups
+to bound ions  in the tested MD simulation models
+qualitatively differs from experiments. The collected experimental dataset and simulation
 results pave the way for improvement of lipid force fields to correctly
-describe negatively charged membranes and their interactions with ions.
-The work is performed in the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
+describe the biologically relevant negatively charged membranes and their interactions with ions.
+This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
 \end{abstract}

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
   % insert abstract here
-Phosphatidylserine (PS) is the most common negatively
-charged lipid in eukaryotic membranes.
-PS lipids interact with signaling and other proteins via
-electrostatic interactions and direct binding, and induce
-membrane fusion and phase separation together with calcium ions.
+Phosphatidylserine (PS) is a negatively
+charged lipid commonly found in eukaryotic membranes, where it interacts with signaling and other proteins via
+electrostatic interactions and direct binding, and can induce
+membrane fusion and phase separation in the presence of calcium ions.
 Molecular details of these phenomena are not well understood,
 because accurate models to interpret the experimental data have not
-been available. Here, we collect a set of experimental NMR data that
-can be used together with molecular dynamics (MD) simulations
-to interpret the lipid headgroup structures and details of ion binding
-in pure PS, and mixed PS:PC, lipid bilayers. We use the open collaboration
-method to collect data from a wide range of available MD models of PS lipids.
-None of the models reproduce the NMR data within experimental accuracy,
-but the best headgroup models suggest that the carboxyl group in the serine headgroup
+been available. Here, we gather a set of experimental NMR data that 
+can, together with molecular dynamics (MD) simulations, be used
+to decipher the lipid headgroup structures and ion binding
+in pure PS and mixed (here PS:PC) lipid bilayers. Using the open collaboration
+method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
+%or: Using the open collaboration method, we collect a database of MD simulations using wide range of available PS models (force fields) and perform a comparison to NMR results
+We find that none of the models reproduce the NMR data within experimental accuracy,
+but the best ones suggest that the carboxyl group in the serine headgroup
 does not rotate freely. In line with the previous results for PC lipids,
-none of the force fields correctly captures the cation binding affinity.
-In contrast to PC lipids, the response of PS headgroups
-to bound ions  in the tested MD simulation models
-qualitatively differs from experiments. The collected experimental dataset and simulation
+none of the PS force fields correctly captures the cation binding affinity.
+Furthermore, the response of PS headgroups to bound ions
+\emph{qualitatively} differs from experiments. The collected experimental dataset and simulation
 results pave the way for improvement of lipid force fields to correctly

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-06 Manuscript/manuscriptPS.tex
@@
 charged lipid in eukaryotic membranes.
-PS lipids compose 8.5\% of total lipid weight of red blood cells,
-but the abundance varies between different organelles up to
-25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
-Despite of the relatively low abundance \todo{Relatively low compared to what?}, PS lipids
-are important signaling molecules. They interact with
+PS lipids compose 8.5\% of total lipid weight of red blood cells. 
+The abundance, however, varies between different organelles, and up to
+25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
+PS lipids are vastly important signaling molecules. They interact with
 signaling proteins \cite{leventis10}, regulate
@@
 
-Previous experimental studies have concluded that the
+Experimental studies have indicated that the
 PS headgroup is more rigid than the phosphatidylcholine (PC)
@@
 electrostatic interactions \cite{browning80,buldt81}.
-Multivalent cations and Li$^+$ are able to form strong
-dehydrated molecular complexes with PS lipids,
-while monovalent ions interact more weakly with
-PS-containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
-The dehydrated complexes of PS headgroup and calcium ions can also lead to
+While monovalent ions interact weakly with
+PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
+dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+The dehydrated complexes of PS headgroup and calcium ions can even lead to
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
@@
 strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.
-The molecular level interpretation of these observations is,
-however, lacking.
 
@@
 
-Several classical molecular dynamics (MD) simulation studies have been done
-to understand the PS headgroups, their influence on lipid bilayer properties, and their
+The molecular level interpretation of these observations is,
+however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
+to understand the PS headgroup structure, their influence on lipid bilayer properties, and their
 interaction with
 ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
-However, the results have depended strongly on the force field used.
+Unfortunately, the results have depended strongly on the force field used.
 For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
@@
 calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
+
 We have recently demonstrated that such controversies can be resolved by
@@
 The results pave the way
-for development of lipid force fields with realistic description of
+for development of lipid models that correctly describe 
 the headgroup region of negatively charged lipids in physiological salt
-conditions. Such models are expected to be useful in understanding
+conditions. Such force fields are expected to be useful in understanding
 biological function of lipid headgroups and glycerol backbone, as
-these behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.
+these are knwon to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.
 

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-06 Manuscript/manuscriptPS.tex
@@
 dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
-The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate a random
+The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}}\todo{citation needed?} was used to simulate a random
 distribution of bilayer orientations in the samples studied.

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
 dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
-The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}}\todo{citation needed?} was used to simulate a random
+The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate a random
 distribution of bilayer orientations in the samples studied.
@@
 and \ref{mixedIONsystems} (mixed PC:PS bilayers at various salt concentrations).
-Further simulation details are given in the SI.
-The simulation data are also indexed in a
+Further simulation details are given in the SI, and
+the simulation data are indexed in a
 searchable database available at \url{www.nmrlipids.fi},
@@
 MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
-The number density profiles were calculated using the {\tt gmx density} tool
+The ion \todo{There are ion number densities, right?} number density profiles were calculated using the {\tt gmx density} tool
 of the Gromacs sofware package \cite{gromacsMANUAL}.

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
 \subsection{Comparison of ion binding to negatively charged lipid bilayers 
-between simulations and experiments using the molecular electrometer}
+between simulations and experiments using the molecular electrometer concept}
 
-As the order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
+The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
 decrease proportionally to the amount of positive
 charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
-they can be used to measure the ion binding affinity.
+and can therefore be used to measure the ion binding affinity.
 This concept, known as the molecular electrometer, is especially useful for 
@@
 the headgroup order parameters at varying cation
-concentrations can be directly calculated from 
-simulations and compared to the corresponding experimental data~\cite{catte16}. The headgroup order parameters
+concentrations can be easily calculated from
+simulations and compared to experimental data~\cite{catte16}. The headgroup order parameters
 of negatively charged PS and PG lipids also exhibit systematic, but less
 understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
-Therefore, the ion binding affinity to negatively charged bilayers
-can be better characterized by measuring the PC headgroup order parameters from 
-mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.
+Therefore, measuring the PC headgroup order parameters from 
+mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see SI section \ref{electrometerFORmixtures} provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
+% Removed a paragraph here because it was out of place in the molecular electrometer section and contained some results. I  rewrote it and added to the end of next section
 
-Before using the PC headgroup order parameters to quantify the ion binding
-affinity, it is important to quantify their response to a known amount of bound charge~\cite{catte16,melcr18}.
+Important preliminary step for using the molecular electrometer concept is to calibrate the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18}.
 This can be done using experimental data from mixtures of
 monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
-see section \ref{electrometerCALIBRATION} in the supplementary information.
-In this work, we also quantify the response of PC headgroup order parameters
-to the negatively charged PS headgroups, which also follows the molecular electrometer
-concept in experiments~\cite{scherer87},
-see section \ref{electrometerFORmixtures} in the supplementary information.
+(see SI section \ref{electrometerCALIBRATION}). Additionally, we quantify the response of PC headgroup order parameters
+to the negatively charged PS, which also follows the molecular electrometer
+concept in experiments~\cite{scherer87} (see SI section \ref{electrometerFORmixtures}).
 
-In the experimental $^2$H~NMR literature data used in this work~\cite{browning80,roux90},
-the lipids were first solvated to the buffer and then centrifuged to a pellet that
-was used in the measurements. Such samples have a lower lipid concentration
-(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
-gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
-Larger multilamellar repeat distances are expected in the samples with lower lipid
-concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
-However, the PS headgroup order parameters measured from gravimetric samples in this work
-are in good agreement with the results from centrifuged samples in the literature~\cite{browning80} (Fig.~\ref{HGorderParameters}).
-Furthermore, the equilibrium repeat distance rapidly decreases with the addition of monovalent
-salts, and is close to the typical simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
-Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
-in the used simulations and reference experiments.
-
-Two different definitions for the salt concentrations have been used when the molecular
-electrometer concept is applied to study ion binding affinities.
+Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
 The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
@@
 In this work, we use the former definition to be consistent with the reference
-experimental data \cite{roux90}. The choice of definition has only a marginal effect
+experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
 to the results in simulations with realistic ion binding affinity
-(section \ref{concentrationDEFsection} in the supplementary information).
+(see SI section \ref{concentrationDEFsection}).
 

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
     POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
-    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
+    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
   }\todo{
@@
     glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
-    and MD simulations with different force fields (dots).
+    and MD simulations with different force fields (symbols).
     Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
     while all the other data are without additional salt.
-    All the data for DOPS at 303~K and all the data for POPS at 298~K.
+    The data for DOPS is at 303~K and the data for POPS is at 298~K.
     Light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
-    The vertical bars shown for all simulation values except MacRog~K$^+$
+    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
+    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
     are not error bars, but demonstrate that for these systems
@@
 available. 
+%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results.
+This also enables referring to figures in order.
+I also removed the refs which do not seem to refer to the actual data used in the figure.
+%Rewrote the paragraph, see if ok.
+
+We note that the the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating the lipids to the buffer and then centrifuging the sample to a pellet that was used in the measurements.
+Such samples have a lower lipid concentration
+(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
+gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
+Larger multilamellar repeat distances are expected in the samples with lower lipid
+concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
+Yet the PS headgroup order parameters measured from gravimetric samples (POPS) in this work
+are in good agreement with the results from centrifuged samples~\cite{browning80}.
+This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the simulations and reference experiments.
 

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
 and \ref{mixedIONsystems} (mixed PC:PS bilayers at various salt concentrations).
-Further simulation details are given in the SI, and
-the simulation data are indexed in a
+Further simulation details are given in the SI.
+The simulation data are also indexed in a
 searchable database available at \url{www.nmrlipids.fi},
@@
 MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
-The ion \todo{There are ion number densities, right?} number density profiles were calculated using the {\tt gmx density} tool
+The number density profiles were calculated using the {\tt gmx density} tool
 of the Gromacs sofware package \cite{gromacsMANUAL}.
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from CHARMM36 simulations in best agreement with experiments
-    to demonstrate the conformational fluctuations around
+    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
+     demonstrate the conformational fluctuations around
     (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
@@
 
-The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
+The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
 and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
 as previously observed also for PC lipids \cite{botan15}.
+None of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
-Therefore, the interpretation of structural differences between PC and PS headgroups
-from simulations is not straightforward.
+which complicates the interpretation of structural differences between PC and PS headgroups.
 
@@
 
-However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
-the best performing models,
-Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
-Slipids reproduces also the feature that the $\beta$-carbon order parameter
-in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
+However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC forking of the $\alpha$-carbon \todo{observed in the experiments?} and the Slipids force field additionally correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
+\ref{HGorderParametersPS} vs.
+Fig.
+2 in Ref.~\citenum{botan15}) \todo{refer to experiments also?}.
+
 Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
 a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
-in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
-The restricted motion is also
-visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
+in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
+The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
 suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
 In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-and restricted angle distribution for PS than for PC headgroup in
-CHARMM36 simulations in best agreement with experiments
+and restricted\todo{rather: narrow?} angle distribution for PS than for PC headgroup in
+CHARMM36 simulations that have the best agreement with experiments
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
-Also, the sampled conformations of glycerol backbone
-significantly vary between different simulation models 
-(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
-but further analysis is beyond the scope of this work which focuses on the PS headgroup.
+The results might reflect the increased rigidity anticipated\todo{detected, predicted, speculated?} in the early experimental studies \cite{browning80,buldt81}.
+%Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
+\ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup.
+There is very little actual information in this statement.
+The variance was already noted above
 
-The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
-can guide further force field development. However, more accurate MD force fields are
-required to confirm the suggested conformations because the simulated PS headgroup order parameters
-are not within the experimental error in any of the tested models.
-
-%
-%The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
-%for the best performing CHARMM36 model is larger for PS lipids (8)
-%than for PC (3) \cite{botan15}.
+The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
+However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.
 

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
 \centering
-\caption{The list of MD simulations of pure PS bilayers without additional salt.
-  Simulation details are given in the supplementary information
+\caption{The list of MD simulations of pure PS bilayers without additional salt along with the references to the force fields used and the MD trajectories.
+Notation 2$\times$[time] indicates that more than two independent MD runs was conducted.
+Additional simulation details are given in the supplementary information.
     %   The lipid force fields named as in our previous work~\cite{botan15}. 
@@
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c r r r r r c c}
+\begin{tabular}{lcrrrrrcc}
 %\hline
 % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+lipid/counter-ions  & force field for lipids / ions  & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
 \hline 
-    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
-    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
-    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
-    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
-    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
-    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
-    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
-    POPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
-    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
+POPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36POPS298K}  & \tabularnewline
+POPS/K$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36POPS298Kpotassium}  & \tabularnewline
+POPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{macrogPOPS298Kcorrect}  & \tabularnewline
+POPS/K$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 200  & 150  & \cite{macrogPOPS298KwithK}  & \tabularnewline
+POPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
+POPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
+POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{slipidsPOPS298K}  & \tabularnewline
 \hline 
-    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+DOPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
+DOPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
+DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{slipidsDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 288  & 11232  & 303  & 200  & 100  & \cite{slipidsDOPSfiles}  & \tabularnewline
 \end{tabular}
@@
     POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
-    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
+    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
   }\todo{
@@
     glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
-    and MD simulations with different force fields (symbols).
+    and MD simulations with different force fields (dots).
     Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
     while all the other data are without additional salt.
-    The data for DOPS is at 303~K and the data for POPS is at 298~K.
+    All the data for DOPS at 303~K and all the data for POPS at 298~K.
     Light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
-    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
+    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
+    The vertical bars shown for all simulation values except MacRog~K$^+$
     are not error bars, but demonstrate that for these systems
@@
 available.
-%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results.
-This also enables referring to figures in order.
-I also removed the refs which do not seem to refer to the actual data used in the figure.
-%Rewrote the paragraph, see if ok.
-
-We note that the the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating the lipids to the buffer and then centrifuging the sample to a pellet that was used in the measurements.
-Such samples have a lower lipid concentration
-(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
-gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
-Larger multilamellar repeat distances are expected in the samples with lower lipid
-concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
-Yet the PS headgroup order parameters measured from gravimetric samples (POPS) in this work
-are in good agreement with the results from centrifuged samples~\cite{browning80}.
-This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the simulations and reference experiments.
 

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
   reported in the experiments by Roux et al.~\cite{roux90}.
+Notation 2$\times$[time] indicates that more than two independent MD runs was conducted.
 The simulation details are given in the supplementary information.
@@
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c c c c c c c c }
-  lipid/counter-ions & force field for lipids / ions & [CaCl$_2$]\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+\begin{tabular}{lccccccccc}
+lipid/counter-ions  & force field for lipids / ions  & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
 \hline 
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & 11207 & 0  & 298  & 200 & 180   & \cite{POPC5POPS1noCaClCHARMM}  \\
-    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
-    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26 & 250:50  & 11190  & 53  & 298  & 200 & 180 & \cite{POPC5POPS1withCaClCHARMM}   \\
-    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50  & 11174  & 214  & 298  & 200 & 180  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
-    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}        & 0    & 150:150 & 10785    & 0  & 298  & 200 & 180   & \cite{POPC1POPS1noCaClCHARMM}  \\
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \cite{POPC5POPS1noCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36pops+83popcT298Kpiggot} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \cite{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \cite{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \cite{POPC1POPS1noCaClCHARMM} \tabularnewline
 \hline 
-    POPC:POPS (1:0)        & MacRog \cite{maciejewski14} &0    & 120:0  & 5120 & 0    & 298  & 200 & 150 & \cite{macrogPOPC298K}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 0.30 & 120:24 & 5760 & 31   & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 1.00   & 120:24 & 5760 & 104  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \cite{macrogPOPC298K} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 \todo{lenght wrong?
+Check zenodo} & 250 \todo{lenght wrong?
+Check zenodo} &  \todo{this ref is to trajectory with CA}\cite{POPCpopsMACROG}\tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.30  & 120  & 5760  & 31  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 1.00  & 120 & 5760  & 104  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 3.00  & 120 & 5760  & 311  & 298  & 600 & 300 & \cite{POPCpopsMACROG}\tabularnewline
+\hline 
 %    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withKCI} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withNaCI} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 60 & 3600  & 0  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
 \hline 
-    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18}  & 0      & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
-    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} & 0      & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
-    POPC:POPS (5:1)         & Lipid14/17 \cite{dickson14,gould18} &0.50    & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
-    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &1.00    & 120:24 & 5760 & 104   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
-    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &2.00    & 120:24 & 5760 & 208   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
-    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &3.00    & 120:24 & 5760 & 311   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
-    POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}        &4.00    & 120:24 & 5760 & 415   & 298  & 300 & 200 & \cite{POPCpopsLIPID17withCaCl}  \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18}  & 0      & 60:12 & 3600  & 0   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.08   & 60:12 & 3561  & 5   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.13   & 60:12 & 3561  & 8   & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.20   & 60:12 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.41   & 60:12 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
-    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.62   & 60:12 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series}   \\
+POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102 & 4290  & 0  & 310  & 120  & 80  & \cite{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
+POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
+usage scaled ions. Concentration taken as reported in the delivered
+data.}  & 104  & 4306  & 24  & 310  & 300  & 100  & \cite{POPCpopsBERGERwith102mMCa} \tabularnewline
+POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104 & 4306  & 72  & 310  & 300  & 100  & \cite{POPCpopsBERGERwith715mMCa} \tabularnewline
 \hline 
-    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0   & 102:26 & 4290 & 0   & 310  & 120 & 80 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}         & 0.102\footnote{Calculation of concetration complicated due the scaled ions.
-Concentration taken as reported in the delivered data.}   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
-    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}         & 0.715\footnote{Calculation of concetration complicated due the scaled ions.
-Concentration taken as reported in the delivered data.}  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
-    \hline
-    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKP \cite{??}             & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPwithNa}  \\
-    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKPM \cite{??}            & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPMwithNa}  \\
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline
 \end{tabular}

### 2019-02-08 Manuscript/manuscriptPS.tex

--- 2019-02-08 Manuscript/manuscriptPS.tex
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
-     demonstrate the conformational fluctuations around
+    Overlayed snapshots from CHARMM36 simulations in best agreement with experiments
+    to demonstrate the conformational fluctuations around
     (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
@@
 
-The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
+The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
 and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
 as previously observed also for PC lipids \cite{botan15}.
-None of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
-which complicates the interpretation of structural differences between PC and PS headgroups.
+Therefore, the interpretation of structural differences between PC and PS headgroups
+from simulations is not straightforward.
 
@@
 
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC forking of the $\alpha$-carbon \todo{observed in the experiments?} and the Slipids force field additionally correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
-\ref{HGorderParametersPS} vs.
-Fig.
-2 in Ref.~\citenum{botan15}) \todo{refer to experiments also?}.
-
+However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
+the best performing models,
+Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
+Slipids reproduces also the feature that the $\beta$-carbon order parameter
+in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
 Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
 a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
-in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
-The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
+in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
+The restricted motion is also
+visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
 suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
 In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-and restricted\todo{rather: narrow?} angle distribution for PS than for PC headgroup in
-CHARMM36 simulations that have the best agreement with experiments
+and restricted angle distribution for PS than for PC headgroup in
+CHARMM36 simulations in best agreement with experiments
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-The results might reflect the increased rigidity anticipated\todo{detected, predicted, speculated?} in the early experimental studies \cite{browning80,buldt81}.
-%Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
-\ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup.
-There is very little actual information in this statement.
-The variance was already noted above
+The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
+Also, the sampled conformations of glycerol backbone
+significantly vary between different simulation models 
+(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
+but further analysis is beyond the scope of this work which focuses on the PS headgroup.
 
-The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
-However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.
+The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
+can guide further force field development. However, more accurate MD force fields are
+required to confirm the suggested conformations because the simulated PS headgroup order parameters
+are not within the experimental error in any of the tested models.
+
+%
+%The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
+%for the best performing CHARMM36 model is larger for PS lipids (8)
+%than for PC (3) \cite{botan15}.
 

### 2019-02-11 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
   \caption{\label{HGorderparametersPSvsPC}
-    Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
+    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
     Experimental values with the signs are measured for pure POPS system in this work.
@@
 Membranes containing PS lipids are always accompanied with counterions that
-modulate electrostatic interactions between lipids and other biomolecules. 
-Counterions are also suggested to screen the repulsion between charged lipid headgroups 
-in MD simulations and thus to reduce the area per lipid of PS bilayers to be smaller than in PC
-bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
+modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
+that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
+bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
 Counterion density profiles along membrane normal 
-indicate significant differences between force fields in both binding affinity
+show significant differences between force fields in both binding affinity
 and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
@@
 is reproduced only in Gromos-CKP and in the MacRog simulation
-with potassium counterions, while other models give significantly lower areas (Fig. \ref{NAdensPOPS}).
-The counterion binding and the concomitant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values,
-because the MacRog simulation, which has the strongest sodium binding
+with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
+However, the counterion binding and the concomitant electrostatic screening of the headgroup
+repulsion does not fully explain the low area per molecule values.
+since the MacRog simulation, which has the strongest sodium binding
 (the lowest concentrations in bulk water), gives the same area per molecule
 as the CHARMM36ua simulation, which has significantly weaker counterion binding
-affinity. On the other hand, changing counterions from sodium to potassium, having weaker binding
-affinity, increases the area per molecule from 53~\AA$^2$ to 63~\AA$^2$ in MacRog simulations.
+affinity.
+On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than with natrium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
 In conclusion, the results are in line with the previous study suggesting that the
@@
 of both counterion binding and hydrogen bonding network between lipid
-headgroups \cite{petrache04}.
+headgroups \cite{petrache04}. \todo{can we really say this based on our data? We don't discuss hydrogen bonding}
 
@@
   \caption{\label{changesWITHCaClPS}
-    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
     as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
@@
     The order parameter values from systems without calcium are set as the zero point of y-axis,
-    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
-    such that the lower order parameter is zero without additional ions to correctly illustrate
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
+    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
     the forking with different concentrations of calcium.
@@
 
-Binding of cations to zwitterionic PC lipid bilayers has been previously
-evaluated against experiments using the changes of headgroup order parameters
-as a function of salt concentration \cite{catte16}.
-Studying binding of cations to negatively charged lipid bilayers is less straighforward,
-because the cationic counterions are always present and the ion-free reference state does thus not exist.
-In addition, the analysis is complicated by the artificial aggregation of counterions
-in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
+The experimentally observed modulation of headgroup order parameters
+by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
+Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state.
+The analysis is further complicated by the artificial aggregation of counterions
+observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)\todo{Why isn't this a problem with PC}.
 Therefore, we evaluate here the amount of bound charge not by adding salt
-(although also this is discussed in the section \ref{mixtureTOadditionalCIs} in the supplementary information),
+(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
 but by studying the changes of the headgroup order parameters with increasing amount of
 negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
+
 According to the molecular electrometer concept, the headgroup order parameters of POPC
@@
 which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
-The CHARMM36 and Berger simulations predict no change, or a decrease,
-in the POPC headgroup order parameters as a function of increased amount of POPS (Fig. \ref{HGorderparametersPCvsPS}).
+The CHARMM36 and Berger simulations either exhibit no change or show a decrease
+in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
 This can be explained by the stronger counterion binding affinity, which cancels
 the effect of negatively charged headgroups and prevents the experimentally observed
-increase of headgroup order parameters with increasing amount of PS lipids.
+increase of headgroup order parameters with growing amount of PS lipids.
 Therefore, we suggest that the relatively weak binding of potassium
-in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
 realistic surface charge density in membranes containing PS lipids,
 while the other tested simulation models overestimate the counterion
-binding affinity. The results are in line with the changes of headgroup order
+binding affinity. The results are in line with the behaviour of headgroup order
 parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
-in the supplementary information.
+in the SI.
 
@@
 POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
-None of the tested models reproduce the changes of POPS headgroup order
+Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
 parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
-Therefore, we conclude that more accurate force fields are necessary
+and we conclude that more accurate force fields are needed
 to correctly describe the PC-PS headgroup interactions in MD simulations.
+%Having this conclusion in every paragraph gets a bit repetive
 

### 2019-02-11 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
   \caption{\label{HGorderparametersPSvsPC}
-    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
+    Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
     Experimental values with the signs are measured for pure POPS system in this work.
@@
 Membranes containing PS lipids are always accompanied with counterions that
-modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
-that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
-bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
+modulate electrostatic interactions between lipids and other biomolecules. 
+Counterions are also suggested to screen the repulsion between charged lipid headgroups 
+in MD simulations and thus to reduce the area per lipid of PS bilayers to be smaller than in PC
+bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
 Counterion density profiles along membrane normal 
-show significant differences between force fields in both binding affinity
+indicate significant differences between force fields in both binding affinity
 and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
@@
 is reproduced only in Gromos-CKP and in the MacRog simulation
-with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
-However, the counterion binding and the concomitant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values.
-since the MacRog simulation, which has the strongest sodium binding
+with potassium counterions, while other models give significantly lower areas (Fig. \ref{NAdensPOPS}).
+The counterion binding and the concomitant electrostatic screening of the headgroup
+repulsion does not fully explain the low area per molecule values,
+because the MacRog simulation, which has the strongest sodium binding
 (the lowest concentrations in bulk water), gives the same area per molecule
 as the CHARMM36ua simulation, which has significantly weaker counterion binding
-affinity.
-On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than with natrium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
+affinity. On the other hand, changing counterions from sodium to potassium, having weaker binding
+affinity, increases the area per molecule from 53~\AA$^2$ to 63~\AA$^2$ in MacRog simulations.
 In conclusion, the results are in line with the previous study suggesting that the
@@
 of both counterion binding and hydrogen bonding network between lipid
-headgroups \cite{petrache04}. \todo{can we really say this based on our data? We don't discuss hydrogen bonding}
+headgroups \cite{petrache04}.
 
@@
   \caption{\label{changesWITHCaClPS}
-   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
     as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
@@
     The order parameter values from systems without calcium are set as the zero point of y-axis,
-    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
-    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
+    such that the lower order parameter is zero without additional ions to correctly illustrate
     the forking with different concentrations of calcium.
@@
 
-The experimentally observed modulation of headgroup order parameters
-by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
-Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state.
-The analysis is further complicated by the artificial aggregation of counterions
-observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)\todo{Why isn't this a problem with PC}.
+Binding of cations to zwitterionic PC lipid bilayers has been previously
+evaluated against experiments using the changes of headgroup order parameters
+as a function of salt concentration \cite{catte16}.
+Studying binding of cations to negatively charged lipid bilayers is less straighforward,
+because the cationic counterions are always present and the ion-free reference state does thus not exist.
+In addition, the analysis is complicated by the artificial aggregation of counterions
+in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
 Therefore, we evaluate here the amount of bound charge not by adding salt
-(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
+(although also this is discussed in the section \ref{mixtureTOadditionalCIs} in the supplementary information),
 but by studying the changes of the headgroup order parameters with increasing amount of
 negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
-
 According to the molecular electrometer concept, the headgroup order parameters of POPC
@@
 which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
-The CHARMM36 and Berger simulations either exhibit no change or show a decrease
-in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
+The CHARMM36 and Berger simulations predict no change, or a decrease,
+in the POPC headgroup order parameters as a function of increased amount of POPS (Fig. \ref{HGorderparametersPCvsPS}).
 This can be explained by the stronger counterion binding affinity, which cancels
 the effect of negatively charged headgroups and prevents the experimentally observed
-increase of headgroup order parameters with growing amount of PS lipids.
+increase of headgroup order parameters with increasing amount of PS lipids.
 Therefore, we suggest that the relatively weak binding of potassium
-in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
 realistic surface charge density in membranes containing PS lipids,
 while the other tested simulation models overestimate the counterion
-binding affinity. The results are in line with the behaviour of headgroup order
+binding affinity. The results are in line with the changes of headgroup order
 parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
-in the SI.
+in the supplementary information.
 
@@
 POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
-Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
+None of the tested models reproduce the changes of POPS headgroup order
 parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
-and we conclude that more accurate force fields are needed
+Therefore, we conclude that more accurate force fields are necessary
 to correctly describe the PC-PS headgroup interactions in MD simulations.
-%Having this conclusion in every paragraph gets a bit repetive
 
@@
     The smallest simulated CaCl$_2$ concentrations are shown.
-    For the density profiles from all the simulated concentrations see figure \ref{CAdensPCPSmixtureALL}
-    in the supplementary information.
+    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
   }
@@
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
-where the dehydrated lipid-ion complexes and phase separation are 
-not observed~\cite{feigenson86,mattai89,roux90,roux91}.
+where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
@@
 experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
-in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
+in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
 indicating overestimated calcium binding affinity.
 The only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
+interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
 indicating weaker binding affinity than experiments.
@@
 Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
-calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
-data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
-is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
-in all the other tested models. This is evident in the calcium density distributions
-along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
-all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
+is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
+in all the other tested models.
+This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig.
+\ref{CAdensPCPSmixture}).
 
-The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
+Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
 exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
-below 100 mM (Fig. \ref{changesWITHCaClPS}).
-In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
+below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
 whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
 the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
-tested simulation models, including CHARMM36 with underestimated binding affinity.
-In addition, different simulation models predict qualitatively different behaviour
+tested simulation models, including CHARMM36 which underestimated the binding affinity.
+Importantly, the different simulation models predict qualitatively different behaviour
 for the POPS $\alpha$-carbon order parameters with added calcium.
@@
 structure without additional ions~\cite{catte16}. Therefore, we conclude that
-improvement of force fields is necessary to correctly describe interactions between the
+improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.
-
 \section{Conclusions}

### 2019-02-11 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
     The smallest simulated CaCl$_2$ concentrations are shown.
-    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
+    For the density profiles from all the simulated concentrations see figure \ref{CAdensPCPSmixtureALL}
+    in the supplementary information.
   }
@@
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
-where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
+where the dehydrated lipid-ion complexes and phase separation are 
+not observed~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
@@
 experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
-in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
+in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
 indicating overestimated calcium binding affinity.
 The only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
+interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
 indicating weaker binding affinity than experiments.
@@
 Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
-calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
-is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
-in all the other tested models.
-This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig.
-\ref{CAdensPCPSmixture}).
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
+data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
+is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
+in all the other tested models. This is evident in the calcium density distributions
+along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
+all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
 
-Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
+The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
 exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
-below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
+below 100 mM (Fig. \ref{changesWITHCaClPS}).
+In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
 whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
 the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
-tested simulation models, including CHARMM36 which underestimated the binding affinity.
-Importantly, the different simulation models predict qualitatively different behaviour
+tested simulation models, including CHARMM36 with underestimated binding affinity.
+In addition, different simulation models predict qualitatively different behaviour
 for the POPS $\alpha$-carbon order parameters with added calcium.
@@
 structure without additional ions~\cite{catte16}. Therefore, we conclude that
-improvement of force fields is necessary to correctly capture the interactions between the
+improvement of force fields is necessary to correctly describe interactions between the
 PS headgroup and calcium ions in MD simulations.
+
 \section{Conclusions}
@@
 play an important role in lipid-mediated signaling
-processes \cite{leventis10,yeung08}. Recent studies using MD
-simulations to interpret the various spectroscopic data
-give contradictory results for the calcium binding details to PS
+processes \cite{leventis10,yeung08}. Attemps to use MD
+simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
 headgroups \cite{melcrova16,valentine18,hallock18}.  
-Here, as was previously done for PC lipids \cite{botan15,catte16},
+Here (as was previously done for PC lipids \cite{botan15,catte16})
 we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
@@
 
-First, we complemented the available experimental data of PS
-lipid headgroup order parameters \cite{browning80,roux90} by measuring the signs of the order parameters.
+First, we complemented the available experimental data \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters.
 Comparison to these data revealed that none of the available force fields
 %tested using the NMRlipids open collaboration
-was accurate enough to reproduce the PS
-headgroup order parameters within the experimental accuracy. However,
+reproduce the PS headgroup order parameters within the experimental accuracy. However,
 the best models for the serine headgroup suggested a characteristic rigid conformation for its
-carboxyl group. Comparison to the previously
-measured headgroup order parameters from POPC:POPS (5:1) bilayers with different ion
+carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
 concentrations \cite{roux90} then showed that the tested MD force fields
-overestimate the cation binding affinity to the negatively charged bilayers
-containing PS lipids with two exceptions. 1) The apparently most realistic monovalent ion binding
-affinity to PS-containing lipid bilayers was observed in the MacRog
-simulations with potassium counterions. 2) The CHARMM36 force field with the recently introduced
+overestimate the cation binding affinity to these bilayers with two exceptions. 1) The MacRog simulation with potassium counterion appear to produce the most realistic monovalent ion binding
+affinity to PS-containing lipid bilayers. 2) The CHARMM36 force field with the recently introduced
 NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
-The experimentally measured trends of the PS headgroup order parameter response
-to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
-qualitatively reproduced in any of the tested force fields, which indicates that improvements in
-the MD force fields are necessary to study interactions between PS lipids and other biomolecules.
-This is different to the previous results with PC lipids,
+The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
+the MD force fields to study interactions between PS lipids and other biomolecules.
+This is in contrast to the previous results with PC lipids,
 where the experimentally measured headgroup order parameter responses to the bound charge
-were qualitative reproduced even though the headgroup structures themselves were
+were in qualitative agreement even though the headgroup structures themselves were
 incorrect and the cation binding affinities overestimated \cite{catte16}.

### 2019-02-18 Manuscript/manuscriptPS.tex

@@
 play an important role in lipid-mediated signaling
-processes \cite{leventis10,yeung08}. Attemps to use MD
-simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
+processes \cite{leventis10,yeung08}. Recent studies using MD
+simulations to interpret the various spectroscopic data
+give contradictory results for the calcium binding details to PS
 headgroups \cite{melcrova16,valentine18,hallock18}.  
-Here (as was previously done for PC lipids \cite{botan15,catte16})
+Here, as was previously done for PC lipids \cite{botan15,catte16},
 we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
@@
 
-First, we complemented the available experimental data \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters.
+First, we complemented the available experimental data of PS
+lipid headgroup order parameters \cite{browning80,roux90} by measuring the signs of the order parameters.
 Comparison to these data revealed that none of the available force fields
 %tested using the NMRlipids open collaboration
-reproduce the PS headgroup order parameters within the experimental accuracy. However,
+was accurate enough to reproduce the PS
+headgroup order parameters within the experimental accuracy. However,
 the best models for the serine headgroup suggested a characteristic rigid conformation for its
-carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
+carboxyl group. Comparison to the previously
+measured headgroup order parameters from POPC:POPS (5:1) bilayers with different ion
 concentrations \cite{roux90} then showed that the tested MD force fields
-overestimate the cation binding affinity to these bilayers with two exceptions. 1) The MacRog simulation with potassium counterion appear to produce the most realistic monovalent ion binding
-affinity to PS-containing lipid bilayers. 2) The CHARMM36 force field with the recently introduced
+overestimate the cation binding affinity to the negatively charged bilayers
+containing PS lipids with two exceptions. 1) The apparently most realistic monovalent ion binding
+affinity to PS-containing lipid bilayers was observed in the MacRog
+simulations with potassium counterions. 2) The CHARMM36 force field with the recently introduced
 NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
-The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
-qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
-the MD force fields to study interactions between PS lipids and other biomolecules.
-This is in contrast to the previous results with PC lipids,
+The experimentally measured trends of the PS headgroup order parameter response
+to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields, which indicates that improvements in
+the MD force fields are necessary to study interactions between PS lipids and other biomolecules.
+This is different to the previous results with PC lipids,
 where the experimentally measured headgroup order parameter responses to the bound charge
-were in qualitative agreement even though the headgroup structures themselves were
+were qualitative reproduced even though the headgroup structures themselves were
 incorrect and the cation binding affinities overestimated \cite{catte16}.

### 2019-02-18 Manuscript/manuscriptPS.tex
Update manuscript pdf

@@
 POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \cite{macrogPOPC298K} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 \todo{lenght wrong?
-Check zenodo} & 250 \todo{lenght wrong?
-Check zenodo} &  \todo{this ref is to trajectory with CA}\cite{POPCpopsMACROG}\tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 \todoi{lenght wrong?
+Check zenodo} & 250 \todoi{lenght wrong?
+Check zenodo} &  \todoi{this ref is to trajectory with CA}\cite{POPCpopsMACROG}\tabularnewline
 POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline

### 2019-02-18 Manuscript/manuscriptPSsuppl.tex

@@
 \todo{J. Melcr, please write simulation details} \\
+
+\subsection{Slipids}
+\todo{To be written by Piggot and Favela}
+{\it DOPC with 288 lipids.}
+The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions.
+The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
+Simulation was performed for 200~ns, and the last 100~ns were used for the analysis.
+Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}.
+Timestep of 2~fs was used with the leapfrog integrator.
+The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath.
+Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
+barostat~\cite{parrinello81} with a time constant of 10.0~ps.
+Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}.
+A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space.
+Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure.
+All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}.
+Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.
 

### 2019-02-18 Manuscript/manuscriptPSsuppl.tex
Update pdf

@@
 along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
-POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham15}, 
+POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
 out of which 150~ns were used for analyses. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.
@@
 \subsection{Slipids}
-\todo{To be written by Piggot and Favela}
 {\it DOPC with 288 lipids.}
@@
 Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.
+
+\todo{Other systems to be written by Piggot}
 

### 2019-03-05 Manuscript/manuscriptPS.tex
Updates based on comments by Tiago Ferreira

@@
 \author{Tiago M. Ferreira}
-\affiliation{Halle, Germany}
+\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
 \author{Ivan Gushchin}
@@
 because accurate models to interpret the experimental data have not
-been available. Here, we gather a set of experimental NMR data that 
-can, together with molecular dynamics (MD) simulations, be used
-to decipher the lipid headgroup structures and ion binding
-in pure PS and mixed (here PS:PC) lipid bilayers. Using the open collaboration
+been available. Here, we use solid-state NMR S-DROSS experiments to measure the signs of the previous set of experimental NMR
+order parameter magnitudes available for POPS to assess the phosphoserine structure in the different atomistic force-fields presently used.    
+We also gather a set of experimental NMR data that can be used to assess the ion binding to pure PS and mixed PC:PS lipid bilayers
+in molecular dynamics (MD) simulations.
+Using the open collaboration
 method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
@@
 Some protein domains interact specifically with PS lipids,
-while others are attracted by nonspecific electrostatics and the
+while other protein sites are attracted by nonspecific electrostatics and the
 binding can be regulated by calcium \cite{leventis10}.
@@
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-On the other hand, some studies propose that the specific binding
-affinity is similar for the negatively charged and zwitterionic lipids, and
+On the other hand, some studies suggest that the specific binding affinity of Ca$^{2+}$
+to negatively charged and zwiterionic phospholipids is similar, and that
 the increased cation binding to negatively charged lipid bilayers arises only due
@@
 
-Here, we collect a set of experimentally measured lipid headgroup and
-glycerol backbone C--H bond order parameters, which can be used to
-evaluate the quality of headgroup structure and ion binding affinity
-in MD simulations of lipid bilayers containing PS lipids. 
-The available MD simulation models of PS are then compared with
-the collected experimental data.
+Here, we extend the available set of experimentally measured PS lipid headgroup and
+glycerol backbone C--H bond order parameters
+by measuring the signs of the order parameters with S-DROSS solid-state NMR spectroscopy.
+The quality of headgroup structures and ion binding affinity in
+%MD simulations of lipid bilayers containing PS lipids
+the available MD simulation models of PS are then compared with the collected experimental data.
 %using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
@@
 
-\subsection{C--H bond order parameters from the natural abundance $^{13}$C\,NMR}
+\subsection{Experimental C--H bond order parameters}% from the natural abundance $^{13}$C\,NMR}
 
-Headgroup and glycerol backbone C--H bond order parameters of POPS
-were determined from the chemical-shift resolved dipolar splittings
-measured with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
+The headgroup and glycerol backbone C--H bond order parameter magnitudes of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
+were determined by measuring the chemical-shift resolved dipolar splittings
+with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
 The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
@@
 for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
-Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
+The used radio frequency pulses had the following nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
 50 kHz (SPINAL64 $^1$H decoupling pulses).
-\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
 The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
-dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?} of 149.5 ppm.
+dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.
 
@@
 experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
-blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. Radio frequency pulses with the nutation
+blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. The used radio frequency pulses had the nutation
 frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
-\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
 The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
-measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?}  of 149.5 ppm.
+measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.
 
 \emph{Numerical simulations of S-DROSS curves.}
-The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00}
-using as input the $^{13}$C--$^1$H
+The numerical simulations of S-DROSS curves were performed using the SIMPSON simulation package \cite{bak00}
+by inputing the $^{13}$C--$^1$H
 dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
@@
 \emph{Sample preparation.}
-The sample was prepared simply by mixing the POPS \todo{From some company?} with water (lipid:water 60:40\,wt-\%) in an Eppendorf
-tube by mixing and centrifuging the sample repeatedly until a \todo{visually(?)} homogeneous viscous fluid was obtained.
+The sample was prepared simply by mixing POPS (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
+as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample repeatedly until a homogeneous viscous fluid was obtained
+\todo{How long and how many repeats did this take approximately? Was the homogeneity determined visually?}.
 Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.
-\todo{Maybe we need little bit more information about the mixing procedure?}
 
@@
 where $\theta$ is the angle between the C--H bond and the membrane normal
-(taken to align with $z$, with bilayer periodic in the $xy$-plane).
+(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
 Angular brackets denote average over all sampled configurations.
@@
     (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
-    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the $\alpha$-carbon splittings.
+    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
+    for the $\alpha$-carbon.
     Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
-    for the smaller $\alpha$-carbon order parameter.
+    for the smaller $\alpha$-carbon C--H bond order parameter.
   }
   \todo{I think that the peak labeling would be good to show also in (A).}\\
-  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
-  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
+%  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
+%  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
 \end{figure}
@@
 are shown in the supplementary information (Fig. \ref{DPslices}).
+
 Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
@@
 order parameters using the S-DROSS experiment \cite{gross97}.
-The S-DROSS slice clearly shows that the order parameter of
-the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
-Could we explain in a few words how this is clearly seen?}
-which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
-suggests that the larger order parameter of the $\alpha$-carbon 
-is positive and the deviation towards negative values with longer T$_1$ times suggests
-that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
+For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
+of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
+We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C—H
+bond order parameter give rise to profiles that initially increase and decrease, respectively.
+In practice, it is useful to use the known negative sign of the acyl chain carbons as a reference to perform
+the phase correction and interpret the distinct initial slopes of the S-DROSS profiles as shown in figure 1. 
+The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
+For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C—H bonds.
+However, the beginning of its S-DROSS slice suggests that the larger order parameter of the alpha-carbon is positive and the
+decrease towards negative values at longer $t_1$ suggests that the smaller order parameter is negative.    
+%
+%The S-DROSS slice clearly shows that the order parameter of
+%the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
+%Could we explain in a few words how this is clearly seen?}
+%which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
+%suggests that the larger order parameter of the $\alpha$-carbon 
+%is positive and the deviation towards negative values with longer T$_1$ times suggests
+%that the smaller order parameter is negative.
+%
+This is confirmed by a SIMPSON simulation
 using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
 The literature value was used because the
-resolution of our experiment was not sufficient to determine the
+resolution of our R-PDLF experiment was not sufficient to determine the
 small value of the order parameter.
@@
 (dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
-confirming \todo{Or rather 'corroborating'?}
-the interpretation that the smaller order parameter is negative.
+corroborating the interpretation that the smaller order parameter is negative.
 
@@
 
-We note that the the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating the lipids to the buffer and then centrifuging the sample to a pellet that was used in the measurements.
-Such samples have a lower lipid concentration
+We note that the the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
+the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
 (approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
@@
     dihedrals of PS and PC headgroups.
-    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
+    The publicly available trajectories via Zenodo were used for CHARMM36 POPC and Slipids POPC
+    (Refs. \citenum{POPCcharmm36T303K} and \citenum{POPCslipids298K}, respectively).
   }
+  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
 \end{figure}
@@
 and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
-as previously observed also for PC lipids \cite{botan15}.
-None of the models produces a set of order parameters in full agreement with the experiments.
+as previously observed also for PC lipids \cite{botan15}. None of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
@@
   \caption{\label{NAdensPOPS}
-    Counterion densities of POPS lipid bilayer along the membrane normal from
-    simulations with different force fields.
+    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
+    of POPS lipid bilayer from simulations with different force fields.
+    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
   }
@@
 
-Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
-a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
-in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
+%Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
+%a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
+%in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
+Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
+dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
 The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
@@
 and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
-The experimental area per lipid (62.7~\AA$^2$) \cite{pan14} 
+The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
 is reproduced only in Gromos-CKP and in the MacRog simulation
@@
 
-Calcium binding affinity to membranes containing the negative charged PS lipids can be
+Calcium binding affinity to membranes containing the negatively charged PS lipids can be
 experimentally measured by detecting the PC lipid headgroup order parameters

### 2019-03-06 Manuscript/manuscriptPS.tex
Updates related to issues https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/42 , https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/43 and https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34

@@
 \caption{The list of MD simulations of pure PS bilayers without additional salt along with the references to the force fields used and the MD trajectories.
-Notation 2$\times$[time] indicates that more than two independent MD runs was conducted.
-Additional simulation details are given in the supplementary information.
+  Notation 2$\times$[time] indicates that two independent MD runs was conducted. Additional simulation details are given in the supplementary information.
     %   The lipid force fields named as in our previous work~\cite{botan15}. 
@@
   This corresponds the concentration in buffer before solvating lipids, which are
-  reported in the experiments by Roux et al.~\cite{roux90}.
-Notation 2$\times$[time] indicates that more than two independent MD runs was conducted.
+  reported in the experiments by Roux et al.~\cite{roux90}.  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
   The simulation details are given in the supplementary information. 
@@
 \begin{tabular}{lccccccccc}
-lipid/counter-ions  & force field for lipids / ions  & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
+lipid/counter-ions  & force field for lipids / ions  & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
 \hline 
@@
 POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \cite{macrogPOPC298K} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 \todoi{lenght wrong?
-Check zenodo} & 250 \todoi{lenght wrong?
-Check zenodo} &  \todoi{this ref is to trajectory with CA}\cite{POPCpopsMACROG}\tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 & 250 &  \cite{POPCpopsMACROG}\tabularnewline
 POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
@@
     are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
-  }\todo{
-  1) Use diamonds for DOPS, spheres for POPS.
-  2) Error bars?
-  3) Change the y-label to $S_\mathrm{CH}$, and invert the y-axis, as in Fig. 3.
-  4) Make lower g1 POPC visible, e.g., by slightly larger point.
+    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
   }
+  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
+  We can do this just before submission when citation numbers will not change anymore.}
 \end{figure}

### 2019-03-07 Manuscript/manuscriptPS.tex
Update Fig. 4 based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48
Add error bars in figures 7 and 8 as requested in issues https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/45 and https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/46

@@
     of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of PS and PC headgroups.
-    The publicly available trajectories via Zenodo were used for CHARMM36 POPC and Slipids POPC
-    (Refs. \citenum{POPCcharmm36T303K} and \citenum{POPCslipids298K}, respectively).
+    dihedrals of PS and PC headgroups. The publicly available trajectory was used for CHARMM36 POPC~\cite{POPCcharmm36T303K}.
   }
-  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
+%  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
 \end{figure}
@@
   \todo{Commented by M. Javanainen in blog:
-    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready}
+    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
+  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
 \end{figure}
@@
     Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
+    The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
+    the error bar is given as the difference between the results divided by two.
   }
@@
     The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
-%    Since the experimental data of POPS in pure and diluted mixture come from
-%    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
-%    the experimental change of the order parameter is less accurate than in typical measurements
-%    where same technique is used in all conditions, see discussion about qualitative and quantitative 
-%    accuracy in Ref. \citenum{ollila16}.
+    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
+    the error bars of 0.02 are used here \cite{botan15,ollila16}.
     The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
@@
     parameter value from pure POPS is at zero to correctly illustrate the significant forking.
+    The values from CHARMM36 and Gromos simulations are averages of two independent simulations and
+    the error bars are given as the difference between the results divided by two.
   }
@@
   \todo{Should we include also counterions into the plot?} \\
+  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
 \end{figure}

### 2019-03-07 Manuscript/manuscriptPS.tex
Updates in manuscript based on todo points

@@
 the same force field without the NBfix parameters, combined with NMR chemical shifts and
-REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
+rotational-echo double-resonance (REDOR) experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
 On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
@@
 %
-Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
-\todo{Should we leave the mention of ECC out from the Introduction
-(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}
+%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
+%\todo{Should we leave the mention of ECC out from the Introduction
+%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}
 
@@
 MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
-The ion \todo{There are ion number densities, right?} number density profiles were calculated using the {\tt gmx density} tool
+The ion number density profiles were calculated using the {\tt gmx density} tool
 of the Gromacs sofware package \cite{gromacsMANUAL}.
@@
   }
-  \todo{I think that the peak labeling would be good to show also in (A).}\\
-%  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
-%  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
+  \todo{I think that the peak labeling would be good to show also in (A).
+  There is possibly a new version of (C) coming.}\\
 \end{figure}
@@
 
-\begin{figure}[]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
-  \caption{\label{HGorderParametersPS}
-    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
-    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
-    and MD simulations with different force fields (symbols).
-    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
-    while all the other data are without additional salt.
-    The data for DOPS is at 303~K and the data for POPS is at 298~K.
-    Light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
-    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
-    are not error bars, but demonstrate that for these systems
-    we had at least two data sets; the ends of the bars mark the extreme values
-    from the sets, and the dot marks their measurement-time-weighted average.
-  }
-\end{figure}
-
 The headgroup and glycerol backbone order parameters of 
@@
 \subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
+
+\begin{figure}[]
+  \centering
+  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
+  \caption{\label{HGorderParametersPS}
+    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
+    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
+    and MD simulations with different force fields (symbols).
+    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
+    while all the other data are without additional salt.
+    The data for DOPS is at 303~K and the data for POPS is at 298~K.
+    Light blue areas span 0.04 units around the average of the extremal experimental values,
+    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
+    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
+    are not error bars, but demonstrate that for these systems
+    we had at least two data sets; the ends of the bars mark the extreme values
+    from the sets, and the dot marks their measurement-time-weighted average.
+  }
+\end{figure}
 
@@
 
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC forking of the $\alpha$-carbon \todo{observed in the experiments?} and the Slipids force field additionally correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
-\ref{HGorderParametersPS} vs.
-Fig.
-2 in Ref.~\citenum{botan15}) \todo{refer to experiments also?}.
+However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
+smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
+also observed in the experiments (Fig. \ref{HGorderParameters}).
 
-%Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
-%a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
-%in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
 Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
@@
 In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-and restricted\todo{rather: narrow?} angle distribution for PS than for PC headgroup in
+and narrow angle distribution for PS than for PC headgroup in
 CHARMM36 simulations that have the best agreement with experiments
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-The results might reflect the increased rigidity anticipated\todo{detected, predicted, speculated?} in the early experimental studies \cite{browning80,buldt81}.
-%Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
+The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
+% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
 \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup.
@@
 as the CHARMM36ua simulation, which has significantly weaker counterion binding
-affinity.
-On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than with natrium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
+affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
+with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
 In conclusion, the results are in line with the previous study suggesting that the
 low areas per molecule in PS lipid bilayers originate from the combination
-of both counterion binding and hydrogen bonding network between lipid
-headgroups \cite{petrache04}. \todo{can we really say this based on our data? We don't discuss hydrogen bonding}
+of both counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
 
@@
 The analysis is further complicated by the artificial aggregation of counterions
-observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)\todo{Why isn't this a problem with PC}.
+observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
+% \todo{Why isn't this a problem with PC}.
+% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
+% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
+% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
 Therefore, we evaluate here the amount of bound charge not by adding salt

### 2019-03-07 Manuscript/manuscriptPSsuppl.tex
Updates in manuscript based on todo points

@@
 \subsection{Slipids}
-{\it DOPC with 288 lipids.}
+
+\noindent
+{\it DOPS with 288 lipids.}
 The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions.
@@
 \todo{Other systems to be written by Piggot}
+
+%{\it DOPS and POPS bilayers.} \\
+%The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.
 
@@
 \todo{T. Piggot, please write simulation details} \\
-
-\subsection{Slipids}
-
-\noindent
-{\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.
-\todo{T. Piggot and F. Favela, please write simulation details} \\
 

### 2019-03-20 Manuscript/manuscriptPS.tex
Update figures and text related to the experimental data based on emails by Tiago Ferreira

@@
 dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
-The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}}\todo{citation needed?} was used to simulate a random
+The chemical shift anisotropy and homonuclear couplings were neglected, and the SIMPSON input file {\it{rep2000}} was used to simulate a random
 distribution of bilayer orientations in the samples studied.
@@
 \emph{Sample preparation.}
-The sample was prepared simply by mixing POPS (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
-as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample repeatedly until a homogeneous viscous fluid was obtained
-\todo{How long and how many repeats did this take approximately? Was the homogeneity determined visually?}.
-Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.
+The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
+as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
+until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.
 
@@
   }
-  \todo{I think that the peak labeling would be good to show also in (A).
-  There is possibly a new version of (C) coming.}\\
 \end{figure}
@@
 bond order parameter give rise to profiles that initially increase and decrease, respectively.
-In practice, it is useful to use the known negative sign of the acyl chain carbons as a reference to perform
-the phase correction and interpret the distinct initial slopes of the S-DROSS profiles as shown in figure 1. 
+In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
+the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
 The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
@@
 This is confirmed by a SIMPSON simulation
-using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
-The literature value was used because the
+using the order parameter values of +0.09 from dipolar coupling measured here (Fig. \ref{DPslices})
+and -0.02 from the $^2$H\,NMR experiment~\cite{roux91}.
+The literature value for the smaller order parameter was used because the
 resolution of our R-PDLF experiment was not sufficient to determine the

### 2019-03-20 Manuscript/manuscriptPSsuppl.tex
Update figures by Pavel Buslaev http://nmrlipids.blogspot.com/2018/12/nmrlipids-iv-toward-submission-of.html?showComment=1552649629798#c3793703148326511254
Update figures and text related to the experimental data based on emails by Tiago Ferreira

@@
 \pagebreak
-\section{Dipolar slices of the R-PDFL experiment}
+\section{Spectral slices in the indirect dimension from the R-PDLF and SDROSS experiments}
 
-Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
-show a single splitting for the $\beta$-carbon with the order parameter value of 0.11,
-and a superposition of a large and a very small splitting for the $\alpha$-carbon.
-The larger splitting from the $\alpha$-carbon gives a order parameter value of 0.09, while the numerical value
-from the small splitting cannot resolved with the available resolution.
+The C--H bond order parameters of the headgroup and glycerol backbone carbons are determined 
+as $S_{\rm CH}=\Delta\nu/(0.315\times21.5  {\rm kHz})$, where $\Delta\nu$ is the dipolar splitting
+given by the largest peak widths observed in the second dimension of the R-PDLF spectra
+(blue lines in Fig. \ref{DPslices}), 0.315 is the scaling factor of the R18 recoupling sequence 
+and 21.5 kHz is the maximum $^1$H-$^{13}$C dipolar coupling for a C--H bond \cite{dvinskikh04}.
+%Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
+%show a single splitting for the $\beta$-carbon with the order parameter value of 0.11,
+%and a superposition of a large and a very small splitting for the $\alpha$-carbon.
+%The larger splitting from the $\alpha$-carbon gives a order parameter value of 0.09, while the numerical value
+%from the small splitting cannot resolved with the available resolution.
+The resulting order parameters are in good agreement with the previously reported values
+from $^2$H NMR experiments \cite{browning80} (Fig. \ref{HGorderParameters} in the main text).
+However, the resolution in our $^{13}$C NMR experiment was not sufficient to detect the
+the splitting related to the smaller order parameter of the C--H bond in the $\alpha$ carbon
+observed in $^2$H NMR experiments \cite{browning80}. Therefore, the value of  0 $\pm$ 0.02 from our
+$^{13}$C NMR experiments is shown figure \ref{HGorderParameters} 
+and the magnitude of 0.02 from the literature is used in the SIMPSON calculations in the main text.
+
+Interpretation of the order parameter signs of the $\alpha$ carbon from the SDROSS experiment
+is complicated by the presence of distinct order parameters for the two attached hydrogens.
+As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
+simulations gave values +0.09 and -0.02 for the $\alpha$-carbon order parameters.
+To corroborate our interpretation, we measured the SDROSS curve also using the higher 8 kHz
+MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
+Also this experiment indicates a postive value for the larger $\alpha$-carbon order parameter.
 \begin{figure}[]
 %  \centering
-  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
+  \includegraphics[width=\textwidth]{../Figs/SI_man.pdf}
   \caption{\label{DPslices}
-    Dipolar coupling slices from the R-PDFL experiment.
+    Spectral slices in indirect dimension from the R-PDLF (top) and S-DROSS (middle and bottom) experiments at distinct chemical shifts.
+    The chemical shifts and order parameters (calculated from the dipolar splittings indicated with blue lines in the R-PDLF slices)
+    are shown on top each column (chemical shift left, order parameter right).
+    The assignment of columns is given in the middle.  
+    The SDROSS slices were measured using both 5 KHz (middle) and 8 kHz (bottom) because different MAS frequencies enable the
+    higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
+    The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
+    are used to determine the error bars for $\alpha$ and $\beta$ carbons in figure \ref{PShgSIGNSsimpson} in the main text.
+    The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
   }
-  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
-    The splitting corresponding the larger $\alpha$-carbon order parameter should be also shown.
-  Also, the size of the figure in the file may not be optimal.}
 \end{figure}
@@
   \centering
-  \includegraphics[width=16.0cm]{../Figs/figS7PC.png}
+  \includegraphics[width=16.0cm]{../Figs/figS11.png}
   \caption{\label{dihedralsHGpc}
@@
   }
-  \todo{Colours with more contrast would be better. The y-axis scale could be zoomed.}
 \end{figure*}
@@
   \centering
-  \includegraphics[width=15.0cm]{../Figs/figS6PC.png}
+  \includegraphics[width=15.0cm]{../Figs/figS10.png}
   \caption{\label{dihedralsGLYpc}

### 2019-04-12 Manuscript/manuscriptPS.tex
Correct order of authors
Add mentioning about ring like structures observed in Berger headgroup, pointed by T. Piggot in the blog: http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1503251046734#c6788124049222929247

@@
 \affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
-\author{Markus Miettinen}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Josef Melcr}
@@
 Prague 6, Czech Republic}
+\author{Markus Miettinen}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Ricky Nencini}
@@
 in all simulation models, despite significant discrepancies in the headgroup
-structure without additional ions~\cite{catte16}. Therefore, we conclude that
-improvement of force fields is necessary to correctly capture the interactions between the
+structure without additional ions~\cite{catte16}.
+The divergent response of the Berger model may arise from the ring like structures
+observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
+%These ring like structures are a widespread feature of typical Berger based lipid force
+%fields containing explicit hydrogen atoms in the head group
+%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
+We conclude that improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the abstract

--- 2019-03-20 Manuscript/manuscriptPS.tex
@@
 Phosphatidylserine (PS) is a negatively
-charged lipid commonly found in eukaryotic membranes, where it interacts with signaling and other proteins via
+charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
 electrostatic interactions and direct binding, and can induce
@@
 because accurate models to interpret the experimental data have not
-been available. Here, we use solid-state NMR S-DROSS experiments to measure the signs of the previous set of experimental NMR
-order parameter magnitudes available for POPS to assess the phosphoserine structure in the different atomistic force-fields presently used.    
-We also gather a set of experimental NMR data that can be used to assess the ion binding to pure PS and mixed PC:PS lipid bilayers
-in molecular dynamics (MD) simulations.
-Using the open collaboration
-method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
-%or: Using the open collaboration method, we collect a database of MD simulations using wide range of available PS models (force fields) and perform a comparison to NMR results
+been available. Here, we use solid-state NMR S-DROSS to measure the signs of the experimental POPS headgroup
+order parameters, and gather a set of experimental NMR results from pure PS and mixed PC:PS lipid bilayers. Using the open collaboration
+method, we then extract data from a wide range of available PS MD models (force fields) and asses the PS headgroup structures and the ion binding produced by these force fields by comparing to the NMR data.
 We find that none of the models reproduce the NMR data within experimental accuracy,
@@
 does not rotate freely. In line with the previous results for PC lipids,
-none of the PS force fields correctly captures the cation binding affinity.
-Furthermore, the response of PS headgroups to bound ions
-\emph{qualitatively} differs from experiments. The collected experimental dataset and simulation
+none of the PS force fields correctly captures the cation binding affinity---the response of PS headgroups to bound ions can even
+\emph{qualitatively} differ from experiments. The collected experimental dataset and simulation
 results pave the way for improvement of lipid force fields to correctly

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the intro

@@
 Phosphatidylserine (PS) is the most common negatively
-charged lipid in eukaryotic membranes.
-PS lipids compose 8.5\% of total lipid weight of red blood cells. 
+charged lipid in eukaryotic membranes. In red blood cells, for example,
+PS lipids compose 8.5\% of the total lipid weight. 
 The abundance, however, varies between different organelles, and up to
 25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
-PS lipids are vastly important signaling molecules. They interact with
+PS lipids are vastly important biomolecules that interact with
 signaling proteins \cite{leventis10}, regulate
@@
 Some protein domains interact specifically with PS lipids,
-while other protein sites are attracted by nonspecific electrostatics and the
+while other protein sites attract PS lipids by nonspecific electrostatics and the
 binding can be regulated by calcium \cite{leventis10}.
-Therefore, the structural details
+Therefore, the deciphering structural details
 of lipid headgroups and the details of cation binding
-are crucial for the PS-mediated signaling processes.
+are crucial for understanding the PS-mediated processes in the cell membranes.
 
@@
 PS headgroup is more rigid than the phosphatidylcholine (PC)
-due to hydrogen bonding network or
-electrostatic interactions \cite{browning80,buldt81}.
-While monovalent ions interact weakly with
+owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
+While most monovalent ions interact weakly with
 PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
 dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
-The dehydrated complexes of PS headgroup and calcium ions can even lead to
+The dehydrated complexes of PS headgroup with calcium ions can even lead to
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-On the other hand, some studies suggest that the specific binding affinity of Ca$^{2+}$
-to negatively charged and zwiterionic phospholipids is similar, and that
+Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
+That said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
+to negatively charged and zwiterionic phospholipids, and that
 the increased cation binding to negatively charged lipid bilayers arises only due
 to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.
-Dilution of bilayers with PC lipids makes PS headgroups
-less rigid and reduces their propensity to form
-strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.
 
@@
 however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
-to understand the PS headgroup structure, their influence on lipid bilayer properties, and their
+to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
 interaction with
 ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
-Unfortunately, the results have depended strongly on the force field used.
+Unfortunately, the results have depended on the force field used.
 For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
@@
 suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
-the same force field without the NBfix parameters, combined with NMR chemical shifts and
-rotational-echo double-resonance (REDOR) experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
-On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
-combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
+the results from same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
+rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
+Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
+combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
 calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
 
-We have recently demonstrated that such controversies can be resolved by
-comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
-and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
-measured from NMR experiments with high accuracy and compared to simulations
-in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
-Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
+We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$~\cite{botan15,catte16}, can be utilized to resolve such controversies. The $S_\mathrm{CH}$ can be
+measured from NMR experiments with high accuracy and directly compared to simulations
+in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
+it established that the structure of PC lipid headgroup and glycerol backbone are not well
 captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
@@
 glycerol backbone C--H bond order parameters
-by measuring the signs of the order parameters with S-DROSS solid-state NMR spectroscopy.
-The quality of headgroup structures and ion binding affinity in
+by measuring the signs of the order parameters using S-DROSS solid-state NMR spectroscopy.
+The quality of headgroup structures and the ion binding affinity in
 %MD simulations of lipid bilayers containing PS lipids
-the available MD simulation models of PS are then compared with the collected experimental data.
+the available MD simulation models of PS lipids are then asessed based on the collected experimental data.
 %using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
@@
 biological function of lipid headgroups and glycerol backbone, as
-these are knwon to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.
+these are known to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.
 

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
edits to methods (simulation)

@@
 Angular brackets denote average over all sampled configurations.
-The order parameters were first calculated averaging over time separately
-for each lipid in the system. The average and
-the standard error of the mean were then calculated over different lipids.
-Python program ({\tt scripts/calcOrderParameters.py}) that uses the
-MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
+The order parameters were calculated by first averaging over time separately
+for each lipid in the system, and then calculating the average and
+the standard error of the mean over the different lipids. The analysis was conducted using a
+Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
+MDAnalysis library \cite{agrawal11,gowers16}. 
 The ion number density profiles were calculated using the {\tt gmx density} tool
@@
 concentrations can be easily calculated from
-simulations and compared to experimental data~\cite{catte16}. The headgroup order parameters
+simulations~\cite{catte16}. The headgroup order parameters
 of negatively charged PS and PG lipids also exhibit systematic, but less
@@
 Therefore, measuring the PC headgroup order parameters from 
-mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see SI section \ref{electrometerFORmixtures} provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
-% Removed a paragraph here because it was out of place in the molecular electrometer section and contained some results. I  rewrote it and added to the end of next section
+mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
 
-Important preliminary step for using the molecular electrometer concept is to calibrate the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18}.
+Calibrating the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer concept.
 This can be done using experimental data from mixtures of
@@
 Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
-The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
+the concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
 or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the 2 first sections of results

--- 2019-04-13 Manuscript/manuscriptPS.tex
@@
 Phosphatidylserine (PS) is a negatively
-charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
+charged lipid commonly found in eukaryotic membranes, where it interacts with signaling and other proteins via
 electrostatic interactions and direct binding, and can induce
@@
 because accurate models to interpret the experimental data have not
-been available. Here, we use solid-state NMR S-DROSS to measure the signs of the experimental POPS headgroup
-order parameters, and gather a set of experimental NMR results from pure PS and mixed PC:PS lipid bilayers. Using the open collaboration
-method, we then extract data from a wide range of available PS MD models (force fields) and asses the PS headgroup structures and the ion binding produced by these force fields by comparing to the NMR data.
+been available. Here, we use solid-state NMR S-DROSS experiments to measure the signs of the previous set of experimental NMR
+order parameter magnitudes available for POPS to assess the phosphoserine structure in the different atomistic force-fields presently used.    
+We also gather a set of experimental NMR data that can be used to assess the ion binding to pure PS and mixed PC:PS lipid bilayers
+in molecular dynamics (MD) simulations.
+Using the open collaboration
+method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
+%or: Using the open collaboration method, we collect a database of MD simulations using wide range of available PS models (force fields) and perform a comparison to NMR results
 We find that none of the models reproduce the NMR data within experimental accuracy,
@@
 does not rotate freely. In line with the previous results for PC lipids,
-none of the PS force fields correctly captures the cation binding affinity---the response of PS headgroups to bound ions can even
-\emph{qualitatively} differ from experiments. The collected experimental dataset and simulation
+none of the PS force fields correctly captures the cation binding affinity.
+Furthermore, the response of PS headgroups to bound ions
+\emph{qualitatively} differs from experiments. The collected experimental dataset and simulation
 results pave the way for improvement of lipid force fields to correctly
@@
 carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
-The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
-The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13} whereas
+the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
 known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
@@
 the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
+experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
     we had at least two data sets; the ends of the bars mark the extreme values
-    from the sets, and the dot marks their measurement-time-weighted average.
+    from the sets, and the symbol marks their measurement-time-weighted average.
   }
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
-     demonstrate the conformational fluctuations around
+    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
     (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
     of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of PS and PC headgroups. The publicly available trajectory was used for CHARMM36 POPC~\cite{POPCcharmm36T303K}.
+    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
   }
-%  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
+
 \end{figure}
@@
   \caption{\label{comparisonTablePS}
-    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
-    Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity.
-Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red).
+    Rough, subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
+    Here M indicates a magnitude problem, F a forking problem; letter size increases with problem severity.
+Color scheme: within experimental error (dark green), almost within experimental errorÓ (light green), Òclear deviation from experiments (light red), and major deviation from experiments (dark red).
 The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
-For full details of the assessment, see Supplementary Information.
+For full details of the assessment, see SI section~\ref{Ranking}.
   }
@@
 
-The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
-as previously observed also for PC lipids \cite{botan15}. None of the models produces a set of order parameters in full agreement with the experiments.
+The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
+as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
@@
 which complicates the interpretation of structural differences between PC and PS headgroups.
+However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
+smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
+also observed in the experiments (Fig. \ref{HGorderParameters}).
 
@@
 
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
-smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
-also observed in the experiments (Fig. \ref{HGorderParameters}).
-
 Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
@@
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
 % Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
@@
 
-The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
+These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
 However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the counterion binding section

@@
 carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
-The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13} whereas
-the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
+The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
 known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
@@
 the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
+experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
     we had at least two data sets; the ends of the bars mark the extreme values
-    from the sets, and the symbol marks their measurement-time-weighted average.
+    from the sets, and the dot marks their measurement-time-weighted average.
   }
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
+    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
+     demonstrate the conformational fluctuations around
     (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
     of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
+    dihedrals of PS and PC headgroups. The publicly available trajectory was used for CHARMM36 POPC~\cite{POPCcharmm36T303K}.
   }
-
+%  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
 \end{figure}
@@
   \caption{\label{comparisonTablePS}
-    Rough, subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
-    Here M indicates a magnitude problem, F a forking problem; letter size increases with problem severity.
-Color scheme: within experimental error (dark green), almost within experimental errorÓ (light green), Òclear deviation from experiments (light red), and major deviation from experiments (dark red).
+    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
+    Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity.
+Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red).
 The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
-For full details of the assessment, see SI section~\ref{Ranking}.
+For full details of the assessment, see Supplementary Information.
   }
@@
 
-The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
-as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
+The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
+as previously observed also for PC lipids \cite{botan15}. None of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
@@
 which complicates the interpretation of structural differences between PC and PS headgroups.
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
-smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
-also observed in the experiments (Fig. \ref{HGorderParameters}).
 
@@
 
+However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
+smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
+also observed in the experiments (Fig. \ref{HGorderParameters}).
+
 Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
@@
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
+The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
 % Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
@@
 
-These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
+The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
 However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.
@@
     the error bars of 0.02 are used here \cite{botan15,ollila16}.
-    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
+    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
     with the same value for both order parameters such that the lower order
@@
 modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
-that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
+that counterions reduce the area per lipid of PS bilayers compared to PC
 bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
-Counterion density profiles along membrane normal 
-show significant differences between force fields in both binding affinity
-and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
+We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
+The force fields studied show significant differences in both binding affinity
+and distribution of ions at the interface.
 The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
 is reproduced only in Gromos-CKP and in the MacRog simulation
-with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
+with potassium counterions, while other models give considerably smaller values.
 However, the counterion binding and the concomitant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values.
+repulsion does not fully explain the low area per molecule values
 since the MacRog simulation, which has the strongest sodium binding
@@
 with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
-In conclusion, the results are in line with the previous study suggesting that the
+The results are in line with the previous study suggesting that the
 low areas per molecule in PS lipid bilayers originate from the combination
-of both counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
+of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
 
@@
 by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
-Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state.
-The analysis is further complicated by the artificial aggregation of counterions
+Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions (the lack of a ion-free reference state) and the analysis is further complicated by the artificial aggregation of counterions
 observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
@@
 Therefore, we evaluate here the amount of bound charge not by adding salt
-(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
+(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
 but by studying the changes of the headgroup order parameters with increasing amount of
@@
 
-According to the molecular electrometer concept, the headgroup order parameters of POPC
+Experimentally, the headgroup order parameters of POPC
 increase when negatively charged POPS lipids are incorporated in lipid bilayer
@@
 in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
-This can be explained by the stronger counterion binding affinity, which cancels
+Therein, the stronger counterion binding cancels
 the effect of negatively charged headgroups and prevents the experimentally observed
@@
 realistic surface charge density in membranes containing PS lipids,
-while the other tested simulation models overestimate the counterion
+while the other tested models overestimate the counterion
 binding affinity. The results are in line with the behaviour of headgroup order
@@
 The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
-together with other experimental results suggest less rigid structure of PS headgroups when diluted with
+together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
 POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
 Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
-parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
-and we conclude that more accurate force fields are needed
+parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
+and more accurate force fields are needed
 to correctly describe the PC-PS headgroup interactions in MD simulations.

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
edits to the counterion binding section
Update manuscriptPS.tex
Edits to the ca binding section

--- 2019-04-13 Manuscript/manuscriptPS.tex
@@
 Calcium binding affinity to membranes containing the negatively charged PS lipids can be
-experimentally measured by detecting the PC lipid headgroup order parameters
+experimentally quantified by measuring the PC lipid headgroup order parameters
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
-where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
+where the measurement is not compromised by the dehydrated lipid-ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
@@
 bilayers \cite{catte16}, almost all the tested simulation models overestimate the
-experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
-in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
-indicating overestimated calcium binding affinity.
-The only exception is the CHARMM36 model with the NBfix
-interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
-indicating weaker binding affinity than experiments.
+experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
+indicating too strong calcium binding affinity.
+The only exception is the CHARMM36 simulation utilizing the NBfix
+interaction for calcium~\cite{kim16}, wherein the modulation of order parameters is underestimated, 
+indicating weaker binding affinity than in experiments.
 %incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
 Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
-calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
 is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
@@
 Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
-exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
-below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
+exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
 whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
-the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
+the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
 tested simulation models, including CHARMM36 which underestimated the binding affinity.
 Importantly, the different simulation models predict qualitatively different behaviour
-for the POPS $\alpha$-carbon order parameters with added calcium.
+for the two POPS $\alpha$-carbon order parameters with added calcium.
 For example, both order parameters decrease in Berger, but increase
@@
 in all simulation models, despite significant discrepancies in the headgroup
-structure without additional ions~\cite{catte16}. Therefore, we conclude that
+structures produced in salt-free simulations~\cite{catte16}. Therefore, we conclude that
 improvement of force fields is necessary to correctly capture the interactions between the

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the conclusions

@@
 Calcium binding affinity to membranes containing the negatively charged PS lipids can be
-experimentally quantified by measuring the PC lipid headgroup order parameters
+experimentally measured by detecting the PC lipid headgroup order parameters
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
-where the measurement is not compromised by the dehydrated lipid-ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
+where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
@@
 bilayers \cite{catte16}, almost all the tested simulation models overestimate the
-experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
-indicating too strong calcium binding affinity.
-The only exception is the CHARMM36 simulation utilizing the NBfix
-interaction for calcium~\cite{kim16}, wherein the modulation of order parameters is underestimated, 
-indicating weaker binding affinity than in experiments.
+experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
+in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
+indicating overestimated calcium binding affinity.
+The only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
+indicating weaker binding affinity than experiments.
 %incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
 Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
-calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
 is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
@@
 Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
-exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
+exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
+below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
 whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
-the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
+the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
 tested simulation models, including CHARMM36 which underestimated the binding affinity.
 Importantly, the different simulation models predict qualitatively different behaviour
-for the two POPS $\alpha$-carbon order parameters with added calcium.
+for the POPS $\alpha$-carbon order parameters with added calcium.
 For example, both order parameters decrease in Berger, but increase
@@
 in all simulation models, despite significant discrepancies in the headgroup
-structures produced in salt-free simulations~\cite{catte16}. Therefore, we conclude that
+structure without additional ions~\cite{catte16}. Therefore, we conclude that
 improvement of force fields is necessary to correctly capture the interactions between the
@@
 \section{Conclusions}
-Lipids with PS headgroups, and their interactions with ions,
-play an important role in lipid-mediated signaling
-processes \cite{leventis10,yeung08}. Attemps to use MD
-simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
-headgroups \cite{melcrova16,valentine18,hallock18}.  
-Here (as was previously done for PC lipids \cite{botan15,catte16})
-we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
-of the headgroup structure and ion binding affinity to PS lipids
-in available MD force fields.
+
+%Lipids with PS headgroups, and their interactions with ions, play an important role in lipid-mediated signaling processes \cite{leventis10,yeung08}.
+Attemps to use MD simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PSheadgroups \cite{melcrova16,valentine18,hallock18}.
+%Here (as was previously done for PC lipids \cite{botan15,catte16})
+Here, we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
+of the headgroup structure and the ion binding affinity 
+in available PS MD models.
 The main advantage of this approach is the direct connection
-between the accurately measured experimental order parameters and the simulations,
+between the accurately measurable experimental order parameters and the simulations,
 which reduces the ambiguity in the interpretation of experiments.
 
-First, we complemented the available experimental data \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters.
-Comparison to these data revealed that none of the available force fields
+First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters, and the proceeded to compare MD simulation results from several force fields to the experimental data
+This revealed that none of the force fields
 %tested using the NMRlipids open collaboration
@@
 concentrations \cite{roux90} then showed that the tested MD force fields
-overestimate the cation binding affinity to these bilayers with two exceptions. 1) The MacRog simulation with potassium counterion appear to produce the most realistic monovalent ion binding
-affinity to PS-containing lipid bilayers. 2) The CHARMM36 force field with the recently introduced
+overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
+affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
 NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
@@
 qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
-the MD force fields to study interactions between PS lipids and other biomolecules.
+the MD force fields to study the biological functioning of PS lipids.
 This is in contrast to the previous results with PC lipids,
@@
 
-Our results pave the way for the development of better MD force fields
-for PS lipids. Using the headgroup order parameters, we were able to evaluate the
-quality of various conformational ensembles in different force fields.
-This can guide the development of force fields that would correctly reproduce the
-conformations sampled by PS headgroups. The experimental dataset of headgroup order
-parameters from POPC:POPS (5:1) mixture with different cation concentrations can
-be used to improve cation binding details in the force fields, as recently demonstrated
-for POPC using the electronic continuum correction \cite{melcr18}.
-Similar study for POPS is being progressed separately \cite{ECCpops}.
+We expect our results to pave the way for the development of better MD force fields
+for PS lipids.
+The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope will to guide the development of force fields towards models that correctly reproduce the PS headgroup structures.
+The cation binding in the force fields can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations.
+This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a
+similar study for POPS is being progressed separately \cite{ECCpops}.
 

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
My edits to section S3 on calibration of the lipid electrometer

@@
 \section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
-Before using the headgroup order parameters to compare ion binding affinity between simulations
-and experiments, the response of the order parameters to the bound charge in simulations should be quantified
-against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
-the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
-in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
+%HA: I rewrote this paragraph based on discussion with markus to make it more clear. Please verify that the statements are still accurate
+When using the molecular electrometer concept to compare ion binding affinity between simulations
+and experiments, one needs to keep in mind that two things affect how much a given order parameter changes when solution charge content is varied: 1) the change in the amount of bound charge and 2) the sensitivity of the order parameter to bound charge.
+Therefore, the response of the order parameters to the bound charge in simulations needs to be first calibrated
+against experiments~\cite{catte16,melcr18}.
+
+In our previous work~\cite{catte16}, we investigated the change in order parameters under varying concentrations of mono- and divalent salts and concluded that the experimental $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio~\cite{akutsu81} was well reproduced by the Lipid14 model, but underestimated by other force fields.
+In a more recent study~\cite{melcr18},
 the headgroup order parameter responses were compared more carefully with the experiments of
 cationic dihexadecyldimethylammonium bromide (DHAB) surfactants in POPC bilayer~\cite{scherer89}.
-The comparison reveals that the both headgroup order parameters in the Lipid14 model are too sensitive to the bound charge,
-while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
-Therefore, the headgroup order parameter response to the bound charge is actually more realistic
-in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
-sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
-The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
-is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
-into account when analysing the ion binding affinities using headgroup order parameters in simulations.
-However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
+The advantage of this approach is that the amount of DHAB in the bilayer is exactly know in experiments, which can be exploited to extract the sensitivity of the order parameters.
+This revealed that both $S_{\rm CH}^\beta$ and $S_{\rm CH}^\alpha$ in the Lipid14 model are equally oversensitive (and thus giving the correct $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$) to the bound charge,
+whereas CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}), indicating that the headgroup order parameter response to the bound charge is actually more realistic
+in CHARMM36 compared to Lipid14. The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
+is relatively more sensitive than the $\alpha$-carbon order parameter.
+
+That said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
 bound charge are typically smaller than then discrepancies arising from ion binding affinity.
@@
   The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
-}
-\end{figure}
-
-\pagebreak
-\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
-Previous studies using the electrometer concept to assess the ion binding affinity to lipid
-bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
-or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
-In this work, we use the former definition to be consistent with the reference
-experimental data \cite{roux90}. The difference between these two concentrations increases
-with the increasing ion binding affinity. However, the difference is essentially negligible
-in the recent model \cite{melcr18} with realistic ion binding affinity (Fig. \ref{concentrationDEFfigure}).
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
-  \caption{\label{concentrationDEFfigure}
-    Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
-    two possible definitions of ion concentration from the recent force field
-    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
-    together with the experimental data \cite{akutsu81,altenbach84}. 
 }

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
My edits to the ion concentration paragraph, changed the section title too

--- 2019-03-20 Manuscript/manuscriptPSsuppl.tex
@@
 \pagebreak
-\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
+\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}
+%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
 Previous studies using the electrometer concept to assess the ion binding affinity to lipid
@@
 or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
-In this work, we use the former definition to be consistent with the reference
-experimental data \cite{roux90}. The difference between these two concentrations increases
-with the increasing ion binding affinity. However, the difference is essentially negligible
-in the recent model \cite{melcr18} with realistic ion binding affinity (Fig. \ref{concentrationDEFfigure}).
+In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
+with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
+in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
 \begin{figure}[]

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
Added a section label
Update manuscriptPSsuppl.tex
My edits to S7 on counterions in 5:1 mixtures

--- 2019-03-20 Manuscript/manuscriptPSsuppl.tex
@@
 of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
-as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
+in increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
 ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-The different binding behaviour is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
+The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
 but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
@@
 
-In simulations with Berger and CHARMM36 models, the responses of headgroup order parameter
-of both POPC and POPS to the added sodium or potassium are more similar to the experiments
-with LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
+In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
+of to added sodium and potassium are not in line with the experiments.
+Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.
+\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
 in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
@@
 ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
-Appearance of such clusters also in the MagRog simulations with 4 M concentration of KCl
-could explain the unsystematic changes of the order parameters in this model with the added KCl.
+Appearance of such clusters in the MagRog simulations with 4 M KCl
+could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
 In conclusion, the results are in line with the section \ref{ciBINDINGsection}
@@
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
+    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
     parameter value is at zero.

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
My edits to the SI NBfix section

--- 2019-04-15 Manuscript/manuscriptPSsuppl.tex
@@
 \pagebreak
-\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}
-%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
+\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
 Previous studies using the electrometer concept to assess the ion binding affinity to lipid
@@
 or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
-In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
-with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
-in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
+In this work, we use the former definition to be consistent with the reference
+experimental data \cite{roux90}. The difference between these two concentrations increases
+with the increasing ion binding affinity. However, the difference is essentially negligible
+in the recent model \cite{melcr18} with realistic ion binding affinity (Fig. \ref{concentrationDEFfigure}).
 \begin{figure}[]
@@
 The response of POPC headgroup order parameters to the CaCl$_2$ concentration
-are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 containing
-the NBfix for interactions between calicum and lipid oxygens \cite{kim16}
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
 (Fig. \ref{changesWITHCaClPS} in the main text),
 indicating that the calcium binding to the bilayer is too weak with these
-parameters. Without the NBfix term, the calcium binding affinity to pure POPC
-lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
-However, after employing the NBfix term, the response of headgroup order
+parameters. The response of headgroup order
 parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
 (Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
-POPC bilayer are underestimated. Notably, CHARMM36 simulations with the NBfix
-terms \cite{venable13,kim16} employed predict similar binding affinity for sodium and calcium.
-In conclusion, the special NBfix~\cite{kim16} for calcium, 
-incorporated in the parameters give by the CHARMM-GUI at the time
-of running the simulations (January 2018), leads to the underestimation of
-calcium binding affinity to pure PC and mixed PC:PS bilayers.
+POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
+terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium.
+Without the NBfix term, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
+
+%HA: this just repeats the above? 
+%In conclusion, the special NBfix~\cite{kim16} for calcium, 
+%incorporated in the parameters give by the CHARMM-GUI at the time
+%of running the simulations (January 2018), leads to the underestimation of
+%calcium binding affinity to pure PC and mixed PC:PS bilayers.
 
@@
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
-    the NBfix term employed for sodium \cite{venable13} ({\it left})
+    Headgroup order parameters from CHARMM36 simulations of POPC, where the NBfix term was employed employed for sodium \cite{venable13} ({\it left})
     and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}
@@
   }
-\end{figure}
-
-\pagebreak
-\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
-  \caption{\label{CAdensPCPSmixtureALL}
-    Ca2+ density profiles from simulations.
-  }
-  \todo{Should we include also counterions into the plot?} \\
 \end{figure}

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
My edits to the SI section S9 on calcium binding profiles

--- 2019-04-15 Manuscript/manuscriptPSsuppl.tex
@@
 \section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
+\todo{This section could benefit from some clarifying text}
 \begin{figure}[ht]
@@
   \caption{\label{CAdensPCPSmixtureALL}
-    Ca2+ density profiles from simulations.
+    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
   }
   \todo{Should we include also counterions into the plot?} \\
+  \todo{It would be helpful to have the panels in the same order as in the main text}
+  \todo{Concentration for C36 here is smaller than in main text}
+
 \end{figure}

### 2019-04-16 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
My edits to SI chapter S10 on the force field ranking

@@
 \section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
-\todo{This section could benefit from some clarifying text}
 \begin{figure}[ht]
@@
   \caption{\label{CAdensPCPSmixtureALL}
-    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
+    Ca2+ density profiles from simulations.
   }
   \todo{Should we include also counterions into the plot?} \\
-  \todo{It would be helpful to have the panels in the same order as in the main text}
-  \todo{Concentration for C36 here is smaller than in main text}
-
 \end{figure}
@@
 \pagebreak
-\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
 
-The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
+\section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
+In Figure~\ref{HGorderParametersPS}) of main text we present a rough and subjective ranking of the force fields investigated in this work. 
+The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.
 %
-First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
-we looked separately at deviations in magnitude and forking.
+For each carbon (the columns in Fig.~\ref{HGorderParametersPS}),
+we first investigated separately how well a given force field represents the {\bf magnitude} of the order parameters and their {\bf forking}.
 
-{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
-the force-field-produced OPs were.
-%
-For each carbon, the following 5-step scale was used:
+\subsubsection*{Magnitude}
+To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$s) were to the 
+the force-field-produced $S_\mathrm{CH}$s, we assign a number for each carbon based on the following 5-step scale:
+
 %
 \begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
 %
-\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
+\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.03$ units away from the SSP.}
 %
-\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
+\item [2  ({\textsf{\small M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.05$ units away from the SSP.}
 %
-\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
+\item [3 ({\textsf{\large M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.10$ units away from the SSP.}
 %
-\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
+\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated $S_\mathrm{CH}$s  were $> 0.10$ units  away from the SSP.}
 \end{description}
 
-{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
+\subsubsection*{Forking}
+Forking in each force field was assessed based on how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
 Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.
@@
 \begin{description}
-\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
+\item [0 (~):] \noindent {The distance $D$ between the symbols (indicating time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated $S_\mathrm{CH}$s (includes all lipid types investigated for a given force field).}
 %
@@
 
-Based on these assessments of magnitude and forking deviations,
-each carbon was then assigned to one of the following groups:
-"within experimental error"
-(magnitude and forking deviations both on step 0 of the scales described above),
+\noindent Based on these assessments of magnitude and forking deviations from experimental values,
+each carbon was then assigned to one of the following groups: "within experimental error"
+(magnitude and forking deviations both on 0 of the scales described above),
 "almost within experimental error"
-(sum of the magnitude and forking deviation steps 1 or 2),
+(sum of the magnitude and forking deviation 1 or 2),
 "clear deviation from experiments"
-(sum of magnitude and forking deviation steps from 3 to 5), and
+(sum of magnitude and forking deviation from 3 to 5), and
 "major deviation from experiments"
-(sum of magnitude and forking deviation steps from 6 to 8).
+(sum of magnitude and forking deviation from 6 to 8).
 These groups are indicated by colors in Fig.~4.
 (Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
-the corresponding group assigment limits were: 0, 1, 2, and 3.)
+the corresponding group assignment limits were: 0, 1, 2, and 3.)
 
@@
 4 (major deviation from experiments),
-and the weights of the five carbons were summed up.
+and the contributions from the five carbons were summed up.
 The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
-was then used to (roughly and subjectively, as should be clear from the
-above description) rank the force fields.
+was then used to (roughly and subjectively) rank the force fields.
 

### 2019-04-18 Manuscript/manuscriptPS.tex
Update author list

--- 2019-04-13 Manuscript/manuscriptPS.tex
@@
 
+\author{Hanne Antila}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Pavel Buslaev}
@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
-\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
+\author{Thomas Piggot}
 \affiliation{Southampton, United Kingdom}
@@
 in all simulation models, despite significant discrepancies in the headgroup
-structures produced in salt-free simulations~\cite{catte16}. Therefore, we conclude that
-improvement of force fields is necessary to correctly capture the interactions between the
+structure without additional ions~\cite{catte16}.
+The divergent response of the Berger model may arise from the ring like structures
+observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
+%These ring like structures are a widespread feature of typical Berger based lipid force
+%fields containing explicit hydrogen atoms in the head group
+%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
+We conclude that improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.

### 2019-04-20 Manuscript/manuscriptPS.tex
Update Figs. 6 and 10, and update caption for Fig. 5.
Figs. 6 and 10 are formatted in the style of NMRlipids II (with scripts to make the new figures included in the Figs/scripts).

In Fig 5 caption, the M and F are now formatted just like in NMRlipids I. Alternate formatting could be {\textbf{\footnotesize F}} which makes the font bold and a tiny bit smaller.

@@
     Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
-    Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity.
-Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red).
+    Here {\textsf{\small M}} indicates a magnitude problem, {\textsf{\small F}} a forking problem; letter size increases with problem severity.
+Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
 The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
+  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
   \caption{\label{NAdensPOPS}
@@
   \centering
-  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWcons.eps}
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
   \caption{\label{CAdensPCPSmixture}

### 2019-04-23 Manuscript/manuscriptPSsuppl.tex

@@
 \section{Dihedral angle distributions of the headgroup and glycerol backbone
-  regions of PS lipids from different simulation models}
+  regions of PS lipids from different simulation models}\label{Diheds}
 

### 2019-04-23 Manuscript/manuscriptPS.tex

@@
 in all simulation models, despite significant discrepancies in the headgroup
-structure without additional ions~\cite{catte16}.
+structures produced in salt-free simulations~\cite{catte16}.
 The divergent response of the Berger model may arise from the ring like structures
@@
 %[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
-We conclude that improvement of force fields is necessary to correctly capture the interactions between the
+Therefore, we conclude that
+improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.

### 2019-04-23 Manuscript/manuscriptPSsuppl.tex

@@
 \pagebreak
+\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}
+%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
+Previous studies using the electrometer concept to assess the ion binding affinity to lipid
+bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
+or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
+In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
+with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
+in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
+\begin{figure}[]
+  \centering
+  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
+  \caption{\label{concentrationDEFfigure}
+    Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
+    two possible definitions of ion concentration from the recent force field
+    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
+    together with the experimental data \cite{akutsu81,altenbach84}. 
+  }
+\end{figure}
+
+\pagebreak
 \section{Spectral slices in the indirect dimension from the R-PDLF and SDROSS experiments}

### 2019-04-23 Manuscript/manuscriptPSsuppl.tex

@@
 \pagebreak
+\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
+\todo{This section could benefit from some clarifying text}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
+  \caption{\label{CAdensPCPSmixtureALL}
+    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
+  }
+  \todo{Should we include also counterions into the plot?} \\
+  \todo{It would be helpful to have the panels in the same order as in the main text}
+  \todo{Concentration for C36 here is smaller than in main text}
+
+\end{figure}
+
+\pagebreak
 \section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

### 2019-04-23 Manuscript/manuscriptPSsuppl.tex
Fix concentration in Fig. S18 as mentioned here: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/69

@@
 The response of POPC headgroup order parameters to the CaCl$_2$ concentration
-are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM\_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
 (Fig. \ref{changesWITHCaClPS} in the main text),
@@
   \todo{It would be helpful to have the panels in the same order as in the main text}
-  \todo{Concentration for C36 here is smaller than in main text}
-
 \end{figure}

### 2019-04-23 Manuscript/manuscriptPS.tex

@@
   \caption{\label{comparisonTablePS}
-    Rough, subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
-    Here M indicates a magnitude problem, F a forking problem; letter size increases with problem severity.
-Color scheme: within experimental error (dark green), almost within experimental errorÓ (light green), Òclear deviation from experiments (light red), and major deviation from experiments (dark red).
+    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
+    Here {\textsf{\small M}} indicates a magnitude problem, {\textsf{\small F}} a forking problem; letter size increases with problem severity.
+Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
 The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
-For full details of the assessment, see SI section~\ref{Ranking}.
+For full details of the assessment, see Supplementary Information.
   }

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex
Small updates

@@
 
-\pagebreak
+\clearpage
 \section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
@@
 
-\pagebreak
+\clearpage
 \section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
 
-The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
+\section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
+In Figure~\ref{HGorderParametersPS}) of main text we present a rough and subjective ranking of the force fields investigated in this work. 
+The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.
 %
-First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
-we looked separately at deviations in magnitude and forking.
+For each carbon (the columns in Fig.~\ref{HGorderParametersPS}),
+we first investigated separately how well a given force field represents the {\bf magnitude} of the order parameters and their {\bf forking}.
 
-{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
-the force-field-produced OPs were.
-%
-For each carbon, the following 5-step scale was used:
+\subsubsection*{Magnitude}
+To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$s) were to the 
+the force-field-produced $S_\mathrm{CH}$s, we assign a number for each carbon based on the following 5-step scale:
+
 %
 \begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
 %
-\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
+\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.03$ units away from the SSP.}
 %
-\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
+\item [2  ({\textsf{\small M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.05$ units away from the SSP.}
 %
-\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
+\item [3 ({\textsf{\large M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.10$ units away from the SSP.}
 %
-\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
+\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated $S_\mathrm{CH}$s  were $> 0.10$ units  away from the SSP.}
 \end{description}
 
-{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
+\subsubsection*{Forking}
+Forking in each force field was assessed based on how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
 Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.

### 2019-04-24 Manuscript/manuscriptPS.tex
Small updates

@@
 suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
-the results from same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
+the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
 rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
@@
 
-We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$~\cite{botan15,catte16}, can be utilized to resolve such controversies. The $S_\mathrm{CH}$ can be
+We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
+can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
 measured from NMR experiments with high accuracy and directly compared to simulations
@@
     MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
-  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
 \end{figure}
@@
   \todo{Should we include also counterions into the plot?} \\
-  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
 \end{figure}
@@
 Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
-exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
+exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}).
+The $\beta$-carbon order parameter increases with added CaCl$_2$,
 whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
@@
 
-First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters, and the proceeded to compare MD simulation results from several force fields to the experimental data
+First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
+and then proceeded to compare MD simulation results from several force fields to the experimental data
 This revealed that none of the force fields

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex
Small updates

--- 2019-04-23 Manuscript/manuscriptPSsuppl.tex
@@
   reported in the experiments by Roux et al.~\cite{roux90}.
-  The simulation details are given in the supplementary information.
+  % The simulation details are given in the supplementary information.
   % The lipid force fields named as in our previous work~\cite{botan15}.
@@
 {\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
-The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
-
+%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
 The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
@@
 Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
-These bilayers were simulated using Gromacs ?? \cite{check the version} in the presence of counterions only, as well as together with different
+These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
 concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
@@
 The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
-Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from 
-Ref.~\citenum{rog16}. Notably, the chains were reversed in order to modify 
-OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{rog16}, the 
-POPS head group conformations generated by CHARMM-GUI are in the naturally occurring L stereoisomer.
+Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from OPPS topology in
+Ref.~\citenum{rog16} by reversing the order of acyl chains.
+The initial structures from CHARMM-GUI were used because they have the
+naturally occurring L stereoisomer in the POPS head group, instead of the 
+the D stereoisomer present in Ref.~\citenum{rog16}.
 TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
@@
 The simulation parameters were taken from Ref.~\citenum{rog16}. 
-Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step 
-\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
+Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
+%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
 The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
@@
 4480 water molecules resulting in 31.1 water molecules per lipid. For
-each system, 128 NaCl molecules are added to neutralize the system.\\
+each system, 128 NaCl molecules are added to neutralize the system.
+In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
+to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
+and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
+ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
+The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
+the number of molecules in simulations, related simulation files and trajectories.
 
@@
 
-In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
-and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
-ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB~\cite{aqvist90} ion parameters.
-The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain the details of the simulations and the raw trajectories, respectively.
-
 \noindent
@@
 {\it POPC:POPS (5:1) mixtures with additional calcium} \\
-The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl}
+%The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl}
 
@@
 
-\todo{J. Melcr, please write simulation details} \\
+\todo{J. Melcr, please write simulation details about Gromacs simulations with Dang ions.} \\
 
@@
 \noindent
+{\it POPS}
+\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}
+
+\noindent
+{\it DOPS with 128 lipids.}
+\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
+
+\noindent
 {\it DOPS with 288 lipids.}
-The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions.
-The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
-Simulation was performed for 200~ns, and the last 100~ns were used for the analysis.
-Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}.
+The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na ions
+was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
+The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
+ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
+The system was simulated in NPT ensemble for 200~ns using the GROMACS 5.0.4 package~\cite{abraham2015gromacs},
+and the last 100~ns was used for the analysis.
 Timestep of 2~fs was used with the leapfrog integrator.
@@
 Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.
-
-\todo{Other systems to be written by Piggot}
-
-%{\it DOPS and POPS bilayers.} \\
-%The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.
 
@@
 of calcium \cite{melcrova16} available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa}
-were used. The additional sodium ions were removed from the structure file of the simulation
-with excess amount of sodium \cite{POPCpopsBERGERwith1000mMNa} to generate the initial structure for the reference simulation
-with sodium counterions only. The data, input and force field parameter files of the reference simulation
-are available from Ref. \citenum{bergerPOPSPOPC4:1mixtureT310K}. \todo{The simulation details of this are
-  not clearly described in Zenodo repository, but can be read from the mdp file available there. Should we
-write the detailes here or not?}
+were used. To generate the reference system with sodium counterions only,
+the additional ions were removed from the structure file in Ref.~\citenum{POPCpopsBERGERwith1000mMNa}.
+This system was simulated 120~ns using Gromacs 5~\cite{abraham2015gromacs} with parameter (mdp file)
+and force field files (top and itp) available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K},
+the last 80~ns was used for the analysis.
+%\todo{The simulation details of this are not clearly described in Zenodo repository, but can be read from the mdp file available there.
+%  Should we write the detailes here or not?}
 
@@
 \pagebreak
-\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}
+\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}\label{concentrationDEFsection}
 %\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
@@
 MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
-Also this experiment indicates a postive value for the larger $\alpha$-carbon order parameter.
+Also this experiment indicates a positive value for the larger $\alpha$-carbon order parameter.
 \begin{figure}[]
@@
 In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
-of to added sodium and potassium are not in line with the experiments.
+to added sodium and potassium are not in line with the experiments.
 Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.
@@
 The response of POPC headgroup order parameters to the CaCl$_2$ concentration
-are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM\_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM-GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
 (Fig. \ref{changesWITHCaClPS} in the main text),
@@
 
-\pagebreak
+\clearpage
 \section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
-\todo{This section could benefit from some clarifying text}
+%\todo{This section could benefit from some clarifying text}
 \begin{figure}[ht]
@@
   }
-  \todo{Should we include also counterions into the plot?} \\
-  \todo{It would be helpful to have the panels in the same order as in the main text}
+%  \todo{Should we include also counterions into the plot?} \\
 \end{figure}
 
-\pagebreak
+\clearpage
 \section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
@@
 
-\newpage
+\section{Author contributions}
+\noindent
+{\it Hanne Antila}
+
+\noindent
+{\it Pavel Buslaev}
+
+\noindent
+{\it Fernando Favela}
+
+\noindent
+{\it Tiago M. Ferreira}
+
+\noindent
+{\it Ivan Gushchin}
+
+\noindent
+{\it Matti Javanainen}
+
+\noindent
+{\it Batuhan Kav}
+
+\noindent
+{\it Jesper J. Madsen}
+
+\noindent
+{\it Josef Melcr}
+
+\noindent
+{\it Markus Miettinen}
+
+\noindent
+{\it Ricky Nencini}
+
+\noindent
+{\it O. H. Samuli Ollila}
+Designed the project and managed the work.
+Ran and analysed several simulations. Wrote the manuscript.
+
+\noindent
+{\it Thomas Piggot}
+
+\clearpage
 \bibliography{refs.bib}

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex

@@
 
-The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
+\section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
+In Figure~\ref{HGorderParametersPS}) of main text we present a rough and subjective ranking of the force fields investigated in this work. 
+The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.
 %
-First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
-we looked separately at deviations in magnitude and forking.
+For each carbon (the columns in Fig.~\ref{HGorderParametersPS}),
+we first investigated separately how well a given force field represents the {\bf magnitude} of the order parameters and their {\bf forking}.
 
-{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
-the force-field-produced OPs were.
-%
-For each carbon, the following 5-step scale was used:
+\subsubsection*{Magnitude}
+To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$s) were to the 
+the force-field-produced $S_\mathrm{CH}$s, we assign a number for each carbon based on the following 5-step scale:
+
 %
 \begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
 %
-\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
+\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.03$ units away from the SSP.}
 %
-\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
+\item [2  ({\textsf{\small M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.05$ units away from the SSP.}
 %
-\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
+\item [3 ({\textsf{\large M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.10$ units away from the SSP.}
 %
-\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
+\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated $S_\mathrm{CH}$s  were $> 0.10$ units  away from the SSP.}
 \end{description}
 
-{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
+\subsubsection*{Forking}
+Forking in each force field was assessed based on how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
 Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex
Add files via upload
updated supp. and bib files to include Amber 18 reference.

--- 2019-04-24 Manuscript/manuscriptPSsuppl.tex
@@
 In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
-and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
+to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 
+simulation package \cite{amber18md} and lipid parameters were obtained from Amber 
+Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with 
+Joung-Cheatham
 ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB~\cite{aqvist90} ion parameters.
@@
 the Amber Lipid 17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
-Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
-The simulation steps are the same as the the simulations of
+Simulations were ran using Amber18 simulation package \cite{amber18md}. The simulation 
+steps are the same as the the simulations of
 DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns
@@
 \section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
-%HA: I rewrote this paragraph based on discussion with markus to make it more clear. Please verify that the statements are still accurate
-When using the molecular electrometer concept to compare ion binding affinity between simulations
-and experiments, one needs to keep in mind that two things affect how much a given order parameter changes when solution charge content is varied: 1) the change in the amount of bound charge and 2) the sensitivity of the order parameter to bound charge.
-Therefore, the response of the order parameters to the bound charge in simulations needs to be first calibrated
-against experiments~\cite{catte16,melcr18}.
-
-In our previous work~\cite{catte16}, we investigated the change in order parameters under varying concentrations of mono- and divalent salts and concluded that the experimental $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio~\cite{akutsu81} was well reproduced by the Lipid14 model, but underestimated by other force fields.
-In a more recent study~\cite{melcr18},
+Before using the headgroup order parameters to compare ion binding affinity between simulations
+and experiments, the response of the order parameters to the bound charge in simulations should be quantified
+against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
+the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
+in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
 the headgroup order parameter responses were compared more carefully with the experiments of
 cationic dihexadecyldimethylammonium bromide (DHAB) surfactants in POPC bilayer~\cite{scherer89}.
-The advantage of this approach is that the amount of DHAB in the bilayer is exactly know in experiments, which can be exploited to extract the sensitivity of the order parameters.
-This revealed that both $S_{\rm CH}^\beta$ and $S_{\rm CH}^\alpha$ in the Lipid14 model are equally oversensitive (and thus giving the correct $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$) to the bound charge,
-whereas CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}), indicating that the headgroup order parameter response to the bound charge is actually more realistic
-in CHARMM36 compared to Lipid14. The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
-is relatively more sensitive than the $\alpha$-carbon order parameter.
-
-That said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
+The comparison reveals that the both headgroup order parameters in the Lipid14 model are too sensitive to the bound charge,
+while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
+Therefore, the headgroup order parameter response to the bound charge is actually more realistic
+in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
+sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
+The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
+is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
+into account when analysing the ion binding affinities using headgroup order parameters in simulations.
+However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
 bound charge are typically smaller than then discrepancies arising from ion binding affinity.
@@
 \section{Dihedral angle distributions of the headgroup and glycerol backbone
-  regions of PS lipids from different simulation models}\label{Diheds}
+  regions of PS lipids from different simulation models}
 
@@
 of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
-in increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
+as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
 ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
+The different binding behaviour is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
 but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
@@
 
-In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
-of to added sodium and potassium are not in line with the experiments.
-Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.
-\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
+In simulations with Berger and CHARMM36 models, the responses of headgroup order parameter
+of both POPC and POPS to the added sodium or potassium are more similar to the experiments
+with LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
 in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
@@
 ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
-Appearance of such clusters in the MagRog simulations with 4 M KCl
-could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
+Appearance of such clusters also in the MagRog simulations with 4 M concentration of KCl
+could explain the unsystematic changes of the order parameters in this model with the added KCl.
 In conclusion, the results are in line with the section \ref{ciBINDINGsection}
@@
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
+    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
     parameter value is at zero.
@@
 The response of POPC headgroup order parameters to the CaCl$_2$ concentration
-are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM\_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 containing
+the NBfix for interactions between calicum and lipid oxygens \cite{kim16}
 (Fig. \ref{changesWITHCaClPS} in the main text),
 indicating that the calcium binding to the bilayer is too weak with these
-parameters. The response of headgroup order
+parameters. Without the NBfix term, the calcium binding affinity to pure POPC
+lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
+However, after employing the NBfix term, the response of headgroup order
 parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
 (Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
-POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
-terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium.
-Without the NBfix term, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
-
-%HA: this just repeats the above? 
-%In conclusion, the special NBfix~\cite{kim16} for calcium, 
-%incorporated in the parameters give by the CHARMM-GUI at the time
-%of running the simulations (January 2018), leads to the underestimation of
-%calcium binding affinity to pure PC and mixed PC:PS bilayers.
+POPC bilayer are underestimated. Notably, CHARMM36 simulations with the NBfix
+terms \cite{venable13,kim16} employed predict similar binding affinity for sodium and calcium.
+In conclusion, the special NBfix~\cite{kim16} for calcium, 
+incorporated in the parameters give by the CHARMM-GUI at the time
+of running the simulations (January 2018), leads to the underestimation of
+calcium binding affinity to pure PC and mixed PC:PS bilayers.
 
@@
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-    Headgroup order parameters from CHARMM36 simulations of POPC, where the NBfix term was employed employed for sodium \cite{venable13} ({\it left})
+    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
+    the NBfix term employed for sodium \cite{venable13} ({\it left})
     and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}
@@
 
-\clearpage
+\pagebreak
 \section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
-\todo{This section could benefit from some clarifying text}
 \begin{figure}[ht]
@@
   \caption{\label{CAdensPCPSmixtureALL}
-    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
+    Ca2+ density profiles from simulations.
   }
   \todo{Should we include also counterions into the plot?} \\
-  \todo{It would be helpful to have the panels in the same order as in the main text}
 \end{figure}
 
-\clearpage
+\pagebreak
 \section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
 
-\section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
-In Figure~\ref{HGorderParametersPS}) of main text we present a rough and subjective ranking of the force fields investigated in this work. 
-The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.
+The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
 %
-For each carbon (the columns in Fig.~\ref{HGorderParametersPS}),
-we first investigated separately how well a given force field represents the {\bf magnitude} of the order parameters and their {\bf forking}.
+First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
+we looked separately at deviations in magnitude and forking.
 
-\subsubsection*{Magnitude}
-To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$s) were to the 
-the force-field-produced $S_\mathrm{CH}$s, we assign a number for each carbon based on the following 5-step scale:
-
+{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
+the force-field-produced OPs were.
+%
+For each carbon, the following 5-step scale was used:
 %
 \begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
 %
-\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.03$ units away from the SSP.}
+\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
 %
-\item [2  ({\textsf{\small M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.05$ units away from the SSP.}
+\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
 %
-\item [3 ({\textsf{\large M}}):] \noindent {All the calculated $S_\mathrm{CH}$s  were $< 0.10$ units away from the SSP.}
+\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
 %
-\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated $S_\mathrm{CH}$s  were $> 0.10$ units  away from the SSP.}
+\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
 \end{description}
 
-\subsubsection*{Forking}
-Forking in each force field was assessed based on how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
+{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally.
 Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen.
@@
 \begin{description}
-\item [0 (~):] \noindent {The distance $D$ between the symbols (indicating time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated $S_\mathrm{CH}$s (includes all lipid types investigated for a given force field).}
+\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
 %
@@
 
-\noindent Based on these assessments of magnitude and forking deviations from experimental values,
-each carbon was then assigned to one of the following groups: "within experimental error"
-(magnitude and forking deviations both on 0 of the scales described above),
+Based on these assessments of magnitude and forking deviations,
+each carbon was then assigned to one of the following groups:
+"within experimental error"
+(magnitude and forking deviations both on step 0 of the scales described above),
 "almost within experimental error"
-(sum of the magnitude and forking deviation 1 or 2),
+(sum of the magnitude and forking deviation steps 1 or 2),
 "clear deviation from experiments"
-(sum of magnitude and forking deviation from 3 to 5), and
+(sum of magnitude and forking deviation steps from 3 to 5), and
 "major deviation from experiments"
-(sum of magnitude and forking deviation from 6 to 8).
+(sum of magnitude and forking deviation steps from 6 to 8).
 These groups are indicated by colors in Fig.~4.
 (Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
-the corresponding group assignment limits were: 0, 1, 2, and 3.)
+the corresponding group assigment limits were: 0, 1, 2, and 3.)
 
@@
 4 (major deviation from experiments),
-and the contributions from the five carbons were summed up.
+and the weights of the five carbons were summed up.
 The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
-was then used to (roughly and subjectively) rank the force fields.
+was then used to (roughly and subjectively, as should be clear from the
+above description) rank the force fields.
 

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex
Add Amber citation

--- 2019-04-24 Manuscript/manuscriptPSsuppl.tex
@@
 In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
+to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
 and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
@@
 the Amber Lipid 17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
-Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
+Simulations were ran using Amber18 simulation package \cite{amber18md} 
 The simulation steps are the same as the the simulations of
@@
 
-One set of simulations of POPC:POPS (5:1) mixtures with additional calcium ions has been performed using Amber18 simulation package.
+One set of simulations of POPC:POPS (5:1) mixtures with additional calcium ions has been performed using Amber18 simulation package \cite{amber18md}.
 Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.

### 2019-04-25 Manuscript/manuscriptPS.tex
Update title
author contributions for matti, adjustment of affiliation

@@
 
-\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids} %Title of paper
+\title{Headgroup structure and cation binding in phosphatidylserine lipid bilayers} %Title of paper
 
@@
 \author{Matti Javanainen}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{Batuhan Kav}

### 2019-04-25 Manuscript/manuscriptPSsuppl.tex
author contributions for matti, adjustment of affiliation

@@
 
-\title{ Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids}
+\title{ Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers}
 
@@
 {\it Matti Javanainen}
+Performed most of the MacRog simulations and provided comments on the manuscript.
 

### 2019-04-26 Manuscript/manuscriptPSsuppl.tex
Update author contributions
Update manuscriptPSsuppl.tex

@@
 \author{Pavel Buslaev}
-\affiliation{Moscow Institute of Physics and Technology}
+\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 \author{Fernando Favela}
@@
 \author{Ivan Gushchin}
-\affiliation{Moscow Institute of Physics and Technology}
+\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 \author{Matti Javanainen}
@@
 {\it Hanne Antila}
+Contiributed to the development analysis tools used for evaluating the force fields,  provided critical discussion on the manuscript content and edited it for clarity.
 
@@
 {\it Pavel Buslaev}
+Analyzed the dihedral angle distributions in the head group and glycerol backbone regions.
 
@@
 {\it Ivan Gushchin}
+Supervised the work of P.B.
 

### 2019-04-26 Manuscript/manuscriptPSsuppl.tex
Simulation description and author contribution for J. Melcr

--- 2019-04-24 Manuscript/manuscriptPSsuppl.tex
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium} \\
+{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration} \\
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
+We have performed two sets of simulations of POPC:POPS (5:1) mixtures using Amber Lipid17 force field \cite{gould18}
+with additional calcium chloride concentrations, 
+which employed different simulation engines and parameters for ions.
 
-One set of simulations of POPC:POPS (5:1) mixtures with additional calcium ions has been performed using Amber18 simulation package \cite{amber18md}.
+One set of simulations has been performed using Amber18 simulation package \cite{amber18md} and ff99SB~\cite{aqvist90} ion parameters. 
 Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
 The simulation details, minimization, heating, and equilibration runs are the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield.
-For all simulations, ff99SB~\cite{aqvist90} ion parameters and Amber lipid 17 force field were used.
 For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
@@
 
-\todo{J. Melcr, please write simulation details about Gromacs simulations with Dang ions.} \\
+Another set of simulations with Lipid17 force field for lipids \cite{gould18}
+and Dang parameters for ions \cite{smith94, dang06}
+was performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}. 
+We used the time step of 2~fs. 
+The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
+and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
+The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
+Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
+Pressure was coupled to the Parrinello-Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
+The composition of the individual simulations is shown in Table~\ref{mixedIONsystems}. 
+Initial structures obtained from CHARMM-GUI~\cite{lee16,jo18} 
+were equilibrated prior to production simulations used for analysis,
+which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
+Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.
 
@@
 {\it Josef Melcr}
+consulted the project, 
+prepared tools for calculating order parameters
+and performed several simulations.
 

### 2019-05-01 Manuscript/manuscriptPS.tex
Tuned the abstract.

@@
 Prague 6, Czech Republic}
-\author{Markus Miettinen}
+\author{Markus S. Miettinen}
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
@@
 charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
-electrostatic interactions and direct binding, and can induce
-membrane fusion and phase separation in the presence of calcium ions.
-Molecular details of these phenomena are not well understood,
+nonspecific electrostatic interactions as well as specific binding; in the presence of calcium ions, PS can induce
+membrane fusion and phase separation.
+Molecular details of these phenomena remain not well understood, partly
 because accurate models to interpret the experimental data have not
-been available. Here, we use solid-state NMR S-DROSS to measure the signs of the experimental POPS headgroup
-order parameters, and gather a set of experimental NMR results from pure PS and mixed PC:PS lipid bilayers. Using the open collaboration
-method, we then extract data from a wide range of available PS MD models (force fields) and asses the PS headgroup structures and the ion binding produced by these force fields by comparing to the NMR data.
-We find that none of the models reproduce the NMR data within experimental accuracy,
-but the best ones suggest that the carboxyl group in the serine headgroup
-does not rotate freely. In line with the previous results for PC lipids,
-none of the PS force fields correctly captures the cation binding affinity---the response of PS headgroups to bound ions can even
-\emph{qualitatively} differ from experiments. The collected experimental dataset and simulation
+been available. Here we
+gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
+PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
+and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
+We then use this data set
+to assess,
+in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
+the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
+We find that none of the MD models reproduce the NMR data within experimental accuracy.
+The best MD models, however, suggest that the carboxyl group in the PS headgroup
+does not rotate freely. In line with our previous results for PC lipids,
+none of the PS force fields correctly captures the cation binding affinity; moreover, the response of PS headgroups to bound ions can  differ from experiments even \emph{qualitatively}.
+The collected experimental dataset and simulation
 results pave the way for improvement of lipid force fields to correctly

### 2019-05-07 Manuscript/manuscriptPSsuppl.tex

@@
 \author{Ivan Gushchin}
-\affiliation{Moscow Institute of Physics and Technology}
+\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 \author{Matti Javanainen}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{Batuhan Kav}

### 2019-05-07 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex

@@
 The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}
-\todo{J. Madsen, please write simulation details}. \\
+POPC:POPS (both 1:1 and 5:1) mixtures were prepared with 300 lipids in total (150:150 and 250:50, respectively) with the neutralizing ions with the CHARMM-GUI \cite{lee16,jo18}.
+Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K.
+For further details see table \ref{mixedIONsystemsMONOVALENT}.
+ \\
 
@@
 The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
-\todo{J. Madsen, please write simulation details}. \\
+Systems with 0.15 M and 1 M calcium chloride were prepared using the CHARMM-GUI \cite{lee16,jo18}.
+Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K.
+For further details see table S1.
+\\
 

### 2019-05-07 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex

@@
 \author{Jesper J. Madsen}
-\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
+\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
 \author{Josef Melcr}

### 2019-05-07 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex

@@
 with potassium countertions were ran using CHARMM36 force field.
-The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}
+The data are available from Refs.
+\citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
 POPC:POPS (both 1:1 and 5:1) mixtures were prepared with 300 lipids in total (150:150 and 250:50, respectively) with the neutralizing ions with the CHARMM-GUI \cite{lee16,jo18}.

### 2019-05-22 Manuscript/manuscriptPS.tex
Updates based on comments and text from Tom Piggot sent by email

@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
-\author{Thomas Piggot}
-\affiliation{Southampton, United Kingdom}
+\author{Thomas J. Piggot}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
 %\author{NMRlipids collaboration}
@@
 binding can be regulated by calcium \cite{leventis10}.
-Therefore, the deciphering structural details
+Therefore, deciphering the structural details
 of lipid headgroups and the details of cation binding
@@
 dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
-The dehydrated complexes of PS headgroup with calcium ions can even lead to
+The dehydrated complexes of PS headgroups with calcium ions can even lead to
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
@@
   Notation 2$\times$[time] indicates that two independent MD runs was conducted. Additional simulation details are given in the supplementary information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
     %   The lipid force fields named as in our previous work~\cite{botan15}. 
@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field for lipids / ions  & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
+lipid/counter-ions  & force field for lipids & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
 \hline 
@@
 POPS/K$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36POPS298Kpotassium}  & \tabularnewline
-POPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline
 POPS/Na$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{macrogPOPS298Kcorrect}  & \tabularnewline
 POPS/K$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 200  & 150  & \cite{macrogPOPS298KwithK}  & \tabularnewline
-POPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
-POPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
-POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
-POPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
-POPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
+POPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
+POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKPM \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKP \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline
 POPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{slipidsPOPS298K}  & \tabularnewline
@@
 DOPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36DOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
-DOPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
-DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
+DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
+DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline
 DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{slipidsDOPS303K}  & \tabularnewline
@@
 in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
-it established that the structure of PC lipid headgroup and glycerol backbone are not well
-captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
-lipid bilayers is overestimated \cite{catte16}.
+it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
+captured by most MD force fields~\cite{botan15}, that the cation binding to PC
+lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
+carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
 %
@@
   The simulation details are given in the supplementary information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
    % The lipid force fields named as in our previous work~\cite{botan15}.
@@
 \hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline
 \end{tabular}
@@
 for each lipid in the system, and then calculating the average and
-the standard error of the mean over the different lipids. The analysis was conducted using a
+the standard error of the mean over the different lipids. The analysis can be conducted using a
 Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
 MDAnalysis library \cite{agrawal11,gowers16}.
+For united atom models, the positions of hydrogens were generated before the order parameter calculation using {\tt protonate} tool
+of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
 The ion number density profiles were calculated using the {\tt gmx density} tool
@@
 which complicates the interpretation of structural differences between PC and PS headgroups.
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
+However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) do replicate the larger-than-in-PC
 forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
@@
     of POPS lipid bilayer from simulations with different force fields.
+    The vertical green bars indicate the location of the phosphate density peak.
     The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
   }
-  \todo{Commented by M. Javanainen in blog:
-    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
+%  \todo{Commented by M. Javanainen in blog:
+%    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
 \end{figure}
@@
   }
-  \todo{After we know which force field is used for POPC in Gromos-CKP simulations, we might be able to
-    add Gromos-CKP data into this plot.}
+  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
+    add GROMOS-CKP data into this plot.}
 \end{figure}
@@
     parameter value from pure POPS is at zero to correctly illustrate the significant forking.
-    The values from CHARMM36 and Gromos simulations are averages of two independent simulations and
+    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
     the error bars are given as the difference between the results divided by two.
@@
 The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
-is reproduced only in Gromos-CKP and in the MacRog simulation
-with potassium counterions, while other models give considerably smaller values.
+is reproduced only in GROMOS-CKP and in the MacRog simulation
+with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
 However, the counterion binding and the concomitant electrostatic screening of the headgroup
@@
 (the lowest concentrations in bulk water), gives the same area per molecule
-as the CHARMM36ua simulation, which has significantly weaker counterion binding
+as the CHARMM36-UA simulation, which has significantly weaker counterion binding
 affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
@@
   \caption{\label{CAdensPCPSmixture}
-    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
+    Number density profiles of Ca$^{2+}$ (solid line) and Cl$^-$ (dashed line) from POPC:POPS (5:1) mixtures
+    simulated with different force fields. The vertical green bars indicate the location of the phosphate density peak.
     The smallest simulated CaCl$_2$ concentrations are shown.
@@
   }
-  \todo{Should we include also counterions into the plot?} \\
 \end{figure}
@@
 % Put your acknowledgments here.
+  MJ acknowledges financial support from the Emil Aaltonen foundation and
+  and CSC-IT center for science for computational resources.
   OHSO acknowledges financial support from Academy of Finland (315596),
@@
   CSC-IT center for science for computational resources.
-  MJ acknowledges financial support from the Emil Aaltonen foundation and
-  and CSC-IT center for science for computational resources.
+  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton
 \end{acknowledgments}

### 2019-05-22 Manuscript/manuscriptPSsuppl.tex
Updates based on comments and text from Tom Piggot sent by email

@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
-\author{Thomas Piggot}
-\affiliation{Southampton, United Kingdom}
+\author{Thomas J. Piggot}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
 
@@
 \noindent
-{\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
-\todo{T. Piggot, please write simulation details} \\
+{\it DOPS and POPS bilayers.}
+%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
+Starting structures for CHARMM36 DOPS and POPS simulations were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
+The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
+Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
+with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
+%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
+For POPS, an identical system was also constructed in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.
+
+Simulations of these systems were performed for 500 ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
+A 2 fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
+For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
+The simulations were maintained at temperatures of 303 K and 298 K for DOPS and POPS respectively using the Nosé-Hoover thermostat \cite{nose84,hoover85}
+with a coupling constant of 1 ps. A pressure of 1 bar was maintained using the Parrinello-Rahman \cite{parrinello81} method with a coupling constant of 5 ps.
+Pressure coupling was applied in a semi-isotropic manner with the x and y dimensions, in the plane of the bilayer, fluctuating independently of the z dimension.
+Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2 nm with the interactions
+switched off between 1.0 nm and 1.2 nm; Coulombic interactions were truncated at 1.2 nm with long-range interactions treating using PME~\cite{darden93,essman95}.
+
+Analyses on these simulations was performed on the final 100 ns of the simulations.
 
 \noindent
-{\it POPC:POPS mixtures without additional ions}\\
-Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
-with potassium and sodium countertions were ran using CHARMM36 force field.
-The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
-and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.
+{\it POPC:POPS mixtures without additional ions}
+Simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were constructed using the CHARMM-GUI~\cite{lee16,jo18}.
+Two identical systems were constructed, apart from one system was neutralised through the addition of 22 Na$^+$ ions while the other system
+contained 22 K$^+$ ions. As per the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
+and settings were identical to those used in the CHARMM36 POPS simulations.
 
-Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids
-with potassium countertions were ran using CHARMM36 force field.
-The data are available from Refs.
-\citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
-POPC:POPS (both 1:1 and 5:1) mixtures were prepared with 300 lipids in total (150:150 and 250:50, respectively) with the neutralizing ions with the CHARMM-GUI \cite{lee16,jo18}.
-Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K.
-For further details see table \ref{mixedIONsystemsMONOVALENT}.
+%Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
+%with potassium and sodium countertions were ran using CHARMM36 force field.
+%The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
+%and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.
+
+Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
+with neutralizing potassium countertions 
+%ran using CHARMM36 force field.
+%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
+%POPC:POPS (both 1:1 and 5:1) mixtures
+were prepared
+%with 300 lipids in total
+%with the  ions
+using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
+and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
  \\
@@
 POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
-were simulated with the additional calcium concentrations.
-The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
-Systems with 0.15 M and 1 M calcium chloride were prepared using the CHARMM-GUI \cite{lee16,jo18}.
-Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K.
-For further details see table S1.
-\\
+%were simulated with the additional calcium concentrations.
+%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
+and 0.26 M and 1 M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
+Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
+by the CHARMM-GUI at 298 K.
 
@@
 {\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
-\todo{T. Piggot, please write simulation details} \\
+%The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
+Starting structures for CHARMM36-UA DOPS and POPS simulations were taken from those generated for the CHARMM36 simulations (see above),
+with any extraneous hydrogen atoms removed in the lipid tails. Force field parameters for DOPS and POPS were constructed through a
+combination of the published CHARMM36 PS~\cite{venable13} and CHARMM36-UA PC parameters~\cite{lee14}.
+% [REF: https://pubs.acs.org/doi/10.1021/jp410344g].
+All other simulation parameters and conditions were identical to those described for the CHARMM36 all-atom DOPS and POPS simulations.
 
@@
 \noindent
-{\it POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
-\todo{T. Piggot and M. Javanainen, please write simulation details} \\
-\hl{Piggot: please write about system with Na$^+$ counterions}.
+{\it POPS bilayers.}
+%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
+%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
+%\hl{Piggot: please write about system with Na$^+$ counterions}.
+Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
+were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
+not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
+The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na+ ions) and was converted into
+the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
+Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
+van der Waals interactions were truncated at 1.0 nm with a dispersion correction applied to the energy and pressure;
+Coulombic interactions were truncated at 1.0 nm with PME used to treat interactions beyond this cut-off.
+All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
+including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.
 
@@
 \subsection{Slipids}
-
+%
+%\noindent
+%{\it POPS}
+%\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}%%
+%
+%\noindent
+%{\it DOPS with 128 lipids.}
+%\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
+%
 \noindent
-{\it POPS}
-\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}
-
-\noindent
-{\it DOPS with 128 lipids.}
-\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
+{\it POPS and DOPS with 128 lipids}
+Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
+for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model,
+however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
+% [REF: https://pubs.acs.org/doi/10.1021/ct300777p].
+This was due to the use of the older GROMACS group-based cut-off scheme in these previous simulations.
+In the simulations reported in this work, the Verlet cut-off scheme was used. Van der Waals interactions were
+truncated at 1.0 nm with LJ-PME~\cite{wennberg15}
+% [REF: https://pubs.acs.org/doi/10.1021/acs.jctc.5b00726]
+applied to account for the long-range interactions.
+Coulombic interactions were truncated at 1.0 nm with PME used for the long-range interactions.
+Full testing and validation of these settings with the Slipids PS force field will be provided in an additional forthcoming work.
+All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations.
 
@@
 {\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
-\todo{T. Piggot, please write simulation details} \\
+%The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
+%\todo{T. Piggot, please write simulation details} \\
+Starting structures for DOPS and POPS simulations performed with a Berger-based PS force field
+were taken from those produced for the CHARMM36 simulations. A script was used to re-order the
+atoms and to remove any extraneous hydrogen atoms. Force field parameters were taken from those
+published by Mukhopadhyay et al.~\cite{mukhopadhyay04}
+% [REF: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1303994/]
+and the SPC water model was used~\cite{berendsen81}. The simulation cut-offs closely matched
+those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0 nm;
+Coulombic interactions were truncated at 1.0 nm with PME applied to account for the interactions beyond the cut-off.
+All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
+including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
 
@@
 \subsection{GROMOS-CKP and GROMOS-CKPM}
-  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
-   and CKP refers to the version with more Gromos compatible version
-   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
-\todo{T. Piggot, please write the description of force field.}
+%  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+%   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
+%   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
+%   and CKP refers to the version with more Gromos compatible version
+%   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
+%\todo{T. Piggot, please write the description of force field.}
+The GROMOS-CKP force field (CKP stands for Chandrasekhar-Kukol-Piggot) contains a set of GROMOS compatible lipid parameters.
+Parameters have been developed and validated for PC~\cite{Chandrasekhar03,kukol09,piggot12},
+%  [REFS: https://www.ncbi.nlm.nih.gov/pubmed/12632209, https://pubs.acs.org/doi/10.1021/ct8003468 and https://pubs.acs.org/doi/abs/10.1021/ct3003157],
+PE, PG, Cardiolipin~\cite{piggot11}
+%[REF: https://pubs.acs.org/doi/abs/10.1021/jp207013v]
+and PI~\cite{holdbrook10}
+%[REF: https://pubs.acs.org/doi/abs/10.1021/bi101743w]
+lipids, amongst others.
+Within this work we provide an initial validation for simulating PS lipids with two variants of
+this force field (termed GROMOS-CKP and GROMOS-CKPM). Additional forthcoming work will provide additional
+validation and testing of these PS parameters. The two variants of the force field only differ in terms
+of the charges in the head group. The GROMOS-CKPM (M stands for Mukhopadhyay after the first published
+Berger-based PS simulations published; see above for further details) lipid uses Berger/Chiu based charges
+for the NH3 part of the PS head group (i.e. has the same charges as used in the choline part of the Berger PC lipids).
+The GROMOS-CKP parameters use charges compatible with the rest of the GROMOS force fields;
+the  charges of the NH3 part of the PS head group were taken from a lysine side-chain of the GROMOS 54A7 protein
+force field~\cite{schmid11} % [REF: https://link.springer.com/article/10.1007%2Fs00249-011-0700-9].
+The remainder of the parameters were assigned by analogy from either the other GROMOS-CKP lipids or the GROMOS 54A7 force field.
 
 \noindent
-{\it DOPS and POPS bilayers.} \\
-The data in Table \ref{PSsystems} and Refs. \citenum{ckp1POPS303K,ckp2POPS303K,ckp1DOPS303K,ckp2DOPS303K}
-\todo{T. Piggot, please write simulation details} \\
+{\it DOPS and POPS bilayers.} 
+%The data in Table \ref{PSsystems} and Refs. \citenum{ckp1POPS303K,ckp2POPS303K,ckp1DOPS303K,ckp2DOPS303K}
+%\todo{T. Piggot, please write simulation details} \\
+
+Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the CHARMM36 simulations.
+A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. The force fields parameters for DOPS and POPS
+were as discussed above. The SPC water model was used in these simulations~\cite{berendsen81}. Simulation cut-offs were chosen
+to closely match those used in the original GROMOS-CKP force field validation: van der Waals interactions were truncated at 1.4 nm
+with a dispersion correction applied to the energy and pressure; Coulombic interactions were also truncated at 1.4 nm with PME used
+to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
+and POPS simulations bar, including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
 
 \noindent
-    {\it POPC:POPS (5:1) mixtures without additional ions}\\
-    The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
-\todo{T. Piggot, please write simulation details} \\
+{\it POPC:POPS (5:1) mixtures without additional ions}
+% The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
+% \todo{T. Piggot, please write simulation details} \\
+
+Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).
+A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation setting were identical to those as described above
+for the GROMOS-CKP(M) POPS bilayers.
 
@@
 {\it Thomas Piggot}
+Setup, performed and analysed many of the simulations. Contributed to parts of the manuscript.
 

### 2019-05-22 Manuscript/manuscriptPS.tex
Small update in the text considering dihedral distrubutions

@@
 In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-and narrow angle distribution for PS than for PC headgroup in
-CHARMM36 simulations that have the best agreement with experiments
+angle distribution for PS than for PC headgroup in
+CHARMM36 and Slipids simulations that have the best agreement with experiments
 (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).

### 2019-05-22 Manuscript/manuscriptPS.tex
Update simulation details text

@@
 \caption{The list of MD simulations of pure PS bilayers without additional salt along with the references to the force fields used and the MD trajectories.
-  Notation 2$\times$[time] indicates that two independent MD runs was conducted. Additional simulation details are given in the supplementary information.
+  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
+  JC refers to the Joung-Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
+  Additional simulation details are given in the supplementary information.
   $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field for lipids & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
+lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
 \hline 
@@
   This corresponds the concentration in buffer before solvating lipids, which are
-  reported in the experiments by Roux et al.~\cite{roux90}.  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
+  reported in the experiments by Roux et al.~\cite{roux90}.
+  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
+  Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
   The simulation details are given in the supplementary information.
@@
 \begin{tabular}{lccccccccc}
-lipid/counter-ions  & force field for lipids / ions  & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
+lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
 \hline 
@@
 %    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withKCI} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withNaCI} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18}  & 0  & 60 & 3600  & 0  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18,smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1000  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withKCI} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withNaCI} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60 & 3600  & 0  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
 \hline 

### 2019-05-22 Manuscript/manuscriptPSsuppl.tex
Update simulation details text

@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium} \\
+{\it POPC:POPS (5:1) mixtures with additional calcium} 
 POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
@@
 \noindent
-{\it DOPS and POPS bilayers.} \\
+{\it DOPS and POPS bilayers.} 
 %The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
@@
 \noindent
-{\it POPS bilayers.}
+{\it POPS bilayers with Na$^+$ counterions.}
 %The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
@@
 not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
+The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
 The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na+ ions) and was converted into
@@
 
-This membrane was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
+\noindent
+{\it POPS bilayers with K$^+$ counterions.}
+The membrane with sodium counterions was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
 were described by the \AA{}qvist parameters \cite{aqvist90}. Other topologies
@@
 \subsection{Lipid17}
-{\it DOPS and POPS bilayers.} \\
-
+{\it DOPS and POPS bilayers.} 
 The structure of the bilayers with 128 DOPS or 128 POPS lipids were
@@
 boundary conditions, the non-bonded interactions are calculated with
-particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.\\
+particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.
 
@@
 intervals. Two independent trajectories of 500~ns long was generated and
-the last 200~ns portions were used for the analysis.\\
-
-\noindent
-{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} \\
-
+the last 200~ns portions were used for the analysis.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} 
 The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
@@
 was used for the analysis. The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
-The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.\\
-
-\noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration} \\
+The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with default Amber parameters} 
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
-We have performed two sets of simulations of POPC:POPS (5:1) mixtures using Amber Lipid17 force field \cite{gould18}
-with additional calcium chloride concentrations, 
-which employed different simulation engines and parameters for ions.
-
-One set of simulations has been performed using Amber18 simulation package \cite{amber18md} and ff99SB~\cite{aqvist90} ion parameters. 
+%We have performed two sets of simulations of POPC:POPS (5:1) mixtures using Amber Lipid17 force field \cite{gould18}
+%with additional calcium chloride concentrations, 
+%which employed different simulation engines and parameters for ions.
+Simulations were performed using Amber18 simulation package \cite{amber18md} with ff99SB~\cite{aqvist90} ion parameters and Amber Lipid17~\cite{gould18} force field. 
 Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
@@
 For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
-The trajectory files can be found in Ref~\citenum{POPCpopsLIPID17withCaCl}
-
-Another set of simulations with Lipid17 force field for lipids \cite{gould18}
-and Dang parameters for ions \cite{smith94, dang06}
-was performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}. 
-We used the time step of 2~fs. 
+The trajectory files can be found from Ref~\citenum{POPCpopsLIPID17withCaCl}.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with Dang ion parameters} 
+The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
+and converted to Gromacs format using acpype \cite{silva12}.  
+The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
+Dang parameters for ions \cite{smith94, dang06} were used.
+% Another set of simulations with Lipid17 force field for lipids \cite{gould18}
+Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
+and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
+with the time step of 2~fs. 
 The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
@@
 Pressure was coupled to the Parrinello-Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
-The composition of the individual simulations is shown in Table~\ref{mixedIONsystems}. 
-Initial structures obtained from CHARMM-GUI~\cite{lee16,jo18} 
-were equilibrated prior to production simulations used for analysis,
+The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
+The systems were equilibrated 50~ns prior to production simulations used for analysis,
 which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
@@
 Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
-for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model,
+for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model~\cite{jambeck13},
 however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
@@
 {\it DOPS with 288 lipids.}
-The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na ions
+The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na$^+$ ions
 was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
-The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
+The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
 ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
@@
 \noindent
-{\it DOPS and POPS bilayers.} \\
+{\it DOPS and POPS bilayers.} 
 %The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
@@
 %\todo{T. Piggot, please write simulation details} \\
-
 Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the CHARMM36 simulations.
@@
 to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
-and POPS simulations bar, including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
+and POPS simulations, including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
 
@@
 % \todo{T. Piggot, please write simulation details} \\
-
 Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).

### 2019-05-22 Manuscript/manuscriptPS.tex
Hide todo
Update manuscriptPS.tex
Full name, institution and acknowledgments by FFR.

@@
 \affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
-\author{Fernando Favela}
-\affiliation{Mexico}
+\author{Fernando Favela-Rosales}
+\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
 \author{Tiago M. Ferreira}
@@
   }
-  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
-    add GROMOS-CKP data into this plot.}
+%  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
+%    add GROMOS-CKP data into this plot.}
 \end{figure}
@@
   CSC-IT center for science for computational resources.
-  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton
+  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
+  FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
+  Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA)
+  Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n
+  Tecnol\'{o}gica (PAPIIT) IG100416 for financial support and Cl\'{u}ster
+  H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
+  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-
+  Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
+  LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
 \end{acknowledgments}

### 2019-05-23 Manuscript/manuscriptPS.tex
Changes to affiliations, tiny modifications to text and typo corrections

--- 2019-05-22 Manuscript/manuscriptPS.tex
@@
 \author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{Markus S. Miettinen}
@@
 \author{Ricky Nencini}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{O. H. Samuli Ollila}
@@
 %\homepage[]{Your web page}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \affiliation{Institute of Biotechnology, University of Helsinki}
@@
 charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
-nonspecific electrostatic interactions as well as specific binding; in the presence of calcium ions, PS can induce
+nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
 membrane fusion and phase separation.
-Molecular details of these phenomena remain not well understood, partly
+Molecular details of these phenomena remain poorly understood, partly
 because accurate models to interpret the experimental data have not
@@
 does not rotate freely. In line with our previous results for PC lipids,
-none of the PS force fields correctly captures the cation binding affinity; moreover, the response of PS headgroups to bound ions can  differ from experiments even \emph{qualitatively}.
+none of the PS force fields correctly captures the cation binding affinity.
+Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}.
 The collected experimental dataset and simulation
-results pave the way for improvement of lipid force fields to correctly
+results pave the way for the improvement of lipid force fields that correctly
 describe the biologically relevant negatively charged membranes and their interactions with ions.
@@
 The abundance, however, varies between different organelles, and up to
-25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
+25--35\% of the cytosolic leaflet of plasma membranes 
+\cite{lemmon08,leventis10,li14} consists of PS lipids.
 PS lipids are vastly important biomolecules that interact with
@@
 of lipid headgroups and the details of cation binding
-are crucial for understanding the PS-mediated processes in the cell membranes.
+are crucial for understanding PS-mediated processes on cellular membranes.
 
@@
 Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
-That said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
+That being said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
 to negatively charged and zwiterionic phospholipids, and that
 the increased cation binding to negatively charged lipid bilayers arises only due
-to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.
+to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.
 
@@
 \caption{The list of MD simulations of pure PS bilayers without additional salt along with the references to the force fields used and the MD trajectories.
-  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
-  JC refers to the Joung-Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
+  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
+  JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
   Additional simulation details are given in the supplementary information.
@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
+lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
 \hline 
@@
 
-The molecular level interpretation of these observations is,
-however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
+The molecular level interpretation of these observations is currently lacking, and classical molecular dynamics (MD) simulations have been widely used in efforts 
 to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
@@
 CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
-suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
+suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
 the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
@@
 measured from NMR experiments with high accuracy and directly compared to simulations
-in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
+in order to evaluate the quality of the simulation model, or to interpret the experiments \cite{ollila16}. Using this approach,
 it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
@@
 glycerol backbone C--H bond order parameters
-by measuring the signs of the order parameters using S-DROSS solid-state NMR spectroscopy.
+by measuring the signs of these order parameters using S-DROSS solid-state NMR spectroscopy.
 The quality of headgroup structures and the ion binding affinity in
 %MD simulations of lipid bilayers containing PS lipids
-the available MD simulation models of PS lipids are then asessed based on the collected experimental data.
+the available MD simulation models of PS lipids are then assessed based on the collected experimental data.
 %using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
 The results pave the way
-for development of lipid models that correctly describe 
-the headgroup region of negatively charged lipids in physiological salt
-conditions. Such force fields are expected to be useful in understanding
+for the development of lipid models that correctly describe 
+the headgroup region of negatively charged lipids under physiological salt
+conditions. Such force fields are expected to be extremely useful in understanding
 biological function of lipid headgroups and glycerol backbone, as
@@
   reported in the experiments by Roux et al.~\cite{roux90}.
-  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
+  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
@@
 \begin{tabular}{lccccccccc}
-lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
+lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
 \hline 
@@
 using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
-The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
-Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
+The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03~MHz.
+Magic angle spinning (MAS) of the sample was used at a frequency of 5.15~kHz (R-PDLF experiment) and 5~kHz (S-DROSS experiment).
 The following experimental setups were used.
@@
 {\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
-for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
-The used radio frequency pulses had the following nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
-50 kHz (SPINAL64 $^1$H decoupling pulses).
-The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
-dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.
+for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=1.94$~ms and $\tau_2=0.97$~ms.
+The used radio frequency pulses had the following nutation frequencies: 46.35~kHz (R18$^7_1$ pulses), 63.45~kHz ($^{13}$C 90$^{\circ}$ and 180$^{\circ}$),
+50~kHz (SPINAL64 $^1$H decoupling pulses).
+The $t_1$ increment was equal to 10.79~$\mu$s $\times18\times2$, and 32 points in the indirect
+dimension were recorded using 1024 scans for each, with a recycle delay of 5~s and a spectral width of 149.5~ppm.
 
@@
 The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
-experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
-blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. The used radio frequency pulses had the nutation
-frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
-The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
-measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.
+experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19~ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
+blocks $R$ were set as $\tau_1=39.4$~$\mu$s and $\tau_2=89.4$~$\mu$s. The used radio frequency pulses had the nutation
+frequencies: 63.45~kHz ($^{13}$C 90$^{\circ}$ and 180$^{\circ}$), 50~kHz ($^1$H\,SPINAL64 decoupling).
+The $t_1$ increment (dipolar recoupling dimension) was 800~$\mu$s, and a total of 8 points along $t_1$ were
+measured using 1024 scans for each, with a recycle delay of 5~s and a spectral width of 149.5~ppm.
 
@@
 The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
-as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
-until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.
+as sodium salt) with water (lipid:water 60:40~wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
+until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4~mm NMR rotors.
 
@@
 the standard error of the mean over the different lipids. The analysis can be conducted using a
-Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
+Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that utilizes the
 MDAnalysis library \cite{agrawal11,gowers16}.
-For united atom models, the positions of hydrogens were generated before the order parameter calculation using {\tt protonate} tool
+For united atom models, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
 of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
@@
 simulations~\cite{catte16}. The headgroup order parameters
-of negatively charged PS and PG lipids also exhibit systematic, but less
-understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
+of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
+
 Therefore, measuring the PC headgroup order parameters from 
@@
 Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
-the concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
-or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
-In this work, we use the former definition to be consistent with the reference
+the concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
+In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
+Nevertheless, in this work we use this former definition to be consistent with the reference
 experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
@@
     (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
-    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
+    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon, and $+0.09$ and $-0.02$
     for the $\alpha$-carbon.
-    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
+    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
     for the smaller $\alpha$-carbon C--H bond order parameter.
@@
     (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
-    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
-    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
+    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
+    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
     are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
@@
 of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
-We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C—H
+We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C--H
 bond order parameter give rise to profiles that initially increase and decrease, respectively.
@@
 The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
-For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C—H bonds.
+For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
 However, the beginning of its S-DROSS slice suggests that the larger order parameter of the alpha-carbon is positive and the
@@
 This is confirmed by a SIMPSON simulation
-using the order parameter values of +0.09 from dipolar coupling measured here (Fig. \ref{DPslices})
-and -0.02 from the $^2$H\,NMR experiment~\cite{roux91}.
+using the order parameter values of $+0.09$ from dipolar coupling measured here (Fig. \ref{DPslices})
+and $-0.02$ from the $^2$H\,NMR experiment~\cite{roux91}.
 The literature value for the smaller order parameter was used because the
@@
 (approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
-gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
+gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
 Larger multilamellar repeat distances are expected in the samples with lower lipid
@@
 and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
-as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
+as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full agreement with the experiments.
 The models perform generally less well for PS than for PC
-(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
-which complicates the interpretation of structural differences between PC and PS headgroups.
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}).
+This complicates the interpretation of structural differences between PC and PS headgroups.
 However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
-smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
-also observed in the experiments (Fig. \ref{HGorderParameters}).
+forking of the $\alpha$-carbon observed in the experiments.
+Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
+\ref{HGorderParametersPS} vs.
+Fig.
+2 in Ref.~\citenum{botan15}), in line with
+experiments (Fig. \ref{HGorderParameters}).
 
@@
     The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
-    the error bar is given as the difference between the results divided by two.
+    the error bar is given as the difference between the results divided by two. {\color{red} Why divided by two?}
   }
-  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
-    add GROMOS-CKP data into this plot.}
+%  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
+%    add GROMOS-CKP data into this plot.}
 \end{figure}
@@
     The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
-    the error bars are given as the difference between the results divided by two.
+    the error bars are given as the difference between the results divided by two. {\color{red} Why divided by two?}
   }
@@
     as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
-    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
+    at 298~K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
     The order parameter values from systems without calcium are set as the zero point of y-axis,
@@
 and more accurate force fields are needed
-to correctly describe the PC-PS headgroup interactions in MD simulations.
+to correctly describe the PC--PS headgroup interactions in MD simulations.
 %Having this conclusion in every paragraph gets a bit repetive
@@
 
-Calcium binding affinity to membranes containing the negatively charged PS lipids can be
+Calcium binding affinity to membranes containing negatively charged PS lipids can be
 experimentally quantified by measuring the PC lipid headgroup order parameters
 from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
-where the measurement is not compromised by the dehydrated lipid-ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
+where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
@@
 indicating too strong calcium binding affinity.
-The only exception is the CHARMM36 simulation utilizing the NBfix
-interaction for calcium~\cite{kim16}, wherein the modulation of order parameters is underestimated, 
-indicating weaker binding affinity than in experiments.
+The sole exception is the CHARMM36 model when paired with the NBfix
+correction for calcium~\cite{kim16}. For this combination, the modulation of order parameters is underestimated, 
+indicating a weaker binding affinity than in experiments.
 %incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
-Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
-calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
-is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
+Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
+calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
+is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16}, but overestimated 
 in all the other tested models.
@@
 Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
-exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}).
+exhibit a strong dependence of CaCl$_2$ at small concentrations yet they saturate already at a concentration below 100~mM (Fig. \ref{changesWITHCaClPS}).
 The $\beta$-carbon order parameter increases with added CaCl$_2$,
-whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
+whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
 the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
-tested simulation models, including CHARMM36 which underestimated the binding affinity.
+tested simulation models---including CHARMM36---which underestimated the binding affinity.
 Importantly, the different simulation models predict qualitatively different behaviour
-for the two POPS $\alpha$-carbon order parameters with added calcium.
+for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
 For example, both order parameters decrease in Berger, but increase
-in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
-This is in contrast to the PC headgroup, where
-qualitatively correct reponse to bound ions is observed
-in all simulation models, despite significant discrepancies in the headgroup
-structures produced in salt-free simulations~\cite{catte16}.
+in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
+This is in contrast to the PC headgroup, where a
+qualitatively correct response to bound ions is reproduced
+by all simulation models, despite the significant discrepancies in the headgroup
+structures observed in salt-free simulations~\cite{catte16}.
 The divergent response of the Berger model may arise from the ring like structures
@@
 Therefore, we conclude that
-improvement of force fields is necessary to correctly capture the interactions between the
+the improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.
+
 \section{Conclusions}
@@
 of the headgroup structure and the ion binding affinity 
-in available PS MD models.
+in available MD models of PS.
 The main advantage of this approach is the direct connection
-between the accurately measurable experimental order parameters and the simulations,
-which reduces the ambiguity in the interpretation of experiments.
+between the order parameters measured accurately in experiments and calculated readily from simulations. 
+This reduces the ambiguity in the interpretation of experiments.
 
 First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
-and then proceeded to compare MD simulation results from several force fields to the experimental data
+and then proceeded to compare MD simulation results from several force fields with the experimental data.
 This revealed that none of the force fields
@@
 qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
-the MD force fields to study the biological functioning of PS lipids.
+MD force fields to study the biological function of PS lipids.
 This is in contrast to the previous results with PC lipids,
@@
 were in qualitative agreement even though the headgroup structures themselves were
-incorrect and the cation binding affinities overestimated \cite{catte16}.
+incorrect and the cation binding affinities were overestimated \cite{catte16}.
 
@@
 for PS lipids.
-The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope will to guide the development of force fields towards models that correctly reproduce the PS headgroup structures.
+The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures.
 The cation binding in the force fields can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations.
-This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a
-similar study for POPS is being progressed separately \cite{ECCpops}.
+This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.
 

### 2019-05-23 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
Full name, institution and contribution of FFR.
Tiny changes to the SI.

@@
-\documentclass[journal=jpcbfk]{achemso}
+\documentclass[journal=jctcce]{achemso}
 
@@
 \author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{Markus Miettinen}
@@
 
-\newpage
+%\newpage
 %\tableofcontents
@@
 \begin{tabular}{l c c c c c c c c c}
-  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
 
-Simulations of these systems were performed for 500 ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
-A 2 fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
+Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
+A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
 For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
-The simulations were maintained at temperatures of 303 K and 298 K for DOPS and POPS respectively using the Nosé-Hoover thermostat \cite{nose84,hoover85}
-with a coupling constant of 1 ps. A pressure of 1 bar was maintained using the Parrinello-Rahman \cite{parrinello81} method with a coupling constant of 5 ps.
-Pressure coupling was applied in a semi-isotropic manner with the x and y dimensions, in the plane of the bilayer, fluctuating independently of the z dimension.
-Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2 nm with the interactions
-switched off between 1.0 nm and 1.2 nm; Coulombic interactions were truncated at 1.2 nm with long-range interactions treating using PME~\cite{darden93,essman95}.
-
-Analyses on these simulations was performed on the final 100 ns of the simulations.
+The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
+with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
+Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
+Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
+switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.
+
+Analyses on these simulations was performed on the final 100~ns of the simulations.
 
@@
 using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
-and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
+and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
  \\
@@
 POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
-potassium or sodium ions corresponding concentrations of $\sim$450 mM and 890 mM
+potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
 were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
 Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
-parameters given by the CHARMM-GUI at 298 K. For further details, simulation files
+parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
 and data, see table \ref{mixedIONsystemsMONOVALENT} and
@@
 %The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
-and 0.26 M and 1 M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
+and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
 Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
-by the CHARMM-GUI at 298 K.
+by the CHARMM-GUI at 298~K.
 
@@
 The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
-The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na+ ions) and was converted into
+The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
 the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
 Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
-van der Waals interactions were truncated at 1.0 nm with a dispersion correction applied to the energy and pressure;
-Coulombic interactions were truncated at 1.0 nm with PME used to treat interactions beyond this cut-off.
+van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
+Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
 All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
@@
 The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
-\cite{essman95} was used for electrostatics. Dispersion correction \cite{shirts07} 
+\cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
 was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
@@
 to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
-and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
+and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
 ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
@@
 Each bilayer structure was first minimized for 2500 steps with steepest
-descent and an additional 2500 step with the conjugent gradient
-algorithms while restrainig the solvent and ion molecules with a force
-constant 500~kcal~mol$^{-1}$~\AA{}$^{-1}$. An additional minimization
+descent and an additional 2500 step with the conjugate gradient
+algorithms while restraining the solvent and ion molecules with a force
+constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
 step was used with the same parameters after removing the constraints on
@@
 Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
-Pressure was coupled to the Parrinello-Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
+Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
 The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
@@
 The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na$^+$ ions
-was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
+was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
 The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
@@
 The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath.
-Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
+Pressure was kept constant at 1.013~bar using a semi-isotropic Parrinello--Rahman
 barostat~\cite{parrinello81} with a time constant of 10.0~ps.
@@
 Previously published simulation \cite{ollila07a} available from Ref. \citenum{bergerPOPC310K}
-was used for POPC at 310 K.
+was used for POPC at 310~K.
 
@@
 and the SPC water model was used~\cite{berendsen81}. The simulation cut-offs closely matched
-those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0 nm;
-Coulombic interactions were truncated at 1.0 nm with PME applied to account for the interactions beyond the cut-off.
+those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0~nm;
+Coulombic interactions were truncated at 1.0~nm with PME applied to account for the interactions beyond the cut-off.
 All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
@@
 were as discussed above. The SPC water model was used in these simulations~\cite{berendsen81}. Simulation cut-offs were chosen
-to closely match those used in the original GROMOS-CKP force field validation: van der Waals interactions were truncated at 1.4 nm
-with a dispersion correction applied to the energy and pressure; Coulombic interactions were also truncated at 1.4 nm with PME used
+to closely match those used in the original GROMOS-CKP force field validation: van der Waals interactions were truncated at 1.4~nm
+with a dispersion correction applied to the energy and pressure; Coulombic interactions were also truncated at 1.4~nm with PME used
 to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
@@
 Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
-to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
+to the CaCl$_2$ concentrations of $\sim$ 50--300mM, depending on the amount of negatively charged
 lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
@@
     with different mole fractions of negatively charged lipids (left column).
-    Right column shows the same data zoomed to the concentrations below 1M.
+    Right column shows the same data zoomed to the concentrations below 1~M.
     Data for Pure DPPC from Ref.~\citenum{akutsu81},
@@
     for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
-    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from Ref.~\citenum{macdonald87}
+    for POPC:POPG (8:2,1:1) mixtures with 0.1~M NaCl from Ref.~\citenum{macdonald87}
     and for POPC:POPG (1:1) mixture data without NaCl from Ref.\citenum{borle85}.
@@
 
-That said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
+That being said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
 bound charge are typically smaller than then discrepancies arising from ion binding affinity.
@@
 (blue lines in Fig. \ref{DPslices}), 0.315 is the scaling factor of the R18 recoupling sequence 
-and 21.5 kHz is the maximum $^1$H-$^{13}$C dipolar coupling for a C--H bond \cite{dvinskikh04}.
+and 21.5~kHz is the maximum $^1$H--$^{13}$C dipolar coupling for a C--H bond \cite{dvinskikh04}.
 %Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
@@
 As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
-simulations gave values +0.09 and -0.02 for the $\alpha$-carbon order parameters.
-To corroborate our interpretation, we measured the SDROSS curve also using the higher 8 kHz
+simulations gave values $+0.09$ and $-0.02$ for the $\alpha$-carbon order parameters.
+To corroborate our interpretation, we measured the SDROSS curve also using the higher 8~kHz
 MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
@@
     The assignment of columns is given in the middle.  
-    The SDROSS slices were measured using both 5 KHz (middle) and 8 kHz (bottom) because different MAS frequencies enable the
+    The SDROSS slices were measured using both 5~KHz (middle) and 8~kHz (bottom) because different MAS frequencies enable the
     higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
     The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
-    are used to determine the error bars for $\alpha$ and $\beta$ carbons in figure \ref{PShgSIGNSsimpson} in the main text.
+    are used to determine the error bars for $\alpha$ and $\beta$ carbons in Fig.~\ref{PShgSIGNSsimpson} in the main text.
     The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
@@
 The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
-wide variety between different simulation models (Figs. \ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
+wide variety between different simulation models (Figs.~\ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
 Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
@@
 of PS headgroup can be suggested based on the best available models
-(Figs. \ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
+(Figs.~\ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
 The glycerol backbone structures are not discussed in this work because our focus is on PS headgroup.
@@
     compared between the POPC and POPS lipids. 
-    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
+    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
   }
@@
     compared between the POPC and POPS lipids.
-    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
+    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
   }
@@
     from CHARMM36 and Slipids simulations compared between the POPC and POPS lipids.
-    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
+    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
   }
@@
 we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
-as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
+as a function of different monovalent ions added to the buffer (Fig.~\ref{PSresponseTONaCl}). 
 Experimental order parameters of POPC headgroup in the mixture are available as a function
@@
 The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
-but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
+but increase with the addition of sodium or potassium (Fig.~\ref{PSresponseTONaCl}). 
 POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
@@
 to added sodium and potassium are not in line with the experiments.
-Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.
-\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
+Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.~\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
 in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
-(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
-less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
+(Fig.~\ref{CIdensPSOCmixt}), but significantly larger error bars and
+less systematic changes in the order parameters (Fig.~\ref{PSresponseTONaCl}).
 Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
@@
 for which the data is not shown due to the formation of
-ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
-Appearance of such clusters in the MagRog simulations with 4 M KCl
+ion clusters in water with relatively low (1~M) ion concentrations (Fig.~\ref{ionCLUSTERS}).
+Appearance of such clusters in the MagRog simulations with 4~M KCl
 could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
-In conclusion, the results are in line with the section \ref{ciBINDINGsection}
+In conclusion, the results are in line with the Section \ref{ciBINDINGsection}
 in the main text, suggesting that the MacRog simulations with KCl give the most
@@
     (except Berger simulations are (4:1) mixture at 310~K).
-    The experimental data is from Ref. \citenum{roux90}.
+    The experimental data is from Ref.~\citenum{roux90}.
     The values from counterion-only systems are set as a zero point of y-axis.
@@
   \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
-  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
-    and MacRog simulations with 4 M of KCl (right).
+  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1~M of NaCl (left)
+    and MacRog simulations with 4~M of KCl (right).
 \label{ionCLUSTERS}
@@
 are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM-GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
-(Fig. \ref{changesWITHCaClPS} in the main text),
+(Fig.~\ref{changesWITHCaClPS} in the main text),
 indicating that the calcium binding to the bilayer is too weak with these
 parameters. The response of headgroup order
-parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
-(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
+parameters (Fig.~\ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
+(Fig.~\ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
 POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
@@
     Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
-    with the NBfix term in sodium at Ref. \citenum{CHARMM36withCaClNBfix}, 
-    with the NBfix term in calcium at Ref. \citenum{CHARMM36withCaClNBfix} and
-    without the NBfix in calcium at Ref. \citenum{charmmPOPC450mMCaClfiles}.
+    with the NBfix term in sodium at Ref.~\citenum{CHARMM36withCaClNBfix}, 
+    with the NBfix term in calcium at Ref.~\citenum{CHARMM36withCaClNBfix} and
+    without the NBfix in calcium at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
   }
@@
     and without (top) the NBfix term for calcium \cite{kim16} compared to the simulation
-    with the NBfix term for sodium \cite{venable13} (bottom). The simulation data are the same as in figure \ref{OP_CHARMM_CaCl_POPC_NBFix}.
+    with the NBfix term for sodium \cite{venable13} (bottom). The simulation data are the same as in Fig.~\ref{OP_CHARMM_CaCl_POPC_NBFix}.
   }
@@
 \noindent
-{\it Fernando Favela}
+{\it Fernando Favela-Rosales}
+Setup and performed one of the DOPS simulations with Slipids.
 

### 2019-05-23 Manuscript/manuscriptPSsuppl.tex

--- 2019-05-22 Manuscript/manuscriptPSsuppl.tex
@@
 including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.
+Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
+the simulations with Verlet cut-off ($\sim$~52\AA$^2$).
 

### 2019-05-27 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

@@
 PS headgroup is more rigid than the phosphatidylcholine (PC)
-owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
+owing to electrostatic interactions or the formation of a hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
 While most monovalent ions interact weakly with
@@
 ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
-Unfortunately, the results have depended on the force field used.
+Unfortunately, the results have depended on the particular force field used.
 For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
-CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
+the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
 suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
 the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
-rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
+rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also toward the phosphate region \cite{hallock18}.
 Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
 combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
-calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
+calcium binding to the carbonyls in the acyl chains \cite{melcrova16}.
 
@@
 measured from NMR experiments with high accuracy and directly compared to simulations
-in order to evaluate the quality of the simulation model, or to interpret the experiments \cite{ollila16}. Using this approach,
+in order to evaluate the quality of the simulation model and/or to interpret the experiments \cite{ollila16}. Using this approach,
 it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
@@
 conformation, stabilized by hydrogen bonds or electrostatic
-interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
+interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
 available. 
@@
 and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
-as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full agreement with the experiments.
-The models perform generally less well for PS than for PC
+as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full quantitative agreement with the experiments.
+Generally, the models perform less well for PS than for PC
 (Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}).
-This complicates the interpretation of structural differences between PC and PS headgroups.
-However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon observed in the experiments.
+This complicates the interpretation of structural differences between the PC and PS headgroups.
+However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon that is observed in the experiments.
 Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
@@
 These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
-However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.
+However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.
 
@@
 with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
-The results are in line with the previous study suggesting that the
+The results are consistent with the previous study suggesting that the
 low areas per molecule in PS lipid bilayers originate from the combination
 of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
+%Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the specific lipid headgroup/ion configurations that are sampled in simulations.
+My feeling is that "fully hydrated" lipids might carry with them differing amounts of water between the different FFs..
+-JJM
 
@@
 while the other tested models overestimate the counterion
-binding affinity. The results are in line with the behaviour of headgroup order
+binding affinity. The results are consistent with the behavior of headgroup order
 parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
@@
 tested simulation models---including CHARMM36---which underestimated the binding affinity.
-Importantly, the different simulation models predict qualitatively different behaviour
+Importantly, the different simulation models predict qualitatively different behavior
 for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
@@
   and CSC-IT center for science for computational resources.
+  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
   OHSO acknowledges financial support from Academy of Finland (315596),

### 2019-05-27 Manuscript/manuscriptPS.tex
Add GROMOS-CKP POPC results with increasing amount of POPS, delivered by T. Piggot by email

@@
     Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
-    The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
-    the error bar is given as the difference between the results divided by two. {\color{red} Why divided by two?}
+    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
+    the error bar is given as the difference between the results divided by two.
+    % {\color{red} Why divided by two?}
+    % Because then both values fit within the error bar.
   }
-%  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
-%    add GROMOS-CKP data into this plot.}
 \end{figure}
@@
 of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
-%Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the specific lipid headgroup/ion configurations that are sampled in simulations.
-My feeling is that "fully hydrated" lipids might carry with them differing amounts of water between the different FFs..
--JJM
+% Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the
+% specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated"
+% lipids might carry with them differing amounts of water between the different FFs.. -JJM
 
@@
 
-Experimentally, the headgroup order parameters of POPC
+Experimentally, the both headgroup order parameters of POPC
 increase when negatively charged POPS lipids are incorporated in lipid bilayer
@@
 which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
-The CHARMM36 and Berger simulations either exhibit no change or show a decrease
-in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
+The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
+in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
 Therein, the stronger counterion binding cancels

### 2019-05-27 Manuscript/manuscriptPS.tex
Update text based on comments by Markus Miettinen delivered in Skype

@@
 Phosphatidylserine (PS) is the most common negatively
-charged lipid in eukaryotic membranes. In red blood cells, for example,
+charged lipid type in eukaryotic biomembranes. In red blood cells, for example,
 PS lipids compose 8.5\% of the total lipid weight. 
@@
 Experimental studies have indicated that the
-PS headgroup is more rigid than the phosphatidylcholine (PC)
-owing to electrostatic interactions or the formation of a hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
+PS headgroup is more rigid than the phosphatidylcholine (PC),
+owing possibly to electrostatic interactions or formation of a hydrogen bond network between the headgroups~\cite{browning80,buldt81}.
 While most monovalent ions interact weakly with
@@
 Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
-That being said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
-to negatively charged and zwiterionic phospholipids, and that
+That being said, some studies suggest that Ca$^{2+}$ has similar specific binding affinity 
+to both negatively charged and zwitterionic phospholipids, and that
 the increased cation binding to negatively charged lipid bilayers arises only due
@@
 \centering
-\caption{The list of MD simulations of pure PS bilayers without additional salt along with the references to the force fields used and the MD trajectories.
+\caption{List of MD simulations of pure PS bilayers without additional salt. % along with the references to the force fields used and the MD trajectories.
   Notation 2$\times$[time] indicates that two independent MD runs were conducted.
@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
+lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
 \hline 
@@
 
-The molecular level interpretation of these observations is currently lacking, and classical molecular dynamics (MD) simulations have been widely used in efforts 
+The molecular level interpretation of these observations is currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
 to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
@@
 Unfortunately, the results have depended on the particular force field used.
-For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
+For example, recent simulations using the NBfix parameters for Ca$^{2+}$~\cite{kim16} in
 the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
@@
 We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
-can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
+can be used to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
 measured from NMR experiments with high accuracy and directly compared to simulations
-in order to evaluate the quality of the simulation model and/or to interpret the experiments \cite{ollila16}. Using this approach,
+to either evaluate the quality of the simulation model and/or to interpret the experiments \cite{ollila16}. Using this approach,
 it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
@@
 
-Here, we extend the available set of experimentally measured PS lipid headgroup and
+Here, we first extend the available set of experimentally measured PS lipid headgroup and
 glycerol backbone C--H bond order parameters
 by measuring the signs of these order parameters using S-DROSS solid-state NMR spectroscopy.
-The quality of headgroup structures and the ion binding affinity in
+Based on the collected experimental data, we then assess
+the quality of headgroup structures and the ion binding affinity in
 %MD simulations of lipid bilayers containing PS lipids
-the available MD simulation models of PS lipids are then assessed based on the collected experimental data.
+the available MD simulation models of PS lipids.
 %using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
-The results pave the way
+Our results pave the way
 for the development of lipid models that correctly describe 
@@
 conditions. Such force fields are expected to be extremely useful in understanding
-biological function of lipid headgroups and glycerol backbone, as
+the biological functions of lipid headgroups and glycerol backbone, as
 these are known to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.
@@
 \centering
-\caption{The list of POPC:POPS mixtures simulated with different molar fractions and different amounts of added calcium. 
+\caption{List of POPC:POPS mixtures simulated at different molar fractions and different amounts of added CaCl$_2$. 
   The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-  This corresponds the concentration in buffer before solvating lipids, which are
+  This corresponds to the concentrations in buffer before solvating lipids, which were
   reported in the experiments by Roux et al.~\cite{roux90}.
@@
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The simulation details are given in the supplementary information.
+  The additional simulation details are given in the supplementary information.
   $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
@@
 \begin{tabular}{lccccccccc}
-lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
+lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
 \hline 
@@
 POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
-usage scaled ions. Concentration taken as reported in the delivered
+use of scaled ions. Concentration taken as reported in the delivered
 data.}  & 104  & 4306  & 24  & 310  & 300  & 100  & \cite{POPCpopsBERGERwith102mMCa} \tabularnewline
@@
 
-The headgroup and glycerol backbone C--H bond order parameter magnitudes of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
+The magnitudes of headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
 were determined by measuring the chemical-shift resolved dipolar splittings
@@
 The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03~MHz.
-Magic angle spinning (MAS) of the sample was used at a frequency of 5.15~kHz (R-PDLF experiment) and 5~kHz (S-DROSS experiment).
+Magic angle spinning (MAS) of the sample was used at a frequency of 5.15~kHz (R-PDLF) and 5~kHz (S-DROSS).
 The following experimental setups were used.
@@
 the standard error of the mean over the different lipids. The analysis can be conducted using a
-Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that utilizes the
+Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that uses the
 MDAnalysis library \cite{agrawal11,gowers16}.
@@
 
-\subsection{Comparison of ion binding to negatively charged lipid bilayers 
-between simulations and experiments using the molecular electrometer concept}
+\subsection{Using the molecular electrometer concept to compare ion binding to negatively charged lipid bilayers 
+in simulations and in experiments}
 
-The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
+The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
 decrease proportionally to the amount of positive
@@
 comparison between simulations and experiments, as
-the headgroup order parameters at varying cation
+the headgroup $S_{\rm CH}$ at varying cation
 concentrations can be easily calculated from
-simulations~\cite{catte16}. The headgroup order parameters
+simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
 of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
-
-Therefore, measuring the PC headgroup order parameters from 
+Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
 mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
 
-Calibrating the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer concept.
-This can be done using experimental data from mixtures of
-monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
-(see SI section \ref{electrometerCALIBRATION}). Additionally, we quantify the response of PC headgroup order parameters
-to the negatively charged PS, which also follows the molecular electrometer
-concept in experiments~\cite{scherer87} (see SI section \ref{electrometerFORmixtures}).
+Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
+This can be done using experimental data from mixtures of PC and
+monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
+see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
+Additionally, we here quantify the response of PC headgroup $S_{\rm CH}$
+to the negatively charged PS (see SI section \ref{electrometerFORmixtures}), which also follows the molecular electrometer
+in experiments~\cite{scherer87}.
 
-Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
-the concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
+Studies applying the molecular electrometer have used two different definitions for salt concentration:
+The concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
 In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
-Nevertheless, in this work we use this former definition to be consistent with the reference
-experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
+However, the choice of definition has only a marginal effect
 to the results in simulations with realistic ion binding affinity
 (see SI section \ref{concentrationDEFsection}).
+In this work we use this former definition to be consistent with the reference
+experimental data \cite{roux90}.
 
@@
 
-\subsection{Headgroup and glycerol backbone order parameters of POPS from $^{13}$C NMR}
+\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
 
@@
 
-The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
+The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
 carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
-The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13} whereas
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
 the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
-known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
-Slices of the R-PDLF spectra and the resulting order parameter values
+known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
+Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
 are shown in the supplementary information (Fig. \ref{DPslices}).
@@
 Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
-only the absolute values of order parameters, we determined the signs of the PS headgroup
-order parameters using the S-DROSS experiment \cite{gross97}.
+only the absolute values of  $S_{\rm CH}$, we determined the signs of the PS headgroup
+ $S_{\rm CH}$ using the S-DROSS experiment \cite{gross97}.
 For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
 of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
-We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C--H
-bond order parameter give rise to profiles that initially increase and decrease, respectively.
+We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the $S_{\rm CH}$ 
+give rise to profiles that initially increase and decrease, respectively.
 In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
@@
 For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
-However, the beginning of its S-DROSS slice suggests that the larger order parameter of the alpha-carbon is positive and the
-decrease towards negative values at longer $t_1$ suggests that the smaller order parameter is negative.    
+However, the beginning of its S-DROSS slice suggests that the larger $S_{\rm CH}$ of the $\alpha$-carbon is positive and the
+decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
 %
@@
 This is confirmed by a SIMPSON simulation
-using the order parameter values of $+0.09$ from dipolar coupling measured here (Fig. \ref{DPslices})
-and $-0.02$ from the $^2$H\,NMR experiment~\cite{roux91}.
-The literature value for the smaller order parameter was used because the
+using the  $S_{\rm CH}$ values of $+0.09$ from the dipolar coupling measured here (Fig. \ref{DPslices})
+and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
+We used the literature value for the smaller  $S_{\rm CH}$ , because the
 resolution of our R-PDLF experiment was not sufficient to determine the
-small value of the order parameter.
-The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
+magnitude small value.
+The S-DROSS curve from SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
 (dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
-corroborating the interpretation that the smaller order parameter is negative.
+corroborating the interpretation that the smaller  $S_{\rm CH}$  is negative.
 
@@
 When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
-the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
+the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative and $\alpha$-carbon
+experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
 
-We note that the the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
+We note that the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
 the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
 (approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
-gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
+the gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
 Larger multilamellar repeat distances are expected in the samples with lower lipid
 concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
-Yet the PS headgroup order parameters measured from gravimetric samples (POPS) in this work
+Yet the PS headgroup  $S_{\rm CH}$  measured from gravimetric samples (POPS) in this work
 are in good agreement with the results from centrifuged samples~\cite{browning80}.
@@
 
-The different PS MD models produce highly varied headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
-as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full quantitative agreement with the experiments.
-Generally, the models perform less well for PS than for PC
-(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}).
-This complicates the interpretation of structural differences between the PC and PS headgroups.
+The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
+As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
+In fact, the models perform less well for PS than for PC
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
+which complicates the interpretation of structural differences between the PC and PS.
 However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon that is observed in the experiments.
-Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig.
-\ref{HGorderParametersPS} vs.
-Fig.
-2 in Ref.~\citenum{botan15}), in line with
+forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
+Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
+for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
 experiments (Fig. \ref{HGorderParameters}).
@@
 
-Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
-dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
-The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
+Interestingly, in the three models that best fit the experimental data,
+%have a narrower distribution in
+the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
+other models yield binodal distributions (Fig. \ref{dihedralsHG}).
+The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
 suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
-In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-angle distribution for PS than for PC headgroup in
-CHARMM36 and Slipids simulations that have the best agreement with experiments
-(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
+On the other hand, the C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedral angle rotates relatively
+freely between approximately 100$^{\circ}$ and 300$^{\circ}$ in the best three models (as seen also in Fig.~\ref{HGstructuresPSandPC} (B)),
+while other models yield more restricted conformations (Fig. \ref{dihedralsHG}).
+
+In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
+the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
+angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
 % Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
@@
 The variance was already noted above
-
 These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.

### 2019-05-27 Manuscript/manuscriptPS.tex
Update text based on comments by Markus Miettinen delivered by email

@@
   \caption{\label{HGorderParametersPS}
-    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
-    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
+    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
+    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
     and MD simulations with different force fields (symbols).
     Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
-    while all the other data are without additional salt.
-    The data for DOPS is at 303~K and the data for POPS is at 298~K.
+    while all the other data are with counterions only.
+    The data for DOPS is at 303~K and for POPS at 298~K.
     Light blue areas span 0.04 units around the average of the extremal experimental values,
@@
     Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
-    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
-    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
+    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N dihedrals
+    of the PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.
   }
@@
     Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
-    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
+    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
     The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
@@
 Membranes containing PS lipids are always accompanied with counterions that
-modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
+modulate electrostatic interactions between the lipids and other biomolecules. MD simulations have suggested
 that counterions reduce the area per lipid of PS bilayers compared to PC
 bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
-We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
+We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, see Fig.~\ref{NAdensPOPS}.
 The force fields studied show significant differences in both binding affinity
@@
 with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
-The results are consistent with the previous study suggesting that the
+The results are consistent with a previous study suggesting that the
 low areas per molecule in PS lipid bilayers originate from the combination
@@
 by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
-Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions (the lack of a ion-free reference state) and the analysis is further complicated by the artificial aggregation of counterions
-observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
+Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic
+counterions (the lack of an ion-free reference state). The analysis is further complicated by the artificial aggregation of counterions
+observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI).
 % \todo{Why isn't this a problem with PC}.
@@
 (although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
-but by studying the changes of the headgroup order parameters with increasing amount of
+but by studying the changes of the headgroup $S_{\rm CH}$ order parameters with increasing amount of
 negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
@@
 Experimentally, the both headgroup order parameters of POPC
-increase when negatively charged POPS lipids are incorporated in lipid bilayer
+increase when negatively charged POPS lipids are incorporated in the bilayer
 (section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
@@
 Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
-parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
-and more accurate force fields are needed
+parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
+More accurate force fields are needed
 to correctly describe the PC--PS headgroup interactions in MD simulations.
@@
 Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
-exhibit a strong dependence of CaCl$_2$ at small concentrations yet they saturate already at a concentration below 100~mM (Fig. \ref{changesWITHCaClPS}).
+exhibit a strong response to small concentrations of CaCl$_2$, which saturates below 100~mM (Fig. \ref{changesWITHCaClPS}).
 The $\beta$-carbon order parameter increases with added CaCl$_2$,
@@
 for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
-For example, both order parameters decrease in Berger, but increase
+For example, both order parameters decrease in Berger, increase
 in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
@@
 %Here (as was previously done for PC lipids \cite{botan15,catte16})
-Here, we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
+We used the headgroup C--H bond order parameters, $S_{\rm CH}$, and the open collaboration approach to evaluate the quality
 of the headgroup structure and the ion binding affinity 
-in available MD models of PS.
+in available MD models of PS lipid.
 The main advantage of this approach is the direct connection
@@
 reproduce the PS headgroup order parameters within the experimental accuracy. However,
-the best models for the serine headgroup suggested a characteristic rigid conformation for its
+the best models for the PS headgroup suggested a characteristic rigid conformation for its
 carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
@@
 affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
-NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
-The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
-qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
+NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
+The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
 MD force fields to study the biological function of PS lipids.
@@
 The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures.
-The cation binding in the force fields can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations.
+The cation binding can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations.
 This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.

### 2019-05-29 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex

--- 2019-05-23 Manuscript/manuscriptPSsuppl.tex
@@
 {\it Batuhan Kav}
-
+Set up, performed, and analyzed Amber Lipid14/17 simulations with the Amber MD. 
 \noindent
 {\it Jesper J. Madsen}
+Set up, performed and analyzed several of the CHARMM36 simulations. Provided comments on the manuscript.
 

### 2019-06-14 Manuscript/manuscriptPS.tex
Update abstract after discussion in issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/94

@@
 We then use this data set
-to assess,
-in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
-the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
-We find that none of the MD models reproduce the NMR data within experimental accuracy.
-The best MD models, however, suggest that the carboxyl group in the PS headgroup
-does not rotate freely. In line with our previous results for PC lipids,
+to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
+in a wide range of available molecular dynamics (MD) force fields of PS.
+We find that none of the MD simulations reproduce the NMR data within experimental accuracy.
+However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
+In line with our previous results for PC lipids,
 none of the PS force fields correctly captures the cation binding affinity.

### 2019-06-16 Manuscript/manuscriptPS.tex

@@
 \author{Tiago M. Ferreira}
-\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
+\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
 \author{Ivan Gushchin}
@@
 been available. Here we
-gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
+gather a set of experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
 PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
-and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
-We then use this data set
+and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the POPS headgroup using S-DROSS solid-state NMR spectrsocopy.
+This data set is then used
 to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
-in a wide range of available molecular dynamics (MD) force fields of PS.
-We find that none of the MD simulations reproduce the NMR data within experimental accuracy.
+in the most commonly used classical molecular dynamics (MD) models including CHARMM36, Lipid17, Macrog, Slipids, GROMOS-CKP, Berger and variants.
+We show large discrepancies between different force-fields and that none of the models reproduces the NMR data within experimental accuracy.
 However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
@@
 of lipid headgroups and the details of cation binding
-are crucial for understanding PS-mediated processes on cellular membranes.
+is crucial for understanding PS-mediated processes on cellular membranes.
 
 Experimental studies have indicated that the
-PS headgroup is more rigid than the phosphatidylcholine (PC),
+serine headgroup in PS is more rigid than the choline headgroup in phosphatidylcholine (PC),
 owing possibly to electrostatic interactions or formation of a hydrogen bond network between the headgroups~\cite{browning80,buldt81}.
@@
 to both negatively charged and zwitterionic phospholipids, and that
-the increased cation binding to negatively charged lipid bilayers arises only due
-to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.
+the increased cation binding to PS lipid bilayers is non-specific and arises only due to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.
 
@@
 captured by most MD force fields~\cite{botan15}, that the cation binding to PC
-lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
+lipid bilayers is overestimated~\cite{catte16}, and that the inequivalent order parameters of the distinct C--H bonds at
 carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
@@
 The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
-as sodium salt) with water (lipid:water 60:40~wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
-until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4~mm NMR rotors.
+as sodium salt) with water (lipid:water 60:40~wt-\%) in an Eppendorf tube, centrifuging the mixture and stirring with a thin glass rod repeatedly (approximately 5 to 6 times centrifuging/stirring) until a homogeneous viscous fluid was visually observed.
+Then 20 mg of the sample was transferred to an NMR insert suitable for 4~mm NMR rotors.
 
@@
   \caption{\label{PShgSIGNSsimpson}
-    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
+    The headgroup and glycerol backbone region of the (A) Refocused-INEPT spectrum and
     (B) 2D R-PDLF spectra.
@@
 resolution of our R-PDLF experiment was not sufficient to determine the
-magnitude small value.
-The S-DROSS curve from SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
+magnitude of the small value.
+The S-DROSS curve from the SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
 (dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
@@
     of the PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.
-  }
+    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
+{ \todo{TMF: I find these pictures very interesting but at the same time a bit difficult to follow and I am not sure if i get them right.
+In each picture two dihedrals distributions are depicted right?
+so there are two pictures superimposed as far as I understand.
+Maybe this could be shortly explained in the caption or using a link to the methods section?}}
 
@@
     % Because then both values fit within the error bar.
-  }
-\end{figure}
+  }\end{figure}
 
@@
 
-Experimentally, the both headgroup order parameters of POPC
-increase when negatively charged POPS lipids are incorporated in the bilayer
+Experimentally, the $S_{\rm{CH}}$ values of the $\alpha$ and $\beta$ headgroup carbons of POPC increase when negatively charged POPS lipids are incorporated in the bilayer 
 (section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.

### 2019-06-16 Manuscript/manuscriptPSsuppl.tex
Add author contribution by Nencini

@@
 {\it Tiago M. Ferreira}
+Performed the NMR experiments and NMR simulations, processed and analysed the experimental data, prepared the corresponding figures, wrote parts of the manuscript (NMR methods and interpretation of the SDROSS results) and took part in the final revision.
 
@@
 {\it Ricky Nencini}
+Run the simulations of POPC systems using CHARMM36 with and without NB-Fix and contributed to a discussion of system/bulk concentration.
 

### 2019-06-18 Manuscript/manuscriptPS.tex

@@
   }
-  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
-  We can do this just before submission when citation numbers will not change anymore.}
+  %\todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
+  %We can do this just before submission when citation numbers will not change anymore.}
+  %SAMULI: Maybe better to do in revision stage because citations might still change.
 \end{figure}
@@
   \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
-    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N dihedrals
-    of the PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
+    of lipid headgroups.
+    (A) Overlaying  C$_\gamma$-C$_\beta$-C$_\alpha$ carbons demonstrate fluctuations around O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
+    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup and
+    (B) overlaying C$_\beta$-C$_\alpha$-O$_\alpha$ carbonds demonstrate fluctuations around N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
     dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
-{ \todo{TMF: I find these pictures very interesting but at the same time a bit difficult to follow and I am not sure if i get them right.
+  %{ \todo{TMF: I find these pictures very interesting but at the same time a bit difficult to follow and I am not sure if i get them right.
 In each picture two dihedrals distributions are depicted right?
@@
 Maybe this could be shortly explained in the caption or using a link to the methods section?}}
-
+  %SAMULI: I have now updated the caption to contain information about the atoms used in the overlay. Hopefully it is more clear now? 
 \end{figure}
@@
 
-\listoftodos
+%\listoftodos
 

### 2019-06-18 Manuscript/manuscriptPSsuppl.tex
Update pdfs and update Fig. 4 caption based on the question by Tiago Ferreira in pull request https://github.com/NMRLipids/NMRlipidsIVotherHGs/pull/97
Author contribution MSM.

--- 2019-05-29 Manuscript/manuscriptPSsuppl.tex
@@
 {\it Hanne Antila}
-Contiributed to the development analysis tools used for evaluating the force fields,  provided critical discussion on the manuscript content and edited it for clarity.
+contributed to the development of analysis tools used for evaluating the force fields,  provided critical discussion on the manuscript content, and edited it for clarity.
 
@@
 {\it Pavel Buslaev}
-Analyzed the dihedral angle distributions in the head group and glycerol backbone regions.
+analysed the dihedral angle distributions in the head group and glycerol backbone regions.
 
@@
 {\it Fernando Favela-Rosales}
-Setup and performed one of the DOPS simulations with Slipids.
+set up and performed one of the DOPS simulations with Slipids.
 
@@
 {\it Tiago M. Ferreira}
+performed the NMR experiments and NMR simulations, processed and analysed the experimental data, prepared the corresponding figures, wrote parts of the manuscript (NMR methods and interpretation of the SDROSS results), and took part in the final revision.
 
@@
 {\it Ivan Gushchin}
-Supervised the work of P.B.
+supervised the work of P.B.
 
@@
 {\it Matti Javanainen}
-Performed most of the MacRog simulations and provided comments on the manuscript.
+performed most of the MacRog simulations, and provided comments on the manuscript.
 
@@
 {\it Batuhan Kav}
-Set up, performed, and analyzed Amber Lipid14/17 simulations with the Amber MD. 
+set up, performed, and analysed Amber Lipid14/17 simulations with the Amber MD.
+
 \noindent
 {\it Jesper J. Madsen}
-Set up, performed and analyzed several of the CHARMM36 simulations. Provided comments on the manuscript.
+set up, performed, and analysed several of the CHARMM36 simulations. Provided comments on the manuscript.
 
@@
 consulted the project, 
-prepared tools for calculating order parameters
+prepared tools for calculating order parameters,
 and performed several simulations.
@@
 \noindent
-{\it Markus Miettinen}
+{\it Markus S. Miettinen}
+worked with
+H. A. on the analysis tools, and with
+B. K. on the Amber MD simulations.
+Made the subjective force field ranking.
+Edited the manuscript.
 
@@
 {\it Ricky Nencini}
+ran the simulations of POPC systems using CHARMM36 with and without NB-Fix, and contributed to the discussion on system/bulk concentration.
 
@@
 {\it O. H. Samuli Ollila}
-Designed the project and managed the work.
+designed the project and managed the work.
 Ran and analysed several simulations. Wrote the manuscript.
@@
 \noindent
-{\it Thomas Piggot}
-Setup, performed and analysed many of the simulations. Contributed to parts of the manuscript.
+{\it Thomas J. Piggot}
+set up, performed, and analysed many of the simulations. Contributed to parts of the manuscript.
 

### 2019-06-20 Manuscript/manuscriptPSsuppl.tex
Unifying author contributions

@@
 {\it Batuhan Kav}
-set up, performed, and analysed Amber Lipid14/17 simulations with the Amber MD.
+set up, performed, and analysed Amber Lipid14/17 simulations with the Amber MD simulation package.
 
@@
 {\it Josef Melcr}
-consulted the project, 
-prepared tools for calculating order parameters,
-and performed several simulations.
+prepared tools for calculating order parameters, performed Lipid14/17 simulations with Gromacs MD simulation package,
+and provided comments on the manuscript.
 
@@
 {\it Ricky Nencini}
-ran the simulations of POPC systems using CHARMM36 with and without NB-Fix, and contributed to the discussion on system/bulk concentration.
+ran the simulations of POPC systems using CHARMM36 with and without the NB-Fix corrections,
+and analyzed the difference between bulk and buffer concentrations.
 

### 2019-06-22 Manuscript/manuscriptPSsuppl.tex
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/99

@@
   \caption{\label{HGandGLYstructuresPS}
-    Overlayed snapshots of the glycerol backbone and headgroup regions from different POPS simulations.
+    Overlayed snapshots of the headgroup (C$_\gamma$-C$_\beta$-C$_\alpha$ carbons overlayed) and
+    glycerol backbone (C$_{\rm g1}$-C$_{\rm g2}$-C$_{\rm g3}$ carbons overlayed) regions from different POPS simulations.
+    Note that only one of the two O$_\gamma$ atoms in the carboxylate group from PS lipids is shown.
   }

### 2019-06-22 Manuscript/manuscriptPS.tex
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/99
Change manuscript to the ACS format

@@
-%% ****** Start of file aiptemplate.tex ****** %
-%%
-%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
-%%   Version 4.1 of 9 October 2009.
-%%
-%
-% This is a template for producing documents for use with 
-% the REVTEX 4.1 document class and the AIP substyles.
-% 
-% Copy this file to another name and then work on that file.
-% That way, you always have this original template file to use.
+\documentclass[journal=jpcbfk,manuscript=article]{achemso}
 
-%\documentclass[aip,graphicx]{revtex4-1}
-%\documentclass[aip,reprint]{revtex4-1}
-
-%\usepackage{graphicx}
-
-%\draft % marks overfull lines with a black rule on the right
-%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
-%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}
-
-\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}
-
-%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}
+\usepackage[version=3]{mhchem}
+\usepackage[T1]{fontenc}
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+\newcommand{\todo}[1]{\textcolor{red}{#1}}
 
 \usepackage{rotating}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{setspace}
-\usepackage{amsmath}
-\usepackage{epstopdf}
-\usepackage[obeyFinal]{easy-todo}
-\usepackage{csquotes}
+\usepackage{upgreek}				
+\usepackage{xcolor}
+\usepackage{booktabs}
+\usepackage{multirow}
+\usepackage{lmodern}
+\usepackage{microtype}
 \usepackage{xr}
 \externaldocument{manuscriptPSsuppl}
-
-\begin{document}
-
-% Use the \preprint command to place your local institutional report number 
-% on the title page in preprint mode.
-% Multiple \preprint commands are allowed.
-%\preprint{}
-
-\title{Headgroup structure and cation binding in phosphatidylserine lipid bilayers} %Title of paper
-
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
-% Explanatory text should go in the []'s, 
-% actual e-mail address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro for the type of information
-
-% \affiliation command applies to all authors since the last \affiliation command. 
-% The \affiliation command should follow the other information.
+\usepackage{soul} % for highlights with \hl{}
 
@@
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\author{Markus S. Miettinen}
+\author{Markus Miettinen}
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
@@
 \author{O. H. Samuli Ollila}
-\email[]{samuli.ollila@helsinki.fi}
+\email{samuli.ollila@helsinki.fi}
 %\homepage[]{Your web page}
@@
 \affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
-%\author{NMRlipids collaboration}
-%\affiliation{nmrlipids.blogspot.fi}
 
-% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
-% \noaffiliation is required (may also be used with the \author command).
-%\collaboration{}
-%\noaffiliation
+\title{Headgroup structure and cation binding in phosphatidylserine lipid bilayers}
 
-\date{\today}
+\begin{document}
+
+%\begin{tocentry}%
+%
+%Some journals require a graphical entry for the Table of Contents.
+%This should be laid out ``print ready'' so that the sizing of the
+%text is correct.%
+%
+%Inside the \texttt{tocentry} environment, the font used is Helvetica
+%8\,pt, as required by \emph{Journal of the American Chemical                    
+%Society}.
+%
+%The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
+%permitted for  \emph{Journal of the American Chemical Society}
+%graphical table of content entries. The box will not resize if the
+%content is too big: instead it will overflow the edge of the box.
+%
+%This box and the associated title will always be printed on a
+%separate page at the end of the document.
+%
+%\end{tocentry}
 
 \begin{abstract}
-  % insert abstract here
 Phosphatidylserine (PS) is a negatively
@@
 
-% Body of paper goes here. Use proper sectioning commands. 
-% References should be done using the \cite, \ref, and \label commands
-
-%\label{}
 \section{Introduction}
@@
 
-\begin{table*}[htb]
-%\begin{sidewaystable*}[!p]
-\centering
-\caption{List of MD simulations of pure PS bilayers without additional salt. % along with the references to the force fields used and the MD trajectories.
-  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
-  JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  Additional simulation details are given in the supplementary information.
-  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
-    %   The lipid force fields named as in our previous work~\cite{botan15}. 
-}\label{PSsystems}
-%begin{minipage}[t]{\textwidth}
-\begin{tabular}{lcrrrrrcc}
-%\hline
-% some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline
-\hline 
-POPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36POPS298K}  & \tabularnewline
-POPS/K$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36POPS298Kpotassium}  & \tabularnewline
-POPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline
-POPS/Na$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{macrogPOPS298Kcorrect}  & \tabularnewline
-POPS/K$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 200  & 150  & \cite{macrogPOPS298KwithK}  & \tabularnewline
-POPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
-POPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
-POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
-POPS/Na$^{+}$  & GROMOS-CKPM \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
-POPS/Na$^{+}$  & GROMOS-CKP \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline
-POPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{slipidsPOPS298K}  & \tabularnewline
-\hline 
-DOPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36DOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
-DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
-DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{slipidsDOPS303K}  & \tabularnewline
-DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 288  & 11232  & 303  & 200  & 100  & \cite{slipidsDOPSfiles}  & \tabularnewline
-\end{tabular}
-%\end{minipage}
-%\end{sidewaystable*} 
-\end{table*}
-
 The molecular level interpretation of these observations is currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
@@
 lipid bilayers is overestimated~\cite{catte16}, and that the inequivalent order parameters of the distinct C--H bonds at
-carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
-%
-%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
-%\todo{Should we leave the mention of ECC out from the Introduction
-%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}
+carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.
 
@@
 the quality of headgroup structures and the ion binding affinity in
-%MD simulations of lipid bilayers containing PS lipids
 the available MD simulation models of PS lipids.
-%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
 Our results pave the way
@@
 
-\begin{table*}[tb]
-%\begin{sidewaystable*}[!p]
-\centering
-\caption{List of POPC:POPS mixtures simulated at different molar fractions and different amounts of added CaCl$_2$. 
-  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-  This corresponds to the concentrations in buffer before solvating lipids, which were
-  reported in the experiments by Roux et al.~\cite{roux90}.
-  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
-  Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The additional simulation details are given in the supplementary information.
-  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
-   % The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{mixedIONsystems}
-%begin{minipage}[t]{\textwidth}
-\begin{tabular}{lccccccccc}
-lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
-\hline 
-%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \cite{POPC5POPS1noCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36pops+83popcT298Kpiggot} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \cite{POPC5POPS1withCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \cite{POPC5POPS1with1MCaClCHARMM} \tabularnewline
-POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \cite{POPC1POPS1noCaClCHARMM} \tabularnewline
-\hline 
-POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \cite{macrogPOPC298K} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 & 250 &  \cite{POPCpopsMACROG}\tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.30  & 120  & 5760  & 31  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 1.00  & 120 & 5760  & 104  & 298  & 600  & 300  & \cite{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 3.00  & 120 & 5760  & 311  & 298  & 600 & 300 & \cite{POPCpopsMACROG}\tabularnewline
-\hline 
-%    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withKCI} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withNaCI} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \cite{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60 & 3600  & 0  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1050  & 1000  & \cite{lipid17_cacl_series} \tabularnewline
-\hline 
-POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102 & 4290  & 0  & 310  & 120  & 80  & \cite{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
-POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
-use of scaled ions. Concentration taken as reported in the delivered
-data.}  & 104  & 4306  & 24  & 310  & 300  & 100  & \cite{POPCpopsBERGERwith102mMCa} \tabularnewline
-POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104 & 4306  & 72  & 310  & 300  & 100  & \cite{POPCpopsBERGERwith715mMCa} \tabularnewline
-\hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline
-\end{tabular}
-%\end{minipage}
-%end{sidewaystable*}
-
-\end{table*}
-
 \section{Methods}
 
-\subsection{Experimental C--H bond order parameters}% from the natural abundance $^{13}$C\,NMR}
+\subsection{Experimental C--H bond order parameters}
 
@@
 
+%\begin{table*}[htb]
+\begin{sidewaystable*}[!p]
+\centering
+\caption{List of MD simulations of pure PS bilayers without additional salt. 
+  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
+  JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
+  Additional simulation details are given in the supplementary information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
+}\label{PSsystems}
+%begin{minipage}[t]{\textwidth}
+\begin{tabular}{lcrrrrrcc}
+%\hline
+% some footnotes are not visible in typeset-MS (pdf)
+lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline
+\hline 
+POPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36POPS298K}  & \tabularnewline
+POPS/K$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36POPS298Kpotassium}  & \tabularnewline
+POPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36uaPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{macrogPOPS298Kcorrect}  & \tabularnewline
+POPS/K$^{+}$  & MacRog \cite{maciejewski14}  & 128  & 4480  & 298  & 200  & 150  & \citenum{macrogPOPS298KwithK}  & \tabularnewline
+POPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \citenum{lipid17POPSjcions}  & \tabularnewline
+POPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \citenum{lipid17POPSff99ions}  & \tabularnewline
+POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{bergerPOPS298K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKPM \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{ckp1POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & GROMOS-CKP \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{ckp2POPS303K}  & \tabularnewline
+POPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \citenum{slipidsPOPS298K}  & \tabularnewline
+\hline 
+DOPS/Na$^{+}$  & CHARMM36 \cite{venable13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{charmm36DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{charmm36uaDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \citenum{lipid17DOPSjcions}  & \tabularnewline
+DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \citenum{lipid17DOPSff99ions}  & \tabularnewline
+DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{bergerDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{ckp1DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{ckp2DOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \citenum{slipidsDOPS303K}  & \tabularnewline
+DOPS/Na$^{+}$  & Slipids \cite{jambeck13}  & 288  & 11232  & 303  & 200  & 100  & \citenum{slipidsDOPSfiles}  & \tabularnewline
+\end{tabular}
+%\end{minipage}
+\end{sidewaystable*} 
+%\end{table*}
+
+%\begin{table*}[tb]
+\begin{sidewaystable*}[!p]
+\centering
+\caption{List of POPC:POPS mixtures simulated at different molar fractions and different amounts of added CaCl$_2$. 
+  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  This corresponds to the concentrations in buffer before solvating lipids, which were
+  reported in the experiments by Roux et al.~\cite{roux90}.
+  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
+  Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
+  The additional simulation details are given in the supplementary information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
+  Symbols have the same meaning as in Table \ref{PSsystems}.
+}\label{mixedIONsystems}
+%begin{minipage}[t]{\textwidth}
+\begin{tabular}{lccccccccc}
+  %lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
+  lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & N$_{{\rm l}}$  & N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & T (K)  & t$_{{\rm sim}}$(ns)  & t$_{{\rm anal}}$ (ns)  & files\tabularnewline
+\hline 
+%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \citenum{POPC1POPS1noCaClCHARMM} \tabularnewline
+\hline 
+POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \citenum{macrogPOPC298K} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 & 250 &  \citenum{POPCpopsMACROG}\tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.30  & 120  & 5760  & 31  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 1.00  & 120 & 5760  & 104  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 3.00  & 120 & 5760  & 311  & 298  & 600 & 300 & \citenum{POPCpopsMACROG}\tabularnewline
+\hline 
+%    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withKCI} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withNaCI} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60 & 3600  & 0  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+\hline 
+POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102 & 4290  & 0  & 310  & 120  & 80  & \citenum{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
+POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
+use of scaled ions. Concentration taken as reported in the delivered
+data.}  & 104  & 4306  & 24  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith102mMCa} \tabularnewline
+POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104 & 4306  & 72  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith715mMCa} \tabularnewline
+\hline 
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
+\end{tabular}
+%\end{minipage}
+\end{sidewaystable*}
+
+%\end{table*}
+
 The C--H bond order parameters were calculated directly
@@
 (see SI section \ref{concentrationDEFsection}).
-In this work we use this former definition to be consistent with the reference
+In this work we use the former definition (concentration before solvating the lipids) to be consistent with the reference
 experimental data \cite{roux90}.
@@
 \subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
+The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
+carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
+the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
+known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
+Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
+are shown in the supplementary information (Fig. \ref{DPslices}).
 
@@
 
-The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
-carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
-The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
-the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
-known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
-Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
-are shown in the supplementary information (Fig. \ref{DPslices}).
-
 Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
@@
 decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
-%
-%The S-DROSS slice clearly shows that the order parameter of
-%the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
-%Could we explain in a few words how this is clearly seen?}
-%which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
-%suggests that the larger order parameter of the $\alpha$-carbon 
-%is positive and the deviation towards negative values with longer T$_1$ times suggests
-%that the smaller order parameter is negative.
-%
 This is confirmed by a SIMPSON simulation
@@
 available.
-%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results.
-This also enables referring to figures in order.
-I also removed the refs which do not seem to refer to the actual data used in the figure.
-%Rewrote the paragraph, see if ok.
 
@@
 \subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
+
+The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
+As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
+In fact, the models perform less well for PS than for PC
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
+which complicates the interpretation of structural differences between the PC and PS.
+However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
+Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
+for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
+experiments (Fig. \ref{HGorderParameters}).
 
@@
     dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
-  %{ \todo{TMF: I find these pictures very interesting but at the same time a bit difficult to follow and I am not sure if i get them right.
-In each picture two dihedrals distributions are depicted right?
-so there are two pictures superimposed as far as I understand.
-Maybe this could be shortly explained in the caption or using a link to the methods section?}}
-  %SAMULI: I have now updated the caption to contain information about the atoms used in the overlay. Hopefully it is more clear now? 
 \end{figure}
@@
 
-The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
-As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
-In fact, the models perform less well for PS than for PC
-(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
-which complicates the interpretation of structural differences between the PC and PS.
-However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
-Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
-for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
-experiments (Fig. \ref{HGorderParameters}).
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
-  \caption{\label{NAdensPOPS}
-    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
-    of POPS lipid bilayer from simulations with different force fields.
-    The vertical green bars indicate the location of the phosphate density peak.
-    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
-  }
-%  \todo{Commented by M. Javanainen in blog:
-%    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
-\end{figure}
-
 Interestingly, in the three models that best fit the experimental data,
-%have a narrower distribution in
 the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
@@
 angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs.
-\ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup.
-There is very little actual information in this statement.
-The variance was already noted above
 These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
@@
 \subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
-
-\begin{figure}[!tb]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
-  \caption{\label{HGorderparametersPCvsPS}
-    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
-    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
-    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
-    the error bar is given as the difference between the results divided by two.
-    % {\color{red} Why divided by two?}
-    % Because then both values fit within the error bar.
-  }\end{figure}
-
-\begin{figure}[!tb]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
-  \caption{\label{HGorderparametersPSvsPC}
-    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
-    Experimental values with the signs are measured for pure POPS system in this work.
-    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
-    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
-    the error bars of 0.02 are used here \cite{botan15,ollila16}.
-    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
-    with the same value for both order parameters such that the lower order
-    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
-    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
-    the error bars are given as the difference between the results divided by two. {\color{red} Why divided by two?}
-  }
-\end{figure}
 
@@
 of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
-% Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the
-% specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated"
-% lipids might carry with them differing amounts of water between the different FFs.. -JJM
 
-\begin{figure*}[ht]
+\begin{figure}[!htb]
   \centering
-  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
-  \caption{\label{changesWITHCaClPS}
-   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
-    at 298~K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
-    The order parameter values from systems without calcium are set as the zero point of y-axis,
-    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
-    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
-    the forking with different concentrations of calcium.
-    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
-    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
+  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
+  \caption{\label{NAdensPOPS}
+    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
+    of POPS lipid bilayer from simulations with different force fields.
+    The vertical green bars indicate the location of the phosphate density peak.
+    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
   }
-%  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
-%    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
-%  these are not necessary, but could be added here if delivered.} \\
-\end{figure*}
+\end{figure}
+
+\begin{figure}[!tb]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
+  \caption{\label{HGorderparametersPCvsPS}
+    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
+    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
+    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
+    the error bar is given as the difference between the results divided by two.
+  }\end{figure}
+
+\begin{figure}[!tb]
+  \centering
+  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
+  \caption{\label{HGorderparametersPSvsPC}
+    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
+    Experimental values with the signs are measured for pure POPS system in this work.
+    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
+    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
+    the error bars of 0.02 are used here \cite{botan15,ollila16}.
+    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
+    with the same value for both order parameters such that the lower order
+    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
+    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
+    the error bars are given as the difference between the results divided by two. 
+  }
+\end{figure}
 
@@
 observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI).
-% \todo{Why isn't this a problem with PC}.
-% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
-% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
-% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
 Therefore, we evaluate here the amount of bound charge not by adding salt
@@
 to correctly describe the PC--PS headgroup interactions in MD simulations.
-%Having this conclusion in every paragraph gets a bit repetive
 
 \subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
-\begin{figure}[tb]
+
+\begin{figure*}[tbp]
   \centering
-  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
-  \caption{\label{CAdensPCPSmixture}
-    Number density profiles of Ca$^{2+}$ (solid line) and Cl$^-$ (dashed line) from POPC:POPS (5:1) mixtures
-    simulated with different force fields. The vertical green bars indicate the location of the phosphate density peak.
-    The smallest simulated CaCl$_2$ concentrations are shown.
-    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
+  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
+  \caption{\label{changesWITHCaClPS}
+   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
+    at 298~K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
+    The order parameter values from systems without calcium are set as the zero point of y-axis,
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
+    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
+    the forking with different concentrations of calcium.
+    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
+    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
   }
-\end{figure}
+\end{figure*}
 
@@
 indicating a weaker binding affinity than in experiments.
-%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
 Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
@@
 \ref{CAdensPCPSmixture}).
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
+  \caption{\label{CAdensPCPSmixture}
+    Number density profiles of Ca$^{2+}$ (solid line) and Cl$^-$ (dashed line) from POPC:POPS (5:1) mixtures
+    simulated with different force fields. The vertical green bars indicate the location of the phosphate density peak.
+    The smallest simulated CaCl$_2$ concentrations are shown.
+    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
+  }
+\end{figure}
 
@@
 observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
-%These ring like structures are a widespread feature of typical Berger based lipid force
-%fields containing explicit hydrogen atoms in the head group
-%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
 Therefore, we conclude that
@@
 
-%Lipids with PS headgroups, and their interactions with ions, play an important role in lipid-mediated signaling processes \cite{leventis10,yeung08}.
-Attemps to use MD simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PSheadgroups \cite{melcrova16,valentine18,hallock18}.
-%Here (as was previously done for PC lipids \cite{botan15,catte16})
 We used the headgroup C--H bond order parameters, $S_{\rm CH}$, and the open collaboration approach to evaluate the quality
@@
 This revealed that none of the force fields
-%tested using the NMRlipids open collaboration
 reproduce the PS headgroup order parameters within the experimental accuracy. However,
@@
 
-% Tables may be be put in the text as floats.
-% Here is an example of the general form of a table:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
-% \begin{tabular}{} command.
-%
-% \begin{table}
-% \caption{\label{} }
-% \begin{tabular}{}
-% \end{tabular}
-% \end{table}
-
-% If you have acknowledgments, this puts in the proper section head.
-\begin{acknowledgments}
-% Put your acknowledgments here.
+\begin{acknowledgement}
   MJ acknowledges financial support from the Emil Aaltonen foundation and
@@
   LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
-\end{acknowledgments}
+\end{acknowledgement}
 
-\newpage
+%\newpage
 

### 2019-06-26 Manuscript/manuscriptPS.tex
Move table a bit

@@
 \begin{sidewaystable*}[!p]
+\vspace{1cm}
   \centering
@@
   The additional simulation details are given in the supplementary information.
-  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
   Symbols have the same meaning as in Table \ref{PSsystems}.
@@
 \hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
 \end{tabular}

### 2019-08-15 Manuscript/manuscriptPS.tex
Updates based on reviewer's comments

@@
 charged lipid type in eukaryotic biomembranes. In red blood cells, for example,
-PS lipids compose 8.5\% of the total lipid weight. 
-The abundance, however, varies between different organelles, and up to
-25--35\% of the cytosolic leaflet of plasma membranes 
-\cite{lemmon08,leventis10,li14} consists of PS lipids.
+PS lipids compose 8.5\% of the total lipid weight~\cite{lemmon08}. 
+The abundance, however, varies between different cells and organelles, and up to
+25--35\% of the cytosolic leaflet of plasma membranes consists of PS lipids~\cite{leventis10,li14}.
 PS lipids are vastly important biomolecules that interact with
@@
 
-The molecular level interpretation of these observations is currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
+Molecular level interpretations of the rigidity of PS headgroup and its interactions with ions
+are currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
 to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
@@
 and can therefore be used to measure the ion binding affinity.
+In addition to ions, the correlation between bound charge and headgroup order parameter change
+is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
 This concept, known as the molecular electrometer, is especially useful for 
@@
 see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
-Additionally, we here quantify the response of PC headgroup $S_{\rm CH}$
-to the negatively charged PS (see SI section \ref{electrometerFORmixtures}), which also follows the molecular electrometer
-in experiments~\cite{scherer87}.
+Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
+charged PS follows the molecular electrometer in experiments~\cite{scherer87},
+which we also quantify here (see SI section \ref{electrometerFORmixtures}).
 
@@
   \caption{\label{comparisonTablePS}
-    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
-    Here {\textsf{\small M}} indicates a magnitude problem, {\textsf{\small F}} a forking problem; letter size increases with problem severity.
+    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
+    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
+    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
+    in Figure~\ref{HGorderParametersPS}.
+    The letter size increases with problem severity.
 Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
@@
 the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
-angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
-These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments.
-However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.
+angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
+which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
+The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
+however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.
 
@@
 This revealed that none of the force fields
-reproduce the PS headgroup order parameters within the experimental accuracy. However,
-the best models for the PS headgroup suggested a characteristic rigid conformation for its
-carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
+reproduce the PS headgroup order parameters within the experimental accuracy.
+However, the best models captured essential differences with PC and
+suggested characteristic conformations of PS headgroups.
+Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
 concentrations \cite{roux90} then showed that the tested MD force fields
@@
 
-We expect our results to pave the way for the development of better MD force fields
-for PS lipids.
-The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures.
-The cation binding can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations.
-This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.
+We expect our results to pave the way for the development of better MD force fields for PS lipids.
+The quality of conformational ensembles produced by the various models can be evaluated
+based on the headgroup order parameters, which we hope to guide the development of force fields towards
+models that correctly reproduce the PS headgroup structures. Also, the ensembles observed in the simulations
+with the most realistic headgroup conformations (CHARMM36 and Slipids) hint the direction for the force
+field improvement. The cation binding can be improved based on the experimental headgroup order parameter
+data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies showed that the
+cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
+polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
+the electronic polarizability effects are important for lipid-ion interactions but polarizable force fields
+may not be necessary to correctly capture the ion binding to lipid bilayers.
 

### 2019-08-20 Manuscript/manuscriptPSsuppl.tex
Change manuscript to the ACS format

@@
-\documentclass[journal=jctcce]{achemso}
+\documentclass[journal=jpcbfk,manuscript=article]{achemso}
 
@@
 \author{Tiago M. Ferreira}
-\affiliation{Halle, Germany}
+\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
 \author{Ivan Gushchin}
@@
 \email{samuli.ollila@helsinki.fi}
-%\homepage[]{Your web page}
+%%\homepage[]{Your web page}
 \affiliation{Institute of Organic Chemistry and Biochemistry of the 
@@
 \clearpage
-\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
+%\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
 
 \section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
-In Figure~\ref{HGorderParametersPS}) of main text we present a rough and subjective ranking of the force fields investigated in this work. 
+In Figure~\ref{comparisonTablePS} of main text we present a rough and subjective ranking of the force fields investigated in this work. 
 The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.

### 2019-08-20 Manuscript/manuscriptPS.tex

@@
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{lccccccccc}
+\resizebox{0.87\textwidth}{!}{\begin{tabular}{lccccccccc}
   %lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
@@
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
 POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} NBFIX 1 \cite{kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} NBFIX 1 \cite{kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 320  & 400  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} NBFIX 2 \cite{han2018graph}  & 0.14  & 250 & 11190  & 28  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} NBFIX 2 \cite{han2018graph}  & 0.94  & 250 & 11174  & 189  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
 POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \citenum{POPC1POPS1noCaClCHARMM} \tabularnewline
@@
 POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
-\end{tabular}
+\end{tabular}}
 %\end{minipage}
 \end{sidewaystable*} 
-
 %\end{table*}

### 2019-08-20 Manuscript/manuscriptPS.tex
Updates related to the simulations with NBfix2 paraeters

--- 2019-08-15 Manuscript/manuscriptPS.tex
@@
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
 POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 320  & 400  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.14  & 250 & 11190  & 28  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.94  & 250 & 11174  & 189  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
 POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \citenum{POPC1POPS1noCaClCHARMM} \tabularnewline

### 2019-08-20 Manuscript/manuscriptPSsuppl.tex
Updates related to the simulations with NBfix2 paraeters

@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium} 
+{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.} 
 POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
@@
 %The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
-and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
+and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in Janyary 2018 \cite{lee16,jo18}.
+The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
 Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
-by the CHARMM-GUI at 298~K.
+by the CHARMM-GUI at 298~K. \\
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.}
+\todo{Ricky, please update the simulation methods.}
+Systems were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
+The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}.
 
@@
 
-\section{Calcium binding to POPC in CHARMM36 simulation with NBfix}\label{CHARMMcalciumNBfix}
+\clearpage
+\section{Calcium binding to POPC in CHARMM36 simulation with NBfix1 parameters}\label{CHARMMcalciumNBfix}
 
 The response of POPC headgroup order parameters to the CaCl$_2$ concentration
-are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM-GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
-(Fig.~\ref{changesWITHCaClPS} in the main text),
-indicating that the calcium binding to the bilayer is too weak with these
-parameters. The response of headgroup order
+are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing the
+NBfix %(as obtained from CHARMM-GUI in January 2018)
+parameters for the interactions between calcium and lipid oxygens \cite{kim16,han2018graph}
+(Fig.~\ref{changesWITHCaClPS} in the main text), indicating that the calcium binding to the
+bilayer is too weak with these parameters. The response of headgroup order
 parameters (Fig.~\ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
-(Fig.~\ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
-POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
-terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium.
-Without the NBfix term, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
-
-%HA: this just repeats the above? 
-%In conclusion, the special NBfix~\cite{kim16} for calcium, 
-%incorporated in the parameters give by the CHARMM-GUI at the time
-%of running the simulations (January 2018), leads to the underestimation of
-%calcium binding affinity to pure PC and mixed PC:PS bilayers.
+(Fig.~\ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium
+were underestimated also in simulations of pure POPC bilayer with the NBfix1 parameters~\cite{kim16}.
+Notably, CHARMM36 simulations with the NBfix1 terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium.
+Because the results for POPC:POPS (5:1) mixture were very similar with both NBfix parameters~\cite{kim16,han2018graph},
+the NBfix2 parameters were not separately tested with pure POPC bilayers.
+Without the NBfix parameters, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
 
@@
 prepared tools for calculating order parameters, performed Lipid14/17 simulations with Gromacs MD simulation package,
-and provided comments on the manuscript.
+prepared the TOC figure and provided comments on the manuscript.
 
@@
 ran the simulations of POPC systems using CHARMM36 with and without the NB-Fix corrections,
+ran the simulations of POPC:POPS (5:1) systems using CHARMM36 with the NBfix2 correction,
 and analyzed the difference between bulk and buffer concentrations.

### 2019-08-20 Manuscript/manuscriptPS.tex

@@
 The sole exception is the CHARMM36 model when paired with the NBfix
-correction for calcium~\cite{kim16}. For this combination, the modulation of order parameters is underestimated, 
+corrections for calcium~\cite{kim16,han2018graph}. For these combinations, the modulation of order parameters is underestimated, 
 indicating a weaker binding affinity than in experiments.
@@
 calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
-is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16}, but overestimated 
+is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
 in all the other tested models.

### 2019-08-20 Manuscript/manuscriptPSsuppl.tex
change in the numbering of pages of SI completed method section in SI covering CHARMM36 NBFIX2 simulations

@@
 \usepackage{soul} % for highlights with \hl{}
+
+\setcounter{page}{0}
 
@@
 %The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
-and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in Janyary 2018 \cite{lee16,jo18}.
+and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
 The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.}
-\todo{Ricky, please update the simulation methods.}
-Systems were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
+{\it POPC:POPS (5:1) mixtures with  additional calcium using NBfix2 parameters.}
+POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids, neutralizing sodium ions, 
+and 0.14~M or 0.94~M CaCl$_2$ were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
 The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}.
+Systems were simulated using Gromacs 2018.6 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
+by the CHARMM-GUI at 320~K.  \\
 

### 2019-08-21 Manuscript/manuscriptPS.tex
Updates based on editorial comments

@@
 
-\title{Headgroup structure and cation binding in phosphatidylserine lipid bilayers}
+\title{Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers}
 
@@
 searchable database available at \url{www.nmrlipids.fi},
-and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).
+and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH})~\cite{MATCHgit}.
 
@@
 
+\begin{suppinfo}
+
+%A listing of the contents of each file supplied as Supporting Information 
+%should be included. For instructions on what should be included in the 
+%Supporting Information as well as how to prepare this material for 
+%publications, refer to the journal's Instructions for Authors.
+
+The following files are available free of charge. 
+\begin{itemize} 
+  \item SI.pdf: Additional simulation, experimental and methodological details, and further analysis. 
+\end{itemize}
+
+\end{suppinfo}
+
 \begin{acknowledgement}
@@
 % Create the reference section using BibTe
-\bibliography{refs.bib}
+\bibliography{refsU.bib}
 

### 2019-08-21 Manuscript/manuscriptPS.tex
change format for number of lipids in a table in the main text

--- 2019-08-20 Manuscript/manuscriptPS.tex
@@
   %lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
-  lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & N$_{{\rm l}}$  & N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & T (K)  & t$_{{\rm sim}}$(ns)  & t$_{{\rm anal}}$ (ns)  & files\tabularnewline
+  lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & N$_{{\rm l}}$  & N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in the system}N$_{{\rm c}}$  & T (K)  & t$_{{\rm sim}}$(ns)  & t$_{{\rm anal}}$ (ns)  & files\tabularnewline
 \hline 
 %    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 0.26  & 250 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 1.06  & 250 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250  & 11207  & 0  & 320  & 400  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.14  & 250 & 11190  & 28  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.94  & 250 & 11174  & 189  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
-POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150  & 10785  & 0  & 298  & 200  & 180  & \citenum{POPC1POPS1noCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298KpiggotSODIUM} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 110:22  & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{charmm36pops+83popcT298Kpiggot} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250:50  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 0.26  & 250:50 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 1.06  & 250:50 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250:50  & 11207  & 0  & 320  & 400  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.14  & 250:50 & 11190  & 28  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix2 \cite{han2018graph}  & 0.94  & 250:50 & 11174  & 189  & 320  & 440  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
+POPC:POPS (1:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 150:150  & 10785  & 0  & 298  & 200  & 180  & \citenum{POPC1POPS1noCaClCHARMM} \tabularnewline
 \hline 
-POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120 & 5120  & 0  & 298  & 200  & 150  & \citenum{macrogPOPC298K} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120 & 5760  & 0  & 298  & 400 & 250 &  \citenum{POPCpopsMACROG}\tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120  & 5760  & 10  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.30  & 120  & 5760  & 31  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 1.00  & 120 & 5760  & 104  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
-POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 3.00  & 120 & 5760  & 311  & 298  & 600 & 300 & \citenum{POPCpopsMACROG}\tabularnewline
+POPC:POPS (1:0)  & MacRog \cite{maciejewski14}  & 0  & 120:0 & 5120  & 0  & 298  & 200  & 150  & \citenum{macrogPOPC298K} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0  & 120:24 & 5760  & 0  & 298  & 400 & 250 &  \citenum{POPCpopsMACROG}\tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.10  & 120:24  & 5760  & 10  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 0.30  & 120:24  & 5760  & 31  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 1.00  & 120:24 & 5760  & 104  & 298  & 600  & 300  & \citenum{POPCpopsMACROG} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & MacRog \cite{maciejewski14}  & 3.00  & 120:24 & 5760  & 311  & 298  & 600 & 300 & \citenum{POPCpopsMACROG}\tabularnewline
 \hline 
 %    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
-POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withKCI} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withNaCI} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60 & 3600  & 0  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.08  & 60  & 3561  & 5  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.13  & 60  & 3561  & 8  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.20  & 60 & 3561  & 13  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.41  & 60 & 3522  & 26  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.62  & 60 & 3483  & 39  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/K$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120:24  & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withKCI} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120:24 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withNaCI} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120:24 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120:24  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120:24  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120:24 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120:24  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60:12 & 3600  & 0  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.08  & 60:12  & 3561  & 5  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.13  & 60:12  & 3561  & 8  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.20  & 60:12 & 3561  & 13  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.41  & 60:12 & 3522  & 26  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0.62  & 60:12 & 3483  & 39  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
 \hline 
-POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102 & 4290  & 0  & 310  & 120  & 80  & \citenum{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
+POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102:26 & 4290  & 0  & 310  & 120  & 80  & \citenum{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
 POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
 use of scaled ions. Concentration taken as reported in the delivered
-data.}  & 104  & 4306  & 24  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith102mMCa} \tabularnewline
-POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104 & 4306  & 72  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith715mMCa} \tabularnewline
+data.}  & 104:26  & 4306  & 24  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith102mMCa} \tabularnewline
+POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104:26 & 4306  & 72  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith715mMCa} \tabularnewline
 \hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
 \end{tabular}}

### 2019-09-03 Manuscript/manuscriptPS.tex
Edits to Authors

@@
 \author{Hanne Antila}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+
 \author{Pavel Buslaev}
-\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+\affiliation[Moscow Institute of Physics and Technology]{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+
 \author{Fernando Favela-Rosales}
-\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
+\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
+
 \author{Tiago M. Ferreira}
-\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
+\affiliation[Martin-Luther University Halle-Wittenberg]{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg, Germany}
+
 \author{Ivan Gushchin}
-\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+\affiliation[Moscow Institute of Physics and Technology]{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+
 \author{Matti Javanainen}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+
 \author{Batuhan Kav}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+
 \author{Jesper J. Madsen}
-\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
+\alsoaffiliation[University of South Florida]{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+
 \author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\author{Markus Miettinen}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+
+\author{Markus S. Miettinen}
+\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+
+\author{Jukka M{\"a}{\"a}tt{\"a}}
+\affiliation[Aalto University]{Department of Chemistry, Aalto University, Espoo, Finland}
+
 \author{Ricky Nencini}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+
 \author{O. H. Samuli Ollila}
@@
 %\homepage[]{Your web page}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\alsoaffiliation[University of Helsinki]{Institute of Biotechnology, University of Helsinki, Finland}
+
 \author{Thomas J. Piggot}
-\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
+\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
 
@@
 
-%\begin{tocentry}%
+\begin{tocentry}%
 %
@@
 %
-%\end{tocentry}
+  \includegraphics[width=8.5cm]{../Figs/TOC_graphic.pdf}
+\end{tocentry}
 

### 2019-09-03 Manuscript/manuscriptPSsuppl.tex
Updates based on editorial comments

@@
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\author{Jukka M{\"a}{\"a}tt{\"a}}
+\affiliation{Department of Chemistry, Aalto University, Espoo, Finland}
 \author{Ricky Nencini}
@@
 
-\title{ Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers}
+\title{ Supporting Information:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers}
 
@@
 \noindent
+{\it Jukka M{\"a}{\"a}tt{\"a}}
+Did minor analysis for and edited Figures 6 and 10.
+
+\noindent
 {\it Ricky Nencini}
@@
 \clearpage
-\bibliography{refs.bib}
+\bibliography{refsU.bib}
 

### 2019-09-03 Manuscript/manuscriptPS.tex

--- 2019-08-21 Manuscript/manuscriptPS.tex
@@
 of the headgroup structure and the ion binding affinity 
-in available MD models of PS lipid.
-The main advantage of this approach is the direct connection
-between the order parameters measured accurately in experiments and calculated readily from simulations. 
+in available MD models of PS lipids.
+The main advantage of using the $S_\mathrm{CH}$ is the direct connection they provide
+between experiments and simulations: They can be measured accurately in experiments and calculated readily from simulations. 
 This reduces the ambiguity in the interpretation of experiments.
@@
 reproduce the PS headgroup order parameters within the experimental accuracy.
-However, the best models captured essential differences with PC and
+However, the best models captured essential differences between PS and PC, and
 suggested characteristic conformations of PS headgroups.
@@
 concentrations \cite{roux90} then showed that the tested MD force fields
-overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
-affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
-NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
-The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
-qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
-MD force fields to study the biological function of PS lipids.
-This is in contrast to the previous results with PC lipids,
-where the experimentally measured headgroup order parameter responses to the bound charge
-were in qualitative agreement even though the headgroup structures themselves were
+overestimate the cation binding affinity to these bilayers.
+There were two exceptions: 1) the MacRog force field with potassium counterions, which appeared to display a more realistic monovalent ion binding
+affinity to PS-containing lipid bilayers than the other models; and 2) the CHARMM36 force field with the recently introduced
+NBfix correction for calcium \cite{kim16}, which underestimated the calcium binding affinity.
+Importantly, the experimentally measured responses of the PS headgroup $S_\mathrm{CH}$ to bound calcium, as well as to dilution of the bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields.
+This is in contrast to previous results with PC lipids,
+where MD force fields were seen to be at least in qualitative agreement with the experimentally measured headgroup $S_\mathrm{CH}$ responses to bound charge---even
+though the headgroup structures themselves were
 incorrect and the cation binding affinities were overestimated \cite{catte16}.
+This underlines the dire need for more realistic MD force fields to study the biological function of PS lipids.
 
-We expect our results to pave the way for the development of better MD force fields for PS lipids.
-The quality of conformational ensembles produced by the various models can be evaluated
-based on the headgroup order parameters, which we hope to guide the development of force fields towards
-models that correctly reproduce the PS headgroup structures. Also, the ensembles observed in the simulations
-with the most realistic headgroup conformations (CHARMM36 and Slipids) hint the direction for the force
-field improvement. The cation binding can be improved based on the experimental headgroup order parameter
-data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies showed that the
+We expect our results to pave the way for the development of better PS force fields.
+As the quality of the conformational ensembles can be evaluated
+against the headgroup C--H order parameters, we hope that the $S_\mathrm{CH}$ will guide the development of
+models that correctly reproduce the PS headgroup structures. For example, the ensembles we observed in the simulations
+with the most realistic headgroup conformations (CHARMM36 and Slipids) do hint a direction for force
+field improvement. The cation binding could be improved based on the experimental headgroup $S_\mathrm{CH}$
+data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies have shown that 
 cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
 polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
-the electronic polarizability effects are important for lipid-ion interactions but polarizable force fields
-may not be necessary to correctly capture the ion binding to lipid bilayers.
+the electronic polarizability effects are important for lipid--ion interactions but polarizable force fields
+may not be necessary to correctly capture ion binding to lipid bilayers.
 

### 2019-09-03 Manuscript/manuscriptPS.tex
Acknowledgements

--- 2019-08-21 Manuscript/manuscriptPS.tex
@@
 \begin{acknowledgement}
-  MJ acknowledges financial support from the Emil Aaltonen foundation and
-  and CSC-IT center for science for computational resources.
-  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
-  OHSO acknowledges financial support from Academy of Finland (315596),
-  Integrated Structural Biology Research Infrastructure of
-  Helsinki Institute of Life Science (Instruct-HiLIFE), and
-  CSC-IT center for science for computational resources.
-  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
+  HA acknowledges the Osk. Huttunen Foundation and the Finnish Academy of Science and Letters (Foundations' Post Doc Pool) for financial support.
   FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
@@
   H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
-  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-
-  Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
+  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
   LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
+  MJ acknowledges financial support from the Emil Aaltonen Foundation
+  and CSC -- IT Center for Science for computational resources.
+  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
+  OHSO acknowledges financial support from Academy of Finland (315596),
+  Integrated Structural Biology Research Infrastructure of
+  Helsinki Institute of Life Science (Instruct-HiLIFE), and
+  CSC -- IT Center for Science for computational resources.
+  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
 \end{acknowledgement}

### 2019-09-03 Manuscript/manuscriptPS.tex
Add mention about Han NBfix in the conclusions

@@
 affinity to PS-containing lipid bilayers than the other models; and 2) the CHARMM36 force field with the recently introduced
-NBfix correction for calcium \cite{kim16}, which underestimated the calcium binding affinity.
+NBfix corrections for calcium \cite{kim16,han2018graph}, which underestimated the calcium binding affinity.
 Importantly, the experimentally measured responses of the PS headgroup $S_\mathrm{CH}$ to bound calcium, as well as to dilution of the bilayer with zwitterionic PC lipids, were not

### 2019-09-03 Manuscript/manuscriptPS.tex
Abstract

--- 2019-09-03 Manuscript/manuscriptPS.tex
@@
 \author{Hanne Antila}
-\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Pavel Buslaev}
-\affiliation[Moscow Institute of Physics and Technology]{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
-
+\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 \author{Fernando Favela-Rosales}
-\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
-
+\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
 \author{Tiago M. Ferreira}
-\affiliation[Martin-Luther University Halle-Wittenberg]{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg, Germany}
-
+\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
 \author{Ivan Gushchin}
-\affiliation[Moscow Institute of Physics and Technology]{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
-
+\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 \author{Matti Javanainen}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-
 \author{Batuhan Kav}
-\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Jesper J. Madsen}
-\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\alsoaffiliation[University of South Florida]{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
-
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
+\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
 \author{Josef Melcr}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-
-\author{Markus S. Miettinen}
-\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-
-\author{Jukka M{\"a}{\"a}tt{\"a}}
-\affiliation[Aalto University]{Department of Chemistry, Aalto University, Espoo, Finland}
-
+\author{Markus Miettinen}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Ricky Nencini}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-
 \author{O. H. Samuli Ollila}
@@
 %\homepage[]{Your web page}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\alsoaffiliation[University of Helsinki]{Institute of Biotechnology, University of Helsinki, Finland}
-
+\affiliation{Institute of Biotechnology, University of Helsinki}
 \author{Thomas J. Piggot}
-\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
 
@@
 Phosphatidylserine (PS) is a negatively
-charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
-nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
+charged lipid type commonly found in eukaryotic membranes, where it interacts with proteins via
+nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS lipids can induce
 membrane fusion and phase separation.
@@
 been available. Here we
-gather a set of experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
+gather a set of previously published experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
 PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
-and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the POPS headgroup using S-DROSS solid-state NMR spectrsocopy.
-This data set is then used
-to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
-in the most commonly used classical molecular dynamics (MD) models including CHARMM36, Lipid17, Macrog, Slipids, GROMOS-CKP, Berger and variants.
-We show large discrepancies between different force-fields and that none of the models reproduces the NMR data within experimental accuracy.
+and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the PS headgroup using S-DROSS solid-state NMR spectroscopy.
+The augmented data set is then used
+to assess the accuracy of the PS headgroup structures in, and the cation binding to, PS-containing membranes
+in the most commonly used classical molecular dynamics (MD) force fields including CHARMM36, Lipid17, MacRog, Slipids, GROMOS-CKP, Berger, and variants.
+We show large discrepancies between different force fields, and that none of them reproduces the NMR data within experimental accuracy.
 However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
-In line with our previous results for PC lipids,
-none of the PS force fields correctly captures the cation binding affinity.
-Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}.
-The collected experimental dataset and simulation
-results pave the way for the improvement of lipid force fields that correctly
+The cation binding affinity is, in line with our previous results for PC lipids, not captured correctly by any of the PS force fields.
+Moreover, the simulated response of PS headgroup to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
+results will pave the way for development of lipid force fields that correctly
 describe the biologically relevant negatively charged membranes and their interactions with ions.
@@
 of the headgroup structure and the ion binding affinity 
-in available MD models of PS lipids.
-The main advantage of using the $S_\mathrm{CH}$ is the direct connection they provide
-between experiments and simulations: They can be measured accurately in experiments and calculated readily from simulations. 
+in available MD models of PS lipid.
+The main advantage of this approach is the direct connection
+between the order parameters measured accurately in experiments and calculated readily from simulations. 
 This reduces the ambiguity in the interpretation of experiments.
@@
 reproduce the PS headgroup order parameters within the experimental accuracy.
-However, the best models captured essential differences between PS and PC, and
+However, the best models captured essential differences with PC and
 suggested characteristic conformations of PS headgroups.
@@
 concentrations \cite{roux90} then showed that the tested MD force fields
-overestimate the cation binding affinity to these bilayers.
-There were two exceptions: 1) the MacRog force field with potassium counterions, which appeared to display a more realistic monovalent ion binding
-affinity to PS-containing lipid bilayers than the other models; and 2) the CHARMM36 force field with the recently introduced
-NBfix correction for calcium \cite{kim16}, which underestimated the calcium binding affinity.
-Importantly, the experimentally measured responses of the PS headgroup $S_\mathrm{CH}$ to bound calcium, as well as to dilution of the bilayer with zwitterionic PC lipids, were not
-qualitatively reproduced in any of the tested force fields.
-This is in contrast to previous results with PC lipids,
-where MD force fields were seen to be at least in qualitative agreement with the experimentally measured headgroup $S_\mathrm{CH}$ responses to bound charge---even
-though the headgroup structures themselves were
+overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
+affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
+NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
+The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
+qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
+MD force fields to study the biological function of PS lipids.
+This is in contrast to the previous results with PC lipids,
+where the experimentally measured headgroup order parameter responses to the bound charge
+were in qualitative agreement even though the headgroup structures themselves were
 incorrect and the cation binding affinities were overestimated \cite{catte16}.
-This underlines the dire need for more realistic MD force fields to study the biological function of PS lipids.
 
-We expect our results to pave the way for the development of better PS force fields.
-As the quality of the conformational ensembles can be evaluated
-against the headgroup C--H order parameters, we hope that the $S_\mathrm{CH}$ will guide the development of
-models that correctly reproduce the PS headgroup structures. For example, the ensembles we observed in the simulations
-with the most realistic headgroup conformations (CHARMM36 and Slipids) do hint a direction for force
-field improvement. The cation binding could be improved based on the experimental headgroup $S_\mathrm{CH}$
-data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies have shown that 
+We expect our results to pave the way for the development of better MD force fields for PS lipids.
+The quality of conformational ensembles produced by the various models can be evaluated
+based on the headgroup order parameters, which we hope to guide the development of force fields towards
+models that correctly reproduce the PS headgroup structures. Also, the ensembles observed in the simulations
+with the most realistic headgroup conformations (CHARMM36 and Slipids) hint the direction for the force
+field improvement. The cation binding can be improved based on the experimental headgroup order parameter
+data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies showed that the
 cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
 polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
-the electronic polarizability effects are important for lipid--ion interactions but polarizable force fields
-may not be necessary to correctly capture ion binding to lipid bilayers.
+the electronic polarizability effects are important for lipid-ion interactions but polarizable force fields
+may not be necessary to correctly capture the ion binding to lipid bilayers.
 

### 2019-09-04 Manuscript/manuscriptPS.tex
Structure without additional ions

--- 2019-09-03 Manuscript/manuscriptPS.tex
@@
   JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  Additional simulation details are given in the supplementary information.
-  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
+  Additional simulation details are given in the Supporting Information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.
 }\label{PSsystems}
@@
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The additional simulation details are given in the supplementary information.
+  The additional simulation details are given in the Supporting Information.
   Symbols have the same meaning as in Table \ref{PSsystems}.
@@
 \hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
 POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
@@
 Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
-are shown in the supplementary information (Fig. \ref{DPslices}).
+are shown in the Supporting Information (Fig. \ref{DPslices}).
 
@@
 
-The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
-As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
+The different PS MD models produced highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the Supporting Information).
+As was previously observed for PC lipids \cite{botan15}, also none of the PS models produced a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
 In fact, the models perform less well for PS than for PC
-(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\citenum{botan15}),
 which complicates the interpretation of structural differences between the PC and PS.
-However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
-forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
-Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
-for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
-experiments (Fig. \ref{HGorderParameters}).
+However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36, and CHARMM36-UA) indeed do qualitatively replicate the larger-than-in-PC
+forking of the $\alpha$-carbon (these models give forkings of $\sim\!0.01$ or smaller for PC~\cite{botan15} and between $\sim\!0.05$ and $\sim\!0.12$ for PS) that is observed in experiments
+(Fig.~\ref{HGorderParameters}).
+Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter
+for PS ($S^\beta_{\rm CH}\approx-0.09$, see Fig.~\ref{HGorderParametersPS}) compared to PC ($S^\beta_{\rm CH}\approx-0.03$,~see Fig.~2 in Ref.~\citenum{botan15}),
+%Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
+in line with experiments (Fig. \ref{HGorderParameters}).
 
@@
   \caption{\label{HGorderParametersPS}
-    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
+    C--H bond order parameters, $S_{\rm CH}$, of the PS headgroup ($\beta$ and $\alpha$) and
     glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
     and MD simulations with different force fields (symbols).
-    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
+    Experimental data for DOPS had been measured with 0.1\,M of NaCl~\cite{browning80},
     while all the other data are with counterions only.
     The data for DOPS is at 303~K and for POPS at 298~K.
-    Light blue areas span 0.04 units around the average of the extremal experimental values,
+    The light blue areas span 0.04 units around the average of the extremal experimental values,
     in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/structures.eps}
-  \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
-    of lipid headgroups.
-    (A) Overlaying  C$_\gamma$-C$_\beta$-C$_\alpha$ carbons demonstrate fluctuations around O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
-    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup and
-    (B) overlaying C$_\beta$-C$_\alpha$-O$_\alpha$ carbonds demonstrate fluctuations around N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
+  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
+  \caption{\label{comparisonTablePS}
+    Rough subjective ranking of force fields based on how far the simulated order parameters were from the experimental
+    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
+    of two hydrogens attached to the same carbon were reproduced ({\textsf{\small F}} indicates a forking problem)
+    in Fig.~\ref{HGorderParametersPS}.
+    The letter size increases with problem severity.
+Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
+The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
+For full details of the assessment, see section~\ref{Ranking} in the Supporting Information.
+  }
 \end{figure}
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
-  \caption{\label{comparisonTablePS}
-    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
-    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
-    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
-    in Figure~\ref{HGorderParametersPS}.
-    The letter size increases with problem severity.
-Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
-The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
-For full details of the assessment, see Supplementary Information.
-  }
+  \includegraphics[width=9.0cm]{../Figs/structures.eps}
+  \caption{\label{HGstructuresPSandPC}
+    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
+    in lipid headgroups.
+    (A) Overlaying the C$_\gamma$, C$_\beta$, and C$_\alpha$ carbons demonstrates fluctuations around the O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
+    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup.
+    (B) Overlaying the C$_\beta$, C$_\alpha$, and O$_\alpha$ atoms demonstrates fluctuations around the N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
 \end{figure}
@@
 the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
-other models yield binodal distributions (Fig. \ref{dihedralsHG}).
+the other models yield binodal distributions (Fig. \ref{dihedralsHG}).
 The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
@@
 which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
-The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
-however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.
+Indeed, the characteristic conformations of the PS headgroup suggested here can be useful when interpreting experiments---however,
+as none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to reveal the true conformational ensemble.
 

### 2019-09-04 Manuscript/manuscriptPSsuppl.tex
Structure without additional ions

--- 2019-08-20 Manuscript/manuscriptPSsuppl.tex
@@
 
-\title{ Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers}
+\title{SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers}
 
@@
 \pagebreak
-\section{Dihedral angle distributions of the headgroup and glycerol backbone
+\section{Dihedral angle distributions in headgroup and glycerol backbone
   regions of PS lipids from different simulation models}\label{Diheds}
 
-The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
-wide variety between different simulation models (Figs.~\ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
-Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
-headgroup and glycerol backbone order parameters. However, some structural characteristics
-of PS headgroup can be suggested based on the best available models
+The dihedral angle distributions (Figs.~\ref{dihedralsHG} and~\ref{dihedralsGLY}) and structures (Fig.~\ref{HGandGLYstructuresPS}) of the glycerol backbone and headgroup regions of POPS lipids showed a
+wide variety between different MD force fields.
+Detailed discussion on the relevance of the observed structural differences is limited by the lack of a realistic force field (that would correctly reproduce the
+headgroup and glycerol backbone C--H bond order parameters, $S_\mathrm{CH}$). However, some structural characteristics
+of the PS headgroup can be suggested based on the best available models for it
 (Figs.~\ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
-The glycerol backbone structures are not discussed in this work because our focus is on PS headgroup.
+The glycerol backbone structures are not discussed in this work, because our focus is on the PS headgroup.
 However, the data presented here can be useful for future investigations.
@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
-  \includegraphics[width=16.0cm]{../Figs/figS7.png}
+  \includegraphics[width=0.5\textwidth]{../Figs/PS_Labels.png}
+  \includegraphics[width=0.8\textwidth]{../Figs/figS7.png}
   \caption{\label{dihedralsHG}
-    Dihedral angle distributions of the headgroup region of POPS from different simulation models.
+    Dihedral angle distributions in the headgroup region of POPS from different simulation models.
   }
@@
   \centering
-    \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
-  \includegraphics[width=13.0cm]{../Figs/figS6.png}
+    \includegraphics[width=0.5\textwidth]{../Figs/PS_Labels.png}
+  \includegraphics[width=0.8\textwidth]{../Figs/figS6.png}
   \caption{\label{dihedralsGLY}
-    Dihedral angle distributions of the glycerol backbone region of POPS lipids from different simulation models.
+    Dihedral angle distributions in the glycerol backbone region of POPS from different simulation models.
   }
@@
   \centering
-  \includegraphics[width=18.0cm]{../Figs/figS8.png}
+  \includegraphics[width=1.03\textwidth]{../Figs/figS8.png}
   \caption{\label{HGandGLYstructuresPS}
     Overlayed snapshots of the headgroup (C$_\gamma$-C$_\beta$-C$_\alpha$ carbons overlayed) and
-    glycerol backbone (C$_{\rm g1}$-C$_{\rm g2}$-C$_{\rm g3}$ carbons overlayed) regions from different POPS simulations.
-    Note that only one of the two O$_\gamma$ atoms in the carboxylate group from PS lipids is shown.
+    glycerol backbone (C$_{\rm g_1}$-C$_{\rm g_2}$-C$_{\rm g_3}$ carbons overlayed) regions from POPS simulations with different force fields.
+    Note that only one of the two O$_\gamma$ atoms in the carboxylate group of PS lipids is shown.
   }
@@
   \centering
-  \includegraphics[width=16.0cm]{../Figs/figS11.png}
+  \includegraphics[width=0.9\textwidth]{../Figs/figS11.png}
   \caption{\label{dihedralsHGpc}
-    Dihedral angle distributions of the headgroup regions from CHARMM36 and Slipids simulations
+    Dihedral angle distributions in the headgroup region from CHARMM36 and Slipids simulations
     compared between the POPC and POPS lipids. 
@@
   \centering
-  \includegraphics[width=15.0cm]{../Figs/figS10.png}
+  \includegraphics[width=0.9\textwidth]{../Figs/figS10.png}
   \caption{\label{dihedralsGLYpc}
-    Dihedral angle distributions of the glycerol backbone regions from CHARMM36 and Slipids simulations
+    Dihedral angle distributions in the glycerol backbone region from CHARMM36 and Slipids simulations
     compared between the POPC and POPS lipids.
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/figS8_POPC.png}
+  \includegraphics[width=0.9\textwidth]{../Figs/figS8_POPC.png}
   \caption{\label{HGandGLYstructuresPSPC}

### 2019-09-04 Manuscript/manuscriptPSsuppl.tex
Structure without additional ions

@@
 \clearpage
-%\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}
 
 \section{Details of the force field ranking (Fig.~\ref{comparisonTablePS})}\label{Ranking} 
-In Figure~\ref{comparisonTablePS} of main text we present a rough and subjective ranking of the force fields investigated in this work. 
+In Fig.~\ref{comparisonTablePS} of main text we present a rough and subjective ranking of the force fields investigated in this work. 
 The assessment was based on the data presented in Fig.~\ref{HGorderParametersPS}.
@@
 \subsubsection*{Magnitude}
-To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$s) were to the 
-the force-field-produced $S_\mathrm{CH}$s, we assign a number for each carbon based on the following 5-step scale:
+To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$:s)
+the force-field-produced $S_\mathrm{CH}$:s were, we assigned a number for each carbon based on the following 5-step scale:
 
@@
 (sum of magnitude and forking deviation from 6 to 8).
-These groups are indicated by colors in Fig.~4.
+These groups are indicated by colors in Fig.~\ref{comparisonTablePS}.
 (Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
@@
 and the contributions from the five carbons were summed up.
-The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
+The sum, given in the $\Sigma$-column of Fig.~\ref{comparisonTablePS},
 was then used to (roughly and subjectively) rank the force fields.

### 2019-09-04 Manuscript/manuscriptPSsuppl.tex

@@
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
+and The Zernike Institute for Advanced Materials, 
+University of Groningen, 9747 AG Groningen, The Netherlands}
 \author{Markus Miettinen}
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\author{Jukka M{\"a}{\"a}tt{\"a}}
+\affiliation{Department of Chemistry, Aalto University, Espoo, Finland}
 \author{Ricky Nencini}

### 2019-09-09 Manuscript/manuscriptPS.tex
affiliation JM
Ca binding

--- 2019-09-04 Manuscript/manuscriptPS.tex
@@
 Phosphatidylserine (PS) is a negatively
-charged lipid type commonly found in eukaryotic membranes, where it interacts with proteins via
-nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS lipids can induce
+charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
+nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
 membrane fusion and phase separation.
@@
 been available. Here we
-gather a set of previously published experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
+gather a set of experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
 PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
-and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the PS headgroup using S-DROSS solid-state NMR spectroscopy.
-The augmented data set is then used
-to assess the accuracy of the PS headgroup structures in, and the cation binding to, PS-containing membranes
-in the most commonly used classical molecular dynamics (MD) force fields including CHARMM36, Lipid17, MacRog, Slipids, GROMOS-CKP, Berger, and variants.
-We show large discrepancies between different force fields, and that none of them reproduces the NMR data within experimental accuracy.
+and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the POPS headgroup using S-DROSS solid-state NMR spectrsocopy.
+This data set is then used
+to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
+in the most commonly used classical molecular dynamics (MD) models including CHARMM36, Lipid17, Macrog, Slipids, GROMOS-CKP, Berger and variants.
+We show large discrepancies between different force-fields and that none of the models reproduces the NMR data within experimental accuracy.
 However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
-The cation binding affinity is, in line with our previous results for PC lipids, not captured correctly by any of the PS force fields.
-Moreover, the simulated response of PS headgroup to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
-results will pave the way for development of lipid force fields that correctly
+In line with our previous results for PC lipids,
+none of the PS force fields correctly captures the cation binding affinity.
+Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}.
+The collected experimental dataset and simulation
+results pave the way for the improvement of lipid force fields that correctly
 describe the biologically relevant negatively charged membranes and their interactions with ions.
@@
   JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  Additional simulation details are given in the Supporting Information.
-  $^*$ Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.
+  Additional simulation details are given in the supplementary information.
+  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.
 }\label{PSsystems}
@@
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The additional simulation details are given in the Supporting Information.
+  The additional simulation details are given in the supplementary information.
   Symbols have the same meaning as in Table \ref{PSsystems}.
@@
 \hline 
-POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
+POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline
 POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^c$ \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPMwithNa} \tabularnewline
@@
 Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
-are shown in the Supporting Information (Fig. \ref{DPslices}).
+are shown in the supplementary information (Fig. \ref{DPslices}).
 
@@
 
-The different PS MD models produced highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
-and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the Supporting Information).
-As was previously observed for PC lipids \cite{botan15}, also none of the PS models produced a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
+The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
+and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
+As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
 In fact, the models perform less well for PS than for PC
-(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\citenum{botan15}),
+(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
 which complicates the interpretation of structural differences between the PC and PS.
-However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36, and CHARMM36-UA) indeed do qualitatively replicate the larger-than-in-PC
-forking of the $\alpha$-carbon (these models give forkings of $\sim\!0.01$ or smaller for PC~\cite{botan15} and between $\sim\!0.05$ and $\sim\!0.12$ for PS) that is observed in experiments
-(Fig.~\ref{HGorderParameters}).
-Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter
-for PS ($S^\beta_{\rm CH}\approx-0.09$, see Fig.~\ref{HGorderParametersPS}) compared to PC ($S^\beta_{\rm CH}\approx-0.03$,~see Fig.~2 in Ref.~\citenum{botan15}),
-%Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
-in line with experiments (Fig. \ref{HGorderParameters}).
+However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
+forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
+Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
+for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
+experiments (Fig. \ref{HGorderParameters}).
 
@@
   \caption{\label{HGorderParametersPS}
-    C--H bond order parameters, $S_{\rm CH}$, of the PS headgroup ($\beta$ and $\alpha$) and
+    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
     glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
     and MD simulations with different force fields (symbols).
-    Experimental data for DOPS had been measured with 0.1\,M of NaCl~\cite{browning80},
+    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
     while all the other data are with counterions only.
     The data for DOPS is at 303~K and for POPS at 298~K.
-    The light blue areas span 0.04 units around the average of the extremal experimental values,
+    Light blue areas span 0.04 units around the average of the extremal experimental values,
     in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
-  \caption{\label{comparisonTablePS}
-    Rough subjective ranking of force fields based on how far the simulated order parameters were from the experimental
-    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
-    of two hydrogens attached to the same carbon were reproduced ({\textsf{\small F}} indicates a forking problem)
-    in Fig.~\ref{HGorderParametersPS}.
-    The letter size increases with problem severity.
-Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
-The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
-For full details of the assessment, see section~\ref{Ranking} in the Supporting Information.
-  }
+  \includegraphics[width=9.0cm]{../Figs/structures.eps}
+  \caption{\label{HGstructuresPSandPC}
+    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
+    of lipid headgroups.
+    (A) Overlaying  C$_\gamma$-C$_\beta$-C$_\alpha$ carbons demonstrate fluctuations around O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
+    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup and
+    (B) overlaying C$_\beta$-C$_\alpha$-O$_\alpha$ carbonds demonstrate fluctuations around N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
+    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
 \end{figure}
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/structures.eps}
-  \caption{\label{HGstructuresPSandPC}
-    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
-    in lipid headgroups.
-    (A) Overlaying the C$_\gamma$, C$_\beta$, and C$_\alpha$ carbons demonstrates fluctuations around the O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
-    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup.
-    (B) Overlaying the C$_\beta$, C$_\alpha$, and O$_\alpha$ atoms demonstrates fluctuations around the N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
-    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
+  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
+  \caption{\label{comparisonTablePS}
+    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
+    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
+    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
+    in Figure~\ref{HGorderParametersPS}.
+    The letter size increases with problem severity.
+Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red).
+The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
+For full details of the assessment, see Supplementary Information.
+  }
 \end{figure}
@@
 the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
-the other models yield binodal distributions (Fig. \ref{dihedralsHG}).
+other models yield binodal distributions (Fig. \ref{dihedralsHG}).
 The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
@@
 which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
-Indeed, the characteristic conformations of the PS headgroup suggested here can be useful when interpreting experiments---however,
-as none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to reveal the true conformational ensemble.
+The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
+however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.
 
@@
   \centering
-  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
+  \includegraphics[width=\textwidth]{../Figs/CHANGESwithCaClPS.eps}
   \caption{\label{changesWITHCaClPS}
-   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
-    at 298~K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
-    The order parameter values from systems without calcium are set as the zero point of y-axis,
-    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
-    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
+   Variation of the POPC (left) and POPS (right) headgroup order parameters in a POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \cite{roux90} and from different force fields
+    at 298~K (except the data for Berger are from a simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
+    The $S_\mathrm{CH}$ values from systems without calcium are set at the zero point of y-axes,
+    except for the $\alpha$-carbon order parameters of POPS (bottom right), for which both $S^\alpha_\mathrm{CH}$ are shifted
+    such that the lower $S^\alpha_\mathrm{CH}$ is zero without additional ions. This is to correctly illustrate
     the forking with different concentrations of calcium.
-    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
-    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
+    Potassium counterions were used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
+    In CHARMM36 and Berger simulations with added calcium, the charge was neutralized with calcium, and monovalent counterions were not present.
   }
@@
 experimentally quantified by measuring the PC lipid headgroup order parameters
-from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
+from POPC:POPS (5:1) mixtures (see section~\ref{electrometerFORmixtures}),
 where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
 Despite the lack of an ion-free reference state
-in the presence of negatively charged lipids, our simulations give
-coherent results for POPC headgroup order parameters as a function of
-CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
+in the presence of negatively charged lipids, our simulations gave
+coherent results for the POPC headgroup order parameters as a function of
+CaCl$_2$ concentration in POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
 As expected from the previous study of pure PC lipid
-bilayers \cite{catte16}, almost all the tested simulation models overestimate the
-experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
+bilayers \cite{catte16}, almost all the tested simulation models overestimated the
+experimentally observed~\cite{roux90} decrease of the POPC headgroup $S_\mathrm{CH}$ upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
 indicating too strong calcium binding affinity.
-The sole exception is the CHARMM36 model when paired with the NBfix
-corrections for calcium~\cite{kim16,han2018graph}. For these combinations, the modulation of order parameters is underestimated, 
+The sole exception was CHARMM36 when paired with the NBfix
+corrections for calcium~\cite{kim16,han2018graph}; for these combinations, the modulation of order parameters was underestimated, 
 indicating a weaker binding affinity than in experiments.
-Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
-calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
-is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
-in all the other tested models.
-This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig.
+Notably, CHARMM36 with the NBfix corrections \cite{venable13,kim16} suggested similar binding affinities for
+calcium and sodium to a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
+is underestimated in CHARMM36 when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
+in all the other tested force fields.
+This is evident in the calcium density distributions, where almost all the Ca$^{2+}$ ions bind to the membrane interface in all models except CHARMM36 (Fig.
 \ref{CAdensPCPSmixture}).
@@
 exhibit a strong response to small concentrations of CaCl$_2$, which saturates below 100~mM (Fig. \ref{changesWITHCaClPS}).
-The $\beta$-carbon order parameter increases with added CaCl$_2$,
-whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
-the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
-tested simulation models---including CHARMM36---which underestimated the binding affinity.
-Importantly, the different simulation models predict qualitatively different behavior
-for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
-For example, both order parameters decrease in Berger, increase
-in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
+The $\beta$-carbon $S_\mathrm{CH}$ increases with added CaCl$_2$,
+whereas the larger $\alpha$-carbon $S_\mathrm{CH}$ decreases. Moreover, a slight increase is observed in
+the smaller $\alpha$-carbon $S_\mathrm{CH}$.
+In simulations, all these responses were qualitatively correct only in CHARMM36.
+In all force fields, even the qualitatively correct responses were much exaggerated---also in CHARMM36 that underestimated the Ca$^{2+}$ binding affinity---with
+the sole exception of the larger $S^\alpha_\mathrm{CH}$, whose change CHARMM36 underestimated.
+%All these changes were significantly overestimated in the tested simulation models---including CHARMM36 that underestimated the binding affinity.
+Importantly, different force fields predicted qualitatively different behavior
+for the two POPS $\alpha$-carbon order parameters as a function of added calcium:
+Both order parameters decreased in Berger, but increased
+in MacRog, whereas Lipid14/17 and CHARMM36 models exhibited more complex behaviors.
 This is in contrast to the PC headgroup, where a
@@
 structures observed in salt-free simulations~\cite{catte16}.
-The divergent response of the Berger model may arise from the ring like structures
+The divergent response of Berger may arise from the ring-like structures
 observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
 Therefore, we conclude that
-the improvement of force fields is necessary to correctly capture the interactions between the
+improvement of force fields is necessary to correctly capture the interactions between the
 PS headgroup and calcium ions in MD simulations.

### 2019-09-10 Manuscript/manuscriptPS.tex

--- 2019-09-04 Manuscript/manuscriptPS.tex
@@
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
+and The Zernike Institute for Advanced Materials, 
+University of Groningen, 9747 AG Groningen, The Netherlands}
 

### 2019-09-10 Manuscript/manuscriptPS.tex
Clarification concerning systems with Ca2+ counterions.

--- 2019-09-09 Manuscript/manuscriptPS.tex
@@
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The additional simulation details are given in the supplementary information.
+  Additional simulation details are given in the supplementary information.
   Symbols have the same meaning as in Table \ref{PSsystems}.
@@
   %lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in addition to counter ions}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline
-  lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & N$_{{\rm l}}$  & N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in the system}N$_{{\rm c}}$  & T (K)  & t$_{{\rm sim}}$(ns)  & t$_{{\rm anal}}$ (ns)  & files\tabularnewline
+  lipids/counter-ion  & force field & {\footnote{For systems with Ca$^{2+}$ counterions this column gives [Ca$^{2+}$].}[}CaCl$_{2}${]}\,(M)  & N$_{{\rm l}}$  & N$_{{\rm w}}$  & \footnote{Number of Ca$^{2+}$ cations in the system}N$_{{\rm c}}$  & T (K)  & t$_{{\rm sim}}$(ns)  & t$_{{\rm anal}}$ (ns)  & files\tabularnewline
 \hline 
@@
 POPC:POPS (5:1)/K$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250:50  & 11207  & 0  & 298  & 200  & 180  & \citenum{POPC5POPS1noCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 0.26  & 250:50 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
-POPC:POPS (5:1)  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 1.06  & 250:50 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 0.26  & 250:50 & 11190  & 53  & 298  & 200  & 180  & \citenum{POPC5POPS1withCaClCHARMM} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & CHARMM36 \cite{klauda10,venable13} / NBfix1 \cite{kim16}  & 1.06  & 250:50 & 11174  & 214  & 298  & 200  & 180  & \citenum{POPC5POPS1with1MCaClCHARMM} \tabularnewline
 POPC:POPS (5:1)/Na$^{+}$  & CHARMM36 \cite{klauda10,venable13}  & 0  & 250:50  & 11207  & 0  & 320  & 400  & 300  & \citenum{POPC5POPS1CHARMMwithNBfixHan} \tabularnewline
@@
 POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0  & 120:24 & 5760  & 0  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withNaCI} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120:24 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120:24  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120:24  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120:24 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
-POPC:POPS (5:1)  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120:24  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 0.50  & 120:24 & 5760  & 52  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 1.00  & 120:24  & 5760  & 104  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 2.00  & 120:24  & 5760  & 208  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 3.00  & 120:24 & 5760  & 311  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
+POPC:POPS (5:1)/Ca$^{2+}$  & Lipid14/17 \cite{dickson14,gould18} / ff99~\cite{aqvist90}  & 4.00  & 120:24  & 5760  & 415  & 298  & 2$\times$500  & 2$\times$200  & \citenum{POPCpopsLIPID17withCaCl} \tabularnewline
 POPC:POPS (5:1)/Na$^{+}$  & Lipid14/17 \cite{dickson14,gould18} / Dang~\cite{smith94,dang06}  & 0  & 60:12 & 3600  & 0  & 298  & 1050  & 1000  & \citenum{lipid17_cacl_series} \tabularnewline
@@
 POPC:POPS (4:1)/Na$^{+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0  & 102:26 & 4290  & 0  & 310  & 120  & 80  & \citenum{bergerPOPSPOPC4:1mixtureT310K} \tabularnewline
-POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concetration complicated by the
+POPC:POPS (4:1)/Ca$^{2+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.102\footnote{\label{noteBerger}Calculation of concentration complicated by the
 use of scaled ions. Concentration taken as reported in the delivered
 data.}  & 104:26  & 4306  & 24  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith102mMCa} \tabularnewline
-POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104:26 & 4306  & 72  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith715mMCa} \tabularnewline
+POPC:POPS (4:1)/Ca$^{2+}$  & Berger \cite{tieleman99,mukhopadhyay04}  & 0.715\textsuperscript{\ref{noteBerger}}  & 104:26 & 4306  & 72  & 310  & 300  & 100  & \citenum{POPCpopsBERGERwith715mMCa} \tabularnewline
 \hline 
@@
     as a function CaCl$_2$ concentration from experiments \cite{roux90} and from different force fields
-    at 298~K (except the data for Berger are from a simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
+    at 298~K (except the data for Berger are from a simulation of a POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
     The $S_\mathrm{CH}$ values from systems without calcium are set at the zero point of y-axes,
@@
     the forking with different concentrations of calcium.
-    Potassium counterions were used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
-    In CHARMM36 and Berger simulations with added calcium, the charge was neutralized with calcium, and monovalent counterions were not present.
+    Counterions were potassium in MacRog simulations, and sodium in Lipid14/17 Dang and CHARMM36 NBfix2 simulations.
+    In CHARMM36 NBfix1, Lipid14/17 ff99, and Berger simulations with added calcium the lipid charge was neutralized with calcium, and monovalent counterions were not present;
+    for these systems the $x$-axis shows [Ca$^{2+}$].
   }

### 2019-09-10 Manuscript/manuscriptPS.tex

--- 2019-09-10 Manuscript/manuscriptPS.tex
@@
   Dang refers to the ion parameters by Dang and co-workers~\cite{smith94,dang06}, and ff99 to the default Amber ion parameters~\cite{aqvist90}.
-  The additional simulation details are given in the Supporting Information.
+  Additional simulation details are given in the Supporting Information.
   Symbols have the same meaning as in Table \ref{PSsystems}.

### 2019-09-10 Manuscript/manuscriptPS.tex
Counterion binding

@@
 bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
-We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, see Fig.~\ref{NAdensPOPS}.
-The force fields studied show significant differences in both binding affinity
-and distribution of ions at the interface.
+We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, $\langle A_\mathrm{pl} \rangle$, see Fig.~\ref{NAdensPOPS}.
+The force fields studied showed significant differences in both the binding affinity
+and the distribution of ions at the interface.
 The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
-is reproduced only in GROMOS-CKP and in the MacRog simulation
-with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
+was reproduced only in GROMOS-CKP and GROMOS-CKPM, as well as in MacRog
+with potassium counterions, while all other models gave considerably smaller values (Fig.~\ref{NAdensPOPS}).
 However, the counterion binding and the concomitant electrostatic screening of the headgroup
-repulsion does not fully explain the low area per molecule values
-since the MacRog simulation, which has the strongest sodium binding
-(the lowest concentrations in bulk water), gives the same area per molecule
-as the CHARMM36-UA simulation, which has significantly weaker counterion binding
-affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
-with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
-The results are consistent with a previous study suggesting that the
-low areas per molecule in PS lipid bilayers originate from the combination
+repulsion does not fully explain the low area per lipid values
+since MacRog, which has the strongest sodium binding
+(the lowest concentrations in bulk water), gave the same $\langle A_\mathrm{pl} \rangle$
+as CHARMM36-UA, which has significantly weaker counterion binding
+affinity. On the other hand, MacRog simulation with potassium produced a larger area per lipid (63~\AA$^2$) than
+with sodium (53~\AA$^2$), in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
+The results are consistent with a previous study that suggested the
+low $\langle A_\mathrm{pl} \rangle$ in PS lipid bilayers to originate from the combination
 of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
@@
   \caption{\label{NAdensPOPS}
-    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
+    Counterion density profiles along the membrane normal and areas per lipid, $\langle A_{\rm pl} \rangle$,
     of POPS lipid bilayer from simulations with different force fields.
-    The vertical green bars indicate the location of the phosphate density peak.
+    The vertical bars indicate the locations of the phosphate density peaks.
     The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
@@
     Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
-    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
-    the error bar is given as the difference between the results divided by two.
+    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations,
+    the error bars show the difference between the results divided by two.
   }\end{figure}
@@
   \caption{\label{HGorderparametersPSvsPC}
-    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
-    Experimental values with the signs are measured for pure POPS system in this work.
-    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
+    Modulation of the POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
+    Experimental values with the signs were measured for a pure POPS system in this work.
+    The signs are assumed to remain unchanged for the mixture, and the magnitudes are taken from Ref. \citenum{roux90}.
     Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
     the error bars of 0.02 are used here \cite{botan15,ollila16}.
-    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
+    The $y$-axis for the $\alpha$-carbon results (bottom) is shifted
     with the same value for both order parameters such that the lower order
     parameter value from pure POPS is at zero to correctly illustrate the significant forking.
-    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
-    the error bars are given as the difference between the results divided by two. 
+    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations,
+    the error bars show the difference between the results divided by two. 
   }
@@
 The experimentally observed modulation of headgroup order parameters
-by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
-Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic
-counterions (the lack of an ion-free reference state). The analysis is further complicated by the artificial aggregation of counterions
-observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI).
+by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
+Studying the binding of cations to negatively charged lipid bilayers is less straightforward, because an ion-free reference state does not exist due to the ever-present cationic
+counterions. Our analysis was further complicated by the artificial aggregation of counterions
+observed in some simulations (section \ref{mixtureTOadditionalCIs} in the Supporting Information).
 Therefore, we evaluate here the amount of bound charge not by adding salt
-(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
-but by studying the changes of the headgroup $S_{\rm CH}$ order parameters with increasing amount of
-negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.
+(although this is discussed in section \ref{mixtureTOadditionalCIs}),
+but by studying the changes of the headgroup $S_{\rm CH}$ with an increasing amount of
+negatively charged lipids (and thus an increasing amount of cationic counterions) in the bilayer.
 
 Experimentally, the $S_{\rm{CH}}$ values of the $\alpha$ and $\beta$ headgroup carbons of POPC increase when negatively charged POPS lipids are incorporated in the bilayer 
-(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
-This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
-which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
-The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
-in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
-Therein, the stronger counterion binding cancels
-the effect of negatively charged headgroups and prevents the experimentally observed
+(section \ref{electrometerFORmixtures})~\cite{seelig87,scherer87}.
+This was reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
+which had the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
+The CHARMM36, Berger, and GROMOS-CKP simulations either exhibited no change, or showed a decrease
+in one or even both of the POPC headgroup order parameters, as the amount of POPS increased (Fig. \ref{HGorderparametersPCvsPS}).
+Therein, the stronger counterion binding canceled
+the effect of negatively charged headgroups and prevented the experimentally observed
 increase of headgroup order parameters with growing amount of PS lipids.
 Therefore, we suggest that the relatively weak binding of potassium
-in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
-realistic surface charge density in membranes containing PS lipids,
-while the other tested models overestimate the counterion
+in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces a more
+realistic surface charge density in membranes containing PS lipids
+than the other tested models that overestimate the counterion
 binding affinity. The results are consistent with the behavior of headgroup order
-parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
-in the SI.
+parameters as a function of added counterions, see section \ref{mixtureTOadditionalCIs}
+in the Supporting Information.
 
@@
 POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
-Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
-parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
+Unfortunately, none of the tested models correctly reproduced the modulation of POPS headgroup order
+parameters with increasing amount of POPC %in POPC:POPS mixtures
+(Fig. \ref{HGorderparametersPSvsPC}).
 More accurate force fields are needed

### 2019-09-11 Manuscript/manuscriptPSsuppl.tex

@@
 
-\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
+\section{Headgroup response to additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
 To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
-we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
-as a function of different monovalent ions added to the buffer (Fig.~\ref{PSresponseTONaCl}). 
-Experimental order parameters of POPC headgroup in the mixture are available as a function
-of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
-in increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
+Fig.~\ref{PSresponseTONaCl} plots the headgroup order parameters $S^\beta_\mathrm{CH}$ and $S^\alpha_\mathrm{CH}$ measured from a POPC:POPS (5:1) mixture
+as a function of different monovalent ions added to the buffer. 
+Experimental order parameters of the POPC headgroup in the mixture are available as a function
+of LiCl and KCl concentrations, while the POPS headgroup order parameters are measured also
+with increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
 ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
-but increase with the addition of sodium or potassium (Fig.~\ref{PSresponseTONaCl}). 
-POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
-but only modest changes as a function of KCl concentration, indicating singificant 
-Li$^+$ binding but only weak K$^+$ binding to the mixture when interpreted using the
-electrometer concept~\cite{akutsu81,altenbach84,seelig87}.
-
-In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
+The different binding behavior is evident in the responses of the PS headgroup order parameters, as for example
+%$S^\beta_\mathrm{CH}$ increases with Li$^+$, but decreases with K$^+$, Na$^+$ grows initially and then decreases;
+%the smaller $S^\alpha_\mathrm{CH}$ decreases with all three monovalent cations;
+the larger $S^\alpha_\mathrm{CH}$
+decreases with Li$^+$, but increases with K$^+$ and Na$^+$, see Fig.~\ref{PSresponseTONaCl}.
+The PC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
+but only modest changes as a function of KCl concentration, indicating significant 
+Li$^+$ binding, but only weak K$^+$ binding, when interpreted using the
+molecular electrometer concept~\cite{akutsu81,altenbach84,seelig87}.
+
+In simulations with Berger and CHARMM36, the responses of POPC and POPS $S_\mathrm{CH}$
 to added sodium and potassium are not in line with the experiments.
-Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.~\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
+Rather, the simulations produce responses similar to experiments conducted in LiCl (Fig.~\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium
 in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
@@
 less systematic changes in the order parameters (Fig.~\ref{PSresponseTONaCl}).
-Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
+Similar unsystematic behavior was also observed in the simulations of Lipid14/17 model
 with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
-for which the data is not shown due to the formation of
-ion clusters in water with relatively low (1~M) ion concentrations (Fig.~\ref{ionCLUSTERS}).
-Appearance of such clusters in the MagRog simulations with 4~M KCl
+for which the data are not shown due to the formation of
+ion clusters in water at relatively low (1~M) ion concentrations (Fig.~\ref{ionCLUSTERS}).
+Appearance of such clusters in the MagRog simulations at 4~M KCl
 could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
-In conclusion, the results are in line with the Section \ref{ciBINDINGsection}
-in the main text, suggesting that the MacRog simulations with KCl give the most
-realistic surface charge at the lipid bilayer interface among the tested simulation models.
+In conclusion, the results are in line with the section 'Counterion binding and interactions between PC and PS headgroups' % \ref{ciBINDINGsection}
+in the main text, suggesting that the MacRog simulations with KCl gave the most
+realistic surface charge at the POPC:POPS (5:1) lipid bilayer interface among the tested simulation models.
 \begin{figure*}[]
@@
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
+    in the PS headgroup (bottom right), the $y$-axis is shifted with the same value for both $S^\alpha_\mathrm{CH}$ such that the lower order
     parameter value is at zero.
@@
   \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
-  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1~M of NaCl (left)
+  \caption{Ion clusters appearing in POPC:POPS (5:1) Lipid14/17 simulations with 1~M of NaCl (left)
     and MacRog simulations with 4~M of KCl (right).

### 2019-09-11 Manuscript/manuscriptPSsuppl.tex
Section S2

@@
 
-\section{Electrometer concept in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}
-
-The electrometer concept is based on the empirical observations that
-the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
+\section{Molecular electrometer in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}
+
+The molecular electrometer concept is based on the empirical observations that
+the C--H bond order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
 decrease (increase) proportionally to the bound positive (negative) 
@@
 Changes of the headgroup order parameters of negatively charged PS and PG lipids
-are also systematic, but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
-Therefore, the ion binding affinity to negatively charged bilayers
-can be better characterized measuring the PC headgroup order parameters from 
+are also systematic, but as they are less well characterized~\cite{borle85,macdonald87,roux86,roux90} than for PC,
+the ion binding affinity to negatively charged bilayers
+can be better characterized by measuring the PC headgroup order parameters from 
 mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
@@
   \caption{\label{HGorderparametersPCvsPEPSPGchol}
-    Headgroup order parameters of POPC measured from mixtures with
-    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
+    Headgroup order parameters $S^\beta_\mathrm{CH}$ and $S^\alpha_\mathrm{CH}$ of POPC measured from mixtures with
+    PS (bovine brain), POPG, POPE, cholesterol, and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
     Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
@@
 When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
-it is important to note that the order parameters
-increase due to the addition of negative charged lipids according to the electrometer
-concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
+it is important to note that these $S_\mathrm{CH}$
+increase due to the addition of negative charged lipids according to the molecular electrometer~\cite{seelig87, scherer87}
+(Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
 Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
-than in pure lipid bilayers. This is evident also in the headgroup order parameters
-of POPC in bilayers with different amounts anionic lipids without added calcium (Fig. \ref{OrderParametersWithCaCl}).
-Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
-to the CaCl$_2$ concentrations of $\sim$ 50--300mM, depending on the amount of negatively charged
+than in pure PC lipid bilayers. This is demonstrated in Fig. \ref{OrderParametersWithCaCl} by the headgroup $S_\mathrm{CH}$
+of POPC in bilayers with different amounts of anionic lipids without added calcium.
+Upon addition of CaCl$_2$, the $S_\mathrm{CH}$ decrease and reach the values of pure PC bilayers around 
+CaCl$_2$ concentrations of $\sim$50--300~mM, depending on the amount of negatively charged
 lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
-Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
-the negative charge lipids, resulting to a neutral membrane. 
-Above such concentrations, the specific binding of calcium leads to
-the overcharging of the membrane.
+Around these concentrations the positive charge of the bound Ca$^{2+}$ cancels
+the negative charge of the lipids, resulting in a neutral membrane. 
+Above these concentrations, the specific binding of calcium leads to
+overcharging of the membrane.
 \begin{figure}[]
@@
     Right column shows the same data zoomed to the concentrations below 1~M.
-    Data for Pure DPPC from Ref.~\citenum{akutsu81},
+    Data for pure DPPC from Ref.~\citenum{akutsu81},
     for pure POPC from Ref.~\citenum{altenbach84}, 
     for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
-    for POPC:POPG (8:2,1:1) mixtures with 0.1~M NaCl from Ref.~\citenum{macdonald87}
-    and for POPC:POPG (1:1) mixture data without NaCl from Ref.\citenum{borle85}.
-  }
-\end{figure}
-
-Because the POPC headgroup order parameters in mixtures with different amounts of anionic lipids
+    for POPC:POPG (8:2,1:1) mixtures with 0.1~M NaCl from Ref.~\citenum{macdonald87},
+    and for POPC:POPG (1:1) mixture without NaCl from Ref.\citenum{borle85}.
+  }
+\end{figure}
+
+Because the POPC headgroup $S_\mathrm{CH}$ in mixtures with different amounts of anionic lipids
 but without additional salt are not equal, the binding affinity of calcium can be better compared
-by plotting the changes of order parameters as a function of added calcium.
+by plotting the changes of order parameters, $\Delta S_\mathrm{CH}$, as a function of added calcium.
 As expected, such a plot reveals more pronounced order parameter decrease in systems
@@
 from various mixtures of POPC and anionic lipids with added calcium gives physically
-consistent results, suggesting that the electrometer concept can be used to determine
-the cation binding affinity also to the lipid bilayers containing mixtures of PC and anionic lipids.  
+consistent results, suggesting that the molecular electrometer can be used to determine
+the cation binding affinity also to lipid bilayers containing mixtures of PC and anionic lipids.  
 \begin{figure}[]
@@
   \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
-    Changes of POPC headgroup order parameters as a function of CaCl$_2$
+    Changes of POPC headgroup order parameters, $\Delta S^\beta_\mathrm{CH}$ and $\Delta S^\alpha_\mathrm{CH}$, as a function of CaCl$_2$
     measured from mixed bilayers containing different amounts of anionic lipids.
-    The original data is the same as in figure~\ref{OrderParametersWithCaCl}.
+    The original data are the same as in Fig.~\ref{OrderParametersWithCaCl}.
 %    The values are taken from 2H NMR experiments reported in the

### 2019-09-11 Manuscript/manuscriptPS.tex
Molecular electrometer

@@
 
-\subsection{Using the molecular electrometer concept to compare ion binding to negatively charged lipid bilayers 
+\subsection{Using the molecular electrometer to compare ion binding to negatively charged lipid bilayers 
 in simulations and in experiments}
@@
 simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
-of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
-Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
-mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
+of negatively charged PS and PG lipids also exhibit systematic dependencies on the bound charge, but these are less well understood than for PC~\cite{borle85,macdonald87,roux86,roux90}.
+Therefore, measuring the PC headgroup $S_{\rm CH}$ in 
+mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also Supporting Information section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing ion binding to such negatively charged membranes.
 
-Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
+Calibrating the PC headgroup $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
 This can be done using experimental data from mixtures of PC and
@@
 Studies applying the molecular electrometer have used two different definitions for salt concentration:
-The concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
-In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
+The concentrations reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
+In the former case, binding of ions to the lipids leads to a lower bulk concentration than what was present in the original solvent.
 However, the choice of definition has only a marginal effect
@@
 (see SI section \ref{concentrationDEFsection}).
-In this work we use the former definition (concentration before solvating the lipids) to be consistent with the reference
+In this work, we use the concentration before solvating the lipids to be consistent with the reference
 experimental data \cite{roux90}.
@@
   TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
+BK gratefully acknowledges financial support from the International Max Planck Research School (IMPRS) on Multiscale Bio-Systems.
 \end{acknowledgement}

### 2019-09-11 Manuscript/manuscriptPS.tex
MD simulations

@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline
+lipid/counter-ion & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline
 \hline 
@@
 where $\theta$ is the angle between the C--H bond and the membrane normal
-(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
+(taken to align with $z$, with membrane in the $xy$-plane).
 Angular brackets denote average over all sampled configurations.
@@
 for each lipid in the system, and then calculating the average and
-the standard error of the mean over the different lipids. The analysis can be conducted using a
-Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that uses the
+the standard error of the mean over the different lipids. The analysis can be conducted using the
+Python program {\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}, that uses the
 MDAnalysis library \cite{agrawal11,gowers16}.
-For united atom models, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
+For the united atom force fields, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
 of the Gromacs 3 sofware package \cite{gromacsMANUAL}.

### 2019-09-11 Manuscript/manuscriptPS.tex
Experiments

@@
 The magnitudes of headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
-were determined by measuring the chemical-shift resolved dipolar splittings
+were determined by measuring the chemical-shift-resolved dipolar splittings
 with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.

### 2019-09-11 Manuscript/manuscriptPS.tex
Introduction

@@
 25--35\% of the cytosolic leaflet of plasma membranes consists of PS lipids~\cite{leventis10,li14}.
-PS lipids are vastly important biomolecules that interact with
+PS lipids are important biomolecules that interact with
 signaling proteins \cite{leventis10}, regulate
@@
 Experimental studies have indicated that the
-serine headgroup in PS is more rigid than the choline headgroup in phosphatidylcholine (PC),
+serine-containing headgroup of PS is more rigid than the choline-containing headgroup of PC (phosphatidylcholine),
 owing possibly to electrostatic interactions or formation of a hydrogen bond network between the headgroups~\cite{browning80,buldt81}.
@@
 phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
+Mixing PS lipids with PC lipids reduces the propensity of PS to form complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
 That being said, some studies suggest that Ca$^{2+}$ has similar specific binding affinity 
@@
 are currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
-to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
+to understand the PS headgroup structure, its influence on lipid bilayer properties, and PS
 interaction with
@@
 Unfortunately, the results have depended on the particular force field used.
-For example, recent simulations using the NBfix parameters for Ca$^{2+}$~\cite{kim16} in
+For example, recent simulations using the NBfix parameters for calcium~\cite{kim16} in
 the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
-suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
-the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
+suggest that Ca$^{2+}$ ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
+results from the same lipid model but without the NBfix ion parameters, combined with NMR chemical shifts and
 rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also toward the phosphate region \cite{hallock18}.
@@
 can be used to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
-measured from NMR experiments with high accuracy and directly compared to simulations
-to either evaluate the quality of the simulation model and/or to interpret the experiments \cite{ollila16}. Using this approach,
+measured from NMR experiments with high accuracy and directly compared to MD simulations
+to evaluate the quality of the force field and to interpret the experiments \cite{ollila16}. Using this approach,
 it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
-captured by most MD force fields~\cite{botan15}, that the cation binding to PC
-lipid bilayers is overestimated~\cite{catte16}, and that the inequivalent order parameters of the distinct C--H bonds at
-carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.
+captured by most MD force fields~\cite{botan15}, that cation binding to PC
+lipid bilayers is overestimated~\cite{catte16}, and that the inequivalence of the order parameters of the distinct C--H bonds at
+the carbon 2 in {\it sn}-2 lipid tail is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.
 
@@
 Our results pave the way
-for the development of lipid models that correctly describe 
+for development of lipid models that correctly describe 
 the headgroup region of negatively charged lipids under physiological salt

### 2019-09-12 Manuscript/manuscriptPS.tex

--- 2019-09-11 Manuscript/manuscriptPS.tex
@@
 % some footnotes are not visible in typeset-MS (pdf)
-lipid/counter-ion & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline
+lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline
 \hline 
@@
 where $\theta$ is the angle between the C--H bond and the membrane normal
-(taken to align with $z$, with membrane in the $xy$-plane).
+(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
 Angular brackets denote average over all sampled configurations.
@@
 for each lipid in the system, and then calculating the average and
-the standard error of the mean over the different lipids. The analysis can be conducted using the
-Python program {\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}, that uses the
+the standard error of the mean over the different lipids. The analysis can be conducted using a
+Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that uses the
 MDAnalysis library \cite{agrawal11,gowers16}.
-For the united atom force fields, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
+For united atom models, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
 of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
@@
 
-\subsection{Using the molecular electrometer to compare ion binding to negatively charged lipid bilayers 
+\subsection{Using the molecular electrometer concept to compare ion binding to negatively charged lipid bilayers 
 in simulations and in experiments}
@@
 simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
-of negatively charged PS and PG lipids also exhibit systematic dependencies on the bound charge, but these are less well understood than for PC~\cite{borle85,macdonald87,roux86,roux90}.
-Therefore, measuring the PC headgroup $S_{\rm CH}$ in 
-mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also Supporting Information section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing ion binding to such negatively charged membranes.
+of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
+Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
+mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
 
-Calibrating the PC headgroup $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
+Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
 This can be done using experimental data from mixtures of PC and
@@
 Studies applying the molecular electrometer have used two different definitions for salt concentration:
-The concentrations reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
-In the former case, binding of ions to the lipids leads to a lower bulk concentration than what was present in the original solvent.
+The concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
+In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
 However, the choice of definition has only a marginal effect
@@
 (see SI section \ref{concentrationDEFsection}).
-In this work, we use the concentration before solvating the lipids to be consistent with the reference
+In this work we use the former definition (concentration before solvating the lipids) to be consistent with the reference
 experimental data \cite{roux90}.
 
-\section{Results and Discussion}
+\section{Results and discussion}
 
@@
 the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
-known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
+C--H bond order parameters known from previous $^2$H\,NMR experiments~\cite{browning80}.
 Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
@@
     (B) 2D R-PDLF spectra.
-    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
-    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon, and $+0.09$ and $-0.02$
-    for the $\alpha$-carbon.
+    (C) Experimental S-DROSS data (points), and SIMPSON simulations with
+    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon and $+0.09$ and $-0.02$
+    for the $\alpha$-carbon (blue lines).
     Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
@@
     POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
-    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
+    are measured in this work, whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
     The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
@@
 and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
-We used the literature value for the smaller  $S_{\rm CH}$ , because the
+We used the literature value for the smaller  $S_{\rm CH}$, because the
 resolution of our R-PDLF experiment was not sufficient to determine the
 magnitude of the small value.
-The S-DROSS curve from the SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
-(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
-corroborating the interpretation that the smaller  $S_{\rm CH}$  is negative.
+The S-DROSS curve from a SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
+(dashed grey in Fig. \ref{PShgSIGNSsimpson} (C)) did not agree with the experiment, 
+corroborating the interpretation that the smaller  $S_{\rm CH}$ is indeed negative.
 
@@
 When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
-the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative and $\alpha$-carbon
+the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative, and the $\alpha$-carbon
 experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
@@
 are in good agreement with the results from centrifuged samples~\cite{browning80}.
-This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the simulations and reference experiments.
+This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the reference experiments and our simulations.
 

### 2019-09-12 Manuscript/manuscriptPS.tex

--- 2019-09-11 Manuscript/manuscriptPS.tex
@@
 
-\section{Results and Discussion}
+\section{Results and discussion}
 

### 2019-09-12 Manuscript/manuscriptPSsuppl.tex
Update to the submitted revised version

@@
 prepared tools for calculating order parameters, performed Lipid14/17 simulations with Gromacs MD simulation package,
-prepared the TOC figure and provided comments on the manuscript.
+and provided comments on the manuscript.
 
@@
 {\it Jukka M{\"a}{\"a}tt{\"a}}
-Did minor analysis for and edited Figures 6 and 10.
+Prepared figures 6 and 10.
 

### 2019-10-16 Manuscript/manuscriptPSsuppl.tex
Add files via upload
Modified SI and SI bibliography

@@
 Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
-Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
+Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
 switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.
-
-Analyses on these simulations was performed on the final 100~ns of the simulations.
-
-\noindent
-{\it POPC:POPS mixtures without additional ions}
-Simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were constructed using the CHARMM-GUI~\cite{lee16,jo18}.
-Two identical systems were constructed, apart from one system was neutralised through the addition of 22 Na$^+$ ions while the other system
-contained 22 K$^+$ ions. As per the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
+The final 100~ns of the simulations was used for analysis.
+
+\noindent
+{\it POPC:POPS mixtures without additional ions.}
+Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were configured using the CHARMM-GUI~\cite{lee16,jo18}.
+Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
+contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
 and settings were identical to those used in the CHARMM36 POPS simulations.
@@
 and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}.
- \\
 
@@
 and data, see table \ref{mixedIONsystemsMONOVALENT} and
-Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\
+Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}.
 
@@
 Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
-by the CHARMM-GUI at 298~K. \\
+by the CHARMM-GUI at 298~K.
 
@@
 Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
-were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
+were constructed manually using the PS parameters provided in ref.~\citenum{rog16} as a guide. The original parameters were
 not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
+{\it POPC:POPS (5:1) mixtures with additional potassium ions.} 
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
@@
 The simulation parameters were taken from Ref.~\citenum{rog16}. 
-Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
+Periodic boundary conditions were employed in all dimensions. The neighbor list was updated every step. 
 %\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
@@
 obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
-4480 water molecules resulting in 31.1 water molecules per lipid. For
-each system, 128 NaCl molecules are added to neutralize the system.
+4480 water molecules resulting in 31.1 waters per lipid. For
+each system, 128 NaCl molecules were added to neutralize the system.
 In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
+to constrain the bonds in TIP3P water molecules. Simulations were run using Amber18 simulation package \cite{amber18md}
 and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
-ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
+ion parameters~\cite{joung2008determination} and another set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
 The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
@@
 
-Each bilayer structure was first minimized for 2500 steps with steepest
-descent and an additional 2500 step with the conjugate gradient
-algorithms while restraining the solvent and ion molecules with a force
+Each bilayer structure was first minimized for 2500 steps using the steepest
+descent algorithm and then an additional 2500 steps with the conjugate gradient
+algorithms while restraining the solvent and ions with a force
 constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
@@
 algorithm and a time step of 2~fs was used. In all steps with periodic
-boundary conditions, the non-bonded interactions are calculated with
+boundary conditions, the non-bonded interactions were calculated with
 particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.
@@
 time step of 2~fs was used. Trajectories are saved in every 10~ps
-intervals. Two independent trajectories of 500~ns long was generated and
-the last 200~ns portions were used for the analysis.
-
-\noindent
-{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} 
+intervals. Two independent, 500~ns long trajectories were generated, of which
+the last 200~ns was used for the analysis.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions.} 
 The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
 Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}. For the simulations of the lipids and ions,
-the Amber Lipid 17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
+the Amber Lipid17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
 Simulations were ran using Amber18 simulation package \cite{amber18md} 
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with default Amber parameters} 
+{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using default Amber parameters.} 
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
@@
 Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
-The simulation details, minimization, heating, and equilibration runs are the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield. 
+The simulation details, minimization, heating, and equilibration runs were the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield. 
 For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with Dang ion parameters} 
+{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using Dang ion parameters.} 
 The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
@@
 \noindent
-{\it POPS and DOPS with 128 lipids}
+{\it POPS and DOPS with 128 lipids.}
 Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
@@
 This system was simulated 120~ns using Gromacs 5~\cite{abraham2015gromacs} with parameter (mdp file)
-and force field files (top and itp) available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K},
-the last 80~ns was used for the analysis.
+and force field files (top and itp) available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K}.
+The last 80~ns of the trajectory was used for the analysis.
 %\todo{The simulation details of this are not clearly described in Zenodo repository, but can be read from the mdp file available there.
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures without additional ions}
+{\it POPC:POPS (5:1) mixtures without additional ions.}
 % The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
@@
 Changes of the headgroup order parameters of negatively charged PS and PG lipids
-are also systematic, but as they are less well characterized~\cite{borle85,macdonald87,roux86,roux90} than for PC,
+are also systematic, but not as well characterized~\cite{borle85,macdonald87,roux86,roux90} than for PC, and
 the ion binding affinity to negatively charged bilayers
-can be better characterized by measuring the PC headgroup order parameters from 
+can be better quantified by measuring the PC headgroup order parameters from 
 mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
@@
 When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
-it is important to note that these $S_\mathrm{CH}$
+it is important to note that these $S_\mathrm{CH}$s
 increase due to the addition of negative charged lipids according to the molecular electrometer~\cite{seelig87, scherer87}
@@
 Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
-than in pure PC lipid bilayers. This is demonstrated in Fig. \ref{OrderParametersWithCaCl} by the headgroup $S_\mathrm{CH}$
-of POPC in bilayers with different amounts of anionic lipids without added calcium.
+than in pure PC lipid bilayers, as seen from Fig. \ref{OrderParametersWithCaCl} by comparing the measurements without added salt.
 Upon addition of CaCl$_2$, the $S_\mathrm{CH}$ decrease and reach the values of pure PC bilayers around 
@@
 \subsubsection*{Magnitude}
-To quantify how close to the experimentally obtained C--H order parameters ($S_\mathrm{CH}$:s)
-the force-field-produced $S_\mathrm{CH}$:s were, we assigned a number for each carbon based on the following 5-step scale:
+To quantify how close the experimentally obtained C--H order parameters were to the $S_\mathrm{CH}$s obtained from the different MD models, we assigned a number for each carbon based on the following 5-step scale:
 
@@
 %
-For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was used:
+For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was employed:
 %
@@
 %
-For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was used:
+For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was utilized:
 %

### 2019-10-17 Manuscript/manuscriptPSsuppl.tex

@@
 along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
-POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
+POPC/POPS mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
 out of which 150~ns were used for analyses. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.
@@
 All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
-including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.
+including use of the Verlet cut-off scheme~\cite{Pall13}, except that here LINCS was used to constrain all bonds.
 Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
@@
 All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
-including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
+including the use of the Verlet cut-off scheme, except that here LINCS was used to constrain all bonds.
 
@@
 to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
-and POPS simulations, including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.
+and POPS simulations, including the use of the Verlet cut-off scheme, except that here LINCS was used to constrain all bonds.
 
@@
     Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
-    with the NBfix term in sodium at Ref.~\citenum{CHARMM36withCaClNBfix}, 
+    with the NBfix term in sodium at Ref.~\citenum{CHARMM36withNaClNBfix}, 
     with the NBfix term in calcium at Ref.~\citenum{CHARMM36withCaClNBfix} and

### 2019-10-17 Manuscript/manuscriptPSsuppl.tex
Final edits on the SI.

@@
 
+%-- To remove the empty first page that comes from the too large title page:
+\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
+\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}
+
 \setcounter{page}{0}
@@
 \author{Hanne Antila}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Pavel Buslaev}
-\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+\affiliation[Moscow Institute of Physics and Technology]{Research Center for Molecular Mechanisms of Aging and Age-Related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny 141701, Russia}
 \author{Fernando Favela-Rosales}
-\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
+\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente, C. P. 99102, Zacatecas, M\'{e}xico}
 \author{Tiago M. Ferreira}
-\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
+\affiliation[Martin-Luther University Halle--Wittenberg]{NMR Group - Institute for Physics, Martin-Luther University Halle--Wittenberg, 06120 Halle (Saale), Germany}
 \author{Ivan Gushchin}
-\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-Related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny 141701, Russia}
 \author{Matti Javanainen}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{Batuhan Kav}
@@
 \author{Jesper J. Madsen}
-\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, 60637 Chicago, Illinois, United States of America}
+\affiliation[University of South Florida]{Department of Global Health, College of Public Health, University of South Florida, 33612 Tampa, Florida, United States of America}
 \author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
-and The Zernike Institute for Advanced Materials, 
-University of Groningen, 9747 AG Groningen, The Netherlands}
-\author{Markus Miettinen}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\alsoaffiliation[University of Groningen]{Groningen Biomolecular Sciences and Biotechnology Institute and The Zernike Institute for Advanced Materials,  University of Groningen, 9747 AG Groningen, The Netherlands}
+\author{Markus S. Miettinen}
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \author{Jukka M{\"a}{\"a}tt{\"a}}
-\affiliation{Department of Chemistry, Aalto University, Espoo, Finland}
+\affiliation[Aalto University]{Department of Chemistry and Materials Science, Aalto University, 00076 Aalto, Finland}
 \author{Ricky Nencini}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \author{O. H. Samuli Ollila}
@@
 %%\homepage[]{Your web page}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+\affiliation[University of Helsinki]{Institute of Biotechnology, University of Helsinki, 00014 Helsinki, Finland}
 \author{Thomas J. Piggot}
-\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
+\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
 
@@
 
+% \end{table*}
+
+\subsection{CHARMM36}
+\label{sec:CHARMM36}
+\noindent
+    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
+    available from Ref. \citenum{charmmPOPC512files} are used.
+% It should be noted that pure POPC are ran with Gromacs 4, but the results with
+% PS are ran with Gromacs 5. This has significant effect on acyl chains:
+% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
+% but the effect to headgroups is small
+% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
+% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}
+
+\noindent
+{\it DOPS and POPS bilayers.}
+%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
+Starting structures %for CHARMM36 DOPS and POPS simulations
+were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
+The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
+Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
+with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
+%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
+For POPS, an identical system was also constructed, in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.
+
+Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
+A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
+For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
+The simulations were maintained at temperatures of 303~K for DOPS and 298~K for POPS using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
+with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
+Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
+Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
+switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.
+The final 100~ns of the simulations was used for analysis.
+
+\noindent
+{\it POPC:POPS mixtures without additional ions.}
+Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were created using the CHARMM-GUI~\cite{lee16,jo18}.
+Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
+contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 waters per lipid were added. All other system and simulation parameters
+and settings were identical to those used in the CHARMM36 POPS simulations.
+
+%Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
+%with potassium and sodium countertions were ran using CHARMM36 force field.
+%The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
+%and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.
+
+Simulations of POPC:POPS (5:1) and POPC:POPS (1:1) mixtures containing a total of 300 lipids (250:50 and 150:150)
+with neutralizing potassium counterions 
+%ran using CHARMM36 force field.
+%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
+%POPC:POPS (both 1:1 and 5:1) mixtures
+were prepared
+%with 300 lipids in total
+%with the  ions
+using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
+and the CHARMM36 force field with simulation parameters given by the CHARMM-GUI at 298~K. For further details see Table \ref{mixedIONsystemsMONOVALENT}.
+
 %\begin{table*}[!p]
@@
 \centering
-\caption{The list of POPC:POPS mixtures simulated with different amounts of added monovalent ions. 
-  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-  This corresponds the concentration in buffer before solvating lipids, which are
+\caption{POPC:POPS mixtures simulated with different amounts of added monovalent ions. 
+  Salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  This corresponds to the concentration in buffer before solvating lipids, as
   reported in the experiments by Roux et al.~\cite{roux90}.
@@
 \begin{tabular}{l c c c c c c c c c}
-  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  lipids/counter-ion & force field for lipids / ion & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) & \footnote{Reference for simulation files}files\\
   \hline
@@
 
-% \end{table*}
-
-\subsection{CHARMM36}
-
-\noindent
-    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
-    available from Ref. \citenum{charmmPOPC512files} are used.
-% It should be noted that pure POPC are ran with Gromacs 4, but the results with
-% PS are ran with Gromacs 5. This has significant effect on acyl chains:
-% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
-% but the effect to headgroups is small
-% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
-% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}
-
-\noindent
-{\it DOPS and POPS bilayers.}
-%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
-Starting structures for CHARMM36 DOPS and POPS simulations were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
-The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
-Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
-with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
-%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
-For POPS, an identical system was also constructed in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.
-
-Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
-A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
-For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
-The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
-with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
-Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
-Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
-switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.
-The final 100~ns of the simulations was used for analysis.
-
-\noindent
-{\it POPC:POPS mixtures without additional ions.}
-Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were configured using the CHARMM-GUI~\cite{lee16,jo18}.
-Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
-contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
-and settings were identical to those used in the CHARMM36 POPS simulations.
-
-%Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
-%with potassium and sodium countertions were ran using CHARMM36 force field.
-%The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
-%and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.
-
-Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
-with neutralizing potassium countertions 
-%ran using CHARMM36 force field.
-%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
-%POPC:POPS (both 1:1 and 5:1) mixtures
-were prepared
-%with 300 lipids in total
-%with the  ions
-using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
-and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}.
-
 \noindent
 {\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
-POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
-potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
+POPC:POPS (110:22) mixtures containing a total of 132 lipids, with additional
+potassium or sodium ions corresponding to concentrations of $\sim$450~mM and 890~mM
 were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
-Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
+Gromacs 5 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation
 parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
-and data, see table \ref{mixedIONsystemsMONOVALENT} and
+and data, see Table \ref{mixedIONsystemsMONOVALENT} and
 Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}.
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.} 
-POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
+{\it POPC:POPS (5:1) mixtures with additional calcium using the NBfix1 parameters.} 
+POPC:POPS (5:1) mixtures containing a total of 300 (250:50) lipids
 %were simulated with the additional calcium concentrations.
@@
 and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
-The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
-Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
+The ion parameters generated by the CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
+Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation parameters given
 by the CHARMM-GUI at 298~K.
@@
 \noindent
-{\it POPC:POPS (5:1) mixtures with  additional calcium using NBfix2 parameters.}
-POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids, neutralizing sodium ions, 
+{\it POPC:POPS (5:1) mixtures with  additional calcium using the NBfix2 parameters.}
+POPC:POPS (5:1) mixtures containing a total of 300 (250:50) lipids, neutralizing sodium ions, 
 and 0.14~M or 0.94~M CaCl$_2$ were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
-The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}.
-Systems were simulated using Gromacs 2018.6 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
+The ion parameters generated by the CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}.
+Systems were simulated using Gromacs 2018.6 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation parameters given
 by the CHARMM-GUI at 320~K.  \\
 
-\subsection{CHARMM36ua}
+\subsection{CHARMM36-UA}
 \noindent
@@
 %The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
-Starting structures for CHARMM36-UA DOPS and POPS simulations were taken from those generated for the CHARMM36 simulations (see above),
+Starting structures %for CHARMM36-UA DOPS and POPS simulations
+were taken from those generated for the CHARMM36 simulations (see section \ref{sec:CHARMM36}),
 with any extraneous hydrogen atoms removed in the lipid tails. Force field parameters for DOPS and POPS were constructed through a
@@
 % reference. I am not sure how much we should repeat these in here.}
-The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
+The POPC bilayer consisting of a total of 128 lipids was constructed from a single lipid structure taken 
 from Ref.~\citenum{kulig15c}. The bilayer was hydrated by a total of 5120 water 
@@
 along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
-POPC/POPS mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
-out of which 150~ns were used for analyses. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.
+POPC:POPS (5:1) mixtures with the MacRog force field, see further below in this section. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
+out of which the last 150~ns were used for analysis. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.
 
@@
 The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
-The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
+The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions), and was converted into
 the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
@@
 van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
-Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
-All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
-including use of the Verlet cut-off scheme~\cite{Pall13}, except that here LINCS was used to constrain all bonds.
-Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
-the simulations with Verlet cut-off ($\sim$~52\AA$^2$).
+long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
+All other simulation settings were identical to those described above in section \ref{sec:CHARMM36} for the CHARMM36 DOPS and POPS simulations,
+including the use of the Verlet cut-off scheme~\cite{Pall13}, except that here LINCS was used to constrain all bonds.
+Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\,\AA$^2$) than
+the simulations with the Verlet cut-off ($\sim$~52\,\AA$^2$).
 
@@
 were described by the \AA{}qvist parameters \cite{aqvist90}. Other topologies
-were unchanged. The simulation parameters were the same as used for the POPC/POPS mixtures with 
-the MacRog force field, see below. The simulation was run for 200~ns, out of which 150~ns was used 
-in the analyses.
+were unchanged. Simulation parameters were the same as used for the POPC:POPS (5:1) mixtures with 
+the MacRog force field, see further below in this section. The simulation was run for 200~ns, out of which the last 150~ns was used 
+in the analysis.
 
@@
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
-The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
+Bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
 two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
@@
 of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
-were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.
+were considered, corresponding to 52, 104, 208, and 311 pairs of K$^+$ and Cl$^-$ ions.
 
 The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
-Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from OPPS topology in
-Ref.~\citenum{rog16} by reversing the order of acyl chains.
-The initial structures from CHARMM-GUI were used because they have the
+Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from the OPPS topology in
+Ref.~\citenum{rog16} by reversing the order of the acyl chains.
+The initial structures from CHARMM-GUI were used, because they had the
 naturally occurring L stereoisomer in the POPS head group, instead of the 
 the D stereoisomer present in Ref.~\citenum{rog16}.
-TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
+The TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
 \cite{aqvist90}, standard for the OPLS force field, were employed for the ions.
 
-The simulation parameters were taken from Ref.~\citenum{rog16}. 
+Simulation parameters were taken from Ref.~\citenum{rog16}. 
 Periodic boundary conditions were employed in all dimensions. The neighbor list was updated every step. 
@@
 \cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
-was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
+was applied to both energy and pressure. Temperature was kept at 298~K using the 
 Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
@@
 (only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
-250, 200, or 300~ns of the simulations were included in the analyses, respectively.
+250, 200, or 300~ns of the simulations were included in the analysis, respectively.
 
@@
 {\it DOPS and POPS bilayers.} 
-The structure of the bilayers with 128 DOPS or 128 POPS lipids were
-obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
-4480 water molecules resulting in 31.1 waters per lipid. For
-each system, 128 NaCl molecules were added to neutralize the system.
-In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
-to constrain the bonds in TIP3P water molecules. Simulations were run using Amber18 simulation package \cite{amber18md}
-and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
-ion parameters~\cite{joung2008determination} and another set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
-The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
-the number of molecules in simulations, related simulation files and trajectories.
+The structures of the bilayers with 128 DOPS or 128 POPS lipids were
+obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers were solvated with
+4480 water molecules, resulting in 31.1 waters per lipid. For
+each system, 128 Na$^+$ ions were added to neutralize the system.
+In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. The SETTLE~\cite{miyamoto92} algorithm was employed 
+to constrain the bonds in TIP3P water molecules. Simulations were run using the Amber18 simulation package,\cite{amber18md}
+and lipid parameters were obtained from the Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with the Joung--Cheatham
+ion parameters~\cite{joung2008determination} and another set of simulations with the ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
+The data in Table~\ref{PSsystems} and Refs.~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
+the number of molecules in simulations, related simulation files, and trajectories.
 
@@
 descent algorithm and then an additional 2500 steps with the conjugate gradient
-algorithms while restraining the solvent and ions with a force
+algorithm while restraining the solvent and ions with a force
 constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
@@
 non-periodic boundary conditions were applied, no pressure scaling was
-used and none of the bonds were restrained. A three step heating
-procedure was applied: in the first heating step the system temperature
-is increased from 0.0~K to 100.0~K using Langevin thermostat with
-5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
+used and none of the bonds were restrained. A three-step heating
+procedure was applied: In the first heating step, at constant volume without periodic
+boundary conditions, the system temperature
+was increased from 0.0~K to 100.0~K using Langevin thermostat with
+5.0~ps$^{-1}$ collision frequency, while restraining the lipid positions
 with harmonic springs of force constant
-20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
-boundary conditions. In the second heating step, the system temperature
+20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$. In the second heating step,  at constant volume, the system temperature
 was increased from 100.0~K to 200.0~K using Langevin thermostat with
-5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
+5.0~ps$^{-1}$ collision frequency within 10\,000 steps, while restraining
 the lipid positions with harmonic springs of force constant
-10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
-heating step, the system temperature is increased from 200.0~K to
-303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
-collision frequency while restraining the lipid positions with harmonic
+10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$. In the third
+heating step, the system temperature was increased from 200.0~K to
+303.0~K within 10\,000 steps using Langevin thermostat with 5.0~ps$^{-1}$
+collision frequency, while restraining the lipid positions with harmonic
 springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
@@
 From the second heating step onward, periodic boundary conditions in all
-dimensions were applied. The semi-isotropic pressure coupling along
-\textit{xy}-plane was employed and the surface tension was set to
-0.0~dyn explicitely. After the third heating step, each structure is
-equilibriated for 100~ns at 1~bar and 303~K constant pressure using
-Berendsen barostat and temperature using Langevin thermostat,
-respectively. For the equilibriation step, semi-isotropic pressure
-coupling along \textit{xy}-plane was employed and the surface tension
-was set to 0.0~dyn explicitely. For all heating and equilibration steps,
-the lengths of the hydrogen containing bonds were restrained using SHAKE
-algorithm and a time step of 2~fs was used. In all steps with periodic
+dimensions were applied. Semi-isotropic pressure coupling along
+\textit{xy}-plane was employed, and surface tension was set to
+0.0~dyn explicitly. After the third heating step, each structure was
+equilibrated for 100~ns at 1~bar and 303~K using
+Berendsen barostat and Langevin thermostat. For the equilibriation step, semi-isotropic pressure
+coupling along \textit{xy}-plane was employed, and the surface tension
+was set to 0.0~dyn explicitly. For all heating and equilibration steps,
+the lengths of the hydrogen containing-bonds were restrained using the SHAKE
+algorithm; time step of 2~fs was used. In all steps with periodic
 boundary conditions, the non-bonded interactions were calculated with
@@
 
-For the production runs, the temperature and pressure were set to 303~K
+For the production runs, temperature and pressure were set to 303~K
 and 1~bar, using Langevin thermostat with 1.0~ps$^{-1}$ collision
-frequency and Berendsen barostat with 1.0~ps relaxation time,
-respectively. Periodic boundary conditions were applied in all
-dimensions and the non-bonded interactions are calculated with particle
+frequency and Berendsen barostat with 1.0~ps relaxation time. Periodic boundary conditions were applied in all
+dimensions, and the non-bonded interactions calculated with particle
 mesh Ewald~\cite{essman95} using 1.0~nm cutoff. The lengths of the
-hydrogen containing bonds were constrained using SHAKE algorithm and a
-time step of 2~fs was used. Trajectories are saved in every 10~ps
-intervals. Two independent, 500~ns long trajectories were generated, of which
+hydrogen-containing bonds were constrained using the SHAKE algorithm.
+Time step of 2~fs was used. Trajectories were saved at 10~ps
+intervals. Two independent 500~ns long trajectories were generated, of which
 the last 200~ns was used for the analysis.
@@
 {\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions.} 
-The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
+The initial structures of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
 Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}. For the simulations of the lipids and ions,
 the Amber Lipid17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
-Simulations were ran using Amber18 simulation package \cite{amber18md} 
-The simulation steps are the same as the the simulations of
-DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns
+Simulations were run using the Amber18 simulation package \cite{amber18md}. 
+The simulation steps were the same as for the simulations of the
+DOPS and POPS bilayers described above in this section, with the only difference that the total simulation time was 300~ns, of which the last 200~ns
 was used for the analysis. The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
-The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.
-
-\noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using default Amber parameters.} 
+The trajectory files can be found in Refs.~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional CaCl$_2$ using default Amber parameters.} 
 %The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
@@
 %which employed different simulation engines and parameters for ions.
-Simulations were performed using Amber18 simulation package \cite{amber18md} with ff99SB~\cite{aqvist90} ion parameters and Amber Lipid17~\cite{gould18} force field. 
+Simulations were performed using the Amber18 simulation package \cite{amber18md} with the ff99SB~\cite{aqvist90} ion parameters and the Amber Lipid17~\cite{gould18} force field. 
 Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
-The simulation details, minimization, heating, and equilibration runs were the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield. 
+The simulation details, minimization, heating, and equilibration runs were the same as described above in this section for the other POPC:POPS (5:1) simulations with the Amber Lipid17 forcefield. 
 For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
-The trajectory files can be found from Ref~\citenum{POPCpopsLIPID17withCaCl}.
-
-\noindent
-{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using Dang ion parameters.} 
+The trajectory files can be found in Ref.~\citenum{POPCpopsLIPID17withCaCl}.
+
+\noindent
+{\it POPC:POPS (5:1) mixtures with additional CaCl$_2$ using Dang ion parameters.} 
 The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
@@
 % Another set of simulations with Lipid17 force field for lipids \cite{gould18}
-Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
-and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
-with the time step of 2~fs. 
-The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
+Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18}, 
+and simulations were performed using the Gromacs 2018 simulation package~\cite{abraham2015gromacs}
+with time step of 2~fs. 
+The non-bonded interactions were calculated directly within 1.0~nm cutoff; the Verlet scheme was used\cite{Pall13};
 and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
-The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
+The bond lengths of hydrogen atoms were constrained using LINCS~\cite{hess97}.
 Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
@@
 The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
-The systems were equilibrated 50~ns prior to production simulations used for analysis,
-which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
-Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.
+The systems were equilibrated 50~ns prior to production simulations,
+which are available in Ref.~\citenum{lipid17_cacl_series} together with the simulation settings. 
+Topologies in Gromacs format are available in the repository of Ref.~\citenum{ECCpops}.
 
@@
 {\it POPS and DOPS with 128 lipids.}
-Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
-for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model~\cite{jambeck13},
+Starting structures %for the Slipids POPS and DOPS (with 128 lipids) simulations
+were the same as for the CHARMM36 simulations, see section~\ref{sec:CHARMM36}. Force field parameters were taken from the published Slipids PS model~\cite{jambeck13},
 however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
 % [REF: https://pubs.acs.org/doi/10.1021/ct300777p].
-This was due to the use of the older GROMACS group-based cut-off scheme in these previous simulations.
+This was due to the use of the older GROMACS group-based cut-off scheme in those previous simulations.
 In the simulations reported in this work, the Verlet cut-off scheme was used. Van der Waals interactions were
@@
 Full testing and validation of these settings with the Slipids PS force field will be provided in an additional forthcoming work.
-All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations.
+All other simulation settings were identical to those described above (section~\ref{sec:CHARMM36}) for the CHARMM36 DOPS and POPS simulations.
 
@@
 was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
-The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
-ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
+The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system, and
+ions were described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
 The system was simulated in NPT ensemble for 200~ns using the GROMACS 5.0.4 package~\cite{abraham2015gromacs},
@@
 The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath.
-Pressure was kept constant at 1.013~bar using a semi-isotropic Parrinello--Rahman
+Pressure was controlled at 1.013~bar using a semi-isotropic Parrinello--Rahman
 barostat~\cite{parrinello81} with a time constant of 10.0~ps.
@@
 \noindent
-{\it POPC bilayers.} 
-Previously published simulation \cite{ollila07a} available from Ref. \citenum{bergerPOPC310K}
+{\it POPC bilayer.} 
+Previously published simulation \cite{ollila07a} from Ref. \citenum{bergerPOPC310K}
 was used for POPC at 310~K.
@@
 %\todo{T. Piggot, please write simulation details} \\
-Starting structures for DOPS and POPS simulations performed with a Berger-based PS force field
-were taken from those produced for the CHARMM36 simulations. A script was used to re-order the
+Starting structures for %DOPS and POPS
+simulations performed with a Berger-based PS force field
+were taken from those produced for the CHARMM36 simulations, see section~\ref{sec:CHARMM36} above. A script was used to re-order the
 atoms and to remove any extraneous hydrogen atoms. Force field parameters were taken from those
-published by Mukhopadhyay et al.~\cite{mukhopadhyay04}
+published by Mukhopadhyay et al.~\cite{mukhopadhyay04},
 % [REF: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1303994/]
@@
 those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0~nm;
-Coulombic interactions were truncated at 1.0~nm with PME applied to account for the interactions beyond the cut-off.
-All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
+Coulombic interactions were truncated at 1.0~nm, with PME to account for the interactions beyond the cut-off.
+All other simulation settings were identical to those described above (section~\ref{sec:CHARMM36}) for the CHARMM36 DOPS and POPS simulations,
 including the use of the Verlet cut-off scheme, except that here LINCS was used to constrain all bonds.
@@
 {\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions.} 
-Previously published simulations with the additional amount of sodium \cite{jurkiewicz12}
-available from Ref. \citenum{POPCpopsBERGERwith1000mMNa} and with the additional amounts
-of calcium \cite{melcrova16} available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa}
+Previously published simulations with additional amount of sodium~\cite{jurkiewicz12},
+available from Ref. \citenum{POPCpopsBERGERwith1000mMNa}, and with additional amounts
+of calcium~\cite{melcrova16}, available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa},
 were used. To generate the reference system with sodium counterions only,
 the additional ions were removed from the structure file in Ref.~\citenum{POPCpopsBERGERwith1000mMNa}.
-This system was simulated 120~ns using Gromacs 5~\cite{abraham2015gromacs} with parameter (mdp file)
-and force field files (top and itp) available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K}.
+This system was simulated 120~ns using Gromacs 5~\cite{abraham2015gromacs} with parameter (mdp)
+and force field (top and itp) files available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K}.
 The last 80~ns of the trajectory was used for the analysis.
@@
 %\todo{T. Piggot, please write the description of force field.}
-The GROMOS-CKP force field (CKP stands for Chandrasekhar-Kukol-Piggot) contains a set of GROMOS compatible lipid parameters.
+The GROMOS-CKP force field (CKP stands for Chandrasekhar--Kukol--Piggot) contains a set of GROMOS-compatible lipid parameters.
 Parameters have been developed and validated for PC~\cite{Chandrasekhar03,kukol09,piggot12},
 %  [REFS: https://www.ncbi.nlm.nih.gov/pubmed/12632209, https://pubs.acs.org/doi/10.1021/ct8003468 and https://pubs.acs.org/doi/abs/10.1021/ct3003157],
-PE, PG, Cardiolipin~\cite{piggot11}
+PE, PG, cardiolipin~\cite{piggot11}
 %[REF: https://pubs.acs.org/doi/abs/10.1021/jp207013v]
@@
 validation and testing of these PS parameters. The two variants of the force field only differ in terms
-of the charges in the head group. The GROMOS-CKPM (M stands for Mukhopadhyay after the first published
-Berger-based PS simulations published; see above for further details) lipid uses Berger/Chiu based charges
-for the NH3 part of the PS head group (i.e. has the same charges as used in the choline part of the Berger PC lipids).
+of the charges in the head group. The GROMOS-CKPM (M stands for Mukhopadhyay after the first
+Berger-based PS simulations published) %; see above for further details)
+lipid uses Berger/Chiu-based charges
+for the NH$_3$ part of the PS head group (i.e., has the same charges as used in the choline part of the Berger PC lipids).
 The GROMOS-CKP parameters use charges compatible with the rest of the GROMOS force fields;
-the  charges of the NH3 part of the PS head group were taken from a lysine side-chain of the GROMOS 54A7 protein
-force field~\cite{schmid11} % [REF: https://link.springer.com/article/10.1007%2Fs00249-011-0700-9].
+the  charges of the NH$_3$ part of the PS head group were taken from a lysine side-chain of the GROMOS 54A7 protein
+force field~\cite{schmid11}. % [REF: https://link.springer.com/article/10.1007%2Fs00249-011-0700-9].
 The remainder of the parameters were assigned by analogy from either the other GROMOS-CKP lipids or the GROMOS 54A7 force field.
@@
 %\todo{T. Piggot, please write simulation details} \\
-Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the CHARMM36 simulations.
-A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. The force fields parameters for DOPS and POPS
-were as discussed above. The SPC water model was used in these simulations~\cite{berendsen81}. Simulation cut-offs were chosen
+Starting structures %for GROMOS-CKP and GROMOS-CKPM simulations
+were taken from those produced for the CHARMM36 simulations, see section~\ref{sec:CHARMM36} above.
+A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. The force field parameters for DOPS and POPS
+were as discussed in the beginning of the present section. The SPC water model was used in these simulations~\cite{berendsen81}. Simulation cut-offs were chosen
 to closely match those used in the original GROMOS-CKP force field validation: van der Waals interactions were truncated at 1.4~nm
-with a dispersion correction applied to the energy and pressure; Coulombic interactions were also truncated at 1.4~nm with PME used
-to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
+with dispersion corrections applied to energy and pressure; Coulombic interactions were also truncated at 1.4~nm, with PME used
+to treat the long-range interactions. All other simulation settings were identical to those described above (section~\ref{sec:CHARMM36}) for the CHARMM36 DOPS
 and POPS simulations, including the use of the Verlet cut-off scheme, except that here LINCS was used to constrain all bonds.
@@
 % \todo{T. Piggot, please write simulation details} \\
-Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).
-A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation setting were identical to those as described above
-for the GROMOS-CKP(M) POPS bilayers.
+Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see section~\ref{sec:CHARMM36} above).
+A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation settings were identical to those described
+previously in the present section for the GROMOS-CKP(M) POPS bilayers.
 
@@
 mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
-\begin{figure}[]
+
+\begin{figure}[t]
   \centering
@@
 
-When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
-it is important to note that these $S_\mathrm{CH}$s
+When using the PC headgroup $S_\mathrm{CH}$ to evaluate the ion binding affinity to a bilayer containing anionic lipids,
+it is important to note that these $S_\mathrm{CH}$
 increase due to the addition of negative charged lipids according to the molecular electrometer~\cite{seelig87, scherer87}
@@
 overcharging of the membrane.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
-  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=7.5cm]{../Figs/LIPIDSwithCaCl.eps}
+  \includegraphics[width=7.5cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
   \caption{\label{OrderParametersWithCaCl}
@@
 the cation binding affinity also to lipid bilayers containing mixtures of PC and anionic lipids.  
-\begin{figure}[]
+\begin{figure}[t]
   \centering
@@
 \pagebreak
-\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
+\section{Calibration of PC headgroup order parameter response to bound cations}\label{electrometerCALIBRATION}
 %HA: I rewrote this paragraph based on discussion with markus to make it more clear. Please verify that the statements are still accurate
-When using the molecular electrometer concept to compare ion binding affinity between simulations
-and experiments, one needs to keep in mind that two things affect how much a given order parameter changes when solution charge content is varied: 1) the change in the amount of bound charge and 2) the sensitivity of the order parameter to bound charge.
-Therefore, the response of the order parameters to the bound charge in simulations needs to be first calibrated
+When using the molecular electrometer to compare ion binding affinity between simulations
+and experiments, one needs to keep in mind that two things affect how much a given $S_\mathrm{CH}$ changes when the charge content of a solution is varied: 1) the change in the amount of bound charge, and 2) the sensitivity of the $S_\mathrm{CH}$ to bound charge.
+Therefore, the response of the order parameters to bound charge in simulations needs to be first calibrated
 against experiments~\cite{catte16,melcr18}.
 
-In our previous work~\cite{catte16}, we investigated the change in order parameters under varying concentrations of mono- and divalent salts and concluded that the experimental $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio~\cite{akutsu81} was well reproduced by the Lipid14 model, but underestimated by other force fields.
+In our previous work~\cite{catte16}, we investigated the change in order parameters under varying concentrations of monovalent and divalent salts and concluded that the experimental $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio~\cite{akutsu81} was well reproduced by Lipid14, but underestimated by other force fields.
 In a more recent study~\cite{melcr18},
-the headgroup order parameter responses were compared more carefully with the experiments of
+the headgroup order parameter responses were compared more carefully against experiments of
 cationic dihexadecyldimethylammonium bromide (DHAB) surfactants in POPC bilayer~\cite{scherer89}.
 The advantage of this approach is that the amount of DHAB in the bilayer is exactly know in experiments, which can be exploited to extract the sensitivity of the order parameters.
-This revealed that both $S_{\rm CH}^\beta$ and $S_{\rm CH}^\alpha$ in the Lipid14 model are equally oversensitive (and thus giving the correct $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$) to the bound charge,
-whereas CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}), indicating that the headgroup order parameter response to the bound charge is actually more realistic
-in CHARMM36 compared to Lipid14. The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
+This revealed that both $S_{\rm CH}^\beta$ and $S_{\rm CH}^\alpha$ in Lipid14 are equally oversensitive (thus giving the correct $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$) to the bound charge,
+whereas CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}), indicating that the headgroup order parameter response to bound charge is actually more realistic
+in CHARMM36 than in Lipid14. The ratio was overestimated for CHARMM36, because the $\beta$-carbon order parameter
 is relatively more sensitive than the $\alpha$-carbon order parameter.
 
-That being said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
+\begin{figure}[]
+  \centering
+  \includegraphics[width=14.0cm]{../Figs/HGopsDHMDMAB.eps}
+  \caption{\label{CHANGESwithDHMDMAB}
+  Responses of headgroup order parameters to cationic surfactant in
+  POPC bilayer from simulations and experiments \cite{scherer89}.
+  The simulation results for Lipid14 are from Ref.~\citenum{melcr18}.
+  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
+}
+\end{figure}
+
+That being said, in the force fields investigated so far, the discrepancies arising from the sensitivity of lipid headgroup to
 bound charge are typically smaller than then discrepancies arising from ion binding affinity.
+
+\pagebreak
+\section{Sensitivity of molecular electrometer on the definition of ion concentration}\label{concentrationDEFsection}
+%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
+Previous studies using the molecular electrometer to assess ion binding affinity to lipid
+bilayers report ion concentrations either 1) in water before solvating the lipids (buffer concentration)~\cite{akutsu81,roux90,catte16},
+or 2) in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
+In this work, we use the definition 1) to be consistent with the experimental reference data \cite{roux90}. The difference between these two definitions increases
+with increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
+in a force field with a realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
 \begin{figure}[]
   \centering
-  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
-  \caption{\label{CHANGESwithDHMDMAB}
-  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
-  POPC bilayer from simulations and experiments \cite{scherer89}.
-  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
-  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
-}
-\end{figure}
-
-\pagebreak
-\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}\label{concentrationDEFsection}
-%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
-Previous studies using the electrometer concept to assess the ion binding affinity to lipid
-bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
-or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
-In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
-with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
-in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
-\begin{figure}[]
-  \centering
-  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
+  \includegraphics[width=9cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
   \caption{\label{concentrationDEFfigure}
     Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
-    two possible definitions of ion concentration from the recent force field
-    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
-    together with the experimental data \cite{akutsu81,altenbach84}. 
+    two possible definitions of ion concentration. Data from a recent force field
+    with realistic calcium binding affinity to a POPC bilayer~\cite{melcr18}
+    shown together with the experimental data~\cite{akutsu81,altenbach84}. 
   }
@@
 \pagebreak
-\section{Spectral slices in the indirect dimension from the R-PDLF and SDROSS experiments}
-
-The C--H bond order parameters of the headgroup and glycerol backbone carbons are determined 
+\section{Spectral slices in indirect dimension from R-PDLF and S-DROSS experiments}
+
+The C--H bond order parameters of the headgroup and glycerol backbone carbons were determined 
 as $S_{\rm CH}=\Delta\nu/(0.315\times21.5  {\rm kHz})$, where $\Delta\nu$ is the dipolar splitting
@@
 %from the small splitting cannot resolved with the available resolution.
-The resulting order parameters are in good agreement with the previously reported values
+The resulting order parameters were in good agreement with the previously reported values
 from $^2$H NMR experiments \cite{browning80} (Fig. \ref{HGorderParameters} in the main text).
-However, the resolution in our $^{13}$C NMR experiment was not sufficient to detect the
-the splitting related to the smaller order parameter of the C--H bond in the $\alpha$ carbon
+However, the resolution in our $^{13}$C NMR experiment was not sufficient to detect
+the splitting related to the smaller order parameter of the C--H bond in the $\alpha$-carbon
+that has been
 observed in $^2$H NMR experiments \cite{browning80}. Therefore, the value of  0 $\pm$ 0.02 from our
-$^{13}$C NMR experiments is shown figure \ref{HGorderParameters} 
-and the magnitude of 0.02 from the literature is used in the SIMPSON calculations in the main text.
-
-Interpretation of the order parameter signs of the $\alpha$ carbon from the SDROSS experiment
+$^{13}$C NMR experiments is shown in Fig.~\ref{HGorderParameters} in the main text,
+and the magnitude of 0.02 from literature is used in the SIMPSON calculations in the main text.
+
+Interpretation of the $\alpha$-carbon order parameter signs of from the S-DROSS experiment
 is complicated by the presence of distinct order parameters for the two attached hydrogens.
-As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
+As discussed in the main text, interpretation of the S-DROSS curve with the help of SIMPSON
 simulations gave values $+0.09$ and $-0.02$ for the $\alpha$-carbon order parameters.
-To corroborate our interpretation, we measured the SDROSS curve also using the higher 8~kHz
-MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
-Also this experiment indicates a positive value for the larger $\alpha$-carbon order parameter.
+To corroborate our interpretation, we measured the S-DROSS curve also using the higher 8~kHz
+MAS frequency (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
+Also this experiment indicated a positive value for the larger $\alpha$-carbon order parameter.
 \begin{figure}[]
@@
   \caption{\label{DPslices}
-    Spectral slices in indirect dimension from the R-PDLF (top) and S-DROSS (middle and bottom) experiments at distinct chemical shifts.
+    Spectral slices in indirect dimension from the R-PDLF (top row) and S-DROSS (middle and bottom rows) experiments at distinct chemical shifts.
     The chemical shifts and order parameters (calculated from the dipolar splittings indicated with blue lines in the R-PDLF slices)
-    are shown on top each column (chemical shift left, order parameter right).
-    The assignment of columns is given in the middle.  
-    The SDROSS slices were measured using both 5~KHz (middle) and 8~kHz (bottom) because different MAS frequencies enable the
+    are written on top of each column (chemical shift left, order parameter right).
+    The assignment of columns is given with blue text in the middle row panels.  
+    The S-DROSS slices were measured using both the 5~KHz (middle row) and 8~kHz (bottom row) MAS frequencies to enable
     higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
-    The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
-    are used to determine the error bars for $\alpha$ and $\beta$ carbons in Fig.~\ref{PShgSIGNSsimpson} in the main text.
-    The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
+    The background noise taken from chemical shift slices without carbon peaks (grey lines in S-DROSS figures)
+    were used to determine the error bars for the $\alpha$ and $\beta$ carbons in Fig.~\ref{PShgSIGNSsimpson} in the main text.
+    The S-DROSS profile of the acyl chain methyl carbons (leftmost column) was used as reference assuming that these carbons have negative order parameters.
   }
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{../Figs/PS_Labels.png}
-  \includegraphics[width=0.8\textwidth]{../Figs/figS7.png}
+  \includegraphics[width=0.9\textwidth]{../Figs/PS_Labels.png}
+  \includegraphics[width=1.0\textwidth]{../Figs/figS7.png}
   \caption{\label{dihedralsHG}
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{../Figs/figS11.png}
+  \includegraphics[width=\textwidth]{../Figs/figS11.png}
   \caption{\label{dihedralsHGpc}
@@
   \centering
-  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
+  \includegraphics[width=1.02\textwidth]{../Figs/CHANGESwithMONVALENTwithPS.eps}
   \caption{\label{PSresponseTONaCl}
@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
+  \includegraphics[width=0.8\textwidth]{../Figs/CIdensPSOCmixt.eps}
   \caption{  Counterion density distributions from PC:PS mixtures.
@@
   \centering
-  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
-  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
-  \caption{Ion clusters appearing in POPC:POPS (5:1) Lipid14/17 simulations with 1~M of NaCl (left)
-    and MacRog simulations with 4~M of KCl (right).
+  \includegraphics[width=0.6\textwidth]{../Figs/lipid17cluster.png}
+  \includegraphics[width=0.6\textwidth]{../Figs/MacRogIONcluster.png}
+  \caption{Ion clusters appearing in POPC:POPS (5:1) Lipid14/17 simulations with 1~M of NaCl (top)
+    and MacRog simulations with 4~M of KCl (bottom).
 \label{ionCLUSTERS}
@@
 \clearpage
-\section{Calcium binding to POPC in CHARMM36 simulation with NBfix1 parameters}\label{CHARMMcalciumNBfix}
-
-The response of POPC headgroup order parameters to the CaCl$_2$ concentration
+\section{Calcium binding to POPC in CHARMM36 with NBfix1 parameters}\label{CHARMMcalciumNBfix}
+
+The response of POPC headgroup order parameters to CaCl$_2$ concentration
 are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing the
@@
 parameters for the interactions between calcium and lipid oxygens \cite{kim16,han2018graph}
-(Fig.~\ref{changesWITHCaClPS} in the main text), indicating that the calcium binding to the
-bilayer is too weak with these parameters. The response of headgroup order
+(Fig.~\ref{changesWITHCaClPS} in the main text), indicating that calcium binding to the
+bilayer is too weak with these parameters. With the NBfix1 parameters~\cite{kim16} the response of headgroup order
 parameters (Fig.~\ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
-(Fig.~\ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium
-were underestimated also in simulations of pure POPC bilayer with the NBfix1 parameters~\cite{kim16}.
+of calcium (Fig.~\ref{density_profile_CHARMM_CaCl_POPC_NBFix})
+were underestimated also in simulations of pure POPC bilayer.
 Notably, CHARMM36 simulations with the NBfix1 terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium.
@@
   \centering
-  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
+  \includegraphics[width=14.5cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
   \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
-    Headgroup order parameters from CHARMM36 simulations of POPC, where the NBfix term was employed employed for sodium \cite{venable13} ({\it left})
-    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}
-    and simulations without NBfix for the calcium.
+    Headgroup order parameters from CHARMM36 simulations of POPC, where the NBfix term was employed for sodium \cite{venable13} (left)
+    and calcium \cite{kim16} (right), compared with the experimental data \cite{akutsu81,altenbach84}
+    and simulations without NBfix for calcium.
     Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
     with the NBfix term in sodium at Ref.~\citenum{CHARMM36withNaClNBfix}, 
-    with the NBfix term in calcium at Ref.~\citenum{CHARMM36withCaClNBfix} and
+    with the NBfix term in calcium at Ref.~\citenum{CHARMM36withCaClNBfix}, and
     without the NBfix in calcium at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
@@
 \clearpage
-\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
+\section{Calcium density profiles from simulations with \\ POPC:POPS (5:1) mixture}
 %\todo{This section could benefit from some clarifying text}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{../Figs/CAdensPCPSmixture.eps}
   \caption{\label{CAdensPCPSmixtureALL}
-    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
+    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
+The ion densities are taken along the $z$-axis that coincides with the bilayer normal, $z=0$ indicating the bilayer center.
   }
@@
 \begin{description}
-\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
+\item [0 (~):] \noindent {More than half of all the calculated $S_\mathrm{CH}$s (includes all hydrogens bound to that carbon and all lipid tail types investigated for the given force field) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
 %
@@
 \begin{description}
-\item [0 (~):] \noindent {The distance $D$ between the symbols (indicating time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated $S_\mathrm{CH}$s (includes all lipid types investigated for a given force field).}
+\item [0 (~):] \noindent {The distance $D$ between the symbols (indicating time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated $S_\mathrm{CH}$s (includes all lipid tail types investigated for a given force field).}
 %
@@
 {\it Tiago M. Ferreira}
-performed the NMR experiments and NMR simulations, processed and analysed the experimental data, prepared the corresponding figures, wrote parts of the manuscript (NMR methods and interpretation of the SDROSS results), and took part in the final revision.
+performed the NMR experiments and NMR simulations, processed and analysed the experimental data, prepared the corresponding figures, wrote parts of the manuscript (NMR methods and interpretation of the S-DROSS results), and took part in the final revision.
 
@@
 {\it Jukka M{\"a}{\"a}tt{\"a}}
-Prepared figures 6 and 10.
+prepared figures 6 and 10.
 
@@
 {\it Ricky Nencini}
-ran the simulations of POPC systems using CHARMM36 with and without the NB-Fix corrections,
+ran the simulations of POPC systems using CHARMM36 with and without the NBfix corrections,
 ran the simulations of POPC:POPS (5:1) systems using CHARMM36 with the NBfix2 correction,
